### 2015-02-02 paper/document.tex
Added latex repo.

\documentclass{bioinfo}

%\usepackage{doi}

\copyrightyear{2014}
\pubyear{2014}

\begin{document}
\firstpage{1}

\title[lossy-compression]{Lossy compression of DNA sequencing quality data}
\author[El Hadidi \textit{et~al.}]{Mohamed El Hadidi\,$^{1}$\footnote{authors contributed equally}, Christopher M. Hill\,$^{2}$\footnotemark[1], Andr\'{a}s Szolek\,$^{3}$\footnotemark[1], and Michael P. Cummings\,$^4$\footnote{to whom correspondence should be addressed}}
\address{$^{1}$Department of Algorithms in Bioinformatics,  Center for Bioinformatics, University of T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany \\
$^{2}$Department of Computer Science, University of Maryland, College Park,  Maryland, 20742 USA\\
$^{3}$Department of Applied Bioinformatics, Center for Bioinformatics, Quantitative Biology Center, and Department of Computer Science, University of T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany\\
$^{4}$Center for Bioinformatics and Computational Biology, University of Maryland, College Park, 20742 USA}
%$\dagger$authors contributed equally
\history{Received on XXXXX; revised on XXXXX; accepted on XXXX}
% Hack to display authors contribution.
\editor{Associate Editor: XXXXXXX}

\maketitle

\begin{abstract}

\section{Motivation:}
The \textsc{fastq} file format has become the \emph{de facto} standard for storing next-generation sequencing data, containing nucleotide information along with a quantitative measure of the reliability of individual base calls.
As the cost of sequencing continues to decrease, the rate of sequencing data production is increasing, requiring efficient ways of storing and transferring this vast amount of data. Most methods on sequencing data compression focus on compressing nucleotide information without any loss of information.
Quality data, however, have different properties than nucleotide data, and methods compressing nucleotide sequences efficiently do not perform as well on quality sequences. Furthermore, while lossless representation is necessary for nucleotide sequences, it is not an essential requirement for quality values.

Existing methods for compressing quality sequences mostly focused on minimizing the loss of information with less emphasis on effects on subsequent analyses. In this paper, we evaluate several different compression methods for quality values that compromise accuracy for better storage efficiency, and their resulting impact on common bioinformatic analyses using sequence read data.

\section{Results:}
Lossy compression of quality information can greatly decrease storage and memory requirements with little discernable effects on subsequent analysis results.
The four compression strategies in this study were able to produce similar results to those obtained with uncompressed quality sequences in terms of quality control, genome assembly, and alignment of short read to a reference sequence.

\section{Contact:} \href{mike@umiacs.umd.edu}{mike@umiacs.umd.edu}
\end{abstract}

\section{Introduction}

The \emph{de facto} standard for storing next-generation sequencing data is the \textsc{fastq} format. A component of \textsc{fastq} format files~\cite{cock2010sanger} is information for a quantitative measure of quality for each nucleotide based on the \textsc{phred} quality value, which is defined as $Q = -10\ log_{10}\/P$~\cite{ewing1998base}. Depending on the sequencing technology these quality values can range from 0 to 93, and the American Standard Code for Information Interchange \textsc{ascii} characters 33 to 126 (with some offset) are used to represent quality values.

Quality values are typically used throughout bioinformatics pipelines.
Certain tools, such as \textsc{fastqc}~\cite{andrews2010fastqc} and Prinseq~\cite{schmieder2011quality}  provide the user with detailed descriptions of the distribution of sequence quality scores. These analyses can then be used to filter and trim the sequencing data~\cite{martin2011cutadapt}.
The processed sequencing data can then be be used in subsequent analyses.
Certain genome assemblers utilize quality values directly to produce better assemblies~\cite{gnerre2011high}. Genome assemblers work by piecing together overlapping sequences, constructing a larger resulting sequence. Short-read alignment tools, such as Bowtie 2~\cite{langmead2012fast}, use quality values to scale the mismatch penalty in a way that mismatches between poor quality bases of the query and the reference are weighted less than mismatches occurring on high quality bases. Similar to alignment, software for detecting single nucleotide polymorphisms (\textsc{snp}s) can also rely on quality values~\cite{mckenna2010genome}. Detected \textsc{snp}s with high-quality bases are more trustworthy than those with low-quality bases, particularly in low coverage regions.

Although the quality values are useful for later steps, the trade-off between storing effectively double the amount of data and the resulting effects on subsequent analyses has not been well-documented. Previous literature on sequence data compression has largely focused on lossless compression of base calls~\cite{sato2001dna,chen2000compression,chen2002dnacompress}. If a reference exists for the biological data, sequences can be compressed using offsets into the reference~\cite{kozanitis2011compressing}. Typical compression methods fail to utilize the biases inherent in existing sequencing technologies. Depending on the sequencing technology used, the base calling is more accurate in the earlier cycles, gradually losing quality in later cycles~\cite{kozanitis2011compressing}.
To this end, methods for compressing these quality sequences utilize these general quality trends. \textsc{SlimGene}~\cite{kozanitis2011compressing} is one such tool that builds a fixed-state Markov model on adjacent quality scores and uses Huffman Coding for compression based on the transition probabilities.

Little prior work has been done examining the effect of lossy compression of quality sequences in the context of subsequent analyses. Our aim here is to examine the effect of multiple lossy compression methods on quality sequences, focusing not only on the differences between the compressed and original versions, but also the effect of lossy compression on typical subsequent analyses.

We examine four general classes of lossy compression methods: score binning, model-based, profile-based, and data type transformation.

One compression strategy is to look at the distribution of quality values and see if we can summarize the information contained within ranges of values. Specifically, we can reduce the dimensionality of the distribution of quality values by binning together adjacent quality values. We can select the number of bins depending on the desired amount of compression and retention of quality information. Taken to nearly the extreme, a base call can be considered either good or bad. This is the motivation of our approach, deemed quality value binning, where we observe the distribution of quality values across the read set and find a cutoff between good and bad bases.

Another compression strategy involves simplifying the data through modeling of the quality values of a given sequence.
By constructing a compact model of our quality values, we only have to store the parameters of the model to recreate the quality sequences.
Intuitively, the simpler the model, the better the compression, but results in potentially higher differences compared to the original quality sequences.
Another benefit of the model-based compression is the ability to decompress any particular quality value in the sequence with constant time during decompression.
In our approach, we represent the quality values of a sequence as a polynomial function, but it is possible to use other models (such as other basis functions or splines).

Profile-based compression utilizes the fact that in a vast set of quality sequences one can identify large classes of highly similar elements. If one could partition the data into such classes, each of those would be efficiently representable with a single consensus profile.

The last approach we investigated is based on transforming the original data into a wholly different representation for which efficient and proven compression algorithms already exist. In this study, we described a way of reducing the problem of quality compression to image compression via blockwise quality data transformation. This approach is different from the others in that depending on the image compression algorithm used it is not restricted to lossy techniques.

\begin{methods}
\section{Methods}

\subsection{Quality value binning}

%The quality value provides some measure of the accuracy of a given base call. A strategy to reduce the dimensionality of these quality values is to aggregate adjacent quality values together into the same bin. Taken to the extreme, a base call can be considered either good or bad. This is the motivation of our approach, deemed good/bad encoding, where we observe the distribution of quality values across the read set and find a cutoff between good and bad bases.

In one of the more extreme cases of quality value binning, we can bin a quality value into one of two categories: good or bad.
We define good/bad encoding as finding the quality value cutoff to separate between good and bad bases.
%The goal of good/bad encoding is to find the quality value cutoff to separate between 
First, we learn the distribution of quality values across all sequences.
Using this distribution, bases are marked bad if their quality value falls below a certain threshold.
We set the cutoff as the first quartile minus 1.5 * the interquartile range (IQR), which is the difference between the first and third quartile.
1.5 * IQR is the value used by Tukey's box plot~\cite{mcgill1978variations}.
The main benefit of this approach is that it is completely data-dependent. No prior assumptions of the distribution of the quality values need to be made.

Because we only have two possible choices for a quality of base, this makes a straightforward binary encoding possible, allowing us to use a single bit to represent the quality of a base instead of the default 8 bits used to store quality values in \textsc{ascii}.

Although simple and intuitive, considering a base as either good or bad may be too limiting for some users, so we provide a few additional options to preserve more of the defining characteristics of the quality values, but we only analyze good/bad encoding in our results.

\subsection{Polynomial regression}

One approach to compressing quality values is to treat them as bivariate data with the ordered nucleotides (1 to read length) representing the abscissa, and quality values representing the ordinate. Viewed in this way we can ask the question: is there a mathematical function that describes the relationship between read position and quality value sufficiently, albeit imperfectly (i.e., with some loss of information), such that modeled values perform well in subsequent analyses?

To this end, we modeled read quality sequences with zero to fifth-degree polynomials obtained with least-squares fitting. However, despite the fact that polynomial functions have significantly lower number of parameters (i.e. one to six coefficients) than a read-length string of raw quality values, the necessity of using floating point numbers to store coefficients greatly limits the compression potential of the method. In order to get meaningful compression on single-precision four-byte floating point numbers, one would have to relax on the least-squares approximation constraint to obtain compressible values on the byte level which is outside the scope of this study.

Nonetheless, fitting a polynomial and reconstructing the quality values that it would have yielded after decompression still makes some sense as it alters the properties of the quality data, making it highly compressible simply due to the smoothing effect it provides. Therefore the compression ratio and bits per bp values for polynomial regression throughout this manuscript refer to bzipped polynomially approximated quality values.

\subsection{Profile-based compression}

As large sets of quality strings show similar trends of quality over their sequence, it makes sense to identify such common patterns in the data and use them as reference profiles to approximate individual quality sequences. Such patterns can be readily determined by clustering data points (i.e. quality vectors) and using the resulting cluster centers as representative profiles.

K-means clustering is a vector quantization method, partitioning a set of samples into K sets that minimize within-cluster sum of squares~\cite{macqueen1967some}. Using a random subset of read quality values, a compression method can use the computed cluster centers as read quality profiles. As the problem is NP-hard, we use a heuristic iterative refinement approach quickly converging to a locally optimal minimum provided by the Python package scikit-learn 0.15~\cite{scikit-learn}.

%First, the method randomly samples from the set of quality sequences to form the training set. The amount of quality sequences sampled can be adjusted by the user.
First, the method samples an adjustable number of quality sequences randomly from the file to be used as a training set. Quality sequences are represented by vectors containing their \textsc{phred} scores corresponding to each position along the read. Subsequently, K-means clustering is performed on the training set until convergence. The obtained cluster centers will be the quality profile prototypes for the dataset.

Once the K quality profiles are determined, all quality sequences are passed through the trained K-means predictor, with the nearest quality profile in Euclidean space being assigned to every quality sequence as their compressed representation.

The compressed quality file therefore consists of an index enumerating the K quality profiles, and a binary part containing the assigned quality profile index for each read.

Although this approach is not able to capture randomly occuring outlier quality values, it ensures that the overall trends in quality sequences are retained. Quality profiles capture different overall qualities and different drop-off positions and gradients. An example of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
\caption{Quality profiles obtained by K-means clustering on the fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data set using K=128, with each row corresponding to a quality profile. Dark to light colors represent low to high quality values. It is readily visible that the two most distinctive features of quality profiles is their drop-off position and average overall quality. One can also see sporadic low-position values in a handful of profiles, likely capturing intermittent problems in the sequencing process affecting thousands of reads at a time.}\label{fig:profiles_128}
\end{figure}

\subsection{Blockwise image compression-based algorithm}

The principle of utilizing image compression algorithms on seemingly dissimilar data such as electromyography signals is not new~\cite{costa2008compression}. Sequence quality values possess some properties such as smooth overall gradients and sudden drop-offs along a sequence whose accurate and efficient representation are crucial to image compression algorithms. Our aim is therefore to reduce the sequencing quality compression problem to a very well known and extensively researched problem, leveraging the power of mature algorithms on the image compression field. The key steps of the problem reduction are:

\begin{enumerate}
    \item Take a number of quality sequences and represent them as a matrix (rows: reads, columns: base positions).
    \item Sort its rows in a way that maximizes smoothness across the matrix, e.g. apply hierarchical clustering, akin to creating a cluster heat map.
    \item Take the resulting matrix as a grayscale image with each pixel's intensity corresponding to the \textsc{phred} score it represents.
    \item Encode the image with a suitable image compression algorithm.
    \item Either store the sorted indices to recover their original order during decompression, or re-order reads corresponding to quality values instead.
\end{enumerate}

\begin{figure*}[!tpb]%figure2
\centerline{\includegraphics[width=6.9in]{png_pipeline.png}}
\caption{Image compression based approach. Quality values are loaded into a matrix in chunks (256 reads of read-length 101 in the example). Reads are hierarchically clustered, and the rows of the matrix accordingly re-ordered. The new matrix, given its properties such as smoothness and distinct spatial features, lends itself to efficient image compression. Note the uniform dip in quality around position 35: the sequencing process was possibly temporarily compromised, affecting several thousands of reads.}\label{fig:png_method}
\end{figure*}

Lossy image compression algorithms are prone to introducing various artifacts, such as ringing, blockiness, etc. Their implication in this context is unpredictable distortion across multiple positions and even reads. Whether this is an acceptable compromise or not primarily depends on the use case.  In this study, we applied a hybrid approach using lossless Portable Network Graphics (\textsc{png}) encoding on adaptively quantized data, all performed by the \textsc{png} encoder pngquant 2.3.0 (http://pngquant.org/). This method is free of the previously mentioned artifacts, and provides an adjustable level of compression.

The choice of this particular image encoder is not essential to the algorithm, but rather serves as an initial proof of concept implementation. Other image compression algorithms could be explored extensively, and seamlessly substituted into our method for further benchmarking.

Our method uses mean linkage hierarchical clustering for row sorting, which, just like the image encoder, can be replaced with other linkage criteria or even sorting methods to be evaluated for performance.

\subsection{Data sets}

For our analyses, we use sequencing data provided by the \textsc{GAGE} project~\cite{salzberg2012gage}.
The read data for \textit{Rhodobacter sphaeroides} 2.4.1 was downloaded from
\href{http://gage.cbcb.umd.edu/data/Rhodobacter_sphaeroides}{http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides},
%http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides
and the corresponding reference sequence was obtained from the NCBI RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
The \textit{Rhodobacter sphaeroides} 2.4.1 data set consists of 2,050,868 paired-end reads (insert size of 180 bp) and 2,050,868 short-jump reads (insert size of 3,000 bp).

%\begin{table}[!t]
%\processtable{This is table caption\label{Tab:01}}
%{\begin{tabular}{llll}\toprule
%head1 & head2 & head3 & head4\\\midrule
%row1 & row1 & row1 & row1\\
%row2 & row2 & row2 & row2\\
%row3 & row3 & row3 & row3\\
%row4 & row4 & row4 & row4\\\botrule
%\end{tabular}}{This is a footnote}
%\end{table}

\subsection{Evaluating lossy compression algorithms}

Lossy compression compromises representation accuracy for better space efficiency. Different lossy compression methods have different properties, therefore we need to evaluate a number of tradeoffs for each of our proposed methods. More importantly, we need to examine each approach in the context of typical bioinformatics tasks, such as quality control, \emph{de novo} genome assembly, and aligning the sequences to a given reference.

For quality control, we use Autoadapt tool which uses \textsc{fastqc}~\cite{andrews2010fastqc} and \textsc{CutAdapt}~\cite{martin2011cutadapt} packages with the default settings. PrinSeq~\cite{schmieder2011quality} was used to count the number of bases and reads in each \textsc{fastq} file.

For \emph{de novo} genome assembly, we use \textsc{allpaths-lg}~\cite{gnerre2011high} version r50191 with default settings and 32 threads.

For aligning the sequences, we use Bowtie 2~\cite{langmead2012fast} with default settings and modified max- and min-mismatch parameters detailed below.

\end{methods}

\section{Results}

\subsection{Compressibility}

For any lossy compression method, we need to evaluate the tradeoff between its physical compression and the loss in its accuracy.
Here, we use average \emph{bits per bp} to measure the physical compression of the data and mean squared error (MSE) to measure the loss in quality.
Bits per bp is calculated simply as the compressed file size (in bits) divided by the total number of quality values.
MSE is calculated by $\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is the number of sequences, $Q_i$ is the actual quality value associated with sequence position $i$, and $Q_i'$ is the corresponding compressed quality value.

We compare the \textsc{mse} versus \emph{bits per bp} of the fragment and short-jump libraries of the  \textit{Rhodobacter sphaeroides} 2.4.1 data set (Figures \ref{fig:mse_vs_bpbp_frag} and \ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values requires 1 \emph{byte} per basepair because they are stored in \textsc{ascii} format; however, losslessly compressing the values with BZip2 2.112 bits per basepair. 0-order polynomial regression and 32-profile encodings have the lowest bits per bp. 5th-order polynomial regression and \textsc{png} encodings have the highest bits per bp, but have among the lowest \textsc{mse}. Despite having similar bits per bp, \textsc{png} encoding has roughly one third the \textsc{mse} as 5th-order polynomial (3.07 vs 10.27, respectively).

As the order of the polynomial increases, the bits per bp increase and the \textsc{mse} decreases at an exponential rate.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\caption{MSE versus bits per bp for different compression methods applied to the fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256: Quality profile compression with 32 and 256 profiles. R0-R5: Polynomial regression-based compression using 0-5 degrees. PNG: Blockwise image compres the generated SAM file for each compressing approach weresion algorithm. Dashed line corresponds to lossless bzip compression of quality values. \em}\label{fig:mse_vs_bpbp_frag}
\end{figure}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_short.png}}
\caption{MSE versus bits per bp for for different compression methods applied to the short-jump library from \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256: Quality profile compression with 32 and 256 profiles. R0-R5: Polynomial regression-based compression using 0-5 degrees. PNG: Blockwise image compression algorithm. Dashed line corresponds to lossless bzip compression of quality values. \em}\label{fig:mse_vs_bpbp_jump}
\end{figure}

\subsection{Lossy compression effect on preprocessing quality}

One way to evaluate different compression approaches is to perform quality preprocessing on the generated \textsc{fastq} files and compare the number of reads and bases with those of the uncompressed \textsc{fastq} files.
Autoadapt is a tool that uses \textsc{fastqc} and \textsc{Cutadapt}  to filter and trim reads.
We used PrinSeq to calculate the number of reads and bases in each sample. Table \ref{tab:quality_control} shows the proportional numbers with respect to those of the uncompressed \textsc{fastq} files.
In the fragment sample, we retain on average 95.4\% and 94.24\%  of bases and reads, respectively. In the short-jump sample, we retain 90.4\% and 90.98\% of bases and reads, respectively. In some cases, the proportion of reads and/or bases compared to the original is more than 100\%; this means that the number of the reads and bases in the compressed samples are more because altering quality values may prevent the read from being filtered and/or trimmed. This is clear in the good/ bad compression approach of the short-jump sample, where the proportion of bases and reads are 113.09\%  and 103.70\% respectively.

\subsection{Lossy compression effect on assembly quality}

Genome assemblers, such as \textsc{allpaths-lg}~\cite{gnerre2011high}, often utilize quality values to aid in the assembly process. By reducing the information contained within the quality values, we introduce the potential for the resulting assembly to be affected.

\textsc{allpaths-lg} is one of the leading bacterial genome assemblers~\cite{salzberg2012gage}. We examine the effect of our lossy compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data set provided by \textsc{gage} (Genome Assembly Gold-Standard Evaluations)~\cite{salzberg2012gage}. \textsc{allpaths-lg} is used to assemble the original data along with the quality compressed versions using good/bad encoding, 0--5th-order polynomial regression encodings, 32- and 256-profile encodings, and \textsc{png} encoding. We evaluate the assembly based on contiguity statistics, log average read probability (LAP~\cite{ghodsi2013novo}), and a collection of reference-based metrics. The contiguity statistics include number of scaffolds and NG50, which is defined as the weighted median contig size (the length of largest contig $c$ such that the total size of the contigs larger than $c$ exceeds half of the known genome size).  The LAP score can be viewed as a log likelihood score, where a value closer to 0 is better. We use a reference-based evaluation script provided by \textsc{gage} which counts single nucleotide polymorphisms (\textsc{snp}s), relocations, translations, and inversions. The reference-based metrics are normalized by the length of the assembly to aid in comparison. Alongside the raw metrics, we also provide a cummulative sum of the individual ranks of each method for each metric. A similar approach was utilized by the \textsc{Assemblathon1} competition~\cite{earl2011assemblathon}.

The assembly using the uncompressed reads outperforms all compression methods in terms of LAP, NG50, relocations, translations, and inversions (Table \ref{tab:assembly_ranks}). Among the compression methods, the \textsc{png} encoding had the highest rank, followed by 256- and 32-profile encoding, then the 4th-order polynomial regression, the 5th-order polynomial regression, the good/bad encoding, and lastly, the 3rd through 0-order polynomials.

The lossy compression methods largely preserve the contiguity found in the assembly produced using the reads with unmodified quality sequences. All compression methods other than 0-order polynomial regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the similar contiguity statistics, the different compression methods vary noticeably in the amount of SNPs. The order of polynomial has an inverse relationship with the amount of SNPs detected.
The good/bad, profile and PNG methods detected the least amount of SNPs compared to the reference genome, outperforming the assembly using the original quality sequences.
A more indepth evaulation is needed to determine whether these compression methods are missing actual SNPs.

It is important to highlight the result that the 4th-order polynomial regression outperforms the 5th-order in all of the reference-based metrics, but not the \emph{de novo} LAP metrics. The 5th-order polynomial regression assembles approximately 2.5 kb more sequence and uses roughly $1.2\%$ more of the total read set than the 4th-order (Supplemental Table 1), which may explain the discrepency in LAP values because the LAP value is influenced by total coverage~\cite{ghodsi2013novo}.

\begin{table*}[!tpb]
%\processtable{Rankings of different assemblies created using proposed compression methods. Assemblies were created using \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides} 2.4.1. data set provided by \textsc{gage}\citep{salzberg2012gage}.\label{tab:assembly_ranks}}
\centering
\caption[]{Rankings of different assemblies produced using proposed compression methods. Assemblies were constructed using \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides} 2.4.1. data set provided by \textsc{gage}~\cite{salzberg2012gage}. Regression methods are sorted by their overall rank (determined by the sum of individual rankings).}
{\begin{tabular}{lcccccccc}
& Overall & & & & & & LAP & LAP \\
Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
\hline
None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
PNG                    & 2    & 4    & 1          & 2           & 2              & 3      & 4          & 2                    \\
Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
Good/bad               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

\subsection{Lossy compression effect on mapping quality}

Certain short read alignment tools use the quality sequence information when finding potential alignments. Bowtie 2 (version 2.2.3) was used to evaluate the different decompressed \textsc{fastq} files. Bowtie 2 uses quality values written in the \textsc{fastq} files when mapping reads against a reference genome. The original uncompressed and decompressed \textsc{fastq} files were mapped with Bowtie 2 against \textit{Rhodobacter sphaeroides} reference genome. The generated \textsc{sam} file for each compressing approach were compared with the uncompressed \textsc{sam} file. The total, shared and unique proportional numbers of mapped reads are calculated with respect to the uncompressed \textsc{sam} matches as shown in table \ref{tab:aligner}. Additionally, to monitor the effect of quality values on mapping in general, Bowtie 2 was tweaked with the maximum and minimum mismatch penalty equivalent to maximum and minimum quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

Generally, increasing the regression model polynomial order from 0 to 5 results in more mapped reads with the most marked change from order 1 to order 2. The best compression approach is the one that has highest proportion of reads shared between the uncompressed and decompressed files and least number of reads that are unique in both uncompressed and decompressed files. In fragment sample, on average, 99.03\% of reads were mapped using different compression approaches, within which 99\% are shared with the uncompressed alignments and 0.02\% unique mapped reads in the decompressed file. The average missed alignments are 0.99\%. Compression with profile 256 approach was shown to be the best performing approach in the fragment decompressed sample, as a proportional of 99.94\% reads are mapped with respect to the uncompressed mapped reads within which 99.91\% are common and only 0.08\% unique mapped reads in the decompressed file. In short-run sample, the average number of mapped reads is 99.14 \% within which 98.73\% are common and 0.41\% unique mapped reads in the decompressed file.The average missed alignments are 1.2\%. Regression with the 5th polynomial order captured the highest mapped reads of 100\%, within which 99.66\% are common and only 0.33\% unique mapped reads in the decompressed file. In both samples, \textsc{png} compression has the least number of mapped reads that are unique in the decompressed files (only  0.013\% and 0.89\% in the fragments and short-jump samples, respectively).

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files against \textit{Rhodobacter sphaeroides} reference genome using Bowtie 2. Numbers corresponds to the proportion of mapped reads with respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the percentage of mapped reads by both the uncompressed and decompressed data. ``Uncompressed only'' denotes the percentage of reads mapped from the uncompressed data that are not mapped after decompression. ``Compressed only'' denotes the percentage of reads mapped from the decompressed data that were not mapped before compression.}
\begin{tabular}{llcccc}
    & & \multicolumn{4}{c}{Percentage of reads mapped}\\
    \cline{3-6}
	Library & Compression Strategy & Total & Shared & Uncompressed only & Compressed only \\ \hline
	& Maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\
	&Good/Bad & 99.92 & 99.89 & 0.11 & 0.03 \\
	&Regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
	&Regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
	&Regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
	Fragment&Regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
	&Regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
	&Regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
	&Profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
	&Profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\
	&PNG & 99.82 & 99.81 & 0.19 & 0.01 \\

\\ \hline \\

	&Maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
	&Minimum quality & 100 & 100 & 0 & 0 \\ 
	&Good/Bad & 97.38 & 97.38 & 2.62 & 0 \\
	&Regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
	&Regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
	&Regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
	Short-jump&Regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
	&Regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
	&Regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
	&Profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
	&Profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
	&PNG & 99.71 & 99.62 & 0.38 & 0.09 \\ \hline
\end{tabular}

\label{tab:aligner}
\end{table*}

\begin{table}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and decompressed \textsc{fastq} files. Numbers correspond to the proportion of bases/reads that are missed or gained after applying compression with respect to the uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lcccc}
	 & \multicolumn{2}{c}{Fragments} & \multicolumn{2}{c}{Short-Jump} \\
	 & \% Bases & \% Reads & \% Bases & \% Reads \\ \hline
	Good/bad & 98.15 & 98.78 & 113.09 & 103.7 \\ 
	Regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
	Regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
	Regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
	Regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
	Regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
	Regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
	Profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
	Profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
	PNG & 99.62 & 99.75 & 99.84 & 99 \\ \hline
\end{tabular}
\end{table}

\section{Discussion}

\subsection{Lossy compression acceptable for subsequent biological analyses}

The primary concern of using lossy compression methods is naturally the extent of information loss, that we quantified by \textsc{mse} in this study. \textsc{mse} and compressibility provide information in the theoretical context to the methods, but they are not the end-all of evaluation criteria. The performance of compressed datasets in different subsequent analyses and applications are just as important. Our benchmarks showed that some of the compression methods with high error rates are still practical for certain kinds of applications. Many subsequent tools proved to have enough additional redundancy built-in to handle such loss in information.
Passing the decompressed quality sequences through quality control software shows that most methods filter nearly as many bases as using original quality sequences.
Assemblers performing sequencing alignment use percent similarity scores that are typically robust to standard sequencing errors.

\subsection{Extension of good/bad encoding}

Good/bad encoding has the nice property of being simple to compute and has a good bits per bp score. The good/bad encoding suffers from having a high \textsc{mse}, but fortunately, we have shown that in the case of genome assembly, good/bad encoding outperforms all polynomial regressions encodings with degree less than 3. Good/bad encoding of the fragment and short-jump libraries of \textit{Rhodobacter sphaeroides} have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-order polynomial regression encodings, respectively. This further highlights the importance of using additional contextual information of the subsequent analyses when evaluating compressed quality values.

A potential extension to good/bad encoding is to incorporate an additional value \emph{okay}.  The \emph{okay} value can be used where the base qualities fall within a good/bad range. Because the distribution of quality values is skewed towards higher quality, we need to experiment with different cutoffs for the \emph{okay} value and determine if the additional storage is noticeable in subsequent analyses.

\subsection{Potential for operations on compressed data}

Perhaps one of the greatest benefits of compressing quality values is the potential to perform quality control and possibly other operations directly on the compressed representations of the data. The tool \textsc{CutAdapt} trims reads that contains a given number of low quality bases from the ends and discards reads that are below a certain threshold. It is trivial to apply the same rules to a few of our compressed quality sequences. In the good/bad encoding, we store a single bit for each quality value. A read with stretches of 0-bits towards the ends can be trimmed and filtered if there are more 0-bits than 1-bits.

In principle, we can quality filter the polynomial regression compression by interpolating where the polynomials are below a certain threshold and then filtering/trimming that part of the read. Unfortunately, storing the coefficients of the polynomials as a float leads to very poor compression.

A beneficial feature of the profile-based approach is that read quality filtering and/or trimming can be performed on the fly during decompression at virtually no computational cost. As all reads are classified into one of $K$ quality profiles, we can determine which of these profiles are of use for subsequent processes, and only decompress reads assigned to these profiles. Furthermore, it is possible to determine trimming points for each profile before extraction, that can be carried out on-line on reads during decompression.

\subsection{The economics of lossy quality sequence compression}

From an economics perspective, we can examine the fixed and marginal costs associated with our lossy compression techniques with respect to memory and computation.
For our quality value binning, modeling-based, and image-based compression methods there are no explicit fixed costs in terms of memory, unlike the profile--based approach, which has a small fixed cost in storing the profiles.
However, the profile-based approach has a much smaller marginal compared to the rest due to the fact that we are simply storing the profile index for each quality sequence.

Regarding computational costs, only quality binning and the profile--based methods have associated fixed costs.
Before performing compression, we have to determine the optimal thresholding values for quality binning, and the cluster representatives for the profile--based method.

In practice, the specific values of these costs are very sensitive to the implementation details.

\subsection{Future of lossy compression in bioinformatics analyses}

We have simply provided here the initial steps in analyzing the effect of lossy compression on quality sequences using a single, high-coverage bacterial data set. More work needs to be done using additional biological data sets, such as human and mouse, along with different sequencing technologies.
A more direct comparison against related lossy compression tools, such as \textsc{SlimGene}~\cite{kozanitis2011compressing} and \textsc{QualComp}~\cite{ochoa2013qualcomp}, needs to be performed. Additionally, other types of sequencing data can be analyzed apart from the Illumina data examined here. For example, the PacBio sequencing instruments produce very long reads (with average read lengths on the order of 10 kbp), but with the trade-off of having a high error-rate ($\sim$15\%).
Unlike the class of quality values we have examined here, the distribution of erroneous bases is relatively uniform~\cite{ferrarini2013evaluation}.
Koren et al.~\cite{koren2013reducing} have shown that the assembly complexity of bacterial genomes can be greatly simplified, producing near complete genome assemblies, by utilizing a single run of these long reads.
If long read sequencing technologies such as PacBio become more widely adopted, it would be of huge benefit to examine the potential of lossy compression algorithms on not only the quality values, but the biological sequencing data themselves.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
%     as it may ignore the figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}

In this paper we have examined lossy compression on sequence quality values and their effect on subsequent analyses. Although most previous examinations on lossy compression primarily focused on information loss, we have shown that typically used bioinformatics software today have additional built-in sensitivity to handle significant loss of information in the compressed quality values.

\section*{Acknowledgement}
We thank the organizers and participants of the 2014 University of T\"{u}bingen and University of Maryland Bioinformatics Exchange for Students and Teachers (BEST) Summer School.

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
\bibliographystyle{plain}
%
\bibliography{Document}

\end{document}

### 2015-02-02 manuscript/q-compression.tex
Changed paper to manuscript, which is a more appropriate name for the directory. Expanded BibTeX file, added and revised text in the LaTeX file, and updated citations so they conform to the BibTexfile compression.bib.

--- 2015-02-02 paper/document.tex
@@
 \section{Motivation:}
-The \textsc{fastq} file format has become the \emph{de facto} standard for storing next-generation sequencing data, containing nucleotide information along with a quantitative measure of the reliability of individual base calls.
-As the cost of sequencing continues to decrease, the rate of sequencing data production is increasing, requiring efficient ways of storing and transferring this vast amount of data.
-Most methods on sequencing data compression focus on compressing nucleotide information without any loss of information.
-Quality data, however, have different properties than nucleotide data, and methods compressing nucleotide sequences efficiently do not perform as well on quality sequences.
-Furthermore, while lossless representation is necessary for nucleotide sequences, it is not an essential requirement for quality values.
+The \textsc{fastq} file format has become the \emph{de facto} standard
+for storing next-generation sequencing data, containing nucleotide
+information along with a quantitative measure of the reliability of
+individual base calls. As the cost of sequencing continues to
+decrease, the rate of sequencing data production is increasing,
+requiring efficient ways of storing and transferring this vast amount
+of data. Most methods on sequencing data compression focus on
+compressing nucleotide information without any loss of information.
+Quality data, however, have different properties than nucleotide data,
+and methods compressing nucleotide sequences efficiently do not
+perform as well on quality sequences. Furthermore, while lossless
+representation is necessary for nucleotide sequences, it is not an
+essential requirement for quality values.
 
-Existing methods for compressing quality sequences mostly focused on minimizing the loss of information with less emphasis on effects on subsequent analyses.
-In this paper, we evaluate several different compression methods for quality values that compromise accuracy for better storage efficiency, and their resulting impact on common bioinformatic analyses using sequence read data.
+Existing methods for compressing quality sequences mostly focused on
+minimizing the loss of information with less emphasis on effects on
+subsequent analyses. In this paper, we evaluate several different
+compression methods for quality values that compromise accuracy for
+better storage efficiency, and their resulting impact on common
+bioinformatic analyses using sequence read data.
 
 \section{Results:}
-Lossy compression of quality information can greatly decrease storage and memory requirements with little discernable effects on subsequent analysis results.
-The four compression strategies in this study were able to produce similar results to those obtained with uncompressed quality sequences in terms of quality control, genome assembly, and alignment of short read to a reference sequence.
+Lossy compression of quality information can greatly decrease storage
+and memory requirements with little discernable effects on subsequent
+analysis results. The four compression strategies in this study were
+able to produce similar results to those obtained with uncompressed
+quality sequences in terms of quality control, genome assembly, and
+alignment of short read to a reference sequence.
 
@@
 
-The \emph{de facto} standard for storing next-generation sequencing data is the \textsc{fastq} format.
-A component of \textsc{fastq} format files~\cite{cock2010sanger} is information for a quantitative measure of quality for each nucleotide based on the \textsc{phred} quality value, which is defined as $Q = -10\ log_{10}\/P$~\cite{ewing1998base}.
-Depending on the sequencing technology these quality values can range from 0 to 93, and the American Standard Code for Information Interchange \textsc{ascii} characters 33 to 126 (with some offset) are used to represent quality values.
+The \emph{de facto} standard for storing next-generation sequencing
+data is the \textsc{fastq} format. A component of \textsc{fastq}
+format files~\cite{Cock:2010ve} is information for a quantitative
+measure of quality for each nucleotide based on the \textsc{phred}
+quality value, which is defined as $Q =
+-10\ log_{10}\/P$~\cite{Ewing:1998zr,Ewing:1998ly}. Depending on the
+sequencing technology these quality values can range from 0 to 93, and
+the American Standard Code for Information Interchange \textsc{ascii}
+characters 33 to 126 (with some offset) are used to represent quality
+values.
 
 Quality values are typically used throughout bioinformatics pipelines.
-Certain tools, such as \textsc{fastqc}~\cite{andrews2010fastqc} and Prinseq~\cite{schmieder2011quality}  provide the user with detailed descriptions of the distribution of sequence quality scores.
-These analyses can then be used to filter and trim the sequencing data~\cite{martin2011cutadapt}.
-The processed sequencing data can then be be used in subsequent analyses.
-Certain genome assemblers utilize quality values directly to produce better assemblies~\cite{gnerre2011high}.
-Genome assemblers work by piecing together overlapping sequences, constructing a larger resulting sequence.
-Short-read alignment tools, such as Bowtie 2~\cite{langmead2012fast}, use quality values to scale the mismatch penalty in a way that mismatches between poor quality bases of the query and the reference are weighted less than mismatches occurring on high quality bases.
-Similar to alignment, software for detecting single nucleotide polymorphisms (\textsc{snp}s) can also rely on quality values~\cite{mckenna2010genome}.
-Detected \textsc{snp}s with high-quality bases are more trustworthy than those with low-quality bases, particularly in low coverage regions.
+Certain tools, such as \textsc{fastqc}~\cite{citeulike:11583827} and
+Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
+descriptions of the distribution of sequence quality scores. These
+analyses can then be used to filter and trim the sequencing
+data~\cite{EJ200}. The processed sequencing data can then be be used
+in subsequent analyses. Certain genome assemblers utilize quality
+values directly to produce better
+assemblies~\cite{Gnerre:2011kx}. Genome assemblers work by piecing
+together overlapping sequences, constructing a larger resulting
+sequence. Short-read alignment tools, such as Bowtie
+2~\cite{Langmead:2012rw}, use quality values to scale the mismatch
+penalty in a way that mismatches between poor quality bases of the
+query and the reference are weighted less than mismatches occurring on
+high quality bases. Similar to alignment, software for detecting
+single nucleotide polymorphisms (\textsc{snp}s) can also rely on
+quality values~\cite{McKenna:2010bh}. Detected \textsc{snp}s with
+high-quality bases are more trustworthy than those with low-quality
+bases, particularly in low coverage regions.
 
-Although the quality values are useful for later steps, the trade-off between storing effectively double the amount of data and the resulting effects on subsequent analyses has not been well-documented.
-Previous literature on sequence data compression has largely focused on lossless compression of base calls~\cite{sato2001dna,chen2000compression,chen2002dnacompress}.
-If a reference exists for the biological data, sequences can be compressed using offsets into the reference~\cite{kozanitis2011compressing}.
-Typical compression methods fail to utilize the biases inherent in existing sequencing technologies.
-Depending on the sequencing technology used, the base calling is more accurate in the earlier cycles, gradually losing quality in later cycles~\cite{kozanitis2011compressing}.
-To this end, methods for compressing these quality sequences utilize these general quality trends.
-\textsc{SlimGene}~\cite{kozanitis2011compressing} is one such tool that builds a fixed-state Markov model on adjacent quality scores and uses Huffman Coding for compression based on the transition probabilities.
+Although the quality values are useful for later steps, the trade-off
+between storing effectively double the amount of data and the
+resulting effects on subsequent analyses has not been
+well-documented. Previous literature on sequence data compression has
+largely focused on lossless compression of base
+calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Giancarlo:2014rw}.
+If a reference exists for the biological data, sequences can be
+compressed using offsets into the
+reference~\cite{Kozanitis:2011kl}. Typical compression methods fail to
+utilize the biases inherent in existing sequencing
+technologies. Depending on the sequencing technology used, the base
+calling is more accurate in the earlier cycles, gradually losing
+quality in later cycles~\cite{Kozanitis:2011kl}. To this end, methods
+for compressing these quality sequences utilize these general quality
+trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
+that builds a fixed-state Markov model on adjacent quality scores and
+uses Huffman Coding for compression based on the transition
+probabilities.
 
-Little prior work has been done examining the effect of lossy compression of quality sequences in the context of subsequent analyses.
-Our aim here is to examine the effect of multiple lossy compression methods on quality sequences, focusing not only on the differences between the compressed and original versions, but also the effect of lossy compression on typical subsequent analyses.
+Little prior work has been done examining the effect of lossy
+compression of quality sequences in the context of subsequent
+analyses. Our aim here is to examine the effect of multiple lossy
+compression methods on quality sequences, focusing not only on the
+differences between the compressed and original versions, but also the
+effect of lossy compression on typical subsequent analyses.
 
-We examine four general classes of lossy compression methods: score binning, model-based, profile-based, and data type transformation.
+We examine four general classes of lossy compression methods: score
+binning, model-based, profile-based, and data type transformation.
 
-One compression strategy is to look at the distribution of quality values and see if we can summarize the information contained within ranges of values.
-Specifically, we can reduce the dimensionality of the distribution of quality values by binning together adjacent quality values.
-We can select the number of bins depending on the desired amount of compression and retention of quality information.
-Taken to nearly the extreme, a base call can be considered either good or bad.
-This is the motivation of our approach, deemed quality value binning, where we observe the distribution of quality values across the read set and find a cutoff between good and bad bases.
+One compression strategy is to look at the distribution of quality
+values and see if we can summarize the information contained within
+ranges of values. Specifically, we can reduce the dimensionality of
+the distribution of quality values by binning together adjacent
+quality values. We can select the number of bins depending on the
+desired amount of compression and retention of quality
+information. Taken to nearly the extreme, a base call can be
+considered either good or bad. This is the motivation of our approach,
+deemed quality value binning, where we observe the distribution of
+quality values across the read set and find a cutoff between good and
+bad bases.
 
-Another compression strategy involves simplifying the data through modeling of the quality values of a given sequence.
-By constructing a compact model of our quality values, we only have to store the parameters of the model to recreate the quality sequences.
-Intuitively, the simpler the model, the better the compression, but results in potentially higher differences compared to the original quality sequences.
-Another benefit of the model-based compression is the ability to decompress any particular quality value in the sequence with constant time during decompression.
-In our approach, we represent the quality values of a sequence as a polynomial function, but it is possible to use other models (such as other basis functions or splines).
+Another compression strategy involves simplifying the data through
+modeling of the quality values of a given sequence. By constructing a
+compact model of our quality values, we only have to store the
+parameters of the model to recreate the quality sequences.
+Intuitively, the simpler the model, the better the compression, but
+results in potentially higher differences compared to the original
+quality sequences. Another benefit of the model-based compression is
+the ability to decompress any particular quality value in the sequence
+with constant time during decompression. In our approach, we represent
+the quality values of a sequence as a polynomial function, but it is
+possible to use other models (such as other basis functions or
+splines).
 
-Profile-based compression utilizes the fact that in a vast set of quality sequences one can identify large classes of highly similar elements.
-If one could partition the data into such classes, each of those would be efficiently representable with a single consensus profile.
+Profile-based compression utilizes the fact that in a vast set of
+quality sequences one can identify large classes of highly similar
+elements. If one could partition the data into such classes, each of
+those would be efficiently representable with a single consensus
+profile.
 
-The last approach we investigated is based on transforming the original data into a wholly different representation for which efficient and proven compression algorithms already exist.
-In this study, we described a way of reducing the problem of quality compression to image compression via blockwise quality data transformation.
-This approach is different from the others in that depending on the image compression algorithm used it is not restricted to lossy techniques.
+The last approach we investigated is based on transforming the
+original data into a wholly different representation for which
+efficient and proven compression algorithms already exist. In this
+study, we described a way of reducing the problem of quality
+compression to image compression via blockwise quality data
+transformation. This approach is different from the others in that
+depending on the image compression algorithm used it is not restricted
+to lossy techniques.
 
@@
 
-%The quality value provides some measure of the accuracy of a given base call.
-A strategy to reduce the dimensionality of these quality values is to aggregate adjacent quality values together into the same bin.
-Taken to the extreme, a base call can be considered either good or bad.
-This is the motivation of our approach, deemed good/bad encoding, where we observe the distribution of quality values across the read set and find a cutoff between good and bad bases.
+%The quality value provides some measure of the accuracy of a given
+%base call. A strategy to reduce the dimensionality of these quality
+%values is to aggregate adjacent quality values together into the same
+%bin. Taken to the extreme, a base call can be considered either good
+%or bad. This is the motivation of our approach, deemed good/bad
+%encoding, where we observe the distribution of quality values across
+%the read set and find a cutoff between good and bad bases.
 
-In one of the more extreme cases of quality value binning, we can bin a quality value into one of two categories: good or bad.
-We define good/bad encoding as finding the quality value cutoff to separate between good and bad bases.
-%The goal of good/bad encoding is to find the quality value cutoff to separate between 
-First, we learn the distribution of quality values across all sequences.
-Using this distribution, bases are marked bad if their quality value falls below a certain threshold.
-We set the cutoff as the first quartile minus 1.5 * the interquartile range (IQR), which is the difference between the first and third quartile.
-1.5 * IQR is the value used by Tukey's box plot~\cite{mcgill1978variations}.
-The main benefit of this approach is that it is completely data-dependent. No prior assumptions of the distribution of the quality values need to be made.
+In one of the more extreme cases of quality value binning, we can bin
+a quality value into one of two categories: good or bad. We define
+good/bad encoding as finding the quality value cutoff to separate
+between good and bad bases. First, we learn the distribution of
+quality values across all sequences. Using this distribution, bases
+are marked bad if their quality value falls below a certain
+threshold. We set the cutoff as the first quartile minus 1.5 * the
+interquartile range (IQR), which is the difference between the first
+and third quartile. 1.5 * IQR is the value used by Tukey's box
+plot~\cite{mcgill1978variations}. The main benefit of this approach is
+that it is completely data-dependent. No prior assumptions of the
+distribution of the quality values need to be made.
 
-Because we only have two possible choices for a quality of base, this makes a straightforward binary encoding possible, allowing us to use a single bit to represent the quality of a base instead of the default 8 bits used to store quality values in \textsc{ascii}.
+Because we only have two possible choices for a quality of base, this
+makes a straightforward binary encoding possible, allowing us to use a
+single bit to represent the quality of a base instead of the default 8
+bits used to store quality values in \textsc{ascii}.
 
-Although simple and intuitive, considering a base as either good or bad may be too limiting for some users, so we provide a few additional options to preserve more of the defining characteristics of the quality values, but we only analyze good/bad encoding in our results.
+Although simple and intuitive, considering a base as either good or
+bad may be too limiting for some users, so we provide a few additional
+options to preserve more of the defining characteristics of the
+quality values, but we only analyze good/bad encoding in our results.
 
@@
 
-One approach to compressing quality values is to treat them as bivariate data with the ordered nucleotides (1 to read length) representing the abscissa, and quality values representing the ordinate.
-Viewed in this way we can ask the question: is there a mathematical function that describes the relationship between read position and quality value sufficiently, albeit imperfectly (i.e., with some loss of information), such that modeled values perform well in subsequent analyses?
+One approach to compressing quality values is to treat them as
+bivariate data with the ordered nucleotides (1 to read length)
+representing the abscissa, and quality values representing the
+ordinate. Viewed in this way we can ask the question: is there a
+mathematical function that describes the relationship between read
+position and quality value sufficiently, albeit imperfectly (i.e.,
+with some loss of information), such that modeled values perform well
+in subsequent analyses?
 
-To this end, we modeled read quality sequences with zero to fifth-degree polynomials obtained with least-squares fitting.
-However, despite the fact that polynomial functions have significantly lower number of parameters (i.e.
-one to six coefficients) than a read-length string of raw quality values, the necessity of using floating point numbers to store coefficients greatly limits the compression potential of the method.
-In order to get meaningful compression on single-precision four-byte floating point numbers, one would have to relax on the least-squares approximation constraint to obtain compressible values on the byte level which is outside the scope of this study.
+To this end, we modeled read quality sequences with zero to
+fifth-degree polynomials obtained with least-squares fitting. However,
+despite the fact that polynomial functions have significantly lower
+number of parameters (i.e. one to six coefficients) than a read-length
+string of raw quality values, the necessity of using floating point
+numbers to store coefficients greatly limits the compression potential
+of the method. In order to get meaningful compression on
+single-precision four-byte floating point numbers, one would have to
+relax on the least-squares approximation constraint to obtain
+compressible values on the byte level which is outside the scope of
+this study.
 
-Nonetheless, fitting a polynomial and reconstructing the quality values that it would have yielded after decompression still makes some sense as it alters the properties of the quality data, making it highly compressible simply due to the smoothing effect it provides.
-Therefore the compression ratio and bits per bp values for polynomial regression throughout this manuscript refer to bzipped polynomially approximated quality values.
+Nonetheless, fitting a polynomial and reconstructing the quality
+values that it would have yielded after decompression still makes some
+sense as it alters the properties of the quality data, making it
+highly compressible simply due to the smoothing effect it
+provides. Therefore the compression ratio and bits per bp values for
+polynomial regression throughout this manuscript refer to bzipped
+polynomially approximated quality values.
 
@@
 
-As large sets of quality strings show similar trends of quality over their sequence, it makes sense to identify such common patterns in the data and use them as reference profiles to approximate individual quality sequences.
-Such patterns can be readily determined by clustering data points (i.e.
-quality vectors) and using the resulting cluster centers as representative profiles.
+As large sets of quality strings show similar trends of quality over
+their sequence, it makes sense to identify such common patterns in the
+data and use them as reference profiles to approximate individual
+quality sequences. Such patterns can be readily determined by
+clustering data points (i.e. quality vectors) and using the resulting
+cluster centers as representative profiles.
 
-K-means clustering is a vector quantization method, partitioning a set of samples into K sets that minimize within-cluster sum of squares~\cite{macqueen1967some}.
-Using a random subset of read quality values, a compression method can use the computed cluster centers as read quality profiles.
-As the problem is NP-hard, we use a heuristic iterative refinement approach quickly converging to a locally optimal minimum provided by the Python package scikit-learn 0.15~\cite{scikit-learn}.
+K-means clustering is a vector quantization method, partitioning a set
+of samples into K sets that minimize within-cluster sum of
+squares~\cite{macqueen1967some}. Using a random subset of read quality
+values, a compression method can use the computed cluster centers as
+read quality profiles. As the problem is NP-hard, we use a heuristic
+iterative refinement approach quickly converging to a locally optimal
+minimum provided by the Python package scikit-learn
+0.15~\cite{pedregosa2011scikit}.
 
-%First, the method randomly samples from the set of quality sequences to form the training set. The amount of quality sequences sampled can be adjusted by the user.
-First, the method samples an adjustable number of quality sequences randomly from the file to be used as a training set.
-Quality sequences are represented by vectors containing their \textsc{phred} scores corresponding to each position along the read.
-Subsequently, K-means clustering is performed on the training set until convergence.
-The obtained cluster centers will be the quality profile prototypes for the dataset.
+First, the method samples an adjustable number of quality sequences
+randomly from the file to be used as a training set. Quality sequences
+are represented by vectors containing their \textsc{phred} scores
+corresponding to each position along the read. Subsequently, K-means
+clustering is performed on the training set until convergence. The
+obtained cluster centers will be the quality profile prototypes for
+the dataset.
 
-Once the K quality profiles are determined, all quality sequences are passed through the trained K-means predictor, with the nearest quality profile in Euclidean space being assigned to every quality sequence as their compressed representation.
+Once the K quality profiles are determined, all quality sequences are
+passed through the trained K-means predictor, with the nearest quality
+profile in Euclidean space being assigned to every quality sequence as
+their compressed representation.
 
-The compressed quality file therefore consists of an index enumerating the K quality profiles, and a binary part containing the assigned quality profile index for each read.
+The compressed quality file therefore consists of an index enumerating
+the K quality profiles, and a binary part containing the assigned
+quality profile index for each read.
 
-Although this approach is not able to capture randomly occuring outlier quality values, it ensures that the overall trends in quality sequences are retained.
-Quality profiles capture different overall qualities and different drop-off positions and gradients.
-An example of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.
+Although this approach is not able to capture randomly occuring
+outlier quality values, it ensures that the overall trends in quality
+sequences are retained. Quality profiles capture different overall
+qualities and different drop-off positions and gradients. An example
+of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.
 
@@
 \centerline{\includegraphics[width=3.35in]{profiles_128.png}}
-\caption{Quality profiles obtained by K-means clustering on the fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data set using K=128, with each row corresponding to a quality profile.
-Dark to light colors represent low to high quality values.
-It is readily visible that the two most distinctive features of quality profiles is their drop-off position and average overall quality.
-One can also see sporadic low-position values in a handful of profiles, likely capturing intermittent problems in the sequencing process affecting thousands of reads at a time.}\label{fig:profiles_128}
+\caption{Quality profiles obtained by K-means clustering on the
+  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
+  set using K=128, with each row corresponding to a quality
+  profile. Dark to light colors represent low to high quality
+  values. It is readily visible that the two most distinctive features
+  of quality profiles is their drop-off position and average overall
+  quality. One can also see sporadic low-position values in a handful
+  of profiles, likely capturing intermittent problems in the
+  sequencing process affecting thousands of reads at a
+  time.}\label{fig:profiles_128}
 \end{figure}
-
-\subsection{Blockwise image compression-based algorithm}
-
-The principle of utilizing image compression algorithms on seemingly dissimilar data such as electromyography signals is not new~\cite{costa2008compression}.
-Sequence quality values possess some properties such as smooth overall gradients and sudden drop-offs along a sequence whose accurate and efficient representation are crucial to image compression algorithms.
-Our aim is therefore to reduce the sequencing quality compression problem to a very well known and extensively researched problem, leveraging the power of mature algorithms on the image compression field.
-The key steps of the problem reduction are:
-
-\begin{enumerate}
-    \item Take a number of quality sequences and represent them as a matrix (rows: reads, columns: base positions).
-    \item Sort its rows in a way that maximizes smoothness across the matrix, e.g. apply hierarchical clustering, akin to creating a cluster heat map.
-    \item Take the resulting matrix as a grayscale image with each pixel's intensity corresponding to the \textsc{phred} score it represents.
-    \item Encode the image with a suitable image compression algorithm.
-    \item Either store the sorted indices to recover their original order during decompression, or re-order reads corresponding to quality values instead.
-\end{enumerate}
-
-\begin{figure*}[!tpb]%figure2
-\centerline{\includegraphics[width=6.9in]{png_pipeline.png}}
-\caption{Image compression based approach.
-Quality values are loaded into a matrix in chunks (256 reads of read-length 101 in the example).
-Reads are hierarchically clustered, and the rows of the matrix accordingly re-ordered.
-The new matrix, given its properties such as smoothness and distinct spatial features, lends itself to efficient image compression.
-Note the uniform dip in quality around position 35: the sequencing process was possibly temporarily compromised, affecting several thousands of reads.}\label{fig:png_method}
-\end{figure*}
-
-Lossy image compression algorithms are prone to introducing various artifacts, such as ringing, blockiness, etc.
-Their implication in this context is unpredictable distortion across multiple positions and even reads.
-Whether this is an acceptable compromise or not primarily depends on the use case.
-In this study, we applied a hybrid approach using lossless Portable Network Graphics (\textsc{png}) encoding on adaptively quantized data, all performed by the \textsc{png} encoder pngquant 2.3.0 (http://pngquant.org/).
-This method is free of the previously mentioned artifacts, and provides an adjustable level of compression.
-
-The choice of this particular image encoder is not essential to the algorithm, but rather serves as an initial proof of concept implementation.
-Other image compression algorithms could be explored extensively, and seamlessly substituted into our method for further benchmarking.
-
-Our method uses mean linkage hierarchical clustering for row sorting, which, just like the image encoder, can be replaced with other linkage criteria or even sorting methods to be evaluated for performance.
 
@@
 
-For our analyses, we use sequencing data provided by the \textsc{GAGE} project~\cite{salzberg2012gage}.
-The read data for \textit{Rhodobacter sphaeroides} 2.4.1 was downloaded from
+For our analyses, we use sequencing data provided by the \textsc{GAGE}
+project~\cite{Salzberg:2012rc}. The read data for \textit{Rhodobacter
+  sphaeroides} 2.4.1 was downloaded from
 \href{http://gage.cbcb.umd.edu/data/Rhodobacter_sphaeroides}{http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides},
-%http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides
-and the corresponding reference sequence was obtained from the NCBI RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
-The \textit{Rhodobacter sphaeroides} 2.4.1 data set consists of 2,050,868 paired-end reads (insert size of 180 bp) and 2,050,868 short-jump reads (insert size of 3,000 bp).
-
-%\begin{table}[!t]
-%\processtable{This is table caption\label{Tab:01}}
-%{\begin{tabular}{llll}\toprule
-%head1 & head2 & head3 & head4\\\midrule
-%row1 & row1 & row1 & row1\\
-%row2 & row2 & row2 & row2\\
-%row3 & row3 & row3 & row3\\
-%row4 & row4 & row4 & row4\\\botrule
-%\end{tabular}}{This is a footnote}
-%\end{table}
+and the corresponding reference sequence was obtained from the NCBI
+RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1,
+NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1). The
+\textit{Rhodobacter sphaeroides} 2.4.1 data set consists of 2,050,868
+paired-end reads (insert size of 180 bp) and 2,050,868 short-jump
+reads (insert size of 3,000 bp).
 
@@
 
-Lossy compression compromises representation accuracy for better space efficiency.
-Different lossy compression methods have different properties, therefore we need to evaluate a number of tradeoffs for each of our proposed methods.
-More importantly, we need to examine each approach in the context of typical bioinformatics tasks, such as quality control, \emph{de novo} genome assembly, and aligning the sequences to a given reference.
+Lossy compression compromises representation accuracy for better space
+efficiency. Different lossy compression methods have different
+properties, therefore we need to evaluate a number of tradeoffs for
+each of our proposed methods. More importantly, we need to examine
+each approach in the context of typical bioinformatics tasks, such as
+quality control, \emph{de novo} genome assembly, and aligning the
+sequences to a given reference.
 
-For quality control, we use Autoadapt tool which uses \textsc{fastqc}~\cite{andrews2010fastqc} and \textsc{CutAdapt}~\cite{martin2011cutadapt} packages with the default settings.
-PrinSeq~\cite{schmieder2011quality} was used to count the number of bases and reads in each \textsc{fastq} file.
+For quality control, we use Autoadapt tool which uses
+\textsc{fastqc}~\cite{citeulike:11583827} and
+\textsc{CutAdapt}~\cite{EJ200} packages with the default
+settings. PrinSeq~\cite{Schmieder:2011gd} was used to count the number
+of bases and reads in each \textsc{fastq} file.
 
-For \emph{de novo} genome assembly, we use \textsc{allpaths-lg}~\cite{gnerre2011high} version r50191 with default settings and 32 threads.
+For \emph{de novo} genome assembly, we use
+\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
+settings and 32 threads.
 
-For aligning the sequences, we use Bowtie 2~\cite{langmead2012fast} with default settings and modified max- and min-mismatch parameters detailed below.
+For aligning the sequences, we use Bowtie 2~\cite{Langmead:2012rw}
+with default settings and modified max- and min-mismatch parameters
+detailed below.
 
@@
 
-For any lossy compression method, we need to evaluate the tradeoff between its physical compression and the loss in its accuracy.
-Here, we use average \emph{bits per bp} to measure the physical compression of the data and mean squared error (MSE) to measure the loss in quality.
-Bits per bp is calculated simply as the compressed file size (in bits) divided by the total number of quality values.
-MSE is calculated by $\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is the number of sequences, $Q_i$ is the actual quality value associated with sequence position $i$, and $Q_i'$ is the corresponding compressed quality value.
+For any lossy compression method, we need to evaluate the tradeoff
+between its physical compression and the loss in its accuracy. Here,
+we use average \emph{bits per bp} to measure the physical compression
+of the data and mean squared error (MSE) to measure the loss in
+quality. Bits per bp is calculated simply as the compressed file size
+(in bits) divided by the total number of quality values. MSE is
+calculated by $\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is
+the number of sequences, $Q_i$ is the actual quality value associated
+with sequence position $i$, and $Q_i'$ is the corresponding compressed
+quality value.
 
-We compare the \textsc{mse} versus \emph{bits per bp} of the fragment and short-jump libraries of the  \textit{Rhodobacter sphaeroides} 2.4.1 data set (Figures \ref{fig:mse_vs_bpbp_frag} and \ref{fig:mse_vs_bpbp_jump}).
-Storing the uncompressed quality values requires 1 \emph{byte} per basepair because they are stored in \textsc{ascii} format; however, losslessly compressing the values with BZip2 2.112 bits per basepair.
-0-order polynomial regression and 32-profile encodings have the lowest bits per bp.
-5th-order polynomial regression and \textsc{png} encodings have the highest bits per bp, but have among the lowest \textsc{mse}.
-Despite having similar bits per bp, \textsc{png} encoding has roughly one third the \textsc{mse} as 5th-order polynomial (3.07 vs 10.27, respectively).
+We compare the \textsc{mse} versus \emph{bits per bp} of the fragment
+and short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
+data set (Figures \ref{fig:mse_vs_bpbp_frag} and
+\ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values
+requires 1 \emph{byte} per basepair because they are stored in
+\textsc{ascii} format; however, losslessly compressing the values with
+BZip2 2.112 bits per basepair. 0-order polynomial regression and
+32-profile encodings have the lowest bits per bp. 5th-order polynomial
+regression and \textsc{png} encodings have the highest bits per bp,
+but have among the lowest \textsc{mse}. Despite having similar bits
+per bp, \textsc{png} encoding has roughly one third the \textsc{mse}
+as 5th-order polynomial (3.07 vs 10.27, respectively).
 
-As the order of the polynomial increases, the bits per bp increase and the \textsc{mse} decreases at an exponential rate.
+As the order of the polynomial increases, the bits per bp increase and
+the \textsc{mse} decreases at an exponential rate.
 
@@
 \centerline{\includegraphics[width=3.65in]{mse_frag.png}}
-\caption{MSE versus bits per bp for different compression methods applied to the fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data set.
-G/B: Good/Bad encoding.
-P32 and P256: Quality profile compression with 32 and 256 profiles.
-R0-R5: Polynomial regression-based compression using 0-5 degrees.
-PNG: Blockwise image compres the generated SAM file for each compressing approach weresion algorithm.
-Dashed line corresponds to lossless bzip compression of quality values.
-\em}\label{fig:mse_vs_bpbp_frag}
+\caption{MSE versus bits per bp for different compression methods
+  applied to the fragment library from \textit{Rhodobacter
+    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
+  Quality profile compression with 32 and 256 profiles. R0-R5:
+  Polynomial regression-based compression using 0-5 degrees. PNG:
+  Blockwise image compres the generated SAM file for each compressing
+  approach weresion algorithm. Dashed line corresponds to lossless
+  bzip compression of quality values. \em}\label{fig:mse_vs_bpbp_frag}
 \end{figure}
@@
 \centerline{\includegraphics[width=3.65in]{mse_short.png}}
-\caption{MSE versus bits per bp for for different compression methods applied to the short-jump library from \textit{Rhodobacter sphaeroides} 2.4.1 data set.
-G/B: Good/Bad encoding.
-P32 and P256: Quality profile compression with 32 and 256 profiles.
-R0-R5: Polynomial regression-based compression using 0-5 degrees.
-PNG: Blockwise image compression algorithm.
-Dashed line corresponds to lossless bzip compression of quality values.
-\em}\label{fig:mse_vs_bpbp_jump}
+\caption{MSE versus bits per bp for for different compression methods
+  applied to the short-jump library from \textit{Rhodobacter
+    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
+  Quality profile compression with 32 and 256 profiles. R0-R5:
+  Polynomial regression-based compression using 0-5 degrees. PNG:
+  Blockwise image compression algorithm. Dashed line corresponds to
+  lossless bzip compression of quality
+  values. \em}\label{fig:mse_vs_bpbp_jump}
 \end{figure}
@@
 
-One way to evaluate different compression approaches is to perform quality preprocessing on the generated \textsc{fastq} files and compare the number of reads and bases with those of the uncompressed \textsc{fastq} files.
-Autoadapt is a tool that uses \textsc{fastqc} and \textsc{Cutadapt}  to filter and trim reads.
-We used PrinSeq to calculate the number of reads and bases in each sample.
-Table \ref{tab:quality_control} shows the proportional numbers with respect to those of the uncompressed \textsc{fastq} files.
-In the fragment sample, we retain on average 95.4\% and 94.24\%  of bases and reads, respectively.
-In the short-jump sample, we retain 90.4\% and 90.98\% of bases and reads, respectively.
-In some cases, the proportion of reads and/or bases compared to the original is more than 100\%; this means that the number of the reads and bases in the compressed samples are more because altering quality values may prevent the read from being filtered and/or trimmed.
-This is clear in the good/ bad compression approach of the short-jump sample, where the proportion of bases and reads are 113.09\%  and 103.70\% respectively.
+One way to evaluate different compression approaches is to perform
+quality preprocessing on the generated \textsc{fastq} files and
+compare the number of reads and bases with those of the uncompressed
+\textsc{fastq} files. Autoadapt is a tool that uses \textsc{fastqc}
+and \textsc{Cutadapt} to filter and trim reads. We used PrinSeq to
+calculate the number of reads and bases in each sample. Table
+\ref{tab:quality_control} shows the proportional numbers with respect
+to those of the uncompressed \textsc{fastq} files. In the fragment
+sample, we retain on average 95.4\% and 94.24\% of bases and reads,
+respectively. In the short-jump sample, we retain 90.4\% and 90.98\%
+of bases and reads, respectively. In some cases, the proportion of
+reads and/or bases compared to the original is more than 100\%; this
+means that the number of the reads and bases in the compressed samples
+are more because altering quality values may prevent the read from
+being filtered and/or trimmed. This is clear in the good/ bad
+compression approach of the short-jump sample, where the proportion of
+bases and reads are 113.09\% and 103.70\% respectively.
 
@@
 
-Genome assemblers, such as \textsc{allpaths-lg}~\cite{gnerre2011high}, often utilize quality values to aid in the assembly process.
-By reducing the information contained within the quality values, we introduce the potential for the resulting assembly to be affected.
+Genome assemblers, such as \textsc{allpaths-lg}~\cite{Gnerre:2011kx},
+often utilize quality values to aid in the assembly process. By
+reducing the information contained within the quality values, we
+introduce the potential for the resulting assembly to be affected.
 
-\textsc{allpaths-lg} is one of the leading bacterial genome assemblers~\cite{salzberg2012gage}.
-We examine the effect of our lossy compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data set provided by \textsc{gage} (Genome Assembly Gold-Standard Evaluations)~\cite{salzberg2012gage}.
-\textsc{allpaths-lg} is used to assemble the original data along with the quality compressed versions using good/bad encoding, 0--5th-order polynomial regression encodings, 32- and 256-profile encodings, and \textsc{png} encoding.
-We evaluate the assembly based on contiguity statistics, log average read probability (LAP~\cite{ghodsi2013novo}), and a collection of reference-based metrics.
-The contiguity statistics include number of scaffolds and NG50, which is defined as the weighted median contig size (the length of largest contig $c$ such that the total size of the contigs larger than $c$ exceeds half of the known genome size).
-The LAP score can be viewed as a log likelihood score, where a value closer to 0 is better.
-We use a reference-based evaluation script provided by \textsc{gage} which counts single nucleotide polymorphisms (\textsc{snp}s), relocations, translations, and inversions.
-The reference-based metrics are normalized by the length of the assembly to aid in comparison.
-Alongside the raw metrics, we also provide a cummulative sum of the individual ranks of each method for each metric.
-A similar approach was utilized by the \textsc{Assemblathon1} competition~\cite{earl2011assemblathon}.
+\textsc{allpaths-lg} is one of the leading bacterial genome
+assemblers~\cite{Salzberg:2012rc}. We examine the effect of our lossy
+compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data
+set provided by \textsc{gage} (Genome Assembly Gold-Standard
+Evaluations)~\cite{Salzberg:2012rc}. \textsc{allpaths-lg} is used to
+assemble the original data along with the quality compressed versions
+using good/bad encoding, 0--5th-order polynomial regression encodings,
+32- and 256-profile encodings, and \textsc{png} encoding. We evaluate
+the assembly based on contiguity statistics, log average read
+probability (LAP~\cite{Ghodsi:2013hb}), and a collection of
+reference-based metrics. The contiguity statistics include number of
+scaffolds and NG50, which is defined as the weighted median contig
+size (the length of largest contig $c$ such that the total size of the
+contigs larger than $c$â exceeds half of the known genome size). The
+LAP score can be viewed as a log likelihood score, where a value
+closer to 0 is better. We use a reference-based evaluation script
+provided by \textsc{gage} which counts single nucleotide polymorphisms
+(\textsc{snp}s), relocations, translations, and inversions. The
+reference-based metrics are normalized by the length of the assembly
+to aid in comparison. Alongside the raw metrics, we also provide a
+cummulative sum of the individual ranks of each method for each
+metric. A similar approach was utilized by the \textsc{Assemblathon1}
+competition~\cite{Earl:2011fv}.
 
-The assembly using the uncompressed reads outperforms all compression methods in terms of LAP, NG50, relocations, translations, and inversions (Table \ref{tab:assembly_ranks}).
-Among the compression methods, the \textsc{png} encoding had the highest rank, followed by 256- and 32-profile encoding, then the 4th-order polynomial regression, the 5th-order polynomial regression, the good/bad encoding, and lastly, the 3rd through 0-order polynomials.
+The assembly using the uncompressed reads outperforms all compression
+methods in terms of LAP, NG50, relocations, translations, and
+inversions (Table \ref{tab:assembly_ranks}). Among the compression
+methods, the \textsc{png} encoding had the highest rank, followed by
+256- and 32-profile encoding, then the 4th-order polynomial
+regression, the 5th-order polynomial regression, the good/bad
+encoding, and lastly, the 3rd through 0-order polynomials.
 
-The lossy compression methods largely preserve the contiguity found in the assembly produced using the reads with unmodified quality sequences.
-All compression methods other than 0-order polynomial regression produce an NG50 ranging from 3.18--3.19 Mbps.
-Despite the similar contiguity statistics, the different compression methods vary noticeably in the amount of SNPs.
-The order of polynomial has an inverse relationship with the amount of SNPs detected.
-The good/bad, profile and PNG methods detected the least amount of SNPs compared to the reference genome, outperforming the assembly using the original quality sequences.
-A more indepth evaulation is needed to determine whether these compression methods are missing actual SNPs.
+The lossy compression methods largely preserve the contiguity found in
+the assembly produced using the reads with unmodified quality
+sequences. All compression methods other than 0-order polynomial
+regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the
+similar contiguity statistics, the different compression methods vary
+noticeably in the amount of SNPs. The order of polynomial has an
+inverse relationship with the amount of SNPs detected. The good/bad,
+profile and PNG methods detected the least amount of SNPs compared to
+the reference genome, outperforming the assembly using the original
+quality sequences. A more indepth evaulation is needed to determine
+whether these compression methods are missing actual SNPs.
 
-It is important to highlight the result that the 4th-order polynomial regression outperforms the 5th-order in all of the reference-based metrics, but not the \emph{de novo} LAP metrics.
-The 5th-order polynomial regression assembles approximately 2.5 kb more sequence and uses roughly $1.2\%$ more of the total read set than the 4th-order (Supplemental Table 1), which may explain the discrepency in LAP values because the LAP value is influenced by total coverage~\cite{ghodsi2013novo}.
+It is important to highlight the result that the 4th-order polynomial
+regression outperforms the 5th-order in all of the reference-based
+metrics, but not the \emph{de novo} LAP metrics. The 5th-order
+polynomial regression assembles approximately 2.5 kb more sequence and
+uses roughly $1.2\%$ more of the total read set than the 4th-order
+(Supplemental Table 1), which may explain the discrepency in LAP
+values because the LAP value is influenced by total
+coverage~\cite{Ghodsi:2013hb}.
 
 \begin{table*}[!tpb]
-%\processtable{Rankings of different assemblies created using proposed compression methods.
-Assemblies were created using \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides} 2.4.1.
-data set provided by \textsc{gage}\citep{salzberg2012gage}.\label{tab:assembly_ranks}}
 \centering
-\caption[]{Rankings of different assemblies produced using proposed compression methods.
-Assemblies were constructed using \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides} 2.4.1.
-data set provided by \textsc{gage}~\cite{salzberg2012gage}.
-Regression methods are sorted by their overall rank (determined by the sum of individual rankings).}
+\caption[]{Rankings of different assemblies produced using proposed
+  compression methods. Assemblies were constructed using
+  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
+  2.4.1. data set provided by
+  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
+  by their overall rank (determined by the sum of individual
+  rankings).}
 {\begin{tabular}{lcccccccc}
@@
 
-Certain short read alignment tools use the quality sequence information when finding potential alignments.
-Bowtie 2 (version 2.2.3) was used to evaluate the different decompressed \textsc{fastq} files.
-Bowtie 2 uses quality values written in the \textsc{fastq} files when mapping reads against a reference genome.
-The original uncompressed and decompressed \textsc{fastq} files were mapped with Bowtie 2 against \textit{Rhodobacter sphaeroides} reference genome.
-The generated \textsc{sam} file for each compressing approach were compared with the uncompressed \textsc{sam} file.
-The total, shared and unique proportional numbers of mapped reads are calculated with respect to the uncompressed \textsc{sam} matches as shown in table \ref{tab:aligner}.
-Additionally, to monitor the effect of quality values on mapping in general, Bowtie 2 was tweaked with the maximum and minimum mismatch penalty equivalent to maximum and minimum quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).
+Certain short read alignment tools use the quality sequence
+information when finding potential alignments. Bowtie 2 (version
+2.2.3) was used to evaluate the different decompressed \textsc{fastq}
+files. Bowtie 2 uses quality values written in the \textsc{fastq}
+files when mapping reads against a reference genome. The original
+uncompressed and decompressed \textsc{fastq} files were mapped with
+Bowtie 2 against \textit{Rhodobacter sphaeroides} reference
+genome. The generated \textsc{sam} file for each compressing approach
+were compared with the uncompressed \textsc{sam} file. The total,
+shared and unique proportional numbers of mapped reads are calculated
+with respect to the uncompressed \textsc{sam} matches as shown in
+table \ref{tab:aligner}. Additionally, to monitor the effect of
+quality values on mapping in general, Bowtie 2 was tweaked with the
+maximum and minimum mismatch penalty equivalent to maximum and minimum
+quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).
 
-Generally, increasing the regression model polynomial order from 0 to 5 results in more mapped reads with the most marked change from order 1 to order 2.
-The best compression approach is the one that has highest proportion of reads shared between the uncompressed and decompressed files and least number of reads that are unique in both uncompressed and decompressed files.
-In fragment sample, on average, 99.03\% of reads were mapped using different compression approaches, within which 99\% are shared with the uncompressed alignments and 0.02\% unique mapped reads in the decompressed file.
-The average missed alignments are 0.99\%.
-Compression with profile 256 approach was shown to be the best performing approach in the fragment decompressed sample, as a proportional of 99.94\% reads are mapped with respect to the uncompressed mapped reads within which 99.91\% are common and only 0.08\% unique mapped reads in the decompressed file.
-In short-run sample, the average number of mapped reads is 99.14 \% within which 98.73\% are common and 0.41\% unique mapped reads in the decompressed file.The average missed alignments are 1.2\%.
-Regression with the 5th polynomial order captured the highest mapped reads of 100\%, within which 99.66\% are common and only 0.33\% unique mapped reads in the decompressed file.
-In both samples, \textsc{png} compression has the least number of mapped reads that are unique in the decompressed files (only  0.013\% and 0.89\% in the fragments and short-jump samples, respectively).
+Generally, increasing the regression model polynomial order from 0 to
+5 results in more mapped reads with the most marked change from order
+1 to order 2. The best compression approach is the one that has
+highest proportion of reads shared between the uncompressed and
+decompressed files and least number of reads that are unique in both
+uncompressed and decompressed files. In fragment sample, on average,
+99.03\% of reads were mapped using different compression approaches,
+within which 99\% are shared with the uncompressed alignments and
+0.02\% unique mapped reads in the decompressed file. The average
+missed alignments are 0.99\%. Compression with profile 256 approach
+was shown to be the best performing approach in the fragment
+decompressed sample, as a proportional of 99.94\% reads are mapped
+with respect to the uncompressed mapped reads within which 99.91\% are
+common and only 0.08\% unique mapped reads in the decompressed
+file. In short-run sample, the average number of mapped reads is 99.14
+\% within which 98.73\% are common and 0.41\% unique mapped reads in
+the decompressed file.The average missed alignments are
+1.2\%. Regression with the 5th polynomial order captured the highest
+mapped reads of 100\%, within which 99.66\% are common and only 0.33\%
+unique mapped reads in the decompressed file. In both samples,
+\textsc{png} compression has the least number of mapped reads that are
+unique in the decompressed files (only 0.013\% and 0.89\% in the
+fragments and short-jump samples, respectively).
 
@@
 \centering
-\caption[]{Mapping results of decompressed \textsc{fastq} files against \textit{Rhodobacter sphaeroides} reference genome using Bowtie 2.
-Numbers corresponds to the proportion of mapped reads with respect to the uncompressed \textsc{fastq}.
-``Shared'' denotes the percentage of mapped reads by both the uncompressed and decompressed data.
-``Uncompressed only'' denotes the percentage of reads mapped from the uncompressed data that are not mapped after decompression.
-``Compressed only'' denotes the percentage of reads mapped from the decompressed data that were not mapped before compression.}
+\caption[]{Mapping results of decompressed \textsc{fastq} files
+  against \textit{Rhodobacter sphaeroides} reference genome using
+  Bowtie 2. Numbers corresponds to the proportion of mapped reads with
+  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
+  percentage of mapped reads by both the uncompressed and decompressed
+  data. ``Uncompressed only'' denotes the percentage of reads mapped
+  from the uncompressed data that are not mapped after
+  decompression. ``Compressed only'' denotes the percentage of reads
+  mapped from the decompressed data that were not mapped before
+  compression.}
 \begin{tabular}{llcccc}
@@
 \centering
-\caption[]{Quality filtering/trimming for the uncompressed and decompressed \textsc{fastq} files.
-Numbers correspond to the proportion of bases/reads that are missed or gained after applying compression with respect to the uncompressed values.}
+\caption[]{Quality filtering/trimming for the uncompressed and
+  decompressed \textsc{fastq} files. Numbers correspond to the
+  proportion of bases/reads that are missed or gained after applying
+  compression with respect to the uncompressed values.}
 \label{tab:quality_control}
@@
 
-The primary concern of using lossy compression methods is naturally the extent of information loss, that we quantified by \textsc{mse} in this study.
-\textsc{mse} and compressibility provide information in the theoretical context to the methods, but they are not the end-all of evaluation criteria.
-The performance of compressed datasets in different subsequent analyses and applications are just as important.
-Our benchmarks showed that some of the compression methods with high error rates are still practical for certain kinds of applications.
-Many subsequent tools proved to have enough additional redundancy built-in to handle such loss in information.
-Passing the decompressed quality sequences through quality control software shows that most methods filter nearly as many bases as using original quality sequences.
-Assemblers performing sequencing alignment use percent similarity scores that are typically robust to standard sequencing errors.
+The primary concern of using lossy compression methods is naturally
+the extent of information loss, that we quantified by \textsc{mse} in
+this study. \textsc{mse} and compressibility provide information in
+the theoretical context to the methods, but they are not the end-all
+of evaluation criteria. The performance of compressed datasets in
+different subsequent analyses and applications are just as
+important. Our benchmarks showed that some of the compression methods
+with high error rates are still practical for certain kinds of
+applications. Many subsequent tools proved to have enough additional
+redundancy built-in to handle such loss in information. Passing the
+decompressed quality sequences through quality control software shows
+that most methods filter nearly as many bases as using original
+quality sequences. Assemblers performing sequencing alignment use
+percent similarity scores that are typically robust to standard
+sequencing errors.
 
@@
 
-Good/bad encoding has the nice property of being simple to compute and has a good bits per bp score.
-The good/bad encoding suffers from having a high \textsc{mse}, but fortunately, we have shown that in the case of genome assembly, good/bad encoding outperforms all polynomial regressions encodings with degree less than 3.
-Good/bad encoding of the fragment and short-jump libraries of \textit{Rhodobacter sphaeroides} have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-order polynomial regression encodings, respectively.
-This further highlights the importance of using additional contextual information of the subsequent analyses when evaluating compressed quality values.
+Good/bad encoding has the nice property of being simple to compute and
+has a good bits per bp score. The good/bad encoding suffers from
+having a high \textsc{mse}, but fortunately, we have shown that in the
+case of genome assembly, good/bad encoding outperforms all polynomial
+regressions encodings with degree less than 3. Good/bad encoding of
+the fragment and short-jump libraries of \textit{Rhodobacter
+  sphaeroides} have \textsc{mse}s of $2.42\times$ and $10.76\times$
+the 3rd-order polynomial regression encodings, respectively. This
+further highlights the importance of using additional contextual
+information of the subsequent analyses when evaluating compressed
+quality values.
 
-A potential extension to good/bad encoding is to incorporate an additional value \emph{okay}.
-The \emph{okay} value can be used where the base qualities fall within a good/bad range.
-Because the distribution of quality values is skewed towards higher quality, we need to experiment with different cutoffs for the \emph{okay} value and determine if the additional storage is noticeable in subsequent analyses.
+A potential extension to good/bad encoding is to incorporate an
+additional value \emph{okay}. The \emph{okay} value can be used where
+the base qualities fall within a good/bad range. Because the
+distribution of quality values is skewed towards higher quality, we
+need to experiment with different cutoffs for the \emph{okay} value
+and determine if the additional storage is noticeable in subsequent
+analyses.
 
@@
 
-Perhaps one of the greatest benefits of compressing quality values is the potential to perform quality control and possibly other operations directly on the compressed representations of the data.
-The tool \textsc{CutAdapt} trims reads that contains a given number of low quality bases from the ends and discards reads that are below a certain threshold.
-It is trivial to apply the same rules to a few of our compressed quality sequences.
-In the good/bad encoding, we store a single bit for each quality value.
-A read with stretches of 0-bits towards the ends can be trimmed and filtered if there are more 0-bits than 1-bits.
+Perhaps one of the greatest benefits of compressing quality values is
+the potential to perform quality control and possibly other operations
+directly on the compressed representations of the data. The tool
+\textsc{CutAdapt} trims reads that contains a given number of low
+quality bases from the ends and discards reads that are below a
+certain threshold. It is trivial to apply the same rules to a few of
+our compressed quality sequences. In the good/bad encoding, we store a
+single bit for each quality value. A read with stretches of 0-bits
+towards the ends can be trimmed and filtered if there are more 0-bits
+than 1-bits.
 
-In principle, we can quality filter the polynomial regression compression by interpolating where the polynomials are below a certain threshold and then filtering/trimming that part of the read.
-Unfortunately, storing the coefficients of the polynomials as a float leads to very poor compression.
+In principle, we can quality filter the polynomial regression
+compression by interpolating where the polynomials are below a certain
+threshold and then filtering/trimming that part of the
+read. Unfortunately, storing the coefficients of the polynomials as a
+float leads to very poor compression.
 
-A beneficial feature of the profile-based approach is that read quality filtering and/or trimming can be performed on the fly during decompression at virtually no computational cost.
-As all reads are classified into one of $K$ quality profiles, we can determine which of these profiles are of use for subsequent processes, and only decompress reads assigned to these profiles.
-Furthermore, it is possible to determine trimming points for each profile before extraction, that can be carried out on-line on reads during decompression.
+A beneficial feature of the profile-based approach is that read
+quality filtering and/or trimming can be performed on the fly during
+decompression at virtually no computational cost. As all reads are
+classified into one of $K$ quality profiles, we can determine which of
+these profiles are of use for subsequent processes, and only
+decompress reads assigned to these profiles. Furthermore, it is
+possible to determine trimming points for each profile before
+extraction, that can be carried out on-line on reads during
+decompression.
 
@@
 
-From an economics perspective, we can examine the fixed and marginal costs associated with our lossy compression techniques with respect to memory and computation.
-For our quality value binning, modeling-based, and image-based compression methods there are no explicit fixed costs in terms of memory, unlike the profile--based approach, which has a small fixed cost in storing the profiles.
-However, the profile-based approach has a much smaller marginal compared to the rest due to the fact that we are simply storing the profile index for each quality sequence.
+From an economics perspective, we can examine the fixed and marginal
+costs associated with our lossy compression techniques with respect to
+memory and computation. For our quality value binning and model-based
+compression methods there are no explicit fixed costs
+in terms of memory, unlike the profile--based approach, which has a
+small fixed cost in storing the profiles. However, the profile-based
+approach has a much smaller marginal compared to the rest due to the
+fact that we are simply storing the profile index for each quality
+sequence.
 
-Regarding computational costs, only quality binning and the profile--based methods have associated fixed costs.
-Before performing compression, we have to determine the optimal thresholding values for quality binning, and the cluster representatives for the profile--based method.
+Regarding computational costs, only quality binning and the
+profile--based methods have associated fixed costs. Before performing
+compression, we have to determine the optimal thresholding values for
+quality binning, and the cluster representatives for the
+profile--based method.
 
-In practice, the specific values of these costs are very sensitive to the implementation details.
+In practice, the specific values of these costs are very sensitive to
+the implementation details.
 
@@
 
-We have simply provided here the initial steps in analyzing the effect of lossy compression on quality sequences using a single, high-coverage bacterial data set.
-More work needs to be done using additional biological data sets, such as human and mouse, along with different sequencing technologies.
-A more direct comparison against related lossy compression tools, such as \textsc{SlimGene}~\cite{kozanitis2011compressing} and \textsc{QualComp}~\cite{ochoa2013qualcomp}, needs to be performed.
-Additionally, other types of sequencing data can be analyzed apart from the Illumina data examined here.
-For example, the PacBio sequencing instruments produce very long reads (with average read lengths on the order of 10 kbp), but with the trade-off of having a high error-rate ($\sim$15\%).
-Unlike the class of quality values we have examined here, the distribution of erroneous bases is relatively uniform~\cite{ferrarini2013evaluation}.
-Koren et al.~\cite{koren2013reducing} have shown that the assembly complexity of bacterial genomes can be greatly simplified, producing near complete genome assemblies, by utilizing a single run of these long reads.
-If long read sequencing technologies such as PacBio become more widely adopted, it would be of huge benefit to examine the potential of lossy compression algorithms on not only the quality values, but the biological sequencing data themselves.
+We have simply provided here the initial steps in analyzing the effect
+of lossy compression on quality sequences using a single,
+high-coverage bacterial data set. More work needs to be done using
+additional biological data sets, such as human and mouse, along with
+different sequencing technologies. A more direct comparison against
+related lossy compression tools, such as
+\textsc{SlimGene}~\cite{Kozanitis:2011kl} and
+\textsc{QualComp}~\cite{Ochoa:2013rt}, needs to be
+performed. Additionally, other types of sequencing data can be
+analyzed apart from the Illumina data examined here. For example, the
+PacBio sequencing instruments produce very long reads (with average
+read lengths on the order of 10 kbp), but with the trade-off of having
+a high error-rate ($\sim$15\%). Unlike the class of quality values we
+have examined here, the distribution of erroneous bases is relatively
+uniform~\cite{Ferrarini:2013vf}. Koren et al.~\cite{Koren:2013ye} have
+shown that the assembly complexity of bacterial genomes can be greatly
+simplified, producing near complete genome assemblies, by utilizing a
+single run of these long reads. If long read sequencing technologies
+such as PacBio become more widely adopted, it would be of huge benefit
+to examine the potential of lossy compression algorithms on not only
+the quality values, but the biological sequencing data themselves.
 
@@
 
-In this paper we have examined lossy compression on sequence quality values and their effect on subsequent analyses.
-Although most previous examinations on lossy compression primarily focused on information loss, we have shown that typically used bioinformatics software today have additional built-in sensitivity to handle significant loss of information in the compressed quality values.
+In this paper we have examined lossy compression on sequence quality
+values and their effect on subsequent analyses. Although most previous
+examinations on lossy compression primarily focused on information
+loss, we have shown that typically used bioinformatics software today
+have additional built-in sensitivity to handle significant loss of
+information in the compressed quality values.
 
 \section*{Acknowledgement}
-We thank the organizers and participants of the 2014 University of T\"{u}bingen and University of Maryland Bioinformatics Exchange for Students and Teachers (BEST) Summer School.
+We thank the organizers and participants of the 2014 University of
+T\"{u}bingen and University of Maryland Bioinformatics Exchange for
+Students and Teachers (BEST) Summer School.
+
+\section{Some Alternative Text}
+
+The following text is taken from a draft grant proposal, and the
+content overlaps considerably with the other text in this
+document. However, at least in some cases the wording of this
+alternative text might be preferable and serve as a replacement. In
+any event the entire draft manuscript has to be substantially revised.
+
+\subsection{Background}
+
+Sequencing read data constitutes the largest category of sequence data
+in genomics research because of its great redundancy, inclusion of
+quality values, and read-level naming and metadata. In recognition of
+this abundance and importance we place particular emphasis on
+compression of read data, which is the focus of this Aim. Among the
+several challenges for compression of read data is dealing with
+different error profiles resulting from the differences in underlying
+chemistries, signal detection and processing mechanisms, inherent
+biases, and other idiosyncratic properties of distinct high-throughput
+sequencing technologies. Our proposed research includes development of
+compression methods that robustly accommodate different error profiles
+associated with sequence read data, and where advantageous we will
+leverage error profiles to increase compression performance.
+
+At the level of sequence read the probability of error for each
+base-call is typically represented by \textsc{phred} quality value,
+which is defined as $Q =
+-10\,log_{10}P$~\cite{Ewing:1998zr,Ewing:1998ly}.  Depending on the
+sequencing technology these quality values can range from 0 to 93, and
+are represented with the \textsc{ascii} characters 33 to 126 (with
+some offset). Quality values comprise a standard component of
+\textsc{fastq} format files~\cite{Cock:2010ve}. The characteristics of
+different sequencing technology-associated error profiles are
+reflected in the quality values, and can be broadly described for both
+Illumina and PacBio sequencing technologies.
+
+Illumina reads are characterized by having far fewer insertion and
+deletion errors, where substitution (miscall) errors are much more
+frequent and have context-specific patterns. The errors and quality
+values are non-uniformly distributed over the read length (e.g., error
+rates increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and
+32.8 -- 67.9\% of reads have low quality tails at the 3$^{\prime}$
+end~\cite{Minoche:2011km}). Illumina reads have a single quality value
+per base-call.
+
+\subsection{Lossy compression of quality values} \label{sec:qual}
+
+\subsubsection{Pilot Study: profile-based compression of quality 
+values}\label{sec:profile-based}
+
+As described previously, quality values in Illumina reads are
+non-uniformly distributed over the read length and show a general
+pattern of decreasing toward the 3$^{\prime}$ end. As large sets of
+quality strings show similar trends of quality over their sequence,
+one approach would be to identify common patterns in the data,
+generate profiles representing these patterns, and use the profiles as
+proxies for read quality value sequences. We demonstrated the efficacy
+of this approach in a pilot study using test data of Illumina reads
+the \textit{Rhodobacter sphaeroides} 2.4.1 data set provided by
+\textsc{gage} (Genome Assembly Gold-Standard
+Evaluations)~\cite{Salzberg:2012rc}, which consists of 2,050,868
+paired-end reads (fragment library; insert size of 180 bp) and
+2,050,868 reads (short-jump library; insert size of 3,000 bp).
+
+Quality value patterns can be readily determined by clustering
+sequences of quality values and using the resulting cluster centers as
+representative profiles. In our pilot study we used $K$-means
+clustering, a vector quantization method that partitions a set of
+samples into $K$ sets that minimize within-cluster sum of
+squares~\cite{hastie_09_elements-of.statistical-learning}, as follows.
+
+\begin{enumerate}
+\item
+Randomly sample an arbitrary number of quality sequences as a training
+set.
+\item
+Perform $K$-means clustering on the sample, and use the cluster
+centers as profiles for the dataset.
+\item
+Replace each quality value sequence with the index value of the
+profile closest in Euclidean distance.
+\end{enumerate}
+
+Our pilot study results demonstrate that $K$ values as small as
+32--256 sufficiently represent patterns in Illumina read quality
+values such that performance in subsequent analyses is very close to
+uncompressed data (see Section \ref{sec:pilot-evaluation}). The
+resulting compressed quality component of the original set of
+\textsc{fastq} files consists of the $K$ quality profiles (the fixed
+size cost), and a single byte representing the assigned quality
+profile index for each read (the marginal size cost). Therefore the
+compression factor is proportional of read length: the marginal size
+cost of 1 byte/read for Illumina HiSeq read lengths of 100 bp requires
+0.01 bits/base-call, and MiSeq read lengths of 300 bp require 0.0033
+bits/base-call. The fixed cost of profiles of course reduces the
+compression ratio, but this fixed cost is amortized across reads in a
+set. Both fixed and marginal costs can be further reduced with
+subsequent compression of the index and profiles (e.g., using general
+compression algorithms). Our pilot study demonstrates that using 256
+profiles with subsequent bzip2 compression (using a Burrows-Wheeler
+block-sorting compression algorithm, and Huffman coding) represents
+quality values of Illumina 100 bp reads at 0.074 bits/base-call
+compared to the original data compressed with bzip2 only of 2.11
+bits/base-call for the fragment library and 2.70 bits/base-call for
+the short jump library. The very recently published
+Read-Quality-Sparsifier
+(\textsc{rqs};~\cite{DBLP:conf/recomb/YuYB14}), which apparently
+represents the current state-of-the-art for compressing quality
+values, achieved best case compression of 0.254 bits/base-call, and a
+mean compression of 1.841 bits/base-call. Whereas our profile method
+achieves $\sim$3.432$\times$ more compression than textsc{rqs} best
+performance, and $\sim$24.878$\times$ more than average
+performance. Note too that the compression ratio for \textsc{rqs}
+varies with the $k$-mer composition of the data, whereas profile-based
+compression is invariant to $k$-mer composition.
+
+\subsubsection{Evaluating compression effects on subsequent 
+analysis results}\label{sec:pilot-evaluation}
+
+Lossy compression compromises representation accuracy for better space
+efficiency. Different lossy compression methods have different
+properties, therefore we need to evaluate performance in a number of
+analysis contexts, and use multiple performance measures to
+appropriately evaluate lossy compression methods. Important
+bioinformatics contexts include typical tasks, such as quality control
+processing, \textit{de novo} genome assembly, and aligning reads to a
+reference. In our pilot study we again used Illumina reads from
+fragment and short-jump libraries of the \textit{Rhodobacter
+  sphaeroides} as test data, and assessed the effects of lossy
+compression on each of the aforementioned bioinformatics
+operations. The results for the profile-based compression follow.
+
+\paragraph{Read quality processing}
+Among the most fundamental uses of sequence quality values is as part
+of the quality assessment and quality control (\textsc{qa/qc})
+processes prior to subsequent analysis steps. Quality control based on
+quality values generally includes two operations:
+\textit{i}.~filtering, the elimination of reads that on the whole do
+not meet quality standards, which reduces the total number of reads;
+and \textit{ii}.~trimming of low quality base-calls from reads, which
+reduces the number total number of bases.
+
+To evaluate the effect of compression on these quality control
+operations, we used \textsc{autoadapt}, which uses
+\textsc{FastQC}~\cite{citeulike:11583827} and
+\textsc{cutadapt}~\cite{EJ200} packages, all with the default
+settings. \textsc{PrinSeq}~\cite{Schmieder:2011gd} was then used to
+count the number of bases and reads in each quality control-processed
+\textsc{fastq} file. The proportion of reads retained using
+compressed/decompressed data compared to that of the uncompressed
+data: 0.998 for the fragment library reads, and 1.013 for the
+short-jump library reads; and proportion of bases retained 1.0 for the
+fragment library reads, and 1.027 for the short-jump library
+reads. Thus there is very little difference between profile-based
+compressed data compared to that of the original uncompressed data in
+read filtering and trimming.
+
+\paragraph{\textit{De novo} genome assembly}
+For de novo genome assembly, we use
+\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
+settings. We evaluate the assembly based on contiguity statistics, log
+average read probability (\textsc{lap})~\cite{Ghodsi:2013hb}, and a
+collection of reference-based metrics, of which the contiguity
+statistics include number of scaffolds and NG50; the latter statistic
+is defined as the weighted median contig size (the length of largest
+contig $c$ such that the total size of the contigs larger than $c$
+exceeds half of the known genome size). The \textsc{lap} score can be
+viewed as a log likelihood score, where a larger value is better. We
+use a reference-based evaluation script provided by \textsc{gage},
+which counts single nucleotide polymorphisms (\textsc{snp}s),
+relocations, translations, and inversions. The reference- based
+metrics are normalized by the length of the assembly to aid in
+comparison. The differences in assembly statistics between using the
+profile-based compressed/decompressed data compared to using
+uncompressed data are extremely small: total assembly size difference
+of 6.745 kbp ($\sim$0.00015); NG50 size difference of 2.259 kbp
+(0.0007); an \textsc{lap} score difference of 0.002 for the fragment
+library and 0.04 for the short-jump library.
+
+\paragraph{Aligning reads to a reference sequence}
+Another important analysis involving read data are the alignments of
+reads to a reference sequence. Such alignments are performed for
+genome annotation, quantification of gene expression in RNA-Seq
+studies, estimation of sequencing coverage, and other purposes. To
+evaluate the effect of compression on read alignment we used
+\textsc{bowtie} 2~\cite{Langmead:2012rw} with default settings and
+max- and min-mismatch parameters of \texttt{-mp 6,6} and \texttt{-mp
+  2,2} respectively. The proportion of reads mapped using
+compressed/decompressed data was extremely close to that of the
+uncompressed data: 0.999 for the fragment library reads, and 0.998 for
+the short-jump library reads.
 
@@
 %
-\bibliography{Document}
+\bibliography{compression}
 

### 2015-02-02 manuscript/q-compression.tex
Added another review reference, corrected several spelling mistakes in text from original course report.

@@
 Lossy compression of quality information can greatly decrease storage
-and memory requirements with little discernable effects on subsequent
+and memory requirements with little descernable effects on subsequent
 analysis results. The four compression strategies in this study were
@@
 largely focused on lossless compression of base
-calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Giancarlo:2014rw}.
+calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
 If a reference exists for the biological data, sequences can be
@@
 study, we described a way of reducing the problem of quality
-compression to image compression via blockwise quality data
+compression to image compression via block-wise quality data
 transformation. This approach is different from the others in that
@@
 provides. Therefore the compression ratio and bits per bp values for
-polynomial regression throughout this manuscript refer to bzipped
-polynomially approximated quality values.
+polynomial regression throughout this manuscript refer to 
+polynomially approximated quality values compressed with zip.
 
@@
 the reference genome, outperforming the assembly using the original
-quality sequences. A more indepth evaulation is needed to determine
+quality sequences. A more in-depth evaluation is needed to determine
 whether these compression methods are missing actual SNPs.
@@
 uses roughly $1.2\%$ more of the total read set than the 4th-order
-(Supplemental Table 1), which may explain the discrepency in LAP
+(Supplemental Table 1), which may explain the descrepency in LAP
 values because the LAP value is influenced by total

### 2015-02-03 manuscript/q-compression.tex
Revised text, including removal of text regarding image compression.

@@
 profile.
-
-The last approach we investigated is based on transforming the
-original data into a wholly different representation for which
-efficient and proven compression algorithms already exist. In this
-study, we described a way of reducing the problem of quality
-compression to image compression via block-wise quality data
-transformation. This approach is different from the others in that
-depending on the image compression algorithm used it is not restricted
-to lossy techniques.
 
@@
 requires 1 \emph{byte} per basepair because they are stored in
-\textsc{ascii} format; however, losslessly compressing the values with
+\textsc{ascii} format; however, lossless compression of the values with
 BZip2 2.112 bits per basepair. 0-order polynomial regression and
 32-profile encodings have the lowest bits per bp. 5th-order polynomial
-regression and \textsc{png} encodings have the highest bits per bp,
-but have among the lowest \textsc{mse}. Despite having similar bits
-per bp, \textsc{png} encoding has roughly one third the \textsc{mse}
-as 5th-order polynomial (3.07 vs 10.27, respectively).
-
-As the order of the polynomial increases, the bits per bp increase and
+regression has the highest bits per bp,
+but have among the lowest \textsc{mse}. As the order of the polynomial increases, the bits per bp increase and
 the \textsc{mse} decreases at an exponential rate.
@@
   Quality profile compression with 32 and 256 profiles. R0-R5:
-  Polynomial regression-based compression using 0-5 degrees. PNG:
-  Blockwise image compres the generated SAM file for each compressing
-  approach weresion algorithm. Dashed line corresponds to lossless
+  Polynomial regression-based compression using 0-5 degrees. Dashed line corresponds to lossless
   bzip compression of quality values. \em}\label{fig:mse_vs_bpbp_frag}
@@
   Quality profile compression with 32 and 256 profiles. R0-R5:
-  Polynomial regression-based compression using 0-5 degrees. PNG:
-  Blockwise image compression algorithm. Dashed line corresponds to
+  Polynomial regression-based compression using 0-5 degrees. Dashed line corresponds to
   lossless bzip compression of quality
@@
 using good/bad encoding, 0--5th-order polynomial regression encodings,
-32- and 256-profile encodings, and \textsc{png} encoding. We evaluate
+and 32- and 256-profile encodings. We evaluate
 the assembly based on contiguity statistics, log average read
@@
 inversions (Table \ref{tab:assembly_ranks}). Among the compression
-methods, the \textsc{png} encoding had the highest rank, followed by
-256- and 32-profile encoding, then the 4th-order polynomial
+methods, the 
+256- and 32-profile encoding had the highest rank, then the 4th-order polynomial
 regression, the 5th-order polynomial regression, the good/bad
@@
 noticeably in the amount of SNPs. The order of polynomial has an
-inverse relationship with the amount of SNPs detected. The good/bad,
-profile and PNG methods detected the least amount of SNPs compared to
+inverse relationship with the amount of SNPs detected. The good/bad and
+profile methods detected the least amount of SNPs compared to
 the reference genome, outperforming the assembly using the original
@@
 None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
-PNG                    & 2    & 4    & 1          & 2           & 2              & 3      & 4          & 2                    \\
 Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
@@
 mapped reads of 100\%, within which 99.66\% are common and only 0.33\%
-unique mapped reads in the decompressed file. In both samples,
-\textsc{png} compression has the least number of mapped reads that are
-unique in the decompressed files (only 0.013\% and 0.89\% in the
-fragments and short-jump samples, respectively).
+unique mapped reads in the decompressed file.
 
@@
 	&Profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\
-	&PNG & 99.82 & 99.81 & 0.19 & 0.01 \\
 
@@
 	&Profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
-	&PNG & 99.71 & 99.62 & 0.38 & 0.09 \\ \hline
 \end{tabular}
@@
 	Profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
-	PNG & 99.62 & 99.75 & 99.84 & 99 \\ \hline
 \end{tabular}

### 2015-02-03 src/compress.py
Added sickle evaluation.

@@ -101,8 +101,6 @@ def calc_mean_squared_error(options):
     std_err_file = open('mse.log', 'w')
 
     for compression_method in options.compressed_dirs:
-        #if compression_method == 'original':
-        #    continue
 
         for reads_filename in options.reads_filenames:
             MSE_CMD = "python src/evaluate_loss.py " + options.output_dir + '/original/' + os.path.basename(reads_filename) + '.quals' + '\t' + \
@@ -116,7 +114,48 @@ def quality_preprocessing(options):
     """
     Examine the effects of lossy compression on quality preprocessing tools.
     """
-    pass
+    std_err_file = open('preprocessing.log', 'w')
+
+    SICKLE_CMD = "sickle se -f [READ] -t sanger -o [OUTPUT]"
+
+    for compression_method in options.compressed_dirs:
+
+        for reads_filename in options.reads_filenames:
+            output_filename = options.output_dir + '/preprocessing/' + compression_method + '/' + os.path.basename(reads_filename)
+            ensure_dir(output_filename)
+
+            stats_file = open(options.output_dir + '/preprocessing/' + compression_method + '/' + os.path.basename(reads_filename) + '.stats', 'w')
+
+            call_arr = SICKLE_CMD.replace('[READ]', options.output_dir + '/' + compression_method + '/' + reads_filename).replace('[OUTPUT]', output_filename).split()
+            out_cmd("", stats_file.name, call_arr)
+            call(call_arr, stdout=stats_file)
+
+            # Process the stats file to get the relevant information.
+            """Output stats file is in the format of:
+
+            SE input file: tmp/goodbad/frag_small_2.fastq
+
+            Total FastQ records: 200
+            FastQ records kept: 138
+            FastQ records discarded: 62
+            """
+
+            # Parse the above to get how many records were kept.
+            for line in open(stats_file.name, 'r').readlines():
+                if line.startswith('FastQ records kept:'):
+                    records_kept = line.strip().split()[3]
+                    open(stats_file.name + '.records_kept', 'w').write(records_kept + '\n')
+
+            # Find out many bases were kept.
+            line_number = 1
+            bases = 0
+            for line in open(output_filename, 'r'):
+                if (line_number % 4) == 2:
+                    bases += len(line.strip())
+                line_number += 1
+
+            open(stats_file.name + '.bases', 'w').write(str(bases) + '\n')
+
 
[…]

### 2015-02-03 src/compress.py
Fixed bowtie2 pathname issue.

@@ -126,7 +126,7 @@ def quality_preprocessing(options):
 
             stats_file = open(options.output_dir + '/preprocessing/' + compression_method + '/' + os.path.basename(reads_filename) + '.stats', 'w')
 
-            call_arr = SICKLE_CMD.replace('[READ]', options.output_dir + '/' + compression_method + '/' + reads_filename).replace('[OUTPUT]', output_filename).split()
+            call_arr = SICKLE_CMD.replace('[READ]', options.output_dir + '/' + compression_method + '/' + os.path.basename(reads_filename)).replace('[OUTPUT]', output_filename).split()
             out_cmd("", stats_file.name, call_arr)
             call(call_arr, stdout=stats_file)
 
@@ -201,7 +201,7 @@ Illumina_02,\tassembly,\tunknown,\tjumping,\t1,\t,\t,\t3000,\t500,\toutward,\t,\
         call(call_arr, stdout=ploidy_file, stderr=std_err_file)
 
         # Run AllpathsLG
-        ALLPATHS_CMD = "RunAllPathsLG PRE=" + os.path.abspath(options.output_dir) + '/assemble/' + compression_method + " DATA_SUBDIR=. RUN=allpaths SUBDIR=run THREADS=32 OVERWRITE=True REFERENCE_NAME=."
+        ALLPATHS_CMD = "RunAllPathsLG PRE=" + os.path.abspath(options.output_dir) + '/assemble/' + compression_method + " DATA_SUBDIR=. RUN=allpaths SUBDIR=run THREADS=" + options.threads + " OVERWRITE=True REFERENCE_NAME=."
         call_arr = ALLPATHS_CMD.split()
         out_cmd("", std_err_file.name, call_arr)
         call(call_arr, stderr=std_err_file)
@@ -224,7 +224,7 @@ def align_reads(options):
     call(call_arr, stderr=std_err_file)
 
     # Align the reads.
-    BOWTIE2_CMD = "bowtie2 -x " + options.output_dir + "/align/reference -U [READ] "
+    BOWTIE2_CMD = "bowtie2 -x " + options.output_dir + "/align/reference -p " + options.threads + "-U [READ] "
 
     for compression_method in options.compressed_dirs:
         for reads_filename in options.reads_filenames:
@@ -233,7 +233,7 @@ def align_reads(options):
             ensure_dir(alignment_filename)
             alignment_file = open(alignment_filename, 'w')
 
[…]

### 2015-02-03 manuscript/q-compression.tex
Lots of changes: new bib entry for Sickle, many text revisions, good/bad is now 2-bin (to emphasize the concept of binning), revised introduction (though it is not yet done).

@@
 Lossy compression of quality information can greatly decrease storage
-and memory requirements with little descernable effects on subsequent
+and memory requirements with little discernable effects on subsequent
 analysis results. The four compression strategies in this study were
@@
 
-The \emph{de facto} standard for storing next-generation sequencing
-data is the \textsc{fastq} format. A component of \textsc{fastq}
-format files~\cite{Cock:2010ve} is information for a quantitative
-measure of quality for each nucleotide based on the \textsc{phred}
-quality value, which is defined as $Q =
--10\ log_{10}\/P$~\cite{Ewing:1998zr,Ewing:1998ly}. Depending on the
-sequencing technology these quality values can range from 0 to 93, and
-the American Standard Code for Information Interchange \textsc{ascii}
-characters 33 to 126 (with some offset) are used to represent quality
-values.
+Sequencing read data constitutes the largest category of sequence data
+in genomics research because of great redundancy, inclusion of quality
+values, and read-level naming and metadata. Because of this abundance
+effective compression of read data has the potential for substantial
+improvement in data storage and transfer efficiency.  Among the
+several challenges for compression of read data is dealing with
+different error profiles resulting from differences in underlying
+chemistries, signal detection and processing mechanisms, inherent
+biases, and other idiosyncratic properties of distinct high-throughput
+sequencing technologies. Illumina sequencing reads, the focus of this
+research, are characterized by having relatively few insertion and
+deletion errors, but substitution (miscall) errors are much more
+frequent and have context-specific patterns. These errors are
+non-uniformly distributed over the read length (e.g., error rates
+increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
+67.9\% of reads have low quality tails at the 3$^{\prime}$
+end~\cite{Minoche:2011km}).
+
+Quality values comprise a standard component of \textsc{fastq} format
+files~\cite{Cock:2010ve}.  At the level of sequence read the
+probability of error for each base-call is typically represented by
+\textsc{phred} quality value, which is defined as $Q =
+-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
+technology these quality values can range from 0 to 93, and are
+represented with the \textsc{ascii} characters 33 to 126 (with some
+offset). These errors are represented a single quality value per
+base-call for Illumina sequencing reads.
 
 Quality values are typically used throughout bioinformatics pipelines.
-Certain tools, such as \textsc{fastqc}~\cite{citeulike:11583827} and
+Software such as \textsc{fastqc}~\cite{citeulike:11583827} and
 Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
-descriptions of the distribution of sequence quality scores. These
-analyses can then be used to filter and trim the sequencing
-data~\cite{EJ200}. The processed sequencing data can then be be used
-in subsequent analyses. Certain genome assemblers utilize quality
-values directly to produce better
-assemblies~\cite{Gnerre:2011kx}. Genome assemblers work by piecing
-together overlapping sequences, constructing a larger resulting
-sequence. Short-read alignment tools, such as Bowtie
-2~\cite{Langmead:2012rw}, use quality values to scale the mismatch
-penalty in a way that mismatches between poor quality bases of the
-query and the reference are weighted less than mismatches occurring on
-high quality bases. Similar to alignment, software for detecting
-single nucleotide polymorphisms (\textsc{snp}s) can also rely on
-quality values~\cite{McKenna:2010bh}. Detected \textsc{snp}s with
-high-quality bases are more trustworthy than those with low-quality
-bases, particularly in low coverage regions.
+descriptions of the distribution of sequence quality scores, and these
+scores can be used to guide filtering and trimming of sequencing read
+data as processing step before subsequent analyses. Quality values can
+be used by genome assembly software to produce better
+assemblies~\cite[e.g.,][]{Gnerre:2011kx}.  Short-read alignment
+software, such as Bowtie 2~\cite{Langmead:2012rw}, use quality values
+to scale the mismatch penalty in a way that mismatches between poor
+quality bases of the query and the reference are weighted less than
+mismatches occurring on high quality bases. Similar to alignment,
+software for detecting single nucleotide polymorphisms (\textsc{snp}s)
+can use quality values~\cite[e.g.,][]{McKenna:2010bh}, and identified
+\textsc{snp}s with high-quality calls are deemed more reliable than
+those with low-quality calls, particularly in low coverage regions.
 
@@
 calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
-If a reference exists for the biological data, sequences can be
-compressed using offsets into the
-reference~\cite{Kozanitis:2011kl}. Typical compression methods fail to
-utilize the biases inherent in existing sequencing
-technologies. Depending on the sequencing technology used, the base
-calling is more accurate in the earlier cycles, gradually losing
-quality in later cycles~\cite{Kozanitis:2011kl}. To this end, methods
-for compressing these quality sequences utilize these general quality
+Typical compression methods fail to utilize the biases inherent in
+existing sequencing technologies. Depending on the sequencing
+technology used, the base calling is more accurate in the earlier
+cycles, gradually losing quality in later
+cycles~\cite{Kozanitis:2011kl}. To this end, methods for compressing
+these quality sequences utilize these general quality
 trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
@@
 
-We examine four general classes of lossy compression methods: score
-binning, model-based, profile-based, and data type transformation.
+We examine three general classes of lossy compression methods:
+binning, model-based, and profile-based.
 
@@
 
-\subsection{Quality value binning}
+\subsection{Binning}
 
-%The quality value provides some measure of the accuracy of a given
-%base call. A strategy to reduce the dimensionality of these quality
-%values is to aggregate adjacent quality values together into the same
-%bin. Taken to the extreme, a base call can be considered either good
-%or bad. This is the motivation of our approach, deemed good/bad
-%encoding, where we observe the distribution of quality values across
-%the read set and find a cutoff between good and bad bases.
-
-In one of the more extreme cases of quality value binning, we can bin
-a quality value into one of two categories: good or bad. We define
-good/bad encoding as finding the quality value cutoff to separate
-between good and bad bases. First, we learn the distribution of
-quality values across all sequences. Using this distribution, bases
-are marked bad if their quality value falls below a certain
-threshold. We set the cutoff as the first quartile minus 1.5 * the
-interquartile range (IQR), which is the difference between the first
-and third quartile. 1.5 * IQR is the value used by Tukey's box
+Quality values can be binned into a set of bins, and any number of
+bins less than the the number of distinct quality values is
+possible. The minimum number of bins that allows for a any distinction
+among quality values is two, i.e., two categories ``good'' and ``bad''
+quality. We implement 2-bin encoding by setting a quality value
+threshold empirically determined by the distribution of quality values
+across reads. Base-calls are marked ``bad'' if their quality value
+falls below the first quartile minus 1.5 $\times$ the interquartile
+range (IQR), which is the difference between the first and third
+quartile. 1.5 $\times$ IQR is the value used by Tukey's box
 plot~\cite{mcgill1978variations}. The main benefit of this approach is
@@
 options to preserve more of the defining characteristics of the
-quality values, but we only analyze good/bad encoding in our results.
+quality values, but we only analyze 2-bin encoding in our results.
 
@@
 highly compressible simply due to the smoothing effect it
-provides. Therefore the compression ratio and bits per bp values for
-polynomial regression throughout this manuscript refer to 
+provides. Therefore the compression ratio and bits/base-call values
+for polynomial regression throughout this manuscript refer to
 polynomially approximated quality values compressed with zip.
@@
 
-K-means clustering is a vector quantization method, partitioning a set
-of samples into K sets that minimize within-cluster sum of
+$k$-means clustering is a vector quantization method, partitioning a set
+of samples into $k$ sets that minimize within-cluster sum of
 squares~\cite{macqueen1967some}. Using a random subset of read quality
@@
 are represented by vectors containing their \textsc{phred} scores
-corresponding to each position along the read. Subsequently, K-means
+corresponding to each position along the read. Subsequently, $k$-means
 clustering is performed on the training set until convergence. The
@@
 
-Once the K quality profiles are determined, all quality sequences are
-passed through the trained K-means predictor, with the nearest quality
+Once the $k$ quality profiles are determined, all quality sequences are
+passed through the trained $k$-means predictor, with the nearest quality
 profile in Euclidean space being assigned to every quality sequence as
@@
 The compressed quality file therefore consists of an index enumerating
-the K quality profiles, and a binary part containing the assigned
+the $k$ quality profiles, and a binary part containing the assigned
 quality profile index for each read.
@@
 \centerline{\includegraphics[width=3.35in]{profiles_128.png}}
-\caption{Quality profiles obtained by K-means clustering on the
+\caption{Quality profiles obtained by $k$-means clustering on the
   fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
-  set using K=128, with each row corresponding to a quality
+  set using $k$ = 128, with each row corresponding to a quality
   profile. Dark to light colors represent low to high quality
@@
 
-For our analyses, we use sequencing data provided by the \textsc{GAGE}
-project~\cite{Salzberg:2012rc}. The read data for \textit{Rhodobacter
-  sphaeroides} 2.4.1 was downloaded from
-\href{http://gage.cbcb.umd.edu/data/Rhodobacter_sphaeroides}{http://gage.cbcb.umd.edu/data/Rhodobacter\_sphaeroides},
-and the corresponding reference sequence was obtained from the NCBI
-RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1,
-NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1). The
-\textit{Rhodobacter sphaeroides} 2.4.1 data set consists of 2,050,868
-paired-end reads (insert size of 180 bp) and 2,050,868 short-jump
-reads (insert size of 3,000 bp).
+We used several Illumina sequence read data sets in this research as
+follows.
+
+\textit{Rhodobacter sphaeroides} 2.4.1 data from the \textsc{gage}
+(Genome Assembly Gold-Standard Evaluations)~\cite{Salzberg:2012rc},
+which are generated from a fragment library (insert size of 180 nt;
+2,050,868 paired-end reads) and short-jump library (insert size of
+3,000 nt; 2,050,868 reads). The corresponding reference sequence was
+obtained from the NCBI RefSeq database (NC\_007488.1, NC\_007489.1,
+NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
+
+\textit{Homo sapiens} \ldots
+
+\textit{Genus speices} \ldots
 
@@
 
-For quality control, we use Autoadapt tool which uses
-\textsc{fastqc}~\cite{citeulike:11583827} and
-\textsc{CutAdapt}~\cite{EJ200} packages with the default
-settings. PrinSeq~\cite{Schmieder:2011gd} was used to count the number
-of bases and reads in each \textsc{fastq} file.
+For quality control, we use
+Sickle~\cite{sickle}. PrinSeq~\cite{Schmieder:2011gd} was used to
+count the number of bases and reads in each \textsc{fastq} file.
 
@@
 
-For aligning the sequences, we use Bowtie 2~\cite{Langmead:2012rw}
-with default settings and modified max- and min-mismatch parameters
+For aligning the sequences, we use Bowtie2~\cite{Langmead:2012rw} with
+default settings and modified max- and min-mismatch parameters
 detailed below.
@@
 between its physical compression and the loss in its accuracy. Here,
-we use average \emph{bits per bp} to measure the physical compression
-of the data and mean squared error (MSE) to measure the loss in
-quality. Bits per bp is calculated simply as the compressed file size
-(in bits) divided by the total number of quality values. MSE is
-calculated by $\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is
-the number of sequences, $Q_i$ is the actual quality value associated
-with sequence position $i$, and $Q_i'$ is the corresponding compressed
+we use mean bits/base-call to measure the physical compression of the
+data and mean squared error (\textsc{mse}) to measure the loss in
+quality. Bits/base-call is calculated simply as the compressed file
+size (in bits) divided by the total number of quality
+values. \textsc{mse} is calculated as
+$\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is the number of
+sequences, $Q_i$ is the original quality value associated with
+sequence position $i$, and $Q_i'$ is the corresponding compressed
 quality value.
 
-We compare the \textsc{mse} versus \emph{bits per bp} of the fragment
-and short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
+We compare the \textsc{mse} versus bits/base-call of the fragment and
+short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
 data set (Figures \ref{fig:mse_vs_bpbp_frag} and
 \ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values
-requires 1 \emph{byte} per basepair because they are stored in
-\textsc{ascii} format; however, lossless compression of the values with
-BZip2 2.112 bits per basepair. 0-order polynomial regression and
-32-profile encodings have the lowest bits per bp. 5th-order polynomial
-regression has the highest bits per bp,
-but have among the lowest \textsc{mse}. As the order of the polynomial increases, the bits per bp increase and
-the \textsc{mse} decreases at an exponential rate.
+requires 1 byte per base-call because they are stored in
+\textsc{ascii} format; however, lossless compression of the values
+with BZip2 2.112 bits/base-call. 0-order polynomial regression and
+32-profile encodings have the lowest bits/base-call. 5th-order
+polynomial regression has the highest bits/base-call, but have among
+the lowest \textsc{mse}. As the order of the polynomial increases, the
+bits/base-call increase and the \textsc{mse} decreases at an
+exponential rate.
 
@@
 \centerline{\includegraphics[width=3.65in]{mse_frag.png}}
-\caption{MSE versus bits per bp for different compression methods
-  applied to the fragment library from \textit{Rhodobacter
-    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
-  Quality profile compression with 32 and 256 profiles. R0-R5:
-  Polynomial regression-based compression using 0-5 degrees. Dashed line corresponds to lossless
-  bzip compression of quality values. \em}\label{fig:mse_vs_bpbp_frag}
+\caption{Mean squared error versus bits/base-call for different
+  compression methods applied to the fragment library from
+  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
+  encoding. P32 and P256: Quality profile compression with 32 and 256
+  profiles. R0-R5: Polynomial regression-based compression using 0-5
+  degrees. Dashed line corresponds to lossless bzip compression of
+  quality values. \em}\label{fig:mse_vs_bpbp_frag}
 \end{figure}
@@
 \centerline{\includegraphics[width=3.65in]{mse_short.png}}
-\caption{MSE versus bits per bp for for different compression methods
-  applied to the short-jump library from \textit{Rhodobacter
-    sphaeroides} 2.4.1 data set. G/B: Good/Bad encoding. P32 and P256:
-  Quality profile compression with 32 and 256 profiles. R0-R5:
-  Polynomial regression-based compression using 0-5 degrees. Dashed line corresponds to
-  lossless bzip compression of quality
-  values. \em}\label{fig:mse_vs_bpbp_jump}
+\caption{Mean squared error versus bits/base-call for for different
+  compression methods applied to the short-jump library from
+  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
+  encoding. P32 and P256: Quality profile compression with 32 and 256
+  profiles. R0-R5: Polynomial regression-based compression using 0-5
+  degrees. Dashed line corresponds to lossless bzip compression of
+  quality values. \em}\label{fig:mse_vs_bpbp_jump}
 \end{figure}
@@
 compare the number of reads and bases with those of the uncompressed
-\textsc{fastq} files. Autoadapt is a tool that uses \textsc{fastqc}
-and \textsc{Cutadapt} to filter and trim reads. We used PrinSeq to
-calculate the number of reads and bases in each sample. Table
-\ref{tab:quality_control} shows the proportional numbers with respect
-to those of the uncompressed \textsc{fastq} files. In the fragment
-sample, we retain on average 95.4\% and 94.24\% of bases and reads,
-respectively. In the short-jump sample, we retain 90.4\% and 90.98\%
-of bases and reads, respectively. In some cases, the proportion of
-reads and/or bases compared to the original is more than 100\%; this
-means that the number of the reads and bases in the compressed samples
-are more because altering quality values may prevent the read from
-being filtered and/or trimmed. This is clear in the good/ bad
-compression approach of the short-jump sample, where the proportion of
-bases and reads are 113.09\% and 103.70\% respectively.
+\textsc{fastq} files. Sickle is a tool to filter and trim reads. We
+used PrinSeq to calculate the number of reads and bases in each
+sample. Table \ref{tab:quality_control} shows the proportional numbers
+with respect to those of the uncompressed \textsc{fastq} files. In the
+fragment sample, we retain on average 95.4\% and 94.24\% of bases and
+reads, respectively. In the short-jump sample, we retain 90.4\% and
+90.98\% of bases and reads, respectively. In some cases, the
+proportion of reads and/or bases compared to the original is more than
+100\%; this means that the number of the reads and bases in the
+compressed samples are more because altering quality values may
+prevent the read from being filtered and/or trimmed. This is clear in
+the good/ bad compression approach of the short-jump sample, where the
+proportion of bases and reads are 113.09\% and 103.70\% respectively.
 
@@
 assemble the original data along with the quality compressed versions
-using good/bad encoding, 0--5th-order polynomial regression encodings,
-and 32- and 256-profile encodings. We evaluate
-the assembly based on contiguity statistics, log average read
-probability (LAP~\cite{Ghodsi:2013hb}), and a collection of
+using 2-bin encoding, 0--5th-order polynomial regression encodings,
+and 32- and 256-profile encodings. We evaluate the assembly based on
+contiguity statistics, log average read probability
+(\textsc{lap}~\cite{Ghodsi:2013hb}), and a collection of
 reference-based metrics. The contiguity statistics include number of
@@
 contigs larger than $c$â exceeds half of the known genome size). The
-LAP score can be viewed as a log likelihood score, where a value
-closer to 0 is better. We use a reference-based evaluation script
-provided by \textsc{gage} which counts single nucleotide polymorphisms
-(\textsc{snp}s), relocations, translations, and inversions. The
-reference-based metrics are normalized by the length of the assembly
-to aid in comparison. Alongside the raw metrics, we also provide a
-cummulative sum of the individual ranks of each method for each
-metric. A similar approach was utilized by the \textsc{Assemblathon1}
-competition~\cite{Earl:2011fv}.
+\textsc{lap} score can be viewed as a log likelihood score, where a
+value closer to 0 is better. We use a reference-based evaluation
+script provided by \textsc{gage} which counts single nucleotide
+polymorphisms (\textsc{snp}s), relocations, translations, and
+inversions. The reference-based metrics are normalized by the length
+of the assembly to aid in comparison. Alongside the raw metrics, we
+also provide a cummulative sum of the individual ranks of each method
+for each metric. A similar approach was utilized by the
+\textsc{Assemblathon1} competition~\cite{Earl:2011fv}.
 
 The assembly using the uncompressed reads outperforms all compression
-methods in terms of LAP, NG50, relocations, translations, and
+methods in terms of \textsc{lap}, NG50, relocations, translations, and
 inversions (Table \ref{tab:assembly_ranks}). Among the compression
-methods, the 
-256- and 32-profile encoding had the highest rank, then the 4th-order polynomial
-regression, the 5th-order polynomial regression, the good/bad
-encoding, and lastly, the 3rd through 0-order polynomials.
+methods, the 256- and 32-profile encoding had the highest rank, then
+the 4th-order polynomial regression, the 5th-order polynomial
+regression, the 2-bin encoding, and lastly, the 3rd through 0-order
+polynomials.
 
@@
 similar contiguity statistics, the different compression methods vary
-noticeably in the amount of SNPs. The order of polynomial has an
-inverse relationship with the amount of SNPs detected. The good/bad and
-profile methods detected the least amount of SNPs compared to
-the reference genome, outperforming the assembly using the original
-quality sequences. A more in-depth evaluation is needed to determine
-whether these compression methods are missing actual SNPs.
+noticeably in the amount of \textsc{snp}s. The order of polynomial has
+an inverse relationship with the amount of \textsc{snp}s detected. The
+2-bin and profile methods detected the least amount of \textsc{snp}s
+compared to the reference genome, outperforming the assembly using the
+original quality sequences. A more in-depth evaluation is needed to
+determine whether these compression methods are missing actual
+\textsc{snp}s.
 
@@
 regression outperforms the 5th-order in all of the reference-based
-metrics, but not the \emph{de novo} LAP metrics. The 5th-order
-polynomial regression assembles approximately 2.5 kb more sequence and
-uses roughly $1.2\%$ more of the total read set than the 4th-order
-(Supplemental Table 1), which may explain the descrepency in LAP
-values because the LAP value is influenced by total
-coverage~\cite{Ghodsi:2013hb}.
+metrics, but not the \emph{de novo} \textsc{lap} metrics. The
+5th-order polynomial regression assembles approximately 2.5 kb more
+sequence and uses roughly $1.2\%$ more of the total read set than the
+4th-order (Supplemental Table 1), which may explain the discrepency in
+\textsc{lap} values because the \textsc{lap} value is influenced by
+total coverage~\cite{Ghodsi:2013hb}.
 
@@
 {\begin{tabular}{lcccccccc}
-& Overall & & & & & & LAP & LAP \\
+& Overall & & & & & & \textsc{lap} & \textsc{lap} \\
 Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
@@
 Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
-Good/bad               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
+2-bin	               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
 Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
@@
 Certain short read alignment tools use the quality sequence
-information when finding potential alignments. Bowtie 2 (version
-2.2.3) was used to evaluate the different decompressed \textsc{fastq}
-files. Bowtie 2 uses quality values written in the \textsc{fastq}
-files when mapping reads against a reference genome. The original
+information when finding potential alignments. Bowtie2 (version 2.2.3)
+was used to evaluate the different decompressed \textsc{fastq}
+files. Bowtie2 uses quality values written in the \textsc{fastq} files
+when mapping reads against a reference genome. The original
 uncompressed and decompressed \textsc{fastq} files were mapped with
-Bowtie 2 against \textit{Rhodobacter sphaeroides} reference
-genome. The generated \textsc{sam} file for each compressing approach
-were compared with the uncompressed \textsc{sam} file. The total,
-shared and unique proportional numbers of mapped reads are calculated
-with respect to the uncompressed \textsc{sam} matches as shown in
-table \ref{tab:aligner}. Additionally, to monitor the effect of
-quality values on mapping in general, Bowtie 2 was tweaked with the
-maximum and minimum mismatch penalty equivalent to maximum and minimum
-quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).
+Bowtie2 against \textit{Rhodobacter sphaeroides} reference genome. The
+generated \textsc{sam} file for each compressing approach were
+compared with the uncompressed \textsc{sam} file. The total, shared
+and unique proportional numbers of mapped reads are calculated with
+respect to the uncompressed \textsc{sam} matches as shown in table
+\ref{tab:aligner}. Additionally, to monitor the effect of quality
+values on mapping in general, Bowtie2 was tweaked with the maximum and
+minimum mismatch penalty equivalent to maximum and minimum quality
+scores (with parameters: --mp 6,6 and --mp 2,2 respectively).
 
@@
   against \textit{Rhodobacter sphaeroides} reference genome using
-  Bowtie 2. Numbers corresponds to the proportion of mapped reads with
+  Bowtie2. Numbers corresponds to the proportion of mapped reads with
   respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
@@
 	&Minimum quality & 100 & 100 & 0 & 0 \\
-	&Good/Bad & 99.92 & 99.89 & 0.11 & 0.03 \\
+	&2-bin & 99.92 & 99.89 & 0.11 & 0.03 \\
 	&Regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
@@
 	&Minimum quality & 100 & 100 & 0 & 0 \\ 
-	&Good/Bad & 97.38 & 97.38 & 2.62 & 0 \\
+	&2-bin & 97.38 & 97.38 & 2.62 & 0 \\
 	&Regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
@@
 	 & \% Bases & \% Reads & \% Bases & \% Reads \\ \hline
-	Good/bad & 98.15 & 98.78 & 113.09 & 103.7 \\ 
+	2-bin & 98.15 & 98.78 & 113.09 & 103.7 \\ 
 	Regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
@@
 
-\subsection{Extension of good/bad encoding}
+\subsection{Extension of 2-bin encoding}
 
-Good/bad encoding has the nice property of being simple to compute and
-has a good bits per bp score. The good/bad encoding suffers from
-having a high \textsc{mse}, but fortunately, we have shown that in the
-case of genome assembly, good/bad encoding outperforms all polynomial
-regressions encodings with degree less than 3. Good/bad encoding of
-the fragment and short-jump libraries of \textit{Rhodobacter
-  sphaeroides} have \textsc{mse}s of $2.42\times$ and $10.76\times$
-the 3rd-order polynomial regression encodings, respectively. This
-further highlights the importance of using additional contextual
-information of the subsequent analyses when evaluating compressed
-quality values.
+2-bin encoding has the nice property of being simple to compute and
+has good bits/base-call values. The 2-bin encoding suffers from having
+a high \textsc{mse}, but fortunately, we have shown that in the case
+of genome assembly, 2-bin encoding outperforms all polynomial
+regressions encodings with degree less than 3. 2-bin encoding of the
+fragment and short-jump libraries of \textit{Rhodobacter sphaeroides}
+have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-order
+polynomial regression encodings, respectively. This further highlights
+the importance of using additional contextual information of the
+subsequent analyses when evaluating compressed quality values.
 
-A potential extension to good/bad encoding is to incorporate an
-additional value \emph{okay}. The \emph{okay} value can be used where
-the base qualities fall within a good/bad range. Because the
-distribution of quality values is skewed towards higher quality, we
-need to experiment with different cutoffs for the \emph{okay} value
-and determine if the additional storage is noticeable in subsequent
+A potential extension to 2-bin encoding is to incorporate an
+additional bin (\emph{okay}). The \emph{okay} value can be used where
+the base qualities fall within a 2-bin range. Because the distribution
+of quality values is skewed towards higher quality, we need to
+experiment with different cutoffs for the \emph{okay} value and
+determine if the additional storage is noticeable in subsequent
 analyses.
@@
 
-Perhaps one of the greatest benefits of compressing quality values is
-the potential to perform quality control and possibly other operations
-directly on the compressed representations of the data. The tool
-\textsc{CutAdapt} trims reads that contains a given number of low
-quality bases from the ends and discards reads that are below a
-certain threshold. It is trivial to apply the same rules to a few of
-our compressed quality sequences. In the good/bad encoding, we store a
-single bit for each quality value. A read with stretches of 0-bits
-towards the ends can be trimmed and filtered if there are more 0-bits
-than 1-bits.
-
-In principle, we can quality filter the polynomial regression
-compression by interpolating where the polynomials are below a certain
-threshold and then filtering/trimming that part of the
-read. Unfortunately, storing the coefficients of the polynomials as a
-float leads to very poor compression.
-
-A beneficial feature of the profile-based approach is that read
-quality filtering and/or trimming can be performed on the fly during
-decompression at virtually no computational cost. As all reads are
-classified into one of $K$ quality profiles, we can determine which of
-these profiles are of use for subsequent processes, and only
-decompress reads assigned to these profiles. Furthermore, it is
-possible to determine trimming points for each profile before
-extraction, that can be carried out on-line on reads during
-decompression.
+Perhaps one of the greatest potential benefits of compressing quality
+values is the potential to perform quality control and possibly other
+operations directly on the compressed representations of the
+data. This is easiest to to consider for profile-based
+compression. The $k$ profiles can be evaluated for (pre-)processing
+operations such as filtering and trimming, and the operations
+transitively applied to the entire set of reads, thus saving
+substantial computation associated with evaluating the full set of
+reads.
 
@@
 memory and computation. For our quality value binning and model-based
-compression methods there are no explicit fixed costs
-in terms of memory, unlike the profile--based approach, which has a
-small fixed cost in storing the profiles. However, the profile-based
-approach has a much smaller marginal compared to the rest due to the
-fact that we are simply storing the profile index for each quality
-sequence.
+compression methods there are no explicit fixed costs in terms of
+memory, unlike the profile--based approach, which has a small fixed
+cost in storing the profiles. However, the profile-based approach has
+a much smaller marginal compared to the rest due to the fact that we
+are simply storing the profile index for each quality sequence.
 
@@
 
-\subsection{Background}
-
-Sequencing read data constitutes the largest category of sequence data
-in genomics research because of its great redundancy, inclusion of
-quality values, and read-level naming and metadata. In recognition of
-this abundance and importance we place particular emphasis on
-compression of read data, which is the focus of this Aim. Among the
-several challenges for compression of read data is dealing with
-different error profiles resulting from the differences in underlying
-chemistries, signal detection and processing mechanisms, inherent
-biases, and other idiosyncratic properties of distinct high-throughput
-sequencing technologies. Our proposed research includes development of
-compression methods that robustly accommodate different error profiles
-associated with sequence read data, and where advantageous we will
-leverage error profiles to increase compression performance.
-
-At the level of sequence read the probability of error for each
-base-call is typically represented by \textsc{phred} quality value,
-which is defined as $Q =
--10\,log_{10}P$~\cite{Ewing:1998zr,Ewing:1998ly}.  Depending on the
-sequencing technology these quality values can range from 0 to 93, and
-are represented with the \textsc{ascii} characters 33 to 126 (with
-some offset). Quality values comprise a standard component of
-\textsc{fastq} format files~\cite{Cock:2010ve}. The characteristics of
-different sequencing technology-associated error profiles are
-reflected in the quality values, and can be broadly described for both
-Illumina and PacBio sequencing technologies.
-
-Illumina reads are characterized by having far fewer insertion and
-deletion errors, where substitution (miscall) errors are much more
-frequent and have context-specific patterns. The errors and quality
-values are non-uniformly distributed over the read length (e.g., error
-rates increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and
-32.8 -- 67.9\% of reads have low quality tails at the 3$^{\prime}$
-end~\cite{Minoche:2011km}). Illumina reads have a single quality value
-per base-call.
-
-\subsection{Lossy compression of quality values} \label{sec:qual}
-
 \subsubsection{Pilot Study: profile-based compression of quality 
@@
 generate profiles representing these patterns, and use the profiles as
-proxies for read quality value sequences. We demonstrated the efficacy
-of this approach in a pilot study using test data of Illumina reads
-the \textit{Rhodobacter sphaeroides} 2.4.1 data set provided by
-\textsc{gage} (Genome Assembly Gold-Standard
-Evaluations)~\cite{Salzberg:2012rc}, which consists of 2,050,868
-paired-end reads (fragment library; insert size of 180 bp) and
-2,050,868 reads (short-jump library; insert size of 3,000 bp).
+proxies for read quality value sequences.
 
@@
 sequences of quality values and using the resulting cluster centers as
-representative profiles. In our pilot study we used $K$-means
+representative profiles. In our pilot study we used $k$-means
 clustering, a vector quantization method that partitions a set of
-samples into $K$ sets that minimize within-cluster sum of
+samples into $k$ sets that minimize within-cluster sum of
 squares~\cite{hastie_09_elements-of.statistical-learning}, as follows.
@@
 \item
-Perform $K$-means clustering on the sample, and use the cluster
+Perform $k$-means clustering on the sample, and use the cluster
 centers as profiles for the dataset.
@@
 
-Our pilot study results demonstrate that $K$ values as small as
+Our pilot study results demonstrate that $k$ values as small as
 32--256 sufficiently represent patterns in Illumina read quality
@@
 resulting compressed quality component of the original set of
-\textsc{fastq} files consists of the $K$ quality profiles (the fixed
+\textsc{fastq} files consists of the $k$ quality profiles (the fixed
 size cost), and a single byte representing the assigned quality
@@
 To evaluate the effect of compression on these quality control
-operations, we used \textsc{autoadapt}, which uses
-\textsc{FastQC}~\cite{citeulike:11583827} and
-\textsc{cutadapt}~\cite{EJ200} packages, all with the default
+operations, we used Sickle~\cite{sickle} with the default
 settings. \textsc{PrinSeq}~\cite{Schmieder:2011gd} was then used to

### 2015-02-03 manuscript/q-compression.tex
More revisions describing the approaches we explore to emphasize the generalize the concepts. The approaches are not called binning, modeling, and profiling.

@@
 We examine three general classes of lossy compression methods:
-binning, model-based, and profile-based.
+binning, modeling, and profiling.
 
@@
 single bit to represent the quality of a base instead of the default 8
-bits used to store quality values in \textsc{ascii}.
+bits used to store quality values in \textsc{ascii}. An additional
+benefit of 2-bin encoding is the potential for increased adjacency of
+identical values and repeating patterns, properties that may increase
+effectiveness of subsequent compression using established
+algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,DBLP:journals/tit/ZivL78}.
 
@@
 
-\subsection{Polynomial regression}
+\subsection{Modeling}
 
-One approach to compressing quality values is to treat them as
-bivariate data with the ordered nucleotides (1 to read length)
-representing the abscissa, and quality values representing the
-ordinate. Viewed in this way we can ask the question: is there a
-mathematical function that describes the relationship between read
-position and quality value sufficiently, albeit imperfectly (i.e.,
-with some loss of information), such that modeled values perform well
-in subsequent analyses?
+If quality values are modeled compression is conceivably possible by
+replacing the original quality values by a representation of the
+model. For example, quality values can be conceptualized as bivariate
+data with the ordered nucleotides (1 to read length) representing the
+abscissa, and quality values representing the ordinate. In this
+research we model read quality values as polynomial functions obtained
+with least-squares fitting, as one approach to compression read
+quality values by modeling.
 
-To this end, we modeled read quality sequences with zero to
-fifth-degree polynomials obtained with least-squares fitting. However,
-despite the fact that polynomial functions have significantly lower
+Despite the fact that polynomial functions have significantly lower
 number of parameters (i.e. one to six coefficients) than a read-length
@@
 
-Nonetheless, fitting a polynomial and reconstructing the quality
-values that it would have yielded after decompression still makes some
-sense as it alters the properties of the quality data, making it
-highly compressible simply due to the smoothing effect it
-provides. Therefore the compression ratio and bits/base-call values
-for polynomial regression throughout this manuscript refer to
-polynomially approximated quality values compressed with zip.
-
-\subsection{Profile-based compression}
+\subsection{Profiling}
 
@@
 
-\subsection{Compressibility}
+\subsection{Compressibility versus error}
 

### 2015-02-03 src/compress.py
Added profile-based compression

@@ -41,6 +41,7 @@ def compress(options):
     std_err_file = open('compress.log', 'w')
 
     GB_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -c 2 -b 0 "
+    PROFILE_COMPRESSION_CMD = "python src/profile_and_compress.py [READ] [OUTPUT] [NUM_PROFILES] [TRAINING_SIZE] [OUTPUT].png "
     
     # Store which compression directories we created.
     options.compressed_dirs = []
@@ -62,7 +63,18 @@ def compress(options):
 
         # TODO: Polynomial regression.
 
-        # TODO: Profile regression.
+        # Profile compression using k-means.
+        for profiles in options.profile_sizes.split(','):
+            ensure_dir(options.output_dir + '/profile_' + profiles + '/')
+            options.compressed_dirs.append('profile_' + profiles)
+
+            call_arr = PROFILE_COMPRESSION_CMD.replace('[READ]', reads_filename)\
+                    .replace('[OUTPUT]', options.output_dir + '/profile_' + profiles + '/' + os.path.basename(reads_filename))\
+                    .replace('[NUM_PROFILES]', profiles)\
+                    .replace('[TRAINING_SIZE]', options.training_size).split()
+
+            out_cmd("", std_err_file.name, call_arr)
+            call(call_arr, stderr=std_err_file)
 
 
     # After we compress/decompress everything, write out the quality values to a separate file and then run bzip on them.
@@ -286,7 +298,11 @@ def get_options():
     parser.add_option("-a", "--assemble", dest="assemble", help="Run assembly evaluation", action='store_true')
     parser.add_option("-p", "--preprocessing", dest="preprocessing", help="Run preprocessing tools evaluation", action='store_true')
     parser.add_option("-b", "--alignment", dest="alignment", help="Run alignment evaluation (using Bowtie2).", action='store_true')
-    
+
+    # Profile-specific compression options.
+    parser.add_option("--training-size", dest="training_size", help="Training size used for clustering.", default = "10000")
+    parser.add_option("--profile-sizes", dest="profile_sizes", help="Comma-separated list of number of profiles to use.", default="256")
+
     # Additional options.
     parser.add_option("-t", "--threads", dest="threads", help="Number of threads (default 32).", default="32")
 

### 2015-02-03 src/profile_and_compress.py

@@ -2,6 +2,7 @@ from sklearn.cluster import KMeans
 from random import shuffle, sample
 from scipy.cluster.hierarchy import linkage, dendrogram
 from scipy.spatial.distance import pdist, squareform
+import scipy.misc
 import numpy as np
 import sys
 
@@ -21,23 +22,27 @@ def sort_rowsum(matrix):
     return aa[aa.sum(axis=1).argsort(),:]
 
 N_CLUSTERS = int(sys.argv[3])
-TRAINING_SIZE = 10000
+TRAINING_SIZE = int(sys.argv[4])
 
 # filename 'gage_rhodo/zzz.quals' 
 
 quals = []
 with open(sys.argv[1], 'r') as zzz:
+    counter = 1
     for line in zzz:
-        quals.append(line.rstrip())
+        if (counter % 4 == 0):
+            quals.append(line.rstrip())
+        counter += 1
 
 profile_training = sample(quals, TRAINING_SIZE)
 
 features = np.array(map(qual_ord, profile_training))
-print 'starting kmeans'
+#print 'starting kmeans'
 km = KMeans(n_clusters=N_CLUSTERS, init='random', max_iter=600)
 km.fit(features)
-print 'finished kmeans'
-#imsave('kmeans_centers_8.png', cluster_rows(km.cluster_centers_))
+#print 'finished kmeans'
+if len(sys.argv) > 5:
+    scipy.misc.imsave(sys.argv[5], cluster_rows(km.cluster_centers_))
 
 #pred = km.predict(features)
 #print ' '.join(map(str,sorted(Counter(pred).values())))
@@ -45,20 +50,22 @@ centers = km.cluster_centers_
 center_qualstrings = [''.join([chr(max(35,int(x)+33)) for x in center]) for center in centers]
 
 
-# #pred = km.predict(quals)
-# print 'starting writing'
-# with open(sys.argv[2], 'w') as outfile_decomp:
-#     with open(sys.argv[2]+'.compressed', 'w') as outfile_comp:
-#         outfile_comp.write('%d\n' % N_CLUSTERS)
-#         for i_center, center_qualstring in enumerate(center_qualstrings):
-#             outfile_comp.write('%d %s\n' % (i_center, center_qualstring))
-        
-#         for i, qq in enumerate(quals):
-#             center_class = km.predict(qual_ord(qq))
-#             #basis = centers[center_class][0]
-#             #qq = np.array(qual_ord(qq))
-#             #diff = np.int32(qq - basis)/10
[…]

### 2015-02-03 manuscript/q-compression.tex
More revisions, particularly to the intro, new references, other changes.

@@
 biases, and other idiosyncratic properties of distinct high-throughput
-sequencing technologies. Illumina sequencing reads, the focus of this
+sequencing technologies. Illumina sequence reads, the focus of this
 research, are characterized by having relatively few insertion and
@@
 offset). These errors are represented a single quality value per
-base-call for Illumina sequencing reads.
+base-call for Illumina sequence reads.
 
@@
 descriptions of the distribution of sequence quality scores, and these
-scores can be used to guide filtering and trimming of sequencing read
+scores can be used to guide filtering and trimming of sequence read
 data as processing step before subsequent analyses. Quality values can
@@
 
-Although the quality values are useful for later steps, the trade-off
-between storing effectively double the amount of data and the
-resulting effects on subsequent analyses has not been
-well-documented. Previous literature on sequence data compression has
-largely focused on lossless compression of base
-calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
+Previous literature on sequence data compression has largely focused
+on lossless compression of base calls~\cite[reviewed
+  in][]{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
+Although some recent work has included or focused on compression of
+quality
+values~\cite[e.g.,][]{Tembe:2010ys,Kozanitis:2011kl,asnani2012lossy,Hach:2012ys,Wan:2012kq,janin2013adaptive,Ochoa:2013rt,Canovas:2014fr,zhou2014compression,DBLP:conf/recomb/YuYB14}.
 Typical compression methods fail to utilize the biases inherent in
@@
 
-Little prior work has been done examining the effect of lossy
-compression of quality sequences in the context of subsequent
-analyses. Our aim here is to examine the effect of multiple lossy
-compression methods on quality sequences, focusing not only on the
-differences between the compressed and original versions, but also the
-effect of lossy compression on typical subsequent analyses.
-
-We examine three general classes of lossy compression methods:
-binning, modeling, and profiling.
+Although we recognize the need for lossless compression for some
+purposes and contexts (e.g., archiving, provenance), our perspective
+is largely pragmatic with a focus on the use of quality values in
+subsequent analyses. From this perspective some absolute loss of
+information is deemed acceptable if the inferences from analyses are
+relatively uneffected. Therefore we here describe our research
+investigating some general strategies for compression of sequence read
+quality values. Our objective provide perspective on several
+strategies logically applied to Illumina sequence read quality values
+rather than to generate a robust high-quality software for
+use. Consistent with our perspective we place emphasis not only on the
+compression efficiency, but also the effect of lossy compression on
+typical subsequent analyses.  We examine three general classes of
+lossy compression methods -- binning, modeling, and profiling -- and
+consider an exemplar of each class.
 
@@
 
-\subsection{Lossy compression effect on preprocessing quality}
+\subsection{Effects on sequence read preprocessing}
 
@@
 
-\subsection{Lossy compression effect on assembly quality}
+\subsection{Effects on genome assembly}
 
@@
 {\begin{tabular}{lcccccccc}
-& Overall & & & & & & \textsc{lap} & \textsc{lap} \\
-Compression            & rank & \textsc{snp}s & Inversions & Relocations & Translocations & Indels & fragment & short-jump \\
+& overall & & & & & & \textsc{lap} & \textsc{lap} \\
+compression            & rank & \textsc{snp}s & inversions & relocations & translocations & indels & fragment & short-jump \\
 \hline
-None                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
-Profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
-Profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
-Regression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
-Regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
+none                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
+profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
+profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
+pegression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
+regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
 2-bin	               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
-Regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
-Regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
-Regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
-Regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
+regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
+regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
+regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
+regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
 \\ \hline
@@
 
-\subsection{Lossy compression effect on mapping quality}
+\subsection{Effects on read mapping}
 
@@
 \begin{tabular}{llcccc}
-    & & \multicolumn{4}{c}{Percentage of reads mapped}\\
+    & & \multicolumn{4}{c}{percentage of reads mapped}\\
     \cline{3-6}
-	Library & Compression Strategy & Total & Shared & Uncompressed only & Compressed only \\ \hline
-	& Maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
-	&Minimum quality & 100 & 100 & 0 & 0 \\
+	library & compression strategy & total & shared & uncompressed only & compressed only \\ \hline
+	& maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
+	&minimum quality & 100 & 100 & 0 & 0 \\
 	&2-bin & 99.92 & 99.89 & 0.11 & 0.03 \\
-	&Regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
-	&Regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
-	&Regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
-	Fragment&Regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
-	&Regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
-	&Regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
-	&Profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
-	&Profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\
+	&regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
+	&regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
+	&regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
+	fragment&regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
+	&regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
+	&regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
+	&profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
+	&profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\
 
@@
 
-	&Maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
-	&Minimum quality & 100 & 100 & 0 & 0 \\ 
+	&maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
+	&minimum quality & 100 & 100 & 0 & 0 \\ 
 	&2-bin & 97.38 & 97.38 & 2.62 & 0 \\
-	&Regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
-	&Regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
-	&Regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
-	Short-jump&Regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
-	&Regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
-	&Regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
-	&Profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
-	&Profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
+	&regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
+	&regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
+	&regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
+	short-jump&regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
+	&regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
+	&regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
+	&profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
+	&profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
 \end{tabular}
@@
 \begin{tabular}{lcccc}
-	 & \multicolumn{2}{c}{Fragments} & \multicolumn{2}{c}{Short-Jump} \\
-	 & \% Bases & \% Reads & \% Bases & \% Reads \\ \hline
+	 & \multicolumn{2}{c}{fragments} & \multicolumn{2}{c}{short-Jump} \\
+	 & \% bases & \% reads & \% bases & \% reads \\ \hline
 	2-bin & 98.15 & 98.78 & 113.09 & 103.7 \\ 
-	Regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
-	Regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
-	Regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
-	Regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
-	Regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
-	Regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
-	Profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
-	Profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
+	regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
+	regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
+	regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
+	regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
+	regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
+	regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
+	profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
+	profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
 \end{tabular}

### 2015-02-04 manuscript/q-compression.tex
Continuing the revision process. Consolidating information in the Methods section, including text that was previously elsewhere. I have changed, slightly, our definition of MSE to be consistent with the standard definition.

@@
 
-Sequencing read data constitutes the largest category of sequence data
-in genomics research because of great redundancy, inclusion of quality
-values, and read-level naming and metadata. Because of this abundance
-effective compression of read data has the potential for substantial
-improvement in data storage and transfer efficiency.  Among the
-several challenges for compression of read data is dealing with
-different error profiles resulting from differences in underlying
-chemistries, signal detection and processing mechanisms, inherent
-biases, and other idiosyncratic properties of distinct high-throughput
-sequencing technologies. Illumina sequence reads, the focus of this
+Read data from high-throughput sequencing constitutes the largest
+category of data in genomics research because of great redundancy,
+inclusion of quality values, and read-level naming and
+metadata. Because of this abundance effective compression of read data
+has the potential for substantial improvement in data storage and
+transfer efficiency. Among the several challenges for compression of
+read data is dealing with different error profiles resulting from
+differences in underlying chemistries, signal detection and processing
+mechanisms, inherent biases, and other idiosyncratic properties of
+distinct high-throughput sequencing technologies. Sequence reads
+generated by instruments such as an Illumina HiSeq, the focus of this
 research, are characterized by having relatively few insertion and
@@
 scores can be used to guide filtering and trimming of sequence read
-data as processing step before subsequent analyses. Quality values can
-be used by genome assembly software to produce better
-assemblies~\cite[e.g.,][]{Gnerre:2011kx}.  Short-read alignment
-software, such as Bowtie 2~\cite{Langmead:2012rw}, use quality values
-to scale the mismatch penalty in a way that mismatches between poor
-quality bases of the query and the reference are weighted less than
-mismatches occurring on high quality bases. Similar to alignment,
-software for detecting single nucleotide polymorphisms (\textsc{snp}s)
-can use quality values~\cite[e.g.,][]{McKenna:2010bh}, and identified
-\textsc{snp}s with high-quality calls are deemed more reliable than
-those with low-quality calls, particularly in low coverage regions.
+data as processing steps before subsequent analyses. Quality values
+can be used by genome assembly software to produce better
+assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
+alignment software, such as Bowtie 2~\cite{Langmead:2012rw}, use
+quality values to weight mismatches between query and reference
+sequences.  Software for detecting single nucleotide polymorphisms
+(\textsc{snp}s) can use quality values, and identified \textsc{snp}s
+with high-quality calls are deemed more reliable than those with
+low-quality calls, particularly in low coverage
+regions~\cite[e.g.,][]{McKenna:2010bh}.
 
@@
   in][]{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
-Although some recent work has included or focused on compression of
+Although some recent work has included, or focused on, compression of
 quality
@@
 information is deemed acceptable if the inferences from analyses are
-relatively uneffected. Therefore we here describe our research
+relatively unaffected. Therefore we here describe our research
 investigating some general strategies for compression of sequence read
@@
 
-\subsection{Evaluating lossy compression algorithms}
+\subsection{Performance evaluation}
+
+As a measure of compression effectiveness we use bits/base-call, and
+define it as the size of the compressed representation of quality
+values (in bits) divided by the number of quality values represented.
+As a measure of information loss we use mean squared error
+(\textsc{mse}) as aloss function, and define it as
+$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
+sequences, $Q_i$ is the original quality value associated with
+sequence position $i$, and $Q_i'$ is the corresponding
+compressed/decompressed quality value.
 
@@
 \subsection{Compressibility versus error}
-
-For any lossy compression method, we need to evaluate the tradeoff
-between its physical compression and the loss in its accuracy. Here,
-we use mean bits/base-call to measure the physical compression of the
-data and mean squared error (\textsc{mse}) to measure the loss in
-quality. Bits/base-call is calculated simply as the compressed file
-size (in bits) divided by the total number of quality
-values. \textsc{mse} is calculated as
-$\frac{1}{n}\sum_{i=1}^{n}{(Q_i-Q_i')^2}$, where $n$ is the number of
-sequences, $Q_i$ is the original quality value associated with
-sequence position $i$, and $Q_i'$ is the corresponding compressed
-quality value.
 

### 2015-02-04 manuscript/q-compression.tex
More revisions to remove from the results that material that more appropriate belongs in the intro and methods sections. Added bonus: methods section is typeset in a smaller font and thus leaves more room for other sections.

@@
 
-\copyrightyear{2014}
-\pubyear{2014}
+\copyrightyear{2015}
+\pubyear{2015}
 
@@
 Quality values are typically used throughout bioinformatics pipelines.
-Software such as \textsc{fastqc}~\cite{citeulike:11583827} and
-Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
-descriptions of the distribution of sequence quality scores, and these
-scores can be used to guide filtering and trimming of sequence read
-data as processing steps before subsequent analyses. Quality values
-can be used by genome assembly software to produce better
+Among the most fundamental uses of sequence quality values is as part
+of the quality assessment and quality control (\textsc{qa/qc})
+processes prior to subsequent analysis steps. Quality control based on
+quality values generally includes two operations:
+\textit{i}.~filtering, the elimination of reads that on the whole do
+not meet quality standards, which reduces the total number of reads;
+and \textit{ii}.~trimming of low quality base-calls from reads, which
+reduces the number total number of bases. Quality values can be used
+by genome assembly software to produce better
 assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
@@
 
+Lossy compression compromises representation accuracy for better space
+efficiency. Different lossy compression methods have different
+properties, therefore we need to evaluate a number of tradeoffs for
+each of our proposed methods. More importantly, we need to examine
+each approach in the context of typical bioinformatics tasks, such as
+quality control, \emph{de novo} genome assembly, and aligning the
+sequences to a given reference.
+
 One compression strategy is to look at the distribution of quality
@@
 
-\subsection{Binning}
+\subsection{Compression strategies}
 
-Quality values can be binned into a set of bins, and any number of
-bins less than the the number of distinct quality values is
-possible. The minimum number of bins that allows for a any distinction
-among quality values is two, i.e., two categories ``good'' and ``bad''
-quality. We implement 2-bin encoding by setting a quality value
-threshold empirically determined by the distribution of quality values
-across reads. Base-calls are marked ``bad'' if their quality value
-falls below the first quartile minus 1.5 $\times$ the interquartile
-range (IQR), which is the difference between the first and third
-quartile. 1.5 $\times$ IQR is the value used by Tukey's box
-plot~\cite{mcgill1978variations}. The main benefit of this approach is
-that it is completely data-dependent. No prior assumptions of the
-distribution of the quality values need to be made.
+\subsubsection{Binning}
+
+Quality values can be binned, and the minimum number of bins that
+allows for a any distinction among quality values is two, i.e., two
+categories ``good'' and ``bad'' quality. We implement 2-bin encoding
+by setting a quality value threshold empirically determined by the
+distribution of quality values across reads. Base-calls are marked
+``bad'' if their quality value falls below the first quartile minus
+1.5 $\times$ the interquartile range (IQR), which is the difference
+between the first and third quartile. 1.5 $\times$ IQR is the value
+used by Tukey's box plot~\cite{mcgill1978variations}. The main benefit
+of this approach is that it is completely data-dependent. No prior
+assumptions of the distribution of the quality values need to be made.
 
@@
 
-\subsection{Modeling}
+\subsubsection{Modeling}
 
@@
 
-\subsection{Profiling}
+\subsubsection{Profiling}
 
@@
 
-$k$-means clustering is a vector quantization method, partitioning a set
-of samples into $k$ sets that minimize within-cluster sum of
+$k$-means clustering is a vector quantization method, partitioning a
+set of samples into $k$ sets that minimize within-cluster sum of
 squares~\cite{macqueen1967some}. Using a random subset of read quality
@@
 
-Once the $k$ quality profiles are determined, all quality sequences are
-passed through the trained $k$-means predictor, with the nearest quality
-profile in Euclidean space being assigned to every quality sequence as
-their compressed representation.
+Once the $k$ quality profiles are determined, all quality sequences
+are passed through the trained $k$-means predictor, with the nearest
+quality profile in Euclidean space being assigned to every quality
+sequence as their compressed representation.
 
@@
 $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
-sequences, $Q_i$ is the original quality value associated with
-sequence position $i$, and $Q_i'$ is the corresponding
-compressed/decompressed quality value.
+sequences, $Q_i'$ is the compressed/decompressed quality value, and
+$Q_i$ is the original quality value associated with sequence position
+$i$.
 
-Lossy compression compromises representation accuracy for better space
-efficiency. Different lossy compression methods have different
-properties, therefore we need to evaluate a number of tradeoffs for
-each of our proposed methods. More importantly, we need to examine
-each approach in the context of typical bioinformatics tasks, such as
-quality control, \emph{de novo} genome assembly, and aligning the
-sequences to a given reference.
+We evaluate effects of information loss from quality value compression
+on quality control steps of read filtering and trimming, which were
+performed using Sickle~\cite{sickle}, quantify read characteristics
+with PrinSeq~\cite{Schmieder:2011gd}, and make comparison to
+uncompressed data.
 
-For quality control, we use
-Sickle~\cite{sickle}. PrinSeq~\cite{Schmieder:2011gd} was used to
-count the number of bases and reads in each \textsc{fastq} file.
-
-For \emph{de novo} genome assembly, we use
+We evaluate effects of information loss from quality value compression
+on \emph{de novo} genome assembly performance using contiguity
+statistics, log average read probability
+(\textsc{lap})~\cite{Ghodsi:2013hb}, and a collection of
+reference-based metrics. The contiguity statistics include number of
+scaffolds and NG50, which is defined as the weighted median contig
+size (the length of largest contig $c$ such that the total size of the
+contigs larger than $c$â exceeds half of the known genome size). The
+\textsc{lap} score can be viewed as a log likelihood score, where a
+value closer to 0 is better. We use a reference-based evaluation
+script provided by \textsc{gage} that counts single nucleotide
+polymorphisms (\textsc{snp}s), relocations, translations, and
+inversions. The reference-based metrics are normalized by the length
+of the assembly to aid in comparison. For the genome assembly we used
+software that makes use quality values in the assembly process:
 \textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
 settings and 32 threads.
-
-For aligning the sequences, we use Bowtie2~\cite{Langmead:2012rw} with
-default settings and modified max- and min-mismatch parameters
-detailed below.
 
@@
 
-\subsection{Compressibility versus error}
+\subsection{Compression effectiveness versus information loss}
 
@@
 
-One way to evaluate different compression approaches is to perform
-quality preprocessing on the generated \textsc{fastq} files and
-compare the number of reads and bases with those of the uncompressed
-\textsc{fastq} files. Sickle is a tool to filter and trim reads. We
-used PrinSeq to calculate the number of reads and bases in each
-sample. Table \ref{tab:quality_control} shows the proportional numbers
-with respect to those of the uncompressed \textsc{fastq} files. In the
+Table \ref{tab:quality_control} shows the proportional numbers with
+respect to those of the uncompressed \textsc{fastq} files. In the
 fragment sample, we retain on average 95.4\% and 94.24\% of bases and
@@
 
-Genome assemblers, such as \textsc{allpaths-lg}~\cite{Gnerre:2011kx},
-often utilize quality values to aid in the assembly process. By
-reducing the information contained within the quality values, we
-introduce the potential for the resulting assembly to be affected.
-
-\textsc{allpaths-lg} is one of the leading bacterial genome
-assemblers~\cite{Salzberg:2012rc}. We examine the effect of our lossy
-compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data
-set provided by \textsc{gage} (Genome Assembly Gold-Standard
-Evaluations)~\cite{Salzberg:2012rc}. \textsc{allpaths-lg} is used to
-assemble the original data along with the quality compressed versions
-using 2-bin encoding, 0--5th-order polynomial regression encodings,
-and 32- and 256-profile encodings. We evaluate the assembly based on
-contiguity statistics, log average read probability
-(\textsc{lap}~\cite{Ghodsi:2013hb}), and a collection of
-reference-based metrics. The contiguity statistics include number of
-scaffolds and NG50, which is defined as the weighted median contig
-size (the length of largest contig $c$ such that the total size of the
-contigs larger than $c$â exceeds half of the known genome size). The
-\textsc{lap} score can be viewed as a log likelihood score, where a
-value closer to 0 is better. We use a reference-based evaluation
-script provided by \textsc{gage} which counts single nucleotide
-polymorphisms (\textsc{snp}s), relocations, translations, and
-inversions. The reference-based metrics are normalized by the length
-of the assembly to aid in comparison. Alongside the raw metrics, we
-also provide a cummulative sum of the individual ranks of each method
-for each metric. A similar approach was utilized by the
-\textsc{Assemblathon1} competition~\cite{Earl:2011fv}.
-
-The assembly using the uncompressed reads outperforms all compression
-methods in terms of \textsc{lap}, NG50, relocations, translations, and
-inversions (Table \ref{tab:assembly_ranks}). Among the compression
-methods, the 256- and 32-profile encoding had the highest rank, then
-the 4th-order polynomial regression, the 5th-order polynomial
-regression, the 2-bin encoding, and lastly, the 3rd through 0-order
-polynomials.
+The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
+using the uncompressed reads outperforms all compression methods in
+terms of \textsc{lap}, NG50, relocations, translations, and inversions
+(Table \ref{tab:assembly_ranks}). Among the compression methods, the
+256- and 32-profile encoding had the highest rank, then the 4th-order
+polynomial regression, the 5th-order polynomial regression, the 2-bin
+encoding, and lastly, the 3rd through 0-order polynomials.
 
@@
 
-As described previously, quality values in Illumina reads are
-non-uniformly distributed over the read length and show a general
-pattern of decreasing toward the 3$^{\prime}$ end. As large sets of
-quality strings show similar trends of quality over their sequence,
-one approach would be to identify common patterns in the data,
-generate profiles representing these patterns, and use the profiles as
-proxies for read quality value sequences.
-
 Quality value patterns can be readily determined by clustering
@@
 
-\paragraph{Read quality processing}
-Among the most fundamental uses of sequence quality values is as part
-of the quality assessment and quality control (\textsc{qa/qc})
-processes prior to subsequent analysis steps. Quality control based on
-quality values generally includes two operations:
-\textit{i}.~filtering, the elimination of reads that on the whole do
-not meet quality standards, which reduces the total number of reads;
-and \textit{ii}.~trimming of low quality base-calls from reads, which
-reduces the number total number of bases.
-
-To evaluate the effect of compression on these quality control
-operations, we used Sickle~\cite{sickle} with the default
-settings. \textsc{PrinSeq}~\cite{Schmieder:2011gd} was then used to
-count the number of bases and reads in each quality control-processed
-\textsc{fastq} file. The proportion of reads retained using
-compressed/decompressed data compared to that of the uncompressed
-data: 0.998 for the fragment library reads, and 1.013 for the
-short-jump library reads; and proportion of bases retained 1.0 for the
-fragment library reads, and 1.027 for the short-jump library
-reads. Thus there is very little difference between profile-based
-compressed data compared to that of the original uncompressed data in
-read filtering and trimming.
-
-\paragraph{\textit{De novo} genome assembly}
-For de novo genome assembly, we use
-\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
-settings. We evaluate the assembly based on contiguity statistics, log
-average read probability (\textsc{lap})~\cite{Ghodsi:2013hb}, and a
-collection of reference-based metrics, of which the contiguity
-statistics include number of scaffolds and NG50; the latter statistic
-is defined as the weighted median contig size (the length of largest
-contig $c$ such that the total size of the contigs larger than $c$
-exceeds half of the known genome size). The \textsc{lap} score can be
-viewed as a log likelihood score, where a larger value is better. We
-use a reference-based evaluation script provided by \textsc{gage},
-which counts single nucleotide polymorphisms (\textsc{snp}s),
-relocations, translations, and inversions. The reference- based
-metrics are normalized by the length of the assembly to aid in
-comparison. The differences in assembly statistics between using the
-profile-based compressed/decompressed data compared to using
-uncompressed data are extremely small: total assembly size difference
-of 6.745 kbp ($\sim$0.00015); NG50 size difference of 2.259 kbp
-(0.0007); an \textsc{lap} score difference of 0.002 for the fragment
-library and 0.04 for the short-jump library.
-
 \paragraph{Aligning reads to a reference sequence}

### 2015-02-04 manuscript/q-compression.tex
Integrated the economic costs of memory into the subsections for each category of compression we examine.

@@
 assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
-alignment software, such as Bowtie 2~\cite{Langmead:2012rw}, use
+alignment software, such as Bowtie2~\cite{Langmead:2012rw}, use
 quality values to weight mismatches between query and reference
@@
 used by Tukey's box plot~\cite{mcgill1978variations}. The main benefit
-of this approach is that it is completely data-dependent. No prior
-assumptions of the distribution of the quality values need to be made.
+of this approach is that it is completely data-dependent, and no
+assumptions regarding the distribution of the quality values need to be made.
 
-Because we only have two possible choices for a quality of base, this
-makes a straightforward binary encoding possible, allowing us to use a
-single bit to represent the quality of a base instead of the default 8
+With 2-bin encoding binary encoding is possible, allowing us to use a
+single bit to represent the quality of a base instead of the standard 8
 bits used to store quality values in \textsc{ascii}. An additional
@@
 
-Although simple and intuitive, considering a base as either good or
-bad may be too limiting for some users, so we provide a few additional
-options to preserve more of the defining characteristics of the
-quality values, but we only analyze 2-bin encoding in our results.
+The economic costs of memory use for binning, in general terms,
+include no fixed costs, and marginal costs that are a function of the
+number of base-call quality values encoded times the cost of the
+encoding.
 
@@
 this study.
+
+The economic costs of memory use for model-based compression, in
+general terms, include no fixed costs, and marginal costs that are a
+function of the number of reads times the cost representing the model
+parameters.
 
@@
 
-Although this approach is not able to capture randomly occuring
+Although this approach is not able to capture randomly occurring
 outlier quality values, it ensures that the overall trends in quality
@@
 of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.
+
+The economic costs of memory use for profile-based compression, in
+general terms, include fixed costs associated with representing the
+profiles, which is a function of the number of profiles times the cost
+of encoding them, and these fixed costs are amortized over the entire
+set of reads to which they apply. Additionally there are marginal
+costs that are a function of the number of reads encoded.
 
@@
 
-\subsection{The economics of lossy quality sequence compression}
-
-From an economics perspective, we can examine the fixed and marginal
-costs associated with our lossy compression techniques with respect to
-memory and computation. For our quality value binning and model-based
-compression methods there are no explicit fixed costs in terms of
-memory, unlike the profile--based approach, which has a small fixed
-cost in storing the profiles. However, the profile-based approach has
-a much smaller marginal compared to the rest due to the fact that we
-are simply storing the profile index for each quality sequence.
-
-Regarding computational costs, only quality binning and the
-profile--based methods have associated fixed costs. Before performing
-compression, we have to determine the optimal thresholding values for
-quality binning, and the cluster representatives for the
-profile--based method.
-
-In practice, the specific values of these costs are very sensitive to
-the implementation details.
-
 \subsection{Future of lossy compression in bioinformatics analyses}

### 2015-02-05 src/compress.py
Added polynomial regression compression.

@@ -40,7 +40,9 @@ def compress(options):
 
     std_err_file = open('compress.log', 'w')
 
+    # Basic command line scripts to run the individual compression schemes.
     GB_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -c 2 -b 0 "
+    POLY_REGRESSION_CMD = "python src/map_fastq.py [READ] [OUTPUT] [DEGREE]"
     PROFILE_COMPRESSION_CMD = "python src/profile_and_compress.py [READ] [OUTPUT] [NUM_PROFILES] [TRAINING_SIZE] [OUTPUT].png "
     
     # Store which compression directories we created.
@@ -61,7 +63,18 @@ def compress(options):
         out_cmd(options.output_dir + '/goodbad/' + os.path.basename(reads_filename), std_err_file.name, call_arr)
         call(call_arr, stdout=output_fp, stderr=std_err_file)
 
-        # TODO: Polynomial regression.
+        # Polynomial regression.
+        for degree in options.poly_degrees.split(','):
+            ensure_dir(options.output_dir + '/degree_' + degree + '/')
+            options.compressed_dirs.append('degree_' + degree)
+
+            call_arr = POLY_REGRESSION_CMD.replace('[READ]', reads_filename)\
+                    .replace('[OUTPUT]', options.output_dir + '/profile_' + degree + '/' + os.path.basename(reads_filename))\
+                    .replace('[DEGREE]', degree).split()
+
+            output_file = open(options.output_dir + '/degree_' + degree + '/' + os.path.basename(reads_filename), 'w')
+            out_cmd(output_file.name, std_err_file.name, call_arr)
+            call(call_arr, stdout=output_file, stderr=std_err_file)
 
         # Profile compression using k-means.
         for profiles in options.profile_sizes.split(','):
@@ -298,6 +311,9 @@ def get_options():
     parser.add_option("-a", "--assemble", dest="assemble", help="Run assembly evaluation", action='store_true')
     parser.add_option("-p", "--preprocessing", dest="preprocessing", help="Run preprocessing tools evaluation", action='store_true')
     parser.add_option("-b", "--alignment", dest="alignment", help="Run alignment evaluation (using Bowtie2).", action='store_true')
+
+    # Polynomial regression specific options.
+    parser.add_option("--poly-degrees", dest="poly_degrees", help="Comma-separated list of polynomial degrees to use for regression.")
 
     # Profile-specific compression options.
[…]

### 2015-02-05 src/map_fastq.py

@@ -12,13 +12,15 @@ def qual_ord(phred_string, offset=33):
 
 def map_qualstring(qual_string):
     x, y = qual_ord(qual_string, offset=0)
-    coeffs = polyfit(x, y, 2)
+    coeffs = polyfit(x, y, int(sys.argv[3]))
     y_out = polyval(coeffs[::-1], x)
     return ''.join([chr(min(max(int(yy),35),74)) for yy in y_out])
 
-def coeffs_for_qualstring(qual_string, degree=5):
+def coeffs_for_qualstring(qual_string, degree=int(sys.argv[3])):
     x, y = qual_ord(qual_string, offset=0)
+    #print degree
     coeffs = polyfit(x, y, degree)
+    #print coeffs
     return coeffs
 
 def process(lines, pool):
@@ -48,17 +50,20 @@ with open(sys.argv[1], 'r') as infile:
     if len(lines) > 0:
         process(lines, pool)
 
+with open(sys.argv[1], 'r') as infile:
     # # use this block to write binary file with coefficients
-    # float_list = []
-    # lines = []
-    # for i_counter, line in enumerate(infile):
-    #     if (i_counter % 1000) == 0:
-    #         process_coeffs_only(lines, pool, float_list)
-    #         lines = []
-    #     lines.append(line.rstrip())
-    # if len(lines) > 0:
-    #     process_coeffs_only(lines, pool, float_list)
-    # with open('coefficients_binary.bin', 'wb') as output_file:
-    #     float_array = array('f', float_list)
-    #     float_array.tofile(output_file)
+    float_list = []
+    lines = []
+    for i_counter, line in enumerate(infile):
+        if (i_counter % 1000) == 0:
+            process_coeffs_only(lines, pool, float_list)
+            lines = []
+        lines.append(line.rstrip())
+    if len(lines) > 0:
+        process_coeffs_only(lines, pool, float_list)
+    with open(sys.argv[2] + '.bin', 'wb') as output_file:
+        #print float_list
+        #print len(float_list)
+        float_array = array('f', np.hstack(float_list))
+        float_array.tofile(output_file)
 

### 2015-02-05 src/profile_and_compress.py

@@ -51,7 +51,6 @@ center_qualstrings = [''.join([chr(max(35,int(x)+33)) for x in center]) for cent
 
 
 #pred = km.predict(quals)
-print 'starting writing'
 with open(sys.argv[1], 'r') as zzz:
     with open(sys.argv[2], 'w') as outfile_decomp:
          with open(sys.argv[2]+'.compressed', 'w') as outfile_comp:

### 2015-02-06 manuscript/q-compression.tex
More polishing of the intro.

@@
 has the potential for substantial improvement in data storage and
-transfer efficiency. Among the several challenges for compression of
-read data is dealing with different error profiles resulting from
-differences in underlying chemistries, signal detection and processing
-mechanisms, inherent biases, and other idiosyncratic properties of
-distinct high-throughput sequencing technologies. Sequence reads
-generated by instruments such as an Illumina HiSeq, the focus of this
-research, are characterized by having relatively few insertion and
-deletion errors, but substitution (miscall) errors are much more
-frequent and have context-specific patterns. These errors are
-non-uniformly distributed over the read length (e.g., error rates
-increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
-67.9\% of reads have low quality tails at the 3$^{\prime}$
-end~\cite{Minoche:2011km}).
+transfer efficiency.
 
-Quality values comprise a standard component of \textsc{fastq} format
-files~\cite{Cock:2010ve}.  At the level of sequence read the
-probability of error for each base-call is typically represented by
-\textsc{phred} quality value, which is defined as $Q =
+Quality values comprise a standard component of \textsc{fastq}
+files~\cite{Cock:2010ve}, a very common format for sequence read data.
+At the level of sequence read the probability of error for each
+base-call is typically represented by \textsc{phred} quality value,
+which is defined as $Q =
 -10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
@@
 represented with the \textsc{ascii} characters 33 to 126 (with some
-offset). These errors are represented a single quality value per
-base-call for Illumina sequence reads.
+offset). There is a single quality value per base-call for Illumina
+sequence reads.
 
-Quality values are typically used throughout bioinformatics pipelines.
-Among the most fundamental uses of sequence quality values is as part
-of the quality assessment and quality control (\textsc{qa/qc})
-processes prior to subsequent analysis steps. Quality control based on
-quality values generally includes two operations:
-\textit{i}.~filtering, the elimination of reads that on the whole do
-not meet quality standards, which reduces the total number of reads;
-and \textit{ii}.~trimming of low quality base-calls from reads, which
+Quality values can be used throughout bioinformatics pipelines.  Among
+the most fundamental uses of sequence quality values is as part of the
+quality assessment and quality control (\textsc{qa/qc}) processes
+prior to subsequent analysis steps. Quality control based on quality
+values generally includes two operations: \textit{i}.~filtering, the
+elimination of reads that on the whole do not meet arbitrary quality
+standards, which reduces the total number of reads; and
+\textit{ii}.~trimming of low quality base-calls from reads, which
 reduces the number total number of bases. Quality values can be used
@@
 alignment software, such as Bowtie2~\cite{Langmead:2012rw}, use
-quality values to weight mismatches between query and reference
+quality values to weight mismatches between read and reference
 sequences.  Software for detecting single nucleotide polymorphisms
-(\textsc{snp}s) can use quality values, and identified \textsc{snp}s
-with high-quality calls are deemed more reliable than those with
-low-quality calls, particularly in low coverage
-regions~\cite[e.g.,][]{McKenna:2010bh}.
+(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
+and identified \textsc{snp}s with high-quality calls are deemed more
+reliable than those with low-quality calls, particularly in low
+coverage regions.
 
@@
 on lossless compression of base calls~\cite[reviewed
-  in][]{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
-Although some recent work has included, or focused on, compression of
-quality
-values~\cite[e.g.,][]{Tembe:2010ys,Kozanitis:2011kl,asnani2012lossy,Hach:2012ys,Wan:2012kq,janin2013adaptive,Ochoa:2013rt,Canovas:2014fr,zhou2014compression,DBLP:conf/recomb/YuYB14}.
-Typical compression methods fail to utilize the biases inherent in
-existing sequencing technologies. Depending on the sequencing
-technology used, the base calling is more accurate in the earlier
-cycles, gradually losing quality in later
-cycles~\cite{Kozanitis:2011kl}. To this end, methods for compressing
-these quality sequences utilize these general quality
-trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
-that builds a fixed-state Markov model on adjacent quality scores and
-uses Huffman Coding for compression based on the transition
-probabilities.
+  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr},
+although some recent work has addressed compression of quality
+values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}.
+Among the several challenges for compression of read data is dealing
+with different error profiles resulting from differences in underlying
+chemistries, signal detection and processing mechanisms, inherent
+biases, and other idiosyncratic properties of distinct high-throughput
+sequencing technologies. Sequence reads generated by instruments such
+as an Illumina HiSeq, the focus of this research, are characterized by
+having relatively few insertion and deletion errors, but substitution
+(miscall) errors are much more frequent and have context-specific
+patterns. These errors are non-uniformly distributed over the read
+length (e.g., error rates increase up to $\sim$16$\times$ at the
+3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low quality tails
+at the 3$^{\prime}$ end~\cite{Minoche:2011km}).
 
@@
 is largely pragmatic with a focus on the use of quality values in
-subsequent analyses. From this perspective some absolute loss of
-information is deemed acceptable if the inferences from analyses are
-relatively unaffected. Therefore we here describe our research
-investigating some general strategies for compression of sequence read
-quality values. Our objective provide perspective on several
-strategies logically applied to Illumina sequence read quality values
-rather than to generate a robust high-quality software for
-use. Consistent with our perspective we place emphasis not only on the
-compression efficiency, but also the effect of lossy compression on
-typical subsequent analyses.  We examine three general classes of
-lossy compression methods -- binning, modeling, and profiling -- and
-consider an exemplar of each class.
-
-Lossy compression compromises representation accuracy for better space
-efficiency. Different lossy compression methods have different
-properties, therefore we need to evaluate a number of tradeoffs for
-each of our proposed methods. More importantly, we need to examine
-each approach in the context of typical bioinformatics tasks, such as
-quality control, \emph{de novo} genome assembly, and aligning the
-sequences to a given reference.
-
-One compression strategy is to look at the distribution of quality
-values and see if we can summarize the information contained within
-ranges of values. Specifically, we can reduce the dimensionality of
-the distribution of quality values by binning together adjacent
-quality values. We can select the number of bins depending on the
-desired amount of compression and retention of quality
-information. Taken to nearly the extreme, a base call can be
-considered either good or bad. This is the motivation of our approach,
-deemed quality value binning, where we observe the distribution of
-quality values across the read set and find a cutoff between good and
-bad bases.
-
-Another compression strategy involves simplifying the data through
-modeling of the quality values of a given sequence. By constructing a
-compact model of our quality values, we only have to store the
-parameters of the model to recreate the quality sequences.
-Intuitively, the simpler the model, the better the compression, but
-results in potentially higher differences compared to the original
-quality sequences. Another benefit of the model-based compression is
-the ability to decompress any particular quality value in the sequence
-with constant time during decompression. In our approach, we represent
-the quality values of a sequence as a polynomial function, but it is
-possible to use other models (such as other basis functions or
-splines).
-
-Profile-based compression utilizes the fact that in a vast set of
-quality sequences one can identify large classes of highly similar
-elements. If one could partition the data into such classes, each of
-those would be efficiently representable with a single consensus
-profile.
+subsequent analyses. From this perspective some loss of information is
+deemed acceptable if the inferences from analyses are relatively
+unaffected. Here we describe our research investigating lossy
+compression of sequence read quality values, specifically those
+associated with Illumina instruments, with the objective to provide
+some perspective on several strategies rather than to develop a robust
+high-quality software for use. Recognizing these properties of
+Illumina sequence reads motivates our exploration of three general
+classes of lossy compression methods -- binning, modeling, and
+profiling -- and consider an exemplar of each class. In addition to
+quantifying compression efficiency, we also assess the effects of
+quality value information loss resulting from compression on
+subsequent genomic analyses including read preprocessing (filtering
+and trimming), genome assembly, and read mapping.
 
@@
 of this approach is that it is completely data-dependent, and no
-assumptions regarding the distribution of the quality values need to be made.
+assumptions regarding the distribution of the quality values need to
+be made.
 
@@
 include no fixed costs, and marginal costs that are a function of the
-number of base-call quality values encoded times the cost of the
-encoding.
+number of base-call quality values times the cost of the encoding.
 
@@
 
-\section{Some Alternative Text}
-
-The following text is taken from a draft grant proposal, and the
-content overlaps considerably with the other text in this
-document. However, at least in some cases the wording of this
-alternative text might be preferable and serve as a replacement. In
-any event the entire draft manuscript has to be substantially revised.
-
-\subsubsection{Pilot Study: profile-based compression of quality 
-values}\label{sec:profile-based}
-
-Quality value patterns can be readily determined by clustering
-sequences of quality values and using the resulting cluster centers as
-representative profiles. In our pilot study we used $k$-means
-clustering, a vector quantization method that partitions a set of
-samples into $k$ sets that minimize within-cluster sum of
-squares~\cite{hastie_09_elements-of.statistical-learning}, as follows.
-
-\begin{enumerate}
-\item
-Randomly sample an arbitrary number of quality sequences as a training
-set.
-\item
-Perform $k$-means clustering on the sample, and use the cluster
-centers as profiles for the dataset.
-\item
-Replace each quality value sequence with the index value of the
-profile closest in Euclidean distance.
-\end{enumerate}
-
-Our pilot study results demonstrate that $k$ values as small as
-32--256 sufficiently represent patterns in Illumina read quality
-values such that performance in subsequent analyses is very close to
-uncompressed data (see Section \ref{sec:pilot-evaluation}). The
-resulting compressed quality component of the original set of
-\textsc{fastq} files consists of the $k$ quality profiles (the fixed
-size cost), and a single byte representing the assigned quality
-profile index for each read (the marginal size cost). Therefore the
-compression factor is proportional of read length: the marginal size
-cost of 1 byte/read for Illumina HiSeq read lengths of 100 bp requires
-0.01 bits/base-call, and MiSeq read lengths of 300 bp require 0.0033
-bits/base-call. The fixed cost of profiles of course reduces the
-compression ratio, but this fixed cost is amortized across reads in a
-set. Both fixed and marginal costs can be further reduced with
-subsequent compression of the index and profiles (e.g., using general
-compression algorithms). Our pilot study demonstrates that using 256
-profiles with subsequent bzip2 compression (using a Burrows-Wheeler
-block-sorting compression algorithm, and Huffman coding) represents
-quality values of Illumina 100 bp reads at 0.074 bits/base-call
-compared to the original data compressed with bzip2 only of 2.11
-bits/base-call for the fragment library and 2.70 bits/base-call for
-the short jump library. The very recently published
-Read-Quality-Sparsifier
-(\textsc{rqs};~\cite{DBLP:conf/recomb/YuYB14}), which apparently
-represents the current state-of-the-art for compressing quality
-values, achieved best case compression of 0.254 bits/base-call, and a
-mean compression of 1.841 bits/base-call. Whereas our profile method
-achieves $\sim$3.432$\times$ more compression than textsc{rqs} best
-performance, and $\sim$24.878$\times$ more than average
-performance. Note too that the compression ratio for \textsc{rqs}
-varies with the $k$-mer composition of the data, whereas profile-based
-compression is invariant to $k$-mer composition.
-
-\subsubsection{Evaluating compression effects on subsequent 
-analysis results}\label{sec:pilot-evaluation}
-
-Lossy compression compromises representation accuracy for better space
-efficiency. Different lossy compression methods have different
-properties, therefore we need to evaluate performance in a number of
-analysis contexts, and use multiple performance measures to
-appropriately evaluate lossy compression methods. Important
-bioinformatics contexts include typical tasks, such as quality control
-processing, \textit{de novo} genome assembly, and aligning reads to a
-reference. In our pilot study we again used Illumina reads from
-fragment and short-jump libraries of the \textit{Rhodobacter
-  sphaeroides} as test data, and assessed the effects of lossy
-compression on each of the aforementioned bioinformatics
-operations. The results for the profile-based compression follow.
-
-\paragraph{Aligning reads to a reference sequence}
-Another important analysis involving read data are the alignments of
-reads to a reference sequence. Such alignments are performed for
-genome annotation, quantification of gene expression in RNA-Seq
-studies, estimation of sequencing coverage, and other purposes. To
-evaluate the effect of compression on read alignment we used
-\textsc{bowtie} 2~\cite{Langmead:2012rw} with default settings and
-max- and min-mismatch parameters of \texttt{-mp 6,6} and \texttt{-mp
-  2,2} respectively. The proportion of reads mapped using
-compressed/decompressed data was extremely close to that of the
-uncompressed data: 0.999 for the fragment library reads, and 0.998 for
-the short-jump library reads.
-
 %\bibliographystyle{natbib}

### 2015-02-06 manuscript/q-compression.tex
Added additional species info to text.

--- 2015-02-04 manuscript/q-compression.tex
@@
 2,050,868 paired-end reads) and short-jump library (insert size of
-3,000 nt; 2,050,868 reads). The corresponding reference sequence was
+3,500 nt; 2,050,868 reads). The corresponding reference sequence was
 obtained from the NCBI RefSeq database (NC\_007488.1, NC\_007489.1,
@@
 
-\textit{Homo sapiens} \ldots
+\textit{Stapylococcus aureus} USA300 data from the \textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a fragment library (insert size of 180 nt; 1,294,104 paired-end reads) and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
+The corresponding reference sequence was obtained from the NCBI RefSeq database (NC\_010063.1, NC\_010079.1, NC\_012417.1).
 
-\textit{Genus speices} \ldots
+\textit{Homo sapiens} chromosome 14 data from the \textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a fragment library (insert size of 155 nt; 36,504,800 paired-end reads) and short-jump library (insert sizes ranging from  2283-2803 nt; 22,669,408 reads).
+The corresponding reference sequence was obtained from the NCBI RefSeq database (NC\_000014.8).
+
+\textit{TODO: MiSeq data set} \ldots
 

### 2015-02-06 manuscript/q-compression.tex
a few small changes

--- 2015-02-06 manuscript/q-compression.tex
@@
 Lossy compression of quality information can greatly decrease storage
-and memory requirements with little discernable effects on subsequent
+and memory requirements with little discernible effects on subsequent
 analysis results. The four compression strategies in this study were

### 2015-02-08 src/compress.py
Added R version of polynomial regression.

@@ -42,7 +42,7 @@ def compress(options):
 
     # Basic command line scripts to run the individual compression schemes.
     GB_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -c 2 -b 0 "
-    POLY_REGRESSION_CMD = "python src/map_fastq.py [READ] [OUTPUT] [DEGREE]"
+    POLY_REGRESSION_CMD = "Rscript src/poly_regression.R [READ] [OUTPUT] [DEGREE]"
     PROFILE_COMPRESSION_CMD = "python src/profile_and_compress.py [READ] [OUTPUT] [NUM_PROFILES] [TRAINING_SIZE] [OUTPUT].png "
     
     # Store which compression directories we created.
@@ -69,12 +69,11 @@ def compress(options):
             options.compressed_dirs.append('degree_' + degree)
 
             call_arr = POLY_REGRESSION_CMD.replace('[READ]', reads_filename)\
-                    .replace('[OUTPUT]', options.output_dir + '/profile_' + degree + '/' + os.path.basename(reads_filename))\
+                    .replace('[OUTPUT]', options.output_dir + '/degree_' + degree + '/' + os.path.basename(reads_filename))\
                     .replace('[DEGREE]', degree).split()
 
-            output_file = open(options.output_dir + '/degree_' + degree + '/' + os.path.basename(reads_filename), 'w')
-            out_cmd(output_file.name, std_err_file.name, call_arr)
-            call(call_arr, stdout=output_file, stderr=std_err_file)
+            out_cmd("", std_err_file.name, call_arr)
+            call(call_arr, stderr=std_err_file)
 
         # Profile compression using k-means.
         for profiles in options.profile_sizes.split(','):

### 2015-02-08 src/poly_regression.R

@@ -1,0 +1,44 @@
+#source("http://bioconductor.org/biocLite.R")
+#biocLite("ShortRead")
+library("ShortRead")
+
+args <- commandArgs(trailingOnly = TRUE)
+
+# Input FASTQ.
+fl <- file.path(args[1])
+
+# Output FASTQ
+output_file <- file(args[2], "w")
+
+# Set the min-max quality values.
+min_max <- function(x) {
+  if (x > 74)
+    return(74)
+  else if (x < 35)
+    return(35)
+  else
+    return(x)
+}
+
+## Iterating over the FASTQ file.
+f <- FastqStreamer(fl, 1)
+while (length(fq <- yield(f))) {
+  
+  # Get the quality values.
+  quals = as.numeric(as.vector(quality(fq)[[1]]))
+  x = seq(1,length(quals))
+  
+  # Fit the polynomial function.
+  fit = lm(quals ~ poly(x, strtoi(args[3]), raw=TRUE))
+  
+  # Predict the quality values using the above equation.
+  predicted_quals = rawToChar(as.raw(unlist(lapply(round(predict(fit)), min_max))))
+  #lines(x, round(predict(fit)))
+
+  # Write out the fastq file to disk.
+  # TODO: Replace with writeFastQ.
+  writeLines(c(paste("@",as.vector(id(fq)), sep = ""), as.vector(sread(fq)), "+", predicted_quals), con=output_file)
+}
+
+close(output_file)
+close(f)

### 2015-02-09 src/compress.py
Added profile compression tool using R. Removed scipy/numpy dependencies.

@@ -43,8 +43,8 @@ def compress(options):
     # Basic command line scripts to run the individual compression schemes.
     GB_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -c 2 -b 0 "
     POLY_REGRESSION_CMD = "Rscript src/poly_regression.R [READ] [OUTPUT] [DEGREE]"
-    PROFILE_COMPRESSION_CMD = "python src/profile_and_compress.py [READ] [OUTPUT] [NUM_PROFILES] [TRAINING_SIZE] [OUTPUT].png "
-    
+    PROFILE_COMPRESSION_CMD = "Rscript src/profile.R [READ] [OUTPUT] [TRAINING_SIZE] [NUM_PROFILES]"
+
     # Store which compression directories we created.
     options.compressed_dirs = []
     options.compressed_dirs.append('original')

### 2015-02-09 src/poly_regression.R

@@ -12,10 +12,10 @@ output_file <- file(args[2], "w")
 
 # Set the min-max quality values.
 min_max <- function(x) {
-  if (x > 74)
-    return(74)
-  else if (x < 35)
-    return(35)
+  if (x > 73)
+    return(73)
+  else if (x < 33)
+    return(33)
   else
     return(x)
 }

### 2015-02-09 src/profile.R
Added missing R file.

@@ -1,0 +1,44 @@
+#source("http://bioconductor.org/biocLite.R")
+#biocLite("ShortRead")
+library("ShortRead")
+
+args <- commandArgs(trailingOnly = TRUE)
+
+output_file <- file(args[2], "a")
+
+training_set_size <- strtoi(args[3])
+
+num_profiles <- strtoi(args[4])
+
+# Taken from:
+# http://stats.stackexchange.com/questions/78322/is-there-a-function-in-r-that-takes-the-centers-of-clusters-that-were-found-and
+assign_to_cluster <- function(x, centers) {
+  # compute squared euclidean distance from each sample to each cluster center
+  tmp <- sapply(seq_len(nrow(x)),
+                function(i) apply(centers, 1,
+                                  function(v) sum((x[i, ]-v)^2)))
+  max.col(-t(tmp))  # find index of min distance
+}
+
+# Read an entire fastq file
+records <- readFastq(args[1])
+
+# Take a training set, where we will run k-means on.
+training_set = sample(records, size = training_set_size)
+
+original_reads <- as(quality(records),'matrix')
+
+# Run k-means.
+mat = as(quality(training_set), 'matrix')
+fit <- kmeans(mat, centers = num_profiles, iter.max = 600)
+
+# Assign the original sequences to their closest cluster centers.
+cluster_assignments = assign_to_cluster(original_reads, fit$centers)
+
+# Write out each FASTQ record with its new profile quality values.
+for (num in seq(1, length(records))) {
+  predicted_quals = rawToChar(as.raw(round(fit$centers[cluster_assignments[num],] + 33)))
+  writeLines(c(paste("@",as.vector(id(records[num])), sep = ""), as.vector(sread(records[num])), "+", as.vector(predicted_quals)), con=output_file)
+}
+
+close(output_file)

### 2015-02-10 src/compress.py
Added parallel poly regression.

@@ -42,7 +42,7 @@ def compress(options):
 
     # Basic command line scripts to run the individual compression schemes.
     GB_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -c 2 -b 0 "
-    POLY_REGRESSION_CMD = "Rscript src/poly_regression.R [READ] [OUTPUT] [DEGREE]"
+    POLY_REGRESSION_CMD = "Rscript src/poly_regression_parallel.R [READ] [OUTPUT] [DEGREE] [NUM_THREADS]"
     PROFILE_COMPRESSION_CMD = "Rscript src/profile.R [READ] [OUTPUT] [TRAINING_SIZE] [NUM_PROFILES]"
 
     # Store which compression directories we created.
@@ -70,7 +70,8 @@ def compress(options):
 
             call_arr = POLY_REGRESSION_CMD.replace('[READ]', reads_filename)\
                     .replace('[OUTPUT]', options.output_dir + '/degree_' + degree + '/' + os.path.basename(reads_filename))\
-                    .replace('[DEGREE]', degree).split()
+                    .replace('[DEGREE]', degree)\
+                    .replace('[NUM_THREADS]', options.threads).split()
 
             out_cmd("", std_err_file.name, call_arr)
             call(call_arr, stderr=std_err_file)

### 2015-02-10 src/poly_regression_parallel.R

@@ -1,0 +1,78 @@
+#source("http://bioconductor.org/biocLite.R")
+#biocLite("ShortRead")
+library("ShortRead")
+library(parallel)
+
+args <- commandArgs(trailingOnly = TRUE)
+
+# Input FASTQ.
+fl <- file.path(args[1])
+
+# Output FASTQ.
+output_file <- file(args[2], "w")
+
+# Polynomial degree.
+degree <- args[3]
+
+# Set the min-max quality values.
+min_max <- function(x) {
+  if (x > 73)
+    return(73)
+  else if (x < 33)
+    return(33)
+  else
+    return(x)
+}
+
+poly_regression <- function(quals) {
+  # Get the quality values. Use a hardcoded quality offset.
+  quals = quals + 33
+  x = seq(1,length(quals))
+  
+  # Fit the polynomial function.
+  fit = lm(quals ~ poly(x, strtoi(degree), raw=TRUE))
+  
+  # Predict the quality values using the above equation.
+  predicted_quals = rawToChar(as.raw(unlist(lapply(round(predict(fit)), min_max))))
+  return(predicted_quals)
+}
+
+
+# Read an entire fastq file
+records <- readFastq(args[1])
+
+original_reads <- as(quality(records),'matrix')
+
+threads <- args[4]
+
+cl <- makeCluster(strtoi(threads))
+clusterExport(cl=cl, varlist=c("min_max", "degree"))
+
+# ptm <- proc.time()
+results <- parApply(cl = cl, original_reads, MARGIN = 1, poly_regression)
+
+# proc.time() - ptm
+# poly_regression
+# fptm <- proc.time()
+# results2 <- apply(original_reads, MARGIN = 1, poly_regression)
+# proc.time() - ptm
+
+stopCluster(cl)
+
+## Iterating over the FASTQ file again and print out the new quality values.
+f <- FastqStreamer(fl, 1)
+counter <- 1
+while (length(fq <- yield(f))) {
+  
+  # Predict the quality values using the above equation.
+  predicted_quals <- results[[counter]]
+  
+  # Write out the fastq file to disk.
+  # TODO: Replace with writeFastQ.
+  writeLines(c(paste("@",as.vector(id(fq)), sep = ""), as.vector(sread(fq)), "+", predicted_quals), con=output_file)
+  
+  counter <- counter + 1
+}
+
+close(output_file)
+close(f)

### 2015-02-10 manuscript/q-compression.tex

@@
 
-\textit{Stapylococcus aureus} USA300 data from the \textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a fragment library (insert size of 180 nt; 1,294,104 paired-end reads) and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
-The corresponding reference sequence was obtained from the NCBI RefSeq database (NC\_010063.1, NC\_010079.1, NC\_012417.1).
+\textit{Stapylococcus aureus} USA300 data from the
+\textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a
+fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
+and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
+The corresponding reference sequence was obtained from the NCBI RefSeq
+database (NC\_010063.1, NC\_010079.1, NC\_012417.1).
 
-\textit{Homo sapiens} chromosome 14 data from the \textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a fragment library (insert size of 155 nt; 36,504,800 paired-end reads) and short-jump library (insert sizes ranging from  2283-2803 nt; 22,669,408 reads).
-The corresponding reference sequence was obtained from the NCBI RefSeq database (NC\_000014.8).
+\textit{Homo sapiens} chromosome 14 data from the
+\textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a
+fragment library (insert size of 155 nt; 36,504,800 paired-end reads)
+and short-jump library (insert sizes ranging from 2283-2803 nt;
+22,669,408 reads).  The corresponding reference sequence was obtained
+from the NCBI RefSeq database (NC\_000014.8).
 
@@
 As a measure of information loss we use mean squared error
-(\textsc{mse}) as aloss function, and define it as
+(\textsc{mse}) as a loss function, and define it as
 $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
@@
 
-%\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
+\bibliographystyle{natbib}
+\bibliographystyle{achemnat}
+\bibliographystyle{plainnat}
+\bibliographystyle{abbrv}
+\bibliographystyle{bioinformatics}
 %
@@
 %
+
 \bibliography{compression}

### 2015-02-10 src/poly_regression_parallel.R
Fixed [I hope] write bottleneck.

@@ -30,13 +30,12 @@ poly_regression <- function(quals) {
   x = seq(1,length(quals))
   
   # Fit the polynomial function.
-  fit = lm(quals ~ poly(x, strtoi(degree), raw=TRUE))
+  fit = lm(quals ~ poly(x, strtoi(degree)))
   
   # Predict the quality values using the above equation.
   predicted_quals = rawToChar(as.raw(unlist(lapply(round(predict(fit)), min_max))))
   return(predicted_quals)
 }
-
 
 # Read an entire fastq file
 records <- readFastq(args[1])
@@ -48,10 +47,10 @@ threads <- args[4]
 cl <- makeCluster(strtoi(threads))
 clusterExport(cl=cl, varlist=c("min_max", "degree"))
 
-# ptm <- proc.time()
-results <- parApply(cl = cl, original_reads, MARGIN = 1, poly_regression)
+#ptm <- proc.time()
+results <- parRapply(cl = cl, original_reads, poly_regression)
 
-# proc.time() - ptm
+#print(proc.time() - ptm)
 # poly_regression
 # fptm <- proc.time()
 # results2 <- apply(original_reads, MARGIN = 1, poly_regression)
@@ -59,20 +58,24 @@ results <- parApply(cl = cl, original_reads, MARGIN = 1, poly_regression)
 
 stopCluster(cl)
 
-## Iterating over the FASTQ file again and print out the new quality values.
-f <- FastqStreamer(fl, 1)
+con  <- file(args[1], open = "r")
 counter <- 1
-while (length(fq <- yield(f))) {
+while (length(oneLine <- readLines(con, n = 1, warn = FALSE)) > 0) {
+  # Write the header...
+  writeLines(oneLine, con=output_file)
+  oneLine <- readLines(con, n = 1, warn = FALSE)
   
-  # Predict the quality values using the above equation.
-  predicted_quals <- results[[counter]]
+  # ... then the DNA sequence ...
+  writeLines(oneLine, con=output_file)
+  oneLine <- readLines(con, n = 1, warn = FALSE)
   
-  # Write out the fastq file to disk.
-  # TODO: Replace with writeFastQ.
-  writeLines(c(paste("@",as.vector(id(fq)), sep = ""), as.vector(sread(fq)), "+", predicted_quals), con=output_file)
+  # ... then the + ...
+  writeLines(oneLine, con=output_file)
+  oneLine <- readLines(con, n = 1, warn = FALSE)
   
[…]

### 2015-02-11 src/compress.py
Parallelized profile compression.

@@ -43,7 +43,7 @@ def compress(options):
     # Basic command line scripts to run the individual compression schemes.
     GB_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -c 2 -b 0 "
     POLY_REGRESSION_CMD = "Rscript src/poly_regression_parallel.R [READ] [OUTPUT] [DEGREE] [NUM_THREADS]"
-    PROFILE_COMPRESSION_CMD = "Rscript src/profile.R [READ] [OUTPUT] [TRAINING_SIZE] [NUM_PROFILES]"
+    PROFILE_COMPRESSION_CMD = "Rscript src/profile_parallel.R [READ] [OUTPUT] [TRAINING_SIZE] [NUM_PROFILES] [NUM_THREADS]"
 
     # Store which compression directories we created.
     options.compressed_dirs = []
@@ -84,7 +84,8 @@ def compress(options):
             call_arr = PROFILE_COMPRESSION_CMD.replace('[READ]', reads_filename)\
                     .replace('[OUTPUT]', options.output_dir + '/profile_' + profiles + '/' + os.path.basename(reads_filename))\
                     .replace('[NUM_PROFILES]', profiles)\
-                    .replace('[TRAINING_SIZE]', options.training_size).split()
+                    .replace('[TRAINING_SIZE]', options.training_size)\
+                    .replace('[NUM_THREADS]', options.threads).split()
 
             out_cmd("", std_err_file.name, call_arr)
             call(call_arr, stderr=std_err_file)

### 2015-02-11 src/good_bad_coding.py

@@ -33,7 +33,8 @@ def main():
     for line in reads_file:
         line_counter += 1
 
-        if line_counter % 4 == 0:
+        # Read every 10th line.
+        if line_counter % 40 == 0:
             
             raw_values = map(ord, line.strip())
             normalized_values = map(lambda x: int(x) - 33, raw_values)

### 2015-02-11 src/poly_regression.R

@@ -33,7 +33,7 @@ while (length(fq <- yield(f))) {
   
   # Predict the quality values using the above equation.
   predicted_quals = rawToChar(as.raw(unlist(lapply(round(predict(fit)), min_max))))
-  #lines(x, round(predict(fit)))
+  #lines(x, round(predict(fit)))0
 
   # Write out the fastq file to disk.
   # TODO: Replace with writeFastQ.

### 2015-02-11 src/profile_parallel.R

@@ -1,0 +1,67 @@
+#source("http://bioconductor.org/biocLite.R")
+#biocLite("ShortRead")
+library("ShortRead")
+library(parallel)
+
+args <- commandArgs(trailingOnly = TRUE)
+output_file <- file(args[2], "a")
+
+training_set_size <- strtoi(args[3])
+
+num_profiles <- strtoi(args[4])
+
+# Taken from:
+# http://stats.stackexchange.com/questions/78322/is-there-a-function-in-r-that-takes-the-centers-of-clusters-that-were-found-and
+assign_to_cluster <- function(x) {
+  # compute squared euclidean distance from each sample to each cluster center
+   tmp <- apply(centers, 1, function(v) sum((x-v)^2))
+   return(max.col(-t(tmp)))  # find index of min distance
+}
+
+# Read an entire fastq file
+records <- readFastq(args[1])
+
+# Take a training set, where we will run k-means on.
+training_set = sample(records, size = training_set_size)
+
+original_reads <- as(quality(records),'matrix')
+
+# Run k-means.
+mat = as(quality(training_set), 'matrix')
+fit <- kmeans(mat, centers = num_profiles, iter.max = 600)
+centers <- fit$centers
+
+threads <- args[5]
+
+cl <- makeCluster(strtoi(threads))
+clusterExport(cl=cl, varlist=c("assign_to_cluster", "centers"))
+
+#ptm <- proc.time()
+# Assign the original sequences to their closest cluster centers.
+cluster_assignments <- parRapply(cl = cl, original_reads, assign_to_cluster)
+
+#print(proc.time() - ptm)
+stopCluster(cl)
+
+con  <- file(args[1], open = "r")
+counter <- 1
+while (length(oneLine <- readLines(con, n = 1, warn = FALSE)) > 0) {
+  # Write the header...
+  writeLines(oneLine, con=output_file)
+  oneLine <- readLines(con, n = 1, warn = FALSE)
+  
+  # ... then the DNA sequence ...
+  writeLines(oneLine, con=output_file)
+  oneLine <- readLines(con, n = 1, warn = FALSE)
+  
+  # ... then the + ...
+  writeLines(oneLine, con=output_file)
+  oneLine <- readLines(con, n = 1, warn = FALSE)
+  
+  # ... and finally the new quality values.
+  #writeLines(results[[counter]], con=output_file)
[…]

### 2015-04-07 manuscript/q-compression.tex
Added placeholder plots.

@@
 
+\begin{figure}[!tpb]%figure2
+\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
+\caption{Quality profiles obtained by $k$-means clustering on the
+  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
+  set using $k$ = 128, with each row corresponding to a quality
+  profile. Dark to light colors represent low to high quality
+  values. It is readily visible that the two most distinctive features
+  of quality profiles is their drop-off position and average overall
+  quality. One can also see sporadic low-position values in a handful
+  of profiles, likely capturing intermittent problems in the
+  sequencing process affecting thousands of reads at a
+  time.}\label{fig:profiles_128}
+\end{figure}
+
+\begin{figure*}[!tb]%figure2
+%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
+\centerline{\includegraphics[width=7in]{compression_results.pdf}}
+\caption{Mean squared error versus bits/base-call for different
+  compression methods applied to the fragment library from
+  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
+  encoding. P32 and P256: Quality profile compression with 32 and 256
+  profiles. R0-R5: Polynomial regression-based compression using 0-5
+  degrees. Dashed line corresponds to lossless bzip compression of
+  quality values. \em}\label{fig:mse_vs_bpbp_frag}
+\end{figure*}
+
 As large sets of quality strings show similar trends of quality over
@@
 costs that are a function of the number of reads encoded.
-
-\begin{figure}[!tpb]%figure2
-\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
-\caption{Quality profiles obtained by $k$-means clustering on the
-  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
-  set using $k$ = 128, with each row corresponding to a quality
-  profile. Dark to light colors represent low to high quality
-  values. It is readily visible that the two most distinctive features
-  of quality profiles is their drop-off position and average overall
-  quality. One can also see sporadic low-position values in a handful
-  of profiles, likely capturing intermittent problems in the
-  sequencing process affecting thousands of reads at a
-  time.}\label{fig:profiles_128}
-\end{figure}
 
@@
 
-\begin{figure}[!tpb]%figure2
-\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
-\caption{Mean squared error versus bits/base-call for different
-  compression methods applied to the fragment library from
-  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
-  encoding. P32 and P256: Quality profile compression with 32 and 256
-  profiles. R0-R5: Polynomial regression-based compression using 0-5
-  degrees. Dashed line corresponds to lossless bzip compression of
-  quality values. \em}\label{fig:mse_vs_bpbp_frag}
-\end{figure}
-
-\begin{figure}[!tpb]%figure2
-\centerline{\includegraphics[width=3.65in]{mse_short.png}}
-\caption{Mean squared error versus bits/base-call for for different
-  compression methods applied to the short-jump library from
-  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
-  encoding. P32 and P256: Quality profile compression with 32 and 256
-  profiles. R0-R5: Polynomial regression-based compression using 0-5
-  degrees. Dashed line corresponds to lossless bzip compression of
-  quality values. \em}\label{fig:mse_vs_bpbp_jump}
-\end{figure}
+%\begin{figure}[!tpb]%figure3
+%\centerline{\includegraphics[width=3.65in]{mse_short.png}}
+%\caption{Mean squared error versus bits/base-call for for different
+%  compression methods applied to the short-jump library from
+%  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
+%  encoding. P32 and P256: Quality profile compression with 32 and 256
+%  profiles. R0-R5: Polynomial regression-based compression using 0-5
+%  degrees. Dashed line corresponds to lossless bzip compression of
+%  quality values. \em}\label{fig:mse_vs_bpbp_jump}
+%\end{figure}
 
@@
 
-\begin{table*}[!tpb]
-\centering
-\caption[]{Rankings of different assemblies produced using proposed
+\begin{figure}[!tbp]%figure2
+%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
+\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
+\caption{Rankings of different assemblies produced using proposed
   compression methods. Assemblies were constructed using
@@
   by their overall rank (determined by the sum of individual
-  rankings).}
-{\begin{tabular}{lcccccccc}
-& overall & & & & & & \textsc{lap} & \textsc{lap} \\
-compression            & rank & \textsc{snp}s & inversions & relocations & translocations & indels & fragment & short-jump \\
-\hline
-none                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
-profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
-profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
-pegression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
-regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
-2-bin	               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
-regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
-regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
-regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
-regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
-\\ \hline
-\end{tabular}
-}
+  rankings). \em}
   \label{tab:assembly_ranks}
-\end{table*}
+\end{figure}
 
@@
 \centering
+\caption[]{Quality filtering/trimming for the uncompressed and
+  decompressed \textsc{fastq} files. Numbers correspond to the
+  number of reads that are kept or discarded after applying
+  compression with respect to the original, uncompressed values.}
+\label{tab:quality_control}
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
+ & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
+  \cline{2-12}
+& kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
+{\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
+\cline{2-12}
+ & sum & 1025434 &   0 &   0 & 1025434 & 872917 & 152517 & 729824 & 295610 & 880724 & 144710 \\ 
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
+ & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
+\cline{2-12}
+& kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
+{\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
+\cline{2-12}
+&  sum & 882375 & 143059 & 888940 & 136494 & 886623 & 138811 & 891944 & 133490 & 894966 & 130468 \\
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
+ & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
+\cline{2-12}
+& kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
+{\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
+\cline{2-12}
+&  sum & 897200 & 128234 & 892136 & 133298 & 892537 & 132897 & 892820 & 132614 & 889118 & 136316 \\ 
+\end{tabular}
+\end{table*}
+
+\begin{table*}[!tbhp]
+\centering
 \caption[]{Mapping results of decompressed \textsc{fastq} files
@@
   compression.}
-\begin{tabular}{llcccc}
-    & & \multicolumn{4}{c}{percentage of reads mapped}\\
-    \cline{3-6}
-	library & compression strategy & total & shared & uncompressed only & compressed only \\ \hline
-	& maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
-	&minimum quality & 100 & 100 & 0 & 0 \\
-	&2-bin & 99.92 & 99.89 & 0.11 & 0.03 \\
-	&regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
-	&regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
-	&regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
-	fragment&regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
-	&regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
-	&regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
-	&profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
-	&profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
+& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
+\cline{2-12}
+& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
+{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
+\cline{2-12}
+& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
+\end{tabular}
 
-\\ \hline \\
+\bigskip
 
-	&maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
-	&minimum quality & 100 & 100 & 0 & 0 \\ 
-	&2-bin & 97.38 & 97.38 & 2.62 & 0 \\
-	&regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
-	&regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
-	&regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
-	short-jump&regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
-	&regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
-	&regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
-	&profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
-	&profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
+& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
+\cline{2-12}
+& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
+{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
+\cline{2-12}
+& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
+& &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
+\cline{2-12}
+& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
+{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
+\cline{2-12}
+& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\
+
 \end{tabular}
@@
 \end{table*}
-
-\begin{table}[!tbhp]
-\centering
-\caption[]{Quality filtering/trimming for the uncompressed and
-  decompressed \textsc{fastq} files. Numbers correspond to the
-  proportion of bases/reads that are missed or gained after applying
-  compression with respect to the uncompressed values.}
-\label{tab:quality_control}
-\begin{tabular}{lcccc}
-	 & \multicolumn{2}{c}{fragments} & \multicolumn{2}{c}{short-Jump} \\
-	 & \% bases & \% reads & \% bases & \% reads \\ \hline
-	2-bin & 98.15 & 98.78 & 113.09 & 103.7 \\ 
-	regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
-	regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
-	regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
-	regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
-	regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
-	regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
-	profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
-	profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
-\end{tabular}
-\end{table}
 

### 2015-04-07 manuscript/q-compression.tex
Fixed column labeling issue in mapping results table (Table 2).

@@
 &  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
-& &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
+& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
 \cline{2-12}

### 2015-04-07 manuscript/q-compression.tex
Added data sets.

--- 2015-04-07 manuscript/q-compression.tex
@@
 %\usepackage{doi}
+\usepackage{url}
 
@@
 
-\textit{TODO: MiSeq data set} \ldots
+\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded from \url{http://www.illumina.com/systems/miseq/scientific_data.html}, which are generated from a
+fragment library (insert size of 180bp nt; 1,145,8940 paired-end reads). 
+The corresponding reference sequence was obtained
+from the NCBI RefSeq database (NC\_000913.2).
 

### 2015-04-08 manuscript/q-compression.tex
Adjusted citations to accommodate the change from number to Harvard (author year) format. This was mostly changing \cite to \citep where appropriate. A few other small changes as well.

--- 2015-04-07 manuscript/q-compression.tex
@@
 Quality values comprise a standard component of \textsc{fastq}
-files~\cite{Cock:2010ve}, a very common format for sequence read data.
+files~\citep{Cock:2010ve}, a very common format for sequence read data.
 At the level of sequence read the probability of error for each
@@
 which is defined as $Q =
--10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
+-10\,log_{10}P$~\citep{Ewing:1998ly}. Depending on the sequencing
 technology these quality values can range from 0 to 93, and are
@@
 assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
-alignment software, such as Bowtie2~\cite{Langmead:2012rw}, use
+alignment software, such as Bowtie2~\citep{Langmead:2012rw}, use
 quality values to weight mismatches between read and reference
@@
 on lossless compression of base calls~\cite[reviewed
-  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr},
-although some recent work has addressed compression of quality
-values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}.
-Among the several challenges for compression of read data is dealing
-with different error profiles resulting from differences in underlying
+  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,
+  Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
+addressed compression of quality
+values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,
+  janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
+  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}.  Among the
+several challenges for compression of read data is dealing with
+different error profiles resulting from differences in underlying
 chemistries, signal detection and processing mechanisms, inherent
@@
 3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low quality tails
-at the 3$^{\prime}$ end~\cite{Minoche:2011km}).
+at the 3$^{\prime}$ end~\citep{Minoche:2011km}).
 
@@
 between the first and third quartile. 1.5 $\times$ IQR is the value
-used by Tukey's box plot~\cite{mcgill1978variations}. The main benefit
-of this approach is that it is completely data-dependent, and no
-assumptions regarding the distribution of the quality values need to
-be made.
+used by Tukey's box plot~\citep{mcgill1978variations}. The main
+benefit of this approach is that it is completely data-dependent, and
+no assumptions regarding the distribution of the quality values need
+to be made.
 
 With 2-bin encoding binary encoding is possible, allowing us to use a
-single bit to represent the quality of a base instead of the standard 8
-bits used to store quality values in \textsc{ascii}. An additional
+single bit to represent the quality of a base instead of the standard
+8 bits used to store quality values in \textsc{ascii}. An additional
 benefit of 2-bin encoding is the potential for increased adjacency of
@@
 effectiveness of subsequent compression using established
-algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,DBLP:journals/tit/ZivL78}.
+algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
+  DBLP:journals/tit/ZivL78}.
 
@@
 set of samples into $k$ sets that minimize within-cluster sum of
-squares~\cite{macqueen1967some}. Using a random subset of read quality
+squares~\citep{macqueen1967some}. Using a random subset of read quality
 values, a compression method can use the computed cluster centers as
@@
 minimum provided by the Python package scikit-learn
-0.15~\cite{pedregosa2011scikit}.
+0.15~\citep{pedregosa2011scikit}.
 
@@
 
-We used several Illumina sequence read data sets in this research as
-follows.
+We used several Illumina sequence read data sets in this research,
+which are taken from data from the \textsc{gage} (Genome Assembly
+Gold-Standard Evaluations)~\citep{Salzberg:2012rc} except as
+noted. These data sets are as follows.
 
-\textit{Rhodobacter sphaeroides} 2.4.1 data from the \textsc{gage}
-(Genome Assembly Gold-Standard Evaluations)~\cite{Salzberg:2012rc},
-which are generated from a fragment library (insert size of 180 nt;
-2,050,868 paired-end reads) and short-jump library (insert size of
-3,500 nt; 2,050,868 reads). The corresponding reference sequence was
-obtained from the NCBI RefSeq database (NC\_007488.1, NC\_007489.1,
-NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
+\textit{Rhodobacter sphaeroides} 2.4.1, which are generated from a
+fragment library (insert size of 180 nt; 2,050,868 paired-end reads)
+and short-jump library (insert size of 3,500 nt; 2,050,868 reads). The
+corresponding reference sequence was obtained from the NCBI RefSeq
+database (NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1,
+NC\_007494.1, NC\_009007.1, NC\_009008.1).
 
-\textit{Stapylococcus aureus} USA300 data from the
-\textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a
+\textit{Stapylococcus aureus} USA300, which are generated from a
 fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
@@
 
-\textit{Homo sapiens} chromosome 14 data from the
-\textsc{gage}~\cite{Salzberg:2012rc}, which are generated from a
+\textit{Homo sapiens} chromosome 14 data, which are generated from a
 fragment library (insert size of 155 nt; 36,504,800 paired-end reads)
@@
 on quality control steps of read filtering and trimming, which were
-performed using Sickle~\cite{sickle}, quantify read characteristics
-with PrinSeq~\cite{Schmieder:2011gd}, and make comparison to
+performed using Sickle~\citep{sickle}, quantify read characteristics
+with PrinSeq~\citep{Schmieder:2011gd}, and make comparison to
 uncompressed data.
@@
 statistics, log average read probability
-(\textsc{lap})~\cite{Ghodsi:2013hb}, and a collection of
+(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
 reference-based metrics. The contiguity statistics include number of
@@
 software that makes use quality values in the assembly process:
-\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
+\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
 settings and 32 threads.
@@
 \textsc{lap} values because the \textsc{lap} value is influenced by
-total coverage~\cite{Ghodsi:2013hb}.
+total coverage~\citep{Ghodsi:2013hb}.
 
@@
 \ref{tab:aligner}. Additionally, to monitor the effect of quality
-values on mapping in general, Bowtie2 was tweaked with the maximum and
-minimum mismatch penalty equivalent to maximum and minimum quality
-scores (with parameters: --mp 6,6 and --mp 2,2 respectively).
+values on mapping in general, Bowtie2 was adjusted so that the maximum
+and minimum mismatch penalty were equivalent to maximum and minimum
+quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).
 
@@
 related lossy compression tools, such as
-\textsc{SlimGene}~\cite{Kozanitis:2011kl} and
-\textsc{QualComp}~\cite{Ochoa:2013rt}, needs to be
+\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
+\textsc{QualComp}~\citep{Ochoa:2013rt}, needs to be
 performed. Additionally, other types of sequencing data can be
@@
 have examined here, the distribution of erroneous bases is relatively
-uniform~\cite{Ferrarini:2013vf}. Koren et al.~\cite{Koren:2013ye} have
-shown that the assembly complexity of bacterial genomes can be greatly
-simplified, producing near complete genome assemblies, by utilizing a
-single run of these long reads. If long read sequencing technologies
-such as PacBio become more widely adopted, it would be of huge benefit
-to examine the potential of lossy compression algorithms on not only
-the quality values, but the biological sequencing data themselves.
+uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
+genomes can be greatly simplified, producing near complete genome
+assemblies, by utilizing a single run of these long
+reads~\citep{Koren:2013ye}. If long read sequencing technologies such
+as PacBio become more widely adopted, it would be of huge benefit to
+examine the potential of lossy compression algorithms on not only the
+quality values, but the biological sequencing data themselves.
 

### 2015-04-09 manuscript/q-compression.tex
Revised a figure legend.

@@
   compression methods. Assemblies were constructed using
-  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
-  2.4.1. data set provided by
-  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
-  by their overall rank (determined by the sum of individual
-  rankings). \em}
+  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}.
+  Rankings above the median value are in cyan, those below in
+  magenta}
   \label{tab:assembly_ranks}

### 2015-04-09 manuscript/q-compression.tex
Added compression results.

@@
 \caption{Mean squared error versus bits/base-call for different
-  compression methods applied to the fragment library from
-  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
-  encoding. P32 and P256: Quality profile compression with 32 and 256
-  profiles. R0-R5: Polynomial regression-based compression using 0-5
-  degrees. Dashed line corresponds to lossless bzip compression of
-  quality values. \em}\label{fig:mse_vs_bpbp_frag}
+  compression methods applied to the
+  \textit{Rhodobacter sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries, and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus musculus} data sets. 2B: 2-bin
+  encoding. P64, P128, and P256: Quality profile compression with 64, 128, and 256
+  profiles. R0-R7: Polynomial regression-based compression using 0-7
+  degrees. Dashed line corresponds to lossless Bzip2 compression of
+  quality values. \em}\label{fig:mse_vs_bpbp}
 \end{figure*}
@@
 
+\textit{Mus musculus} data was downloaded from \url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}, which consisted of 18,828,274 reads of length 36.
+
 \subsection{Performance evaluation}
@@
 
-We compare the \textsc{mse} versus bits/base-call of the fragment and
-short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
-data set (Figures \ref{fig:mse_vs_bpbp_frag} and
-\ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values
+We compare the \textsc{mse} versus bits/base-call of the \textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens} chromosome 14, \textit{Escherichia coli} str.
+K-12 MG1655, and \textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}).
+We only include the fragment libraries for the \textit{Rhodobacter sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data sets, but the additional short-jump library results are available in the supplementary (TODO:TABLE).
+Storing the uncompressed quality values
 requires 1 byte per base-call because they are stored in
-\textsc{ascii} format; however, lossless compression of the values
-with BZip2 2.112 bits/base-call. 0-order polynomial regression and
-32-profile encodings have the lowest bits/base-call. 5th-order
-polynomial regression has the highest bits/base-call, but have among
-the lowest \textsc{mse}. As the order of the polynomial increases, the
+\textsc{ascii} format and is denoted by the dotted black line in the figure.
+The lossless compression of each data set using BZip2 ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored dotted lines.
+The compression methods tend to cluster together across the different data sets.
+Across all data sets, the 0-order polynomial regression, profile encodings, and QualComp have the lowest bits/base-call.
+
+QualComp with the rate parameter set to 100 bits/read has the lowest \textsc{mse}, but requires 10-15x more storage than the profile encoding methods for only a 2-3x reduction in \textsc{mse}.
+When QualComp's rate parameter is set to match the profile encoding methods, QualComp performs slightly worse in terms of \textsc{mse}.
+In the \textit{Rhodobacter sphaeroides} 2.4.1 fragment library, QualComp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29.
+Using 256-profile encoding requires less storage (0.079 bits/bp) and has a lower \textsc{mse} (11.85).
+
+As the order of the polynomial increases, the
 bits/base-call increase and the \textsc{mse} decreases at an
 exponential rate.
+The 7th-order polynomial regression has the highest bits/base-call and in the \textit{Mus musculus} data set, and requires more storage than the \textsc{ascii} original quality values.
+A 7th-order polynomial requires storing eight floating point values, resulting in 32 bytes per quality sequence.
+The read length of the \textit{Mus musculus} data set is only 26, so the 7th-order polynomial regression is storing six more bytes than necessary to losslessly encode the quality data.
 

### 2015-04-10 manuscript/q-compression.tex
Various small changes in text and a figure legend. More to do.

@@
 \section{Motivation:}
+
 The \textsc{fastq} file format has become the \emph{de facto} standard
@@
 Quality values comprise a standard component of \textsc{fastq}
-files~\citep{Cock:2010ve}, a very common format for sequence read data.
-At the level of sequence read the probability of error for each
+files~\citep{Cock:2010ve}, a very common format for sequence read
+data. At the level of sequence read the probability of error for each
 base-call is typically represented by \textsc{phred} quality value,
@@
 \caption{Mean squared error versus bits/base-call for different
-  compression methods applied to the
-  \textit{Rhodobacter sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries, and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus musculus} data sets. 2B: 2-bin
-  encoding. P64, P128, and P256: Quality profile compression with 64, 128, and 256
-  profiles. R0-R7: Polynomial regression-based compression using 0-7
-  degrees. Dashed line corresponds to lossless Bzip2 compression of
-  quality values. \em}\label{fig:mse_vs_bpbp}
+  compression methods applied to the \textit{Rhodobacter sphaeroides}
+  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
+  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
+    musculus} data sets. 2B --- 2-bin encoding; P$n$ 
+  --- profiling with $n$ profiles; R$n$ --- modeling with polynomial regression
+  models of degree $n$; Q$n$ --- QualComp with rate parameter 
+  of $n$. Dashed line corresponds to lossless Bzip2
+  compression of quality values, and dotted line corresponds to
+  original uncompressed data.}
+\label{fig:mse_vs_bpbp}
 \end{figure*}
@@
 set of samples into $k$ sets that minimize within-cluster sum of
-squares~\citep{macqueen1967some}. Using a random subset of read quality
-values, a compression method can use the computed cluster centers as
-read quality profiles. As the problem is NP-hard, we use a heuristic
-iterative refinement approach quickly converging to a locally optimal
-minimum provided by the Python package scikit-learn
+squares~\citep{macqueen1967some}. Using a random subset of read
+quality values, a compression method can use the computed cluster
+centers as read quality profiles. As the problem is NP-hard, we use a
+heuristic iterative refinement approach quickly converging to a
+locally optimal minimum provided by the Python package scikit-learn
 0.15~\citep{pedregosa2011scikit}.
@@
 
-\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded from \url{http://www.illumina.com/systems/miseq/scientific_data.html}, which are generated from a
-fragment library (insert size of 180bp nt; 1,145,8940 paired-end reads). 
-The corresponding reference sequence was obtained
-from the NCBI RefSeq database (NC\_000913.2).
+\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded
+from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
+which are generated from a fragment library (insert size of 180bp nt;
+1,145,8940 paired-end reads).  The corresponding reference sequence
+was obtained from the NCBI RefSeq database (NC\_000913.2).
 
-\textit{Mus musculus} data was downloaded from \url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}, which consisted of 18,828,274 reads of length 36.
+\textit{Mus musculus} data was downloaded from
+\url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}, which
+consisted of 18,828,274 reads of length 36.
 
@@
 
-We compare the \textsc{mse} versus bits/base-call of the \textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens} chromosome 14, \textit{Escherichia coli} str.
-K-12 MG1655, and \textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}).
-We only include the fragment libraries for the \textit{Rhodobacter sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data sets, but the additional short-jump library results are available in the supplementary (TODO:TABLE).
-Storing the uncompressed quality values
-requires 1 byte per base-call because they are stored in
-\textsc{ascii} format and is denoted by the dotted black line in the figure.
-The lossless compression of each data set using BZip2 ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored dotted lines.
-The compression methods tend to cluster together across the different data sets.
-Across all data sets, the 0-order polynomial regression, profile encodings, and QualComp have the lowest bits/base-call.
+We compare the \textsc{mse} versus bits/base-call of the
+\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
+chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
+\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
+only include the fragment libraries for the \textit{Rhodobacter
+  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
+sets, but the additional short-jump library results are available in
+the supplementary (TODO:TABLE). Storing the uncompressed quality
+values requires 1 byte per base-call because they are stored in
+\textsc{ascii} format and is denoted by the dotted black line in the
+figure.  The lossless compression of each data set using BZip2 ranges
+from 2.19 - 3.10 bits/base-call and is denoted by the colored dotted
+lines.  The compression methods tend to cluster together across the
+different data sets.  Across all data sets, the 0-order polynomial
+regression, profile encodings, and QualComp have the lowest
+bits/base-call.
 
-QualComp with the rate parameter set to 100 bits/read has the lowest \textsc{mse}, but requires 10-15x more storage than the profile encoding methods for only a 2-3x reduction in \textsc{mse}.
-When QualComp's rate parameter is set to match the profile encoding methods, QualComp performs slightly worse in terms of \textsc{mse}.
-In the \textit{Rhodobacter sphaeroides} 2.4.1 fragment library, QualComp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29.
-Using 256-profile encoding requires less storage (0.079 bits/bp) and has a lower \textsc{mse} (11.85).
+QualComp with the rate parameter set to 100 bits/read has the lowest
+\textsc{mse}, but requires 10-15x more storage than the profile
+encoding methods for only a 2-3x reduction in \textsc{mse}.  When
+QualComp's rate parameter is set to match the profile encoding
+methods, QualComp performs slightly worse in terms of \textsc{mse}.
+In the \textit{Rhodobacter sphaeroides} 2.4.1 fragment library,
+QualComp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of
+17.29.  Using 256-profile encoding requires less storage (0.079
+bits/bp) and has a lower \textsc{mse} (11.85).
 
-As the order of the polynomial increases, the
-bits/base-call increase and the \textsc{mse} decreases at an
-exponential rate.
-The 7th-order polynomial regression has the highest bits/base-call and in the \textit{Mus musculus} data set, and requires more storage than the \textsc{ascii} original quality values.
-A 7th-order polynomial requires storing eight floating point values, resulting in 32 bytes per quality sequence.
-The read length of the \textit{Mus musculus} data set is only 26, so the 7th-order polynomial regression is storing six more bytes than necessary to losslessly encode the quality data.
-
-%\begin{figure}[!tpb]%figure3
-%\centerline{\includegraphics[width=3.65in]{mse_short.png}}
-%\caption{Mean squared error versus bits/base-call for for different
-%  compression methods applied to the short-jump library from
-%  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
-%  encoding. P32 and P256: Quality profile compression with 32 and 256
-%  profiles. R0-R5: Polynomial regression-based compression using 0-5
-%  degrees. Dashed line corresponds to lossless bzip compression of
-%  quality values. \em}\label{fig:mse_vs_bpbp_jump}
-%\end{figure}
+As the order of the polynomial increases, the bits/base-call increase
+and the \textsc{mse} decreases at an exponential rate.  The 7th-order
+polynomial regression has the highest bits/base-call and in the
+\textit{Mus musculus} data set, and requires more storage than the
+\textsc{ascii} original quality values.  A 7th-degree polynomial
+requires storing eight floating point values, resulting in 32 bytes
+per quality sequence.  The read length of the \textit{Mus musculus}
+data set is only 26, so the 7th-degree polynomial regression is storing
+six more bytes than necessary for lossless encoding the quality data.
 
@@
 \caption[]{Quality filtering/trimming for the uncompressed and
-  decompressed \textsc{fastq} files. Numbers correspond to the
-  number of reads that are kept or discarded after applying
-  compression with respect to the original, uncompressed values.}
+  decompressed \textsc{fastq} files. Numbers correspond to the number
+  of reads that are kept or discarded after applying compression with
+  respect to the original, uncompressed values.}
 \label{tab:quality_control}

### 2015-04-13 manuscript/q-compression.tex
Added assembly results.

@@
 
-The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
-using the uncompressed reads outperforms all compression methods in
-terms of \textsc{lap}, NG50, relocations, translations, and inversions
-(Table \ref{tab:assembly_ranks}). Among the compression methods, the
-256- and 32-profile encoding had the highest rank, then the 4th-order
-polynomial regression, the 5th-order polynomial regression, the 2-bin
-encoding, and lastly, the 3rd through 0-order polynomials.
+No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set outperforms all others in all metrics (Table \ref{fig:assembly_ranks}).
+%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set using the uncompressed reads outperforms all compression methods in terms of \textsc{lap}, NG50, relocations, translations, and inversions.
+Among the compression methods, the
+256-, 64-, 128-profile encoding had the highest ranks, followed by QualComp with rates 10 bits/read, 30 bits/read, 100 bits/read, then 7th-order polynomial regression, followed by the QualComp with rate 6 bits/read, the 2-bin encoding, and lastly, the 3rd-order, 5th-order, and 0-order polynomials.
 
@@
 sequences. All compression methods other than 0-order polynomial
-regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the
+regression produce an N50 ranging from 3.17--3.22 Mbps (see Supplemental Table TODO). Despite the
 similar contiguity statistics, the different compression methods vary
@@
 
-It is important to highlight the result that the 4th-order polynomial
-regression outperforms the 5th-order in all of the reference-based
-metrics, but not the \emph{de novo} \textsc{lap} metrics. The
-5th-order polynomial regression assembles approximately 2.5 kb more
-sequence and uses roughly $1.2\%$ more of the total read set than the
-4th-order (Supplemental Table 1), which may explain the discrepency in
-\textsc{lap} values because the \textsc{lap} value is influenced by
-total coverage~\citep{Ghodsi:2013hb}.
+It is important to highlight that using uncompressed reads does not produce the best assembly in terms of any of the metrics.
+The uncompressed reads scores worse than the top overall assembly (256-profile encoding) in terms of assembled bases, missing reference bases, N50, \textsc{snp}s, indels $>$5bp, and relocations.
+The assembly using uncompressed reads has an error rate of roughly 8.75 errors per 100 kb of assembled sequence, while the 256-profile encoding has an error rate of 8.02 errors per 100 kb (see Supplemental table TODO).
+
+In general, the greater the polynomial order, the better overall assembly; however, the 5th-order polynomial regression performs slightly worse than the 3rd-order polynomial.
+Their respective ranks in terms of N50 and relocations are fairly distant, which lowers the 5th-order polynomial's overall ranking barely below the 3rd-order's.
+The 1st- and 0-order polynomial regression methods perform poor in all metrics except assembled bases.
+One explanation is that the high error portions of sequences are being marked as high quality, so \textsc{allpaths-lg} is unable to trim or error correct the sequences.
+Assembled sequences that overlap maybe unable to align to the errors at the end of the reads, artificially inflating the assembled genome size.
+
+Among the different QualComp rate parameters, the 10 bits/read rate ranked highest overall, outperforming the other rate parameters in terms of corrected N50, least missing reference bases, \textsc{snp}s, and indels $>$5bp.
+With the exception of the 6 bits/read rate, the assemblies decrease in rank with the increase in the rate parameter in terms of corrected N50, and least missing reference bases.
+This trend runs counter to the the decrease in \textsc{mse} of the different rates.
 
@@
   magenta}
-  \label{tab:assembly_ranks}
+  \label{fig:assembly_ranks}
 \end{figure}

### 2015-04-13 manuscript/q-compression.tex
Changed -order to -degree, as degree is the more proper term.

@@
 lines.  The compression methods tend to cluster together across the
-different data sets.  Across all data sets, the 0-order polynomial
+different data sets.  Across all data sets, the 0-degree polynomial
 regression, profile encodings, and QualComp have the lowest
@@
 As the order of the polynomial increases, the bits/base-call increase
-and the \textsc{mse} decreases at an exponential rate.  The 7th-order
+and the \textsc{mse} decreases at an exponential rate.  The 7th-degree
 polynomial regression has the highest bits/base-call and in the
@@
 Among the compression methods, the
-256-, 64-, 128-profile encoding had the highest ranks, followed by QualComp with rates 10 bits/read, 30 bits/read, 100 bits/read, then 7th-order polynomial regression, followed by the QualComp with rate 6 bits/read, the 2-bin encoding, and lastly, the 3rd-order, 5th-order, and 0-order polynomials.
+256-, 64-, 128-profile encoding had the highest ranks, followed by QualComp with rates 10 bits/read, 30 bits/read, 100 bits/read, then 7th-degree polynomial regression, followed by the QualComp with rate 6 bits/read, the 2-bin encoding, and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.
 
@@
 the assembly produced using the reads with unmodified quality
-sequences. All compression methods other than 0-order polynomial
+sequences. All compression methods other than 0-degree polynomial
 regression produce an N50 ranging from 3.17--3.22 Mbps (see Supplemental Table TODO). Despite the
@@
 
-In general, the greater the polynomial order, the better overall assembly; however, the 5th-order polynomial regression performs slightly worse than the 3rd-order polynomial.
-Their respective ranks in terms of N50 and relocations are fairly distant, which lowers the 5th-order polynomial's overall ranking barely below the 3rd-order's.
-The 1st- and 0-order polynomial regression methods perform poor in all metrics except assembled bases.
+In general, the greater the polynomial order, the better overall assembly; however, the 5th-degree polynomial regression performs slightly worse than the 3rd-degree polynomial.
+Their respective ranks in terms of N50 and relocations are fairly distant, which lowers the 5th-degree polynomial's overall ranking barely below the 3rd-degree's.
+The 1st- and 0-degree polynomial regression methods perform poor in all metrics except assembled bases.
 One explanation is that the high error portions of sequences are being marked as high quality, so \textsc{allpaths-lg} is unable to trim or error correct the sequences.
@@
 fragment and short-jump libraries of \textit{Rhodobacter sphaeroides}
-have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-order
+have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-degree
 polynomial regression encodings, respectively. This further highlights

### 2015-04-13 manuscript/q-compression.tex
Some small revisions to text.

@@
   and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
-    musculus} data sets. 2B --- 2-bin encoding; P$n$ 
-  --- profiling with $n$ profiles; R$n$ --- modeling with polynomial regression
-  models of degree $n$; Q$n$ --- QualComp with rate parameter 
-  of $n$. Dashed line corresponds to lossless Bzip2
-  compression of quality values, and dotted line corresponds to
-  original uncompressed data.}
+    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
+  with $n$ profiles; R$n$ --- modeling with polynomial regression
+  models of degree $n$; Q$n$ --- QualComp with rate parameter of
+  $n$. Dashed line corresponds to lossless Bzip2 compression of
+  quality values, and dotted line corresponds to original uncompressed
+  data.}
 \label{fig:mse_vs_bpbp}
@@
 \textsc{lap} score can be viewed as a log likelihood score, where a
-value closer to 0 is better. We use a reference-based evaluation
-script provided by \textsc{gage} that counts single nucleotide
-polymorphisms (\textsc{snp}s), relocations, translations, and
-inversions. The reference-based metrics are normalized by the length
-of the assembly to aid in comparison. For the genome assembly we used
-software that makes use quality values in the assembly process:
+value closer to 0 is better. We use a script provided by \textsc{gage}
+reference-based evaluation to count single nucleotide polymorphisms
+(\textsc{snp}s), relocations, translations, and inversions. The
+reference-based metrics are normalized by the length of the assembly
+to facilitate comparison. For the genome assembly we used software
+that makes use quality values in the assembly process:
 \textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
@@
 
-No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set outperforms all others in all metrics (Table \ref{fig:assembly_ranks}).
-%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set using the uncompressed reads outperforms all compression methods in terms of \textsc{lap}, NG50, relocations, translations, and inversions.
-Among the compression methods, the
-256-, 64-, 128-profile encoding had the highest ranks, followed by QualComp with rates 10 bits/read, 30 bits/read, 100 bits/read, then 7th-degree polynomial regression, followed by the QualComp with rate 6 bits/read, the 2-bin encoding, and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.
+No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
+outperforms all others in all metrics (Table
+\ref{fig:assembly_ranks}).
+%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
+%using the uncompressed reads outperforms all compression methods in
+%terms of \textsc{lap}, NG50, relocations, translations, and
+%inversions.
+Among the compression methods, the 256-, 64-, 128-profile encoding had
+the highest ranks, followed by QualComp with rates 10 bits/read, 30
+bits/read, 100 bits/read, then 7th-degree polynomial regression,
+followed by the QualComp with rate 6 bits/read, the 2-bin encoding,
+and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.
 
@@
 sequences. All compression methods other than 0-degree polynomial
-regression produce an N50 ranging from 3.17--3.22 Mbps (see Supplemental Table TODO). Despite the
-similar contiguity statistics, the different compression methods vary
-noticeably in the amount of \textsc{snp}s. The order of polynomial has
-an inverse relationship with the amount of \textsc{snp}s detected. The
-2-bin and profile methods detected the least amount of \textsc{snp}s
-compared to the reference genome, outperforming the assembly using the
-original quality sequences. A more in-depth evaluation is needed to
-determine whether these compression methods are missing actual
-\textsc{snp}s.
+regression produce an N50 ranging from 3.17--3.22 Mbps (see
+Supplemental Table TODO). Despite the similar contiguity statistics,
+the different compression methods vary noticeably in the amount of
+\textsc{snp}s. The order of polynomial has an inverse relationship
+with the amount of \textsc{snp}s detected. The 2-bin and profile
+methods detected the least amount of \textsc{snp}s compared to the
+reference genome, outperforming the assembly using the original
+quality sequences. A more in-depth evaluation is needed to determine
+whether these compression methods are missing actual \textsc{snp}s.
 
-It is important to highlight that using uncompressed reads does not produce the best assembly in terms of any of the metrics.
-The uncompressed reads scores worse than the top overall assembly (256-profile encoding) in terms of assembled bases, missing reference bases, N50, \textsc{snp}s, indels $>$5bp, and relocations.
-The assembly using uncompressed reads has an error rate of roughly 8.75 errors per 100 kb of assembled sequence, while the 256-profile encoding has an error rate of 8.02 errors per 100 kb (see Supplemental table TODO).
+It is important to highlight that using uncompressed reads does not
+produce the best assembly in terms of any of the metrics.  The
+uncompressed reads scores worse than the top overall assembly
+(256-profile encoding) in terms of assembled bases, missing reference
+bases, N50, \textsc{snp}s, indels $>$5bp, and relocations.  The
+assembly using uncompressed reads has an error rate of roughly 8.75
+errors per 100 kb of assembled sequence, while the 256-profile
+encoding has an error rate of 8.02 errors per 100 kb (see Supplemental
+table TODO).
 
-In general, the greater the polynomial order, the better overall assembly; however, the 5th-degree polynomial regression performs slightly worse than the 3rd-degree polynomial.
-Their respective ranks in terms of N50 and relocations are fairly distant, which lowers the 5th-degree polynomial's overall ranking barely below the 3rd-degree's.
-The 1st- and 0-degree polynomial regression methods perform poor in all metrics except assembled bases.
-One explanation is that the high error portions of sequences are being marked as high quality, so \textsc{allpaths-lg} is unable to trim or error correct the sequences.
-Assembled sequences that overlap maybe unable to align to the errors at the end of the reads, artificially inflating the assembled genome size.
+In general, the greater the polynomial order, the better overall
+assembly; however, the 5th-degree polynomial regression performs
+slightly worse than the 3rd-degree polynomial. The respective ranks in
+terms of N50 and relocations are fairly distant, which lowers the
+overall ranking of the 5th-degree polynomial slightly below that for
+3rd-degree polynomial model. The 1st- and 0-degree polynomial
+regression methods perform poor in all metrics except assembled bases.
+One explanation is that the high error portions of sequences are being
+marked as high quality, so \textsc{allpaths-lg} is unable to trim or
+error correct the sequences. Assembled sequences that overlap maybe
+unable to align to the errors at the end of the reads, artificially
+inflating the assembled genome size.
 
-Among the different QualComp rate parameters, the 10 bits/read rate ranked highest overall, outperforming the other rate parameters in terms of corrected N50, least missing reference bases, \textsc{snp}s, and indels $>$5bp.
-With the exception of the 6 bits/read rate, the assemblies decrease in rank with the increase in the rate parameter in terms of corrected N50, and least missing reference bases.
-This trend runs counter to the the decrease in \textsc{mse} of the different rates.
+Among the different QualComp rate parameters, the 10 bits/read rate
+ranked highest overall, outperforming the other rate parameters in
+terms of corrected N50, least missing reference bases, \textsc{snp}s,
+and indels $>$5bp.  With the exception of the 6 bits/read rate, the
+assemblies decrease in rank with the increase in the rate parameter in
+terms of corrected N50, and least missing reference bases. This trend
+runs counter to the the decrease in \textsc{mse} of the different
+rates.
 

### 2015-04-13 manuscript/q-compression.tex
Added outline of alignment results.

@@
 
-Generally, increasing the regression model polynomial order from 0 to
-5 results in more mapped reads with the most marked change from order
-1 to order 2. The best compression approach is the one that has
-highest proportion of reads shared between the uncompressed and
-decompressed files and least number of reads that are unique in both
-uncompressed and decompressed files. In fragment sample, on average,
-99.03\% of reads were mapped using different compression approaches,
-within which 99\% are shared with the uncompressed alignments and
-0.02\% unique mapped reads in the decompressed file. The average
-missed alignments are 0.99\%. Compression with profile 256 approach
-was shown to be the best performing approach in the fragment
-decompressed sample, as a proportional of 99.94\% reads are mapped
-with respect to the uncompressed mapped reads within which 99.91\% are
-common and only 0.08\% unique mapped reads in the decompressed
-file. In short-run sample, the average number of mapped reads is 99.14
-\% within which 98.73\% are common and 0.41\% unique mapped reads in
-the decompressed file.The average missed alignments are
-1.2\%. Regression with the 5th polynomial order captured the highest
-mapped reads of 100\%, within which 99.66\% are common and only 0.33\%
-unique mapped reads in the decompressed file.
+We evaluate the compression methods using two approaches.
+In the first approach, we order the compression  methods based on how similar the alignment results are to the uncompressed sequences, i.e., the amount of reads aligned by both the uncompressed and compressed methods minus the amount of reads uniquely aligned by the uncompressed and compression method.
+In the second approach, we order the compression methods by total proportion of aligned reads.
+
+The best compression method in terms of similarity with the uncompressed reads is QualComp with rate 100 bits/read, followed by QualComp with rate 30 bits/read, 256-, and 128-profile encoding, then 2-bin encoding, 64-profile encoding, QualComp with rate 10 bits/read, 7th-degree polynomial regression, QualComp with rate 6 bits/read, and finally, 5th-degree through 0-degree polynomial regression.
+
+Ranking the compression methods by overall alignment rate produces an identical ordering as above.
+Aside from 0-degree polynomial regression (83.1\%), all other compression methods have an alignment rate between 87\% and 86.1\%.
+
+Most of the compression methods did not vary greatly in terms of the number of reads that were mapped \emph{only} by the compression method; however, there is a sizable difference in the amount of reads that are originally mapped, but unmapped by the compressed methods.
+QualComp with rate 100 bits/read results in the fewest missing original read alignments (159).
+Increasing the regression model polynomial degree results in a decreasing amount of reads that are originally mapped, but unmapped by the regression model (40,931 to 1,134 reads for 0-degree and 7th-degree, respectively).
+There is no such trend for reads that are mapped only by the regression model.
+
+Setting all bases as minimum quality results in the highest fraction of mapped reads 88.1\%.
+Conversely, setting all bases as maximum quality results in the lowest fraction of mapped reads 72.8\%.
 
@@
 \cline{2-12}
- & sum & 1025434 &   0 &   0 & 1025434 & 872917 & 152517 & 729824 & 295610 & 880724 & 144710 \\ 
+ & fraction & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
 \end{tabular}
@@
 \cline{2-12}
-&  sum & 882375 & 143059 & 888940 & 136494 & 886623 & 138811 & 891944 & 133490 & 894966 & 130468 \\
+& fraction & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
 \end{tabular}
@@
 \cline{2-12}
-&  sum & 897200 & 128234 & 892136 & 133298 & 892537 & 132897 & 892820 & 132614 & 889118 & 136316 \\ 
+& fraction & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
 \end{tabular}
@@
 \cline{2-12}
-& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
+%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
+& fraction & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
 \end{tabular}
@@
 \cline{2-12}
-& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
+%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
+& fraction & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
 \end{tabular}
@@
 \cline{2-12}
-& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\
+%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
+& fraction & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
 

### 2015-04-13 manuscript/q-compression.tex
A few small changes in text (e.g., fraction -> proportion).

@@
 
-Setting all bases as minimum quality results in the highest fraction of mapped reads 88.1\%.
-Conversely, setting all bases as maximum quality results in the lowest fraction of mapped reads 72.8\%.
+Setting all bases as minimum quality results in the highest proportion of mapped reads 88.1\%.
+Conversely, setting all bases as maximum quality results in the lowest proportion of mapped reads 72.8\%.
 
@@
 \cline{2-12}
- & fraction & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
+ & proportion & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
 \end{tabular}
@@
 \cline{2-12}
-& fraction & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
+& proportion & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
 \end{tabular}
@@
 \cline{2-12}
-& fraction & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
+& proportion & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
 \end{tabular}
@@
 %& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
-& fraction & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
+& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
 \end{tabular}
@@
 %& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
-& fraction & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
+& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
 \end{tabular}
@@
 %& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
-& fraction & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
+& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
 

### 2015-04-14 manuscript/q-compression.tex
Added outline of preprocessing results.

@@
 values requires 1 byte per base-call because they are stored in
-\textsc{ascii} format and is denoted by the dotted black line in the
+\textsc{ascii} format and is denoted by the dotted black asterisk in the
 figure.  The lossless compression of each data set using BZip2 ranges
-from 2.19 - 3.10 bits/base-call and is denoted by the colored dotted
-lines.  The compression methods tend to cluster together across the
+from 2.19 - 3.10 bits/base-call and is denoted by the colored asterisks on the x-axis.  The compression methods tend to cluster together across the
 different data sets.  Across all data sets, the 0-degree polynomial
@@
 
+\begin{figure}[!tbp]%figure2
+%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
+\centerline{\includegraphics[width=3.65in]{preprocessing_results.pdf}}
+\caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
+  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
+  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
+    musculus} data sets.
+Sequences were trimmed using Sickle.
+The total amount of bases filtered by each compression method is compared with the amount of bases filtered using the uncompressed sequences.}
+  \label{fig:preprocessing}
+\end{figure}
+
 \subsection{Effects on sequence read preprocessing}
 
-Table \ref{tab:quality_control} shows the proportional numbers with
-respect to those of the uncompressed \textsc{fastq} files. In the
-fragment sample, we retain on average 95.4\% and 94.24\% of bases and
-reads, respectively. In the short-jump sample, we retain 90.4\% and
-90.98\% of bases and reads, respectively. In some cases, the
-proportion of reads and/or bases compared to the original is more than
-100\%; this means that the number of the reads and bases in the
-compressed samples are more because altering quality values may
-prevent the read from being filtered and/or trimmed. This is clear in
-the good/ bad compression approach of the short-jump sample, where the
-proportion of bases and reads are 113.09\% and 103.70\% respectively.
+The majority of compression methods retain more base-pairs after preprocessing than the uncompressed sequences (Figure \ref{fig:preprocessing}).
+In general, as a given compression model increases in complexity, i.e., as the number of profiles, polynomial degrees, or rate increases, the amount of base-pairs kept approaches the number base-pairs kept using the uncompressed sequences.
+The compression methods on the \emph{Mus Musculus} data set have the greatest proportion of retained base-pairs compared to the uncompressed sequences.
+The \textit{Escherichia coli} str. K-12 MG1655 MiSeq data set has the smallest range.
+
+The 2-bin approach is the only compression method that results in a higher number of filtered base-pairs across all data sets.
+Sickle uses a sliding window approach to smooth the quality sequences before it trims the sequence based on a specific quality threshold.
+In the 2-bin approach, there is not an even distribution of values per bin.
+In other words, \emph{bad} quality values may range from 0-33, whereas \emph{good} values may only range from 34-40.
+Thus, mid-range quality values that are above the threshold (20 by default) are set below the quality threshold when compressed, resulting in an increased number of filtered bases.
+
+The 0-degree polynomial regression results in the highest proportion of bases kept.
+If the average quality value of the sequence is above the filtering threshold, then no base-pairs are trimmed.
+Only sequences that are comprised of mainly low quality values will be filtered.
+
+It is important to highlight that the even though a compression method may result in the same number of filtered base-pairs as the uncompressed sequences, it does not mean the \emph{same} base-pairs were filtered.
+The 1st-degree and 5th-degree polynomial regression of the \textit{Rhodobacter sphaeroides} fragment library filters nearly as many bases as each other.
+However, if we examine the specific sequences filtered, the 5th-degree polynomial regression discards approximately two thirds less sequences than the 1st-degree polynomial regression that are kept by the uncompressed sequences (4,640 and 12,066 sequences, respectively) (TODO: move to Supplemental Table \ref{tab:quality_control}).
 
@@
 We evaluate the compression methods using two approaches.
-In the first approach, we order the compression  methods based on how similar the alignment results are to the uncompressed sequences, i.e., the amount of reads aligned by both the uncompressed and compressed methods minus the amount of reads uniquely aligned by the uncompressed and compression method.
+In the first approach, we order the compression  methods based on how similar the alignment results are to the uncompressed sequences, i.e., the amount of reads aligned by both the uncompressed and compressed methods plus the amount of reads unaligned by both methods minus the amount of reads uniquely aligned by the uncompressed and compression method.
 In the second approach, we order the compression methods by total proportion of aligned reads.
@@
 Aside from 0-degree polynomial regression (83.1\%), all other compression methods have an alignment rate between 87\% and 86.1\%.
+The alignment rate of the uncompressed reads is 87\%.
 

### 2015-04-15 manuscript/q-compression.tex
Some refinement of the text and figure legends.

@@
   models of degree $n$; Q$n$ --- QualComp with rate parameter of
-  $n$. Dashed line corresponds to lossless Bzip2 compression of
-  quality values, and dotted line corresponds to original uncompressed
+  $n$. Asterisks denote the corresponding lossless compression using
+  Bzip2, with the black asterisk corresponds to original uncompressed
   data.}
@@
 values requires 1 byte per base-call because they are stored in
-\textsc{ascii} format and is denoted by the dotted black asterisk in the
-figure.  The lossless compression of each data set using BZip2 ranges
-from 2.19 - 3.10 bits/base-call and is denoted by the colored asterisks on the x-axis.  The compression methods tend to cluster together across the
-different data sets.  Across all data sets, the 0-degree polynomial
-regression, profile encodings, and QualComp have the lowest
-bits/base-call.
+\textsc{ascii} format and is denoted by the dotted black asterisk in
+the figure.  The lossless compression of each data set using BZip2
+ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored
+asterisks on the abscissa. The compression methods tend to cluster
+together across the different data sets. Across all data sets, the
+0-degree polynomial regression, profile encodings, and QualComp have
+the lowest bits/base-call.
 
@@
   and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
-    musculus} data sets.
-Sequences were trimmed using Sickle.
-The total amount of bases filtered by each compression method is compared with the amount of bases filtered using the uncompressed sequences.}
+    musculus} data sets. Sequences were trimmed using Sickle. The
+  total amount of bases filtered by each compression method is
+  compared with the amount of bases filtered using the uncompressed
+  sequences.}
   \label{fig:preprocessing}
@@
 
-The majority of compression methods retain more base-pairs after preprocessing than the uncompressed sequences (Figure \ref{fig:preprocessing}).
-In general, as a given compression model increases in complexity, i.e., as the number of profiles, polynomial degrees, or rate increases, the amount of base-pairs kept approaches the number base-pairs kept using the uncompressed sequences.
-The compression methods on the \emph{Mus Musculus} data set have the greatest proportion of retained base-pairs compared to the uncompressed sequences.
-The \textit{Escherichia coli} str. K-12 MG1655 MiSeq data set has the smallest range.
+The majority of compression methods retain more base-pairs after
+preprocessing than the uncompressed sequences (Figure
+\ref{fig:preprocessing}).  In general, as a given compression model
+increases in complexity, i.e., as the number of profiles, polynomial
+degrees, or rate increases, the amount of base-pairs kept approaches
+the number base-pairs kept using the uncompressed sequences.  The
+compression methods on the \textit{Mus Musculus} data set have the
+greatest proportion of retained base-pairs compared to the
+uncompressed sequences.  The \textit{Escherichia coli} str. K-12
+MG1655 MiSeq data set has the smallest range.
 
-The 2-bin approach is the only compression method that results in a higher number of filtered base-pairs across all data sets.
-Sickle uses a sliding window approach to smooth the quality sequences before it trims the sequence based on a specific quality threshold.
-In the 2-bin approach, there is not an even distribution of values per bin.
-In other words, \emph{bad} quality values may range from 0-33, whereas \emph{good} values may only range from 34-40.
-Thus, mid-range quality values that are above the threshold (20 by default) are set below the quality threshold when compressed, resulting in an increased number of filtered bases.
+The 2-bin approach is the only compression method that results in a
+higher number of filtered base-pairs across all data sets.  Sickle
+uses a sliding window approach to smooth the quality sequences before
+it trims the sequence based on a specific quality threshold.  In the
+2-bin approach, there is not an even distribution of values per bin.
+In other words, \emph{bad} quality values may range from 0-33, whereas
+\emph{good} values may only range from 34-40.  Thus, mid-range quality
+values that are above the threshold (20 by default) are set below the
+quality threshold when compressed, resulting in an increased number of
+filtered bases.
 
-The 0-degree polynomial regression results in the highest proportion of bases kept.
-If the average quality value of the sequence is above the filtering threshold, then no base-pairs are trimmed.
-Only sequences that are comprised of mainly low quality values will be filtered.
+The 0-degree polynomial regression results in the highest proportion
+of bases kept.  If the average quality value of the sequence is above
+the filtering threshold, then no base-pairs are trimmed.  Only
+sequences that are comprised of mainly low quality values will be
+filtered.
 
-It is important to highlight that the even though a compression method may result in the same number of filtered base-pairs as the uncompressed sequences, it does not mean the \emph{same} base-pairs were filtered.
-The 1st-degree and 5th-degree polynomial regression of the \textit{Rhodobacter sphaeroides} fragment library filters nearly as many bases as each other.
-However, if we examine the specific sequences filtered, the 5th-degree polynomial regression discards approximately two thirds less sequences than the 1st-degree polynomial regression that are kept by the uncompressed sequences (4,640 and 12,066 sequences, respectively) (TODO: move to Supplemental Table \ref{tab:quality_control}).
+It is important to highlight that the even though a compression method
+may result in the same number of filtered base-pairs as the
+uncompressed sequences, it does not mean the \emph{same} base-pairs
+were filtered.  The 1st-degree and 5th-degree polynomial regression of
+the \textit{Rhodobacter sphaeroides} fragment library filters nearly
+as many bases as each other.  However, if we examine the specific
+sequences filtered, the 5th-degree polynomial regression discards
+approximately two thirds less sequences than the 1st-degree polynomial
+regression that are kept by the uncompressed sequences (4,640 and
+12,066 sequences, respectively) (TODO: move to Supplemental Table
+\ref{tab:quality_control}).
 
@@
 
-We evaluate the compression methods using two approaches.
-In the first approach, we order the compression  methods based on how similar the alignment results are to the uncompressed sequences, i.e., the amount of reads aligned by both the uncompressed and compressed methods plus the amount of reads unaligned by both methods minus the amount of reads uniquely aligned by the uncompressed and compression method.
-In the second approach, we order the compression methods by total proportion of aligned reads.
+We evaluate the compression methods using two approaches.  In the
+first approach, we order the compression methods based on how similar
+the alignment results are to the uncompressed sequences, i.e., the
+amount of reads aligned by both the uncompressed and compressed
+methods plus the amount of reads unaligned by both methods minus the
+amount of reads uniquely aligned by the uncompressed and compression
+method.  In the second approach, we order the compression methods by
+total proportion of aligned reads.
 
-The best compression method in terms of similarity with the uncompressed reads is QualComp with rate 100 bits/read, followed by QualComp with rate 30 bits/read, 256-, and 128-profile encoding, then 2-bin encoding, 64-profile encoding, QualComp with rate 10 bits/read, 7th-degree polynomial regression, QualComp with rate 6 bits/read, and finally, 5th-degree through 0-degree polynomial regression.
+The best compression method in terms of similarity with the
+uncompressed reads is QualComp with rate 100 bits/read, followed by
+QualComp with rate 30 bits/read, 256-, and 128-profile encoding, then
+2-bin encoding, 64-profile encoding, QualComp with rate 10 bits/read,
+7th-degree polynomial regression, QualComp with rate 6 bits/read, and
+finally, 5th-degree through 0-degree polynomial regression.
 
-Ranking the compression methods by overall alignment rate produces an identical ordering as above.
-Aside from 0-degree polynomial regression (83.1\%), all other compression methods have an alignment rate between 87\% and 86.1\%.
-The alignment rate of the uncompressed reads is 87\%.
+Ranking the compression methods by overall alignment rate produces an
+identical ordering as above.  Aside from 0-degree polynomial
+regression (83.1\%), all other compression methods have an alignment
+rate between 87\% and 86.1\%. The alignment rate of the uncompressed
+reads is 87\%.
 
-Most of the compression methods did not vary greatly in terms of the number of reads that were mapped \emph{only} by the compression method; however, there is a sizable difference in the amount of reads that are originally mapped, but unmapped by the compressed methods.
-QualComp with rate 100 bits/read results in the fewest missing original read alignments (159).
-Increasing the regression model polynomial degree results in a decreasing amount of reads that are originally mapped, but unmapped by the regression model (40,931 to 1,134 reads for 0-degree and 7th-degree, respectively).
-There is no such trend for reads that are mapped only by the regression model.
+Most of the compression methods did not vary greatly in terms of the
+number of reads that were mapped \emph{only} by the compression
+method; however, there is a sizable difference in the amount of reads
+that are originally mapped, but unmapped by the compressed methods.
+QualComp with rate 100 bits/read results in the fewest missing
+original read alignments (159).  Increasing the regression model
+polynomial degree results in a decreasing amount of reads that are
+originally mapped, but unmapped by the regression model (40,931 to
+1,134 reads for 0-degree and 7th-degree, respectively).  There is no
+such trend for reads that are mapped only by the regression model.
 
-Setting all bases as minimum quality results in the highest proportion of mapped reads 88.1\%.
-Conversely, setting all bases as maximum quality results in the lowest proportion of mapped reads 72.8\%.
+Setting all bases as minimum quality results in the highest proportion
+of mapped reads 88.1\%.  Conversely, setting all bases as maximum
+quality results in the lowest proportion of mapped reads 72.8\%.
 
@@
 \section*{Acknowledgement}
-We thank the organizers and participants of the 2014 University of
-T\"{u}bingen and University of Maryland Bioinformatics Exchange for
-Students and Teachers (BEST) Summer School.
+We thank the organizers and participants of the 2014 Bioinformatics
+Exchange for Students and Teachers (BEST) Summer School, which was
+funded by the offices of the Dean of The Graduate School at University
+of Maryland and the Rektor of University of T\"{u}bingen.
 

### 2015-04-15 manuscript/q-compression.tex
Various changes in wording and formatting.

@@
   with $n$ profiles; R$n$ --- modeling with polynomial regression
-  models of degree $n$; Q$n$ --- QualComp with rate parameter of
+  models of degree $n$; Q$n$ --- \textsc{q}ual\textsc{c}omp with rate parameter of
   $n$. Asterisks denote the corresponding lossless compression using
-  Bzip2, with the black asterisk corresponds to original uncompressed
+  \textsc{bz}ip2, with the black asterisk corresponds to original uncompressed
   data.}
@@
 \textsc{ascii} format and is denoted by the dotted black asterisk in
-the figure.  The lossless compression of each data set using BZip2
+the figure.  The lossless compression of each data set using \textsc{bz}ip2
 ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored
@@
 together across the different data sets. Across all data sets, the
-0-degree polynomial regression, profile encodings, and QualComp have
+0-degree polynomial regression, profile encodings, and \textsc{q}ual\textsc{c}omp have
 the lowest bits/base-call.
 
-QualComp with the rate parameter set to 100 bits/read has the lowest
+\textsc{q}ual\textsc{c}omp with the rate parameter set to 100 bits/read has the lowest
 \textsc{mse}, but requires 10-15x more storage than the profile
 encoding methods for only a 2-3x reduction in \textsc{mse}.  When
-QualComp's rate parameter is set to match the profile encoding
-methods, QualComp performs slightly worse in terms of \textsc{mse}.
+\textsc{q}ual\textsc{c}omp's rate parameter is set to match the profile encoding
+methods, \textsc{q}ual\textsc{c}omp performs slightly worse in terms of \textsc{mse}.
 In the \textit{Rhodobacter sphaeroides} 2.4.1 fragment library,
-QualComp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of
+\textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of
 17.29.  Using 256-profile encoding requires less storage (0.079
@@
 the number base-pairs kept using the uncompressed sequences.  The
-compression methods on the \textit{Mus Musculus} data set have the
+compression methods on the \textit{Mus musculus} data set have the
 greatest proportion of retained base-pairs compared to the
@@
 Among the compression methods, the 256-, 64-, 128-profile encoding had
-the highest ranks, followed by QualComp with rates 10 bits/read, 30
+the highest ranks, followed by \textsc{q}ual\textsc{c}omp with rates 10 bits/read, 30
 bits/read, 100 bits/read, then 7th-degree polynomial regression,
-followed by the QualComp with rate 6 bits/read, the 2-bin encoding,
+followed by the \textsc{q}ual\textsc{c}omp with rate 6 bits/read, the 2-bin encoding,
 and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.
@@
 
-Among the different QualComp rate parameters, the 10 bits/read rate
+Among the different \textsc{q}ual\textsc{c}omp rate parameters, the 10 bits/read rate
 ranked highest overall, outperforming the other rate parameters in
@@
 \centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
-\caption{Rankings of different assemblies produced using proposed
-  compression methods. Assemblies were constructed using
-  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}.
-  Rankings above the median value are in cyan, those below in
-  magenta}
+\caption{Rankings of compression methods based on \textit{Rhodobacter
+    sphaeroides} assembly attributes sorted by overall
+  rank. Assemblies were constructed using \textsc{allpaths-lg}.
+  Rankings above the median value are in cyan, those below the median
+  value in magenta.}
   \label{fig:assembly_ranks}
@@
 The best compression method in terms of similarity with the
-uncompressed reads is QualComp with rate 100 bits/read, followed by
-QualComp with rate 30 bits/read, 256-, and 128-profile encoding, then
-2-bin encoding, 64-profile encoding, QualComp with rate 10 bits/read,
-7th-degree polynomial regression, QualComp with rate 6 bits/read, and
+uncompressed reads is \textsc{q}ual\textsc{c}omp with rate 100 bits/read, followed by
+\textsc{q}ual\textsc{c}omp with rate 30 bits/read, 256-, and 128-profile encoding, then
+2-bin encoding, 64-profile encoding, \textsc{q}ual\textsc{c}omp with rate 10 bits/read,
+7th-degree polynomial regression, \textsc{q}ual\textsc{c}omp with rate 6 bits/read, and
 finally, 5th-degree through 0-degree polynomial regression.
@@
 that are originally mapped, but unmapped by the compressed methods.
-QualComp with rate 100 bits/read results in the fewest missing
+\textsc{q}ual\textsc{c}omp with rate 100 bits/read results in the fewest missing
 original read alignments (159).  Increasing the regression model
@@
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
+ & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
  & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
@@
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
+ & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
  & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
@@
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
-&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
+&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
  & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
@@
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
+ & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
 & & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
@@
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
+ & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
 & & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
@@
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
-&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
+&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
 & &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
@@
 \textsc{SlimGene}~\citep{Kozanitis:2011kl} and
-\textsc{QualComp}~\citep{Ochoa:2013rt}, needs to be
+\textsc{\textsc{q}ual\textsc{c}omp}~\citep{Ochoa:2013rt}, needs to be
 performed. Additionally, other types of sequencing data can be
@@
 \section*{Acknowledgement}
-We thank the organizers and participants of the 2014 Bioinformatics
-Exchange for Students and Teachers (BEST) Summer School, which was
+We thank the 2014 Bioinformatics
+Exchange for Students and Teachers (\textsc{best}) Summer School, 
 funded by the offices of the Dean of The Graduate School at University
-of Maryland and the Rektor of University of T\"{u}bingen.
+of Maryland and the Rektor of University of T\"{u}bingen, where this 
+research was initiated.
 

### 2015-04-20 manuscript/q-compression.tex
Added related work into methods section.

@@
 classes of lossy compression methods -- binning, modeling, and
-profiling -- and consider an exemplar of each class. In addition to
-quantifying compression efficiency, we also assess the effects of
-quality value information loss resulting from compression on
+profiling -- and consider an exemplar of each class.
+~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated the effects of lossy compression on identifying variants within a data set.
+We build on these prior works and access the effects of
+% quantifying compression efficiency, we also assess the effects of
+quality value information loss resulting from compression on additional
 subsequent genomic analyses including read preprocessing (filtering
@@
 
+\cite{Wan:2012kq} provide three similar lossy compression strategies based on binning the base error probability distribution: UniBinning, Truncating, and LogBinning.
+UniBinning evenly splits the error probability distribution into a user-defined number of partitions.
+Truncating treats a user-defined number of highest quality values as a single bin.
+LogBinning works similar to UniBinning, except with the \emph{log} of the error probability distribution, which effectively bins the \textsc{ascii} quality values evenly.
+Our 2-bin encoding is a combination of LogBinning and Truncating in that we are placing the highest quality values (as defined above) of the log of the error probability distribution into a single bin.
+
+\begin{figure}[!tpb]%figure2
+\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
+\caption{Quality profiles obtained by $k$-means clustering on the
+  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
+  set using $k$ = 128, with each row corresponding to a quality
+  profile. Dark to light colors represent low to high quality
+  values. It is readily visible that the two most distinctive features
+  of quality profiles is their drop-off position and average overall
+  quality. One can also see sporadic low-position values in a handful
+  of profiles, likely capturing intermittent problems in the
+  sequencing process affecting thousands of reads at a
+  time.}\label{fig:profiles_128}
+\end{figure}
+
 \subsubsection{Modeling}
@@
 
+\textsc{QualComp} is a lossy compression tool that models the quality sequence as a multivariate Gaussian distribution, computing the mean and covariance for each position in the read\citep{Ochoa:2013rt}.
+Once the model parameters are calculated, they are stored by the decoder to later reconstruct a \emph{representative} quality value.
+\textsc{QualComp} takes as input a user-specified rate (bits per read) and then poses an optimization problem of how to allot these bits for a given position in a quality sequence while minimizing the \textsc{mse}.
+The quality sequences can be clustered before hand to produce more accurate models.
+
+% they are stored for use by the decoder, and used to derive the number of bits used to encode each quality value, by minimizing the mean square error of a rate distortion problem.
+This optimization problem is parameterized by a user input, which is the total number of output bits the quality scores can consume, expressed as a rate r.
+The solution to the problem specifies, for position i in a read, the number of bits, 􏰗i, that should be used to encode the quality value in position i.
+For each quality value, if it is to be coded in 􏰗i bits, the Gaussian for that position is split into 2􏰗i regions of equal probability, and a bit pattern selected to indicate the region in which the quality value lies.
+The decoder reads in the Gaussian parameters and r, performs the optimization to recover the 􏰗i values and then emits a representative correspond- ing to the region specified by each code word.
+
 \subsubsection{Profiling}
-
-\begin{figure}[!tpb]%figure2
-\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
-\caption{Quality profiles obtained by $k$-means clustering on the
-  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
-  set using $k$ = 128, with each row corresponding to a quality
-  profile. Dark to light colors represent low to high quality
-  values. It is readily visible that the two most distinctive features
-  of quality profiles is their drop-off position and average overall
-  quality. One can also see sporadic low-position values in a handful
-  of profiles, likely capturing intermittent problems in the
-  sequencing process affecting thousands of reads at a
-  time.}\label{fig:profiles_128}
-\end{figure}
 
@@
 
-\textit{Stapylococcus aureus} USA300, which are generated from a
-fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
-and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
-The corresponding reference sequence was obtained from the NCBI RefSeq
-database (NC\_010063.1, NC\_010079.1, NC\_012417.1).
+% Will be moved to supplemental.
+%\textit{Stapylococcus aureus} USA300, which are generated from a
+%fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
+%and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
+%The corresponding reference sequence was obtained from the NCBI RefSeq
+%database (NC\_010063.1, NC\_010079.1, NC\_012417.1).
 
@@
 on quality control steps of read filtering and trimming, which were
-performed using Sickle~\citep{sickle}, quantify read characteristics
-with PrinSeq~\citep{Schmieder:2011gd}, and make comparison to
+performed using Sickle~\citep{sickle}, and make comparison to
 uncompressed data.
@@
 (\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
-reference-based metrics. The contiguity statistics include number of
-scaffolds and NG50, which is defined as the weighted median contig
+reference-based metrics. The contiguity statistics include N50, which is defined as the median contig
 size (the length of largest contig $c$ such that the total size of the
-contigs larger than $c$â exceeds half of the known genome size). The
+contigs larger than $c$ exceeds half of the assembly size) and corrected N50, which is the recalculated N50 size after the contigs are broken apart at locations of errors.  The
 \textsc{lap} score can be viewed as a log likelihood score, where a
@@
 
-\begin{table*}[!tbhp]
-\centering
-\caption[]{Quality filtering/trimming for the uncompressed and
-  decompressed \textsc{fastq} files. Numbers correspond to the number
-  of reads that are kept or discarded after applying compression with
-  respect to the original, uncompressed values.}
-\label{tab:quality_control}
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
- & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
-  \cline{2-12}
-& kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
-{\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
-\cline{2-12}
- & proportion & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
-\end{tabular}
+% \begin{table*}[!tbhp]
+% \centering
+% \caption[]{Quality filtering/trimming for the uncompressed and
+%   decompressed \textsc{fastq} files. Numbers correspond to the number
+%   of reads that are kept or discarded after applying compression with
+%   respect to the original, uncompressed values.}
+% \label{tab:quality_control}
+% \begin{tabular}{lr|cc|cc|cc|cc|cc}
+%  & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
+%  & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
+%   \cline{2-12}
+% & kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
+% {\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
+% \cline{2-12}
+%  & proportion & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
+% \end{tabular}
 
-\bigskip
+% \bigskip
 
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
- & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
-\cline{2-12}
-& kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
-{\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
-\cline{2-12}
-& proportion & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
-\end{tabular}
+% \begin{tabular}{lr|cc|cc|cc|cc|cc}
+%  & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
+%  & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
+% \cline{2-12}
+% & kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
+% {\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
+% \cline{2-12}
+% & proportion & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
+% \end{tabular}
 
-\bigskip
+% \bigskip
 
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
-&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
- & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
-\cline{2-12}
-& kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
-{\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
-\cline{2-12}
-& proportion & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
-\end{tabular}
-\end{table*}
+% \begin{tabular}{lr|cc|cc|cc|cc|cc}
+% &  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
+%  & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
+% \cline{2-12}
+% & kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
+% {\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
+% \cline{2-12}
+% & proportion & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
+% \end{tabular}
+% \end{table*}
 

### 2015-04-21 manuscript/q-compression.tex
Added section on extending polynomial regression.  Text clean up.

@@
   models of degree $n$; Q$n$ --- \textsc{q}ual\textsc{c}omp with rate parameter of
-  $n$. Asterisks denote the corresponding lossless compression using
+  $n$. All compression methods have been losslessly compressed with \textsc{bz}ip2. Asterisks denote the corresponding lossless compression of the original data using
   \textsc{bz}ip2, with the black asterisk corresponds to original uncompressed
@@
 
+It is important to highlight the result that the 2-bin method aligns a greater portion of reads than the various regression models.
+If a poor quality base is flanked by high quality sequence, then the low degree polynomial regression models tend to smooth out the base quality, erroneously marking the low quality base as higher quality.
+During alignment, Bowtie2 penalizes mismatches at high quality bases more so than mismatches at lower quality bases.
+Thus, the polynomial regression models incur a high penalty for these bases, resulting in less alignments, despite having a better \textsc{mse} than 2-bin in most cases.
+
 Setting all bases as minimum quality results in the highest proportion
@@
 
+\subsection{Extension of polynomial regression}
+
+The downside of modeling quality sequences using polynomial regression is that the model often requires a high number of degrees to achieve the same \textsc{mse} as the profile and \textsc{QualComp} methods.
+However, storing a high number of coefficients requires more storage than losslessly compressing the original data.
+In order to increase the compressibility of modeling, we can attempt to store the \emph{profiles} of certain polynomial functions.
+In other words, we can use a spline (a function that is piecewise-defined by polynomial functions) to represent a given quality sequence.
+Similar to our profile-encoding method, the user can specify how many polynomial functions they wish to store.
+Then a quality sequence can be divided evenly into a given number of segments, and each segment can be annotated with a polynomial function profile that closely matches its quality sequence.
+
 \subsection{Potential for operations on compressed data}
@@
 We have simply provided here the initial steps in analyzing the effect
-of lossy compression on quality sequences using a single,
-high-coverage bacterial data set. More work needs to be done using
-additional biological data sets, such as human and mouse, along with
-different sequencing technologies. A more direct comparison against
-related lossy compression tools, such as
-\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
-\textsc{\textsc{q}ual\textsc{c}omp}~\citep{Ochoa:2013rt}, needs to be
-performed. Additionally, other types of sequencing data can be
+of lossy compression on quality sequences using two
+high-coverage bacterial, human, and mouse data sets.
+A more direct comparison against additional lossy compression tools, such as
+\textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
+performed. Furthermore, other types of sequencing data can be
 analyzed apart from the Illumina data examined here. For example, the

### 2015-04-21 manuscript/q-compression.tex
Some small edits.

--- 2015-04-20 manuscript/q-compression.tex
@@
 classes of lossy compression methods -- binning, modeling, and
-profiling -- and consider an exemplar of each class.
-~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated the effects of lossy compression on identifying variants within a data set.
+profiling -- and consider an exemplar of each class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated the effects of lossy compression on identifying variants within a data set.
 We build on these prior works and access the effects of
@@
 
-\textsc{QualComp} is a lossy compression tool that models the quality sequence as a multivariate Gaussian distribution, computing the mean and covariance for each position in the read\citep{Ochoa:2013rt}.
+\textsc{QualComp} is a lossy compression tool that models quality values as a multivariate Gaussian distribution, computing the mean and covariance for each position in the read\citep{Ochoa:2013rt}.
 Once the model parameters are calculated, they are stored by the decoder to later reconstruct a \emph{representative} quality value.
-\textsc{QualComp} takes as input a user-specified rate (bits per read) and then poses an optimization problem of how to allot these bits for a given position in a quality sequence while minimizing the \textsc{mse}.
-The quality sequences can be clustered before hand to produce more accurate models.
+\textsc{QualComp} takes as input a user-specified rate (bits per read) and then poses an optimization problem of how to allot these bits for a given position while minimizing the \textsc{mse}.
+The quality values can be clustered before hand to produce more accurate models.
 
@@
 As large sets of quality strings show similar trends of quality over
-their sequence, it makes sense to identify such common patterns in the
+their length, it makes sense to identify such common patterns in the
 data and use them as reference profiles to approximate individual
-quality sequences. Such patterns can be readily determined by
+sequences of quality values. Such patterns can be readily determined by
 clustering data points (i.e. quality vectors) and using the resulting
@@
 
-First, the method samples an adjustable number of quality sequences
-randomly from the file to be used as a training set. Quality sequences
+First, the method samples an adjustable number of reads
+randomly from the file to be used as a training set. Quality values
 are represented by vectors containing their \textsc{phred} scores
@@
 
-Once the $k$ quality profiles are determined, all quality sequences
+Once the $k$ quality profiles are determined, all reads
 are passed through the trained $k$-means predictor, with the nearest
-quality profile in Euclidean space being assigned to every quality
-sequence as their compressed representation.
+quality profile in Euclidean space being assigned to every read
+as their compressed representation.
 
@@
 outlier quality values, it ensures that the overall trends in quality
-sequences are retained. Quality profiles capture different overall
+value patterns are retained. Quality profiles capture different overall
 qualities and different drop-off positions and gradients. An example
@@
 requires storing eight floating point values, resulting in 32 bytes
-per quality sequence.  The read length of the \textit{Mus musculus}
+per sequence of quality values. The read length of the \textit{Mus musculus}
 data set is only 26, so the 7th-degree polynomial regression is storing
@@
 higher number of filtered base-pairs across all data sets. Sickle
-uses a sliding window approach to smooth the quality sequences before
+uses a sliding window approach to smooth the read quality values before
 it trims the sequence based on a specific quality threshold. In the
@@
 The 0-degree polynomial regression results in the highest proportion
-of bases kept.  If the average quality value of the sequence is above
+of bases kept. If the mean quality value of the read is above
 the filtering threshold, then no base-pairs are trimmed. Only
-sequences that are comprised of mainly low quality values will be
+reads that are comprised of mainly low quality values will be
 filtered.
@@
 as many bases as each other. However, if we examine the specific
-sequences filtered, the 5th-degree polynomial regression discards
-approximately two thirds less sequences than the 1st-degree polynomial
-regression that are kept by the uncompressed sequences (4,640 and
-12,066 sequences, respectively) (TODO: move to Supplemental Table
+reads filtered, the 5th-degree polynomial regression discards
+approximately two thirds less reads than the 1st-degree polynomial
+regression that are kept by the uncompressed reads (4,640 and
+12,066 reads, respectively) (TODO: move to Supplemental Table
 \ref{tab:quality_control}).
@@
 the assembly produced using the reads with unmodified quality
-sequences. All compression methods other than 0-degree polynomial
+values. All compression methods other than 0-degree polynomial
 regression produce an N50 ranging from 3.17--3.22 Mbps (see
@@
 reference genome, outperforming the assembly using the original
-quality sequences. A more in-depth evaluation is needed to determine
+quality values. A more in-depth evaluation is needed to determine
 whether these compression methods are missing actual \textsc{snp}s.
@@
 regression methods perform poor in all metrics except assembled bases.
-One explanation is that the high error portions of sequences are being
+One explanation is that the high error portions of reads are being
 marked as high quality, so \textsc{allpaths-lg} is unable to trim or
@@
 
-Certain short read alignment tools use the quality sequence
+Certain short read alignment tools use the quality value
 information when finding potential alignments. Bowtie2 (version 2.2.3)
@@
 first approach, we order the compression methods based on how similar
-the alignment results are to the uncompressed sequences, i.e., the
+the alignment results are using the original uncompressed quality values, i.e., the
 amount of reads aligned by both the uncompressed and compressed
@@
 redundancy built-in to handle such loss in information. Passing the
-decompressed quality sequences through quality control software shows
+decompressed quality values through quality control software shows
 that most methods filter nearly as many bases as using original
-quality sequences. Assemblers performing sequencing alignment use
+quality sequences. Assemblers performing sequence alignment use
 percent similarity scores that are typically robust to standard
@@
 We have simply provided here the initial steps in analyzing the effect
-of lossy compression on quality sequences using a single,
+of lossy compression on quality values using a single,
 high-coverage bacterial data set. More work needs to be done using

### 2015-04-21 manuscript/q-compression.tex
Text clean up.

--- 2015-04-21 manuscript/q-compression.tex
@@
 and memory requirements with little discernible effects on subsequent
-analysis results. The four compression strategies in this study were
+analysis results. The three compression strategies in this study were
 able to produce similar results to those obtained with uncompressed
@@
 
-If quality values are modeled compression is conceivably possible by
+If quality values are modeled, compression is conceivably possible by
 replacing the original quality values by a representation of the
@@
 centers as read quality profiles. As the problem is NP-hard, we use a
-heuristic iterative refinement approach quickly converging to a
-locally optimal minimum provided by the Python package scikit-learn
-0.15~\citep{pedregosa2011scikit}.
+heuristic iterative refinement approach by quickly converging to a
+locally optimal minimum provided by R~\citep{hartigan1979algorithm}.
 
@@
 error correct the sequences. Assembled sequences that overlap maybe
-unable to align to the errors at the end of the reads, artificially
+unable to align across the errors at the end of the reads, artificially
 inflating the assembled genome size.

### 2015-04-21 manuscript/q-compression.tex
Added extension to polynomial regression section.

@@
 The quality values can be clustered before hand to produce more accurate models.
-
-% they are stored for use by the decoder, and used to derive the number of bits used to encode each quality value, by minimizing the mean square error of a rate distortion problem.
-This optimization problem is parameterized by a user input, which is the total number of output bits the quality scores can consume, expressed as a rate r.
-The solution to the problem specifies, for position i in a read, the number of bits, 􏰗i, that should be used to encode the quality value in position i.
-For each quality value, if it is to be coded in 􏰗i bits, the Gaussian for that position is split into 2􏰗i regions of equal probability, and a bit pattern selected to indicate the region in which the quality value lies.
-The decoder reads in the Gaussian parameters and r, performs the optimization to recover the 􏰗i values and then emits a representative correspond- ing to the region specified by each code word.
 

### 2015-04-21 manuscript/q-compression.tex
Added missing text.

@@
 
+It is important to highlight the result that the 2-bin method aligns a greater portion of reads than the various regression models.   
+If a poor quality base is flanked by high quality sequence, then the low degree polynomial regression models tend to smooth out the base quality, erroneously marking the low quality base as higher quality.
+During alignment, Bowtie2 penalizes mismatches at high quality bases more so than mismatches at lower quality bases.   
+Thus, the polynomial regression models incur a high penalty for these bases, resulting in less alignments, despite having a better \textsc{mse} than 2-bin in most cases.
+
 Setting all bases as minimum quality results in the highest proportion
@@
 We have simply provided here the initial steps in analyzing the effect
-of lossy compression on quality values using a single,
-high-coverage bacterial data set. More work needs to be done using
-additional biological data sets, such as human and mouse, along with
-different sequencing technologies. A more direct comparison against
-related lossy compression tools, such as
-\textsc{SlimGene}~\citep{Kozanitis:2011kl} and
-\textsc{\textsc{q}ual\textsc{c}omp}~\citep{Ochoa:2013rt}, needs to be
-performed. Additionally, other types of sequencing data can be
+of lossy compression on quality sequences using two
+high-coverage bacterial, human, and mouse data sets.
+A more direct comparison against additional lossy compression tools, such as
+\textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
+performed. Furthermore, other types of sequencing data can be
 analyzed apart from the Illumina data examined here. For example, the

### 2015-04-22 manuscript/q-compression.tex
Added beginning supplemental spreadsheet.

@@
 sets, but the additional short-jump library results are available in
-the supplementary (TODO:TABLE). Storing the uncompressed quality
+the Supplemental Table 1. Storing the uncompressed quality
 values requires 1 byte per base-call because they are stored in
@@
 regression that are kept by the uncompressed reads (4,640 and
-12,066 reads, respectively) (TODO: move to Supplemental Table
-\ref{tab:quality_control}).
+12,066 reads, respectively) (Supplemental Table 2).
 
@@
 regression produce an N50 ranging from 3.17--3.22 Mbps (see
-Supplemental Table TODO). Despite the similar contiguity statistics,
+Supplemental Table 3). Despite the similar contiguity statistics,
 the different compression methods vary noticeably in the amount of
@@
 errors per 100 kb of assembled sequence, while the 256-profile
-encoding has an error rate of 8.02 errors per 100 kb (see Supplemental
-table TODO).
+encoding has an error rate of 8.02 errors per 100 kbp (Supplemental
+Table 3).
 

### 2015-04-23 manuscript/q-compression.tex
Numerous .small edits

@@
 
-\title[lossy-compression]{Lossy compression of DNA sequencing quality data}
-\author[El Hadidi \textit{et~al.}]{Mohamed El Hadidi\,$^{1}$\footnote{authors contributed equally}, Christopher M.
-Hill\,$^{2}$\footnotemark[1], Andr\'{a}s Szolek\,$^{3}$\footnotemark[1], and Michael P.
-Cummings\,$^4$\footnote{to whom correspondence should be addressed}}
-\address{$^{1}$Department of Algorithms in Bioinformatics,  Center for Bioinformatics, University of T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany \\
-$^{2}$Department of Computer Science, University of Maryland, College Park,  Maryland, 20742 USA\\
-$^{3}$Department of Applied Bioinformatics, Center for Bioinformatics, Quantitative Biology Center, and Department of Computer Science, University of T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany\\
-$^{4}$Center for Bioinformatics and Computational Biology, University of Maryland, College Park, 20742 USA}
+\title[lossy-compression]{Lossy compression of DNA sequencing quality
+  data} \author[El Hadidi \textit{et~al.}]{Mohamed El
+  Hadidi\,$^{1}$\footnote{authors contributed equally}, Christopher
+  M. Hill\,$^{2}$\footnotemark[1], Andr\'{a}s
+  Szolek\,$^{3}$\footnotemark[1], and Michael
+  P. Cummings\,$^4$\footnote{to whom correspondence should be
+    addressed}} \address{$^{1}$Department of Algorithms in
+  Bioinformatics, Center for Bioinformatics, University of
+  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany
+  \\ $^{2}$Department of Computer Science, University of Maryland,
+  College Park, Maryland, 20742 USA\\ $^{3}$Department of Applied
+  Bioinformatics, Center for Bioinformatics, Quantitative Biology
+  Center, and Department of Computer Science, University of
+  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany\\ $^{4}$Center
+  for Bioinformatics and Computational Biology, University of
+  Maryland, College Park, 20742 USA}
 %$\dagger$authors contributed equally
@@
 classes of lossy compression methods -- binning, modeling, and
-profiling -- and consider an exemplar of each class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated the effects of lossy compression on identifying variants within a data set.
-We build on these prior works and access the effects of
+profiling -- and consider an exemplar of each
+class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated
+the effects of lossy compression on identifying variants within a data
+set. We build on these prior works and access the effects of
 % quantifying compression efficiency, we also assess the effects of
-quality value information loss resulting from compression on additional
-subsequent genomic analyses including read preprocessing (filtering
-and trimming), genome assembly, and read mapping.
+quality value information loss resulting from compression on
+additional subsequent genomic analyses including read preprocessing
+(filtering and trimming), genome assembly, and read mapping.
 
@@
 
-\cite{Wan:2012kq} provide three similar lossy compression strategies based on binning the base error probability distribution: UniBinning, Truncating, and LogBinning.
-UniBinning evenly splits the error probability distribution into a user-defined number of partitions.
-Truncating treats a user-defined number of highest quality values as a single bin.
-LogBinning works similar to UniBinning, except with the \emph{log} of the error probability distribution, which effectively bins the \textsc{ascii} quality values evenly.
-Our 2-bin encoding is a combination of LogBinning and Truncating in that we are placing the highest quality values (as defined above) of the log of the error probability distribution into a single bin.
+\cite{Wan:2012kq} provide three similar lossy compression strategies
+based on binning the base error probability distribution: UniBinning,
+Truncating, and LogBinning.  UniBinning evenly splits the error
+probability distribution into a user-defined number of partitions.
+Truncating treats a user-defined number of highest quality values as a
+single bin.  LogBinning works similar to UniBinning, except with the
+\emph{log} of the error probability distribution, which effectively
+bins the \textsc{ascii} quality values evenly.  Our 2-bin encoding is
+a combination of LogBinning and Truncating in that we are placing the
+highest quality values (as defined above) of the log of the error
+probability distribution into a single bin.
 
@@
 
-\textsc{QualComp} is a lossy compression tool that models quality values as a multivariate Gaussian distribution, computing the mean and covariance for each position in the read\citep{Ochoa:2013rt}.
-Once the model parameters are calculated, they are stored by the decoder to later reconstruct a \emph{representative} quality value.
-\textsc{QualComp} takes as input a user-specified rate (bits per read) and then poses an optimization problem of how to allot these bits for a given position while minimizing the \textsc{mse}.
-The quality values can be clustered before hand to produce more accurate models.
+\textsc{QualComp} is a lossy compression tool that models quality
+values as a multivariate Gaussian distribution, computing the mean and
+covariance for each position in the read\citep{Ochoa:2013rt}. Once the
+model parameters are calculated, they are stored by the decoder to
+later reconstruct a \emph{representative} quality value.
+\textsc{QualComp} takes as input a user-specified rate (bits per read)
+and then poses an optimization problem of how to allot these bits for
+a given position while minimizing the \textsc{mse}. The quality values
+can be clustered before hand to produce more accurate models.
 
@@
 \centerline{\includegraphics[width=7in]{compression_results.pdf}}
-\caption{Mean squared error versus bits/base-call for different
-  compression methods applied to the \textit{Rhodobacter sphaeroides}
-  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
-  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
-    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
-  with $n$ profiles; R$n$ --- modeling with polynomial regression
-  models of degree $n$; Q$n$ --- \textsc{q}ual\textsc{c}omp with rate parameter of
+\caption{The relationship of bits/base-call and mean squared error for
+  quality value compression methods applied to four data sets:
+  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
+  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
+  \textit{Mus musculus}. Point labels correspond to different
+  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
+  profiles; R$n$, modeling with polynomial regression of degree $n$;
+  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
   $n$. Asterisks denote the corresponding lossless compression using
-  \textsc{bz}ip2, with the black asterisk corresponds to original uncompressed
+  \textsc{bz}ip2, with the black asterisk corresponding to original
   data.}
@@
 data and use them as reference profiles to approximate individual
-sequences of quality values. Such patterns can be readily determined by
-clustering data points (i.e. quality vectors) and using the resulting
-cluster centers as representative profiles.
+sequences of quality values. Such patterns can be readily determined
+by clustering data points (i.e. quality vectors) and using the
+resulting cluster centers as representative profiles.
 
@@
 
-First, the method samples an adjustable number of reads
-randomly from the file to be used as a training set. Quality values
-are represented by vectors containing their \textsc{phred} scores
-corresponding to each position along the read. Subsequently, $k$-means
-clustering is performed on the training set until convergence. The
-obtained cluster centers will be the quality profile prototypes for
-the dataset.
+First, the method samples an adjustable number of reads randomly from
+the file to be used as a training set. Quality values are represented
+by vectors containing their \textsc{phred} scores corresponding to
+each position along the read. Subsequently, $k$-means clustering is
+performed on the training set until convergence. The obtained cluster
+centers will be the quality profile prototypes for the dataset.
 
-Once the $k$ quality profiles are determined, all reads
-are passed through the trained $k$-means predictor, with the nearest
-quality profile in Euclidean space being assigned to every read
-as their compressed representation.
+Once the $k$ quality profiles are determined, all reads are passed
+through the trained $k$-means predictor, with the nearest quality
+profile in Euclidean space being assigned to every read as their
+compressed representation.
 
@@
 outlier quality values, it ensures that the overall trends in quality
-value patterns are retained. Quality profiles capture different overall
-qualities and different drop-off positions and gradients. An example
-of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.
+value patterns are retained. Quality profiles capture different
+overall qualities and different drop-off positions and gradients. An
+example of 128 quality profiles are shown on Figure
+\ref{fig:profiles_128}.
 
@@
 which are generated from a fragment library (insert size of 180bp nt;
-1,145,8940 paired-end reads). The corresponding reference sequence
-was obtained from the NCBI RefSeq database (NC\_000913.2).
+1,145,8940 paired-end reads). The corresponding reference sequence was
+obtained from the NCBI RefSeq database (NC\_000913.2).
 
@@
 (\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
-reference-based metrics. The contiguity statistics include N50, which is defined as the median contig
-size (the length of largest contig $c$ such that the total size of the
-contigs larger than $c$ exceeds half of the assembly size) and corrected N50, which is the recalculated N50 size after the contigs are broken apart at locations of errors. The
+reference-based metrics. The contiguity statistics include N50, which
+is defined as the median contig size (the length of largest contig $c$
+such that the total size of the contigs larger than $c$ exceeds half
+of the assembly size) and corrected N50, which is the recalculated N50
+size after the contigs are broken apart at locations of errors. The
 \textsc{lap} score can be viewed as a log likelihood score, where a
@@
 sets, but the additional short-jump library results are available in
-the Supplemental Table 1. Storing the uncompressed quality
-values requires 1 byte per base-call because they are stored in
+the Supplemental Table 1. Storing the uncompressed quality values
+requires 1 byte per base-call because they are stored in
 \textsc{ascii} format and is denoted by the dotted black asterisk in
-the figure. The lossless compression of each data set using \textsc{bz}ip2
-ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored
-asterisks on the abscissa. The compression methods tend to cluster
-together across the different data sets. Across all data sets, the
-0-degree polynomial regression, profile encodings, and \textsc{q}ual\textsc{c}omp have
-the lowest bits/base-call.
+the figure. The lossless compression of each data set using
+\textsc{bz}ip2 ranges from 2.19 - 3.10 bits/base-call and is denoted
+by the colored asterisks on the abscissa. The compression methods tend
+to cluster together across the different data sets. Across all data
+sets, the 0-degree polynomial regression, profile encodings, and
+\textsc{q}ual\textsc{c}omp have the lowest bits/base-call.
 
-\textsc{q}ual\textsc{c}omp with the rate parameter set to 100 bits/read has the lowest
-\textsc{mse}, but requires 10-15x more storage than the profile
-encoding methods for only a 2-3x reduction in \textsc{mse}. When
-\textsc{q}ual\textsc{c}omp's rate parameter is set to match the profile encoding
-methods, \textsc{q}ual\textsc{c}omp performs slightly worse in terms of \textsc{mse}.
-In the \textit{Rhodobacter sphaeroides} 2.4.1 fragment library,
-\textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of
-17.29. Using 256-profile encoding requires less storage (0.079
-bits/bp) and has a lower \textsc{mse} (11.85).
+\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
+bits/read has the lowest \textsc{mse}, but requires 10-15x more
+storage than the profile encoding methods for only a 2-3x reduction in
+\textsc{mse}. When \textsc{q}ual\textsc{c}omp's rate parameter is set
+to match the profile encoding methods, \textsc{q}ual\textsc{c}omp
+performs slightly worse in terms of \textsc{mse}.  In the
+\textit{Rhodobacter sphaeroides} 2.4.1 fragment library,
+\textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has
+a \textsc{mse} of 17.29. Using 256-profile encoding requires less
+storage (0.079 bits/bp) and has a lower \textsc{mse} (11.85).
 
-As the order of the polynomial increases, the bits/base-call increase
+As the degree of the polynomial increases, the bits/base-call increase
 and the \textsc{mse} decreases at an exponential rate. The 7th-degree
@@
 requires storing eight floating point values, resulting in 32 bytes
-per sequence of quality values. The read length of the \textit{Mus musculus}
-data set is only 26, so the 7th-degree polynomial regression is storing
-six more bytes than necessary for lossless encoding the quality data.
+per sequence of quality values. The read length of the \textit{Mus
+  musculus} data set is only 26, so the 7th-degree polynomial
+regression is storing six more bytes than necessary for lossless
+encoding the quality data.
 
@@
 greatest proportion of retained base-pairs compared to the
-uncompressed sequences. The \textit{Escherichia coli} str. K-12
-MG1655 MiSeq data set has the smallest range.
+uncompressed sequences. The \textit{Escherichia coli} str. K-12 MG1655
+MiSeq data set has the smallest range.
 
 The 2-bin approach is the only compression method that results in a
-higher number of filtered base-pairs across all data sets. Sickle
-uses a sliding window approach to smooth the read quality values before
-it trims the sequence based on a specific quality threshold. In the
-2-bin approach, there is not an even distribution of values per bin.
-In other words, \emph{bad} quality values may range from 0-33, whereas
+higher number of filtered base-pairs across all data sets. Sickle uses
+a sliding window approach to smooth the read quality values before it
+trims the sequence based on a specific quality threshold. In the 2-bin
+approach, there is not an even distribution of values per bin.  In
+other words, \emph{bad} quality values may range from 0-33, whereas
 \emph{good} values may only range from 34-40. Thus, mid-range quality
@@
 The 0-degree polynomial regression results in the highest proportion
-of bases kept. If the mean quality value of the read is above
-the filtering threshold, then no base-pairs are trimmed. Only
-reads that are comprised of mainly low quality values will be
-filtered.
+of bases kept. If the mean quality value of the read is above the
+filtering threshold, then no base-pairs are trimmed. Only reads that
+are comprised of mainly low quality values will be filtered.
 
 It is important to highlight that the even though a compression method
-may result in the same number of filtered base-pairs as the
-uncompressed sequences, it does not mean the \emph{same} base-pairs
+may result in the same number of filtered base calls as the
+uncompressed sequences, it does not mean the \emph{same} base calls
 were filtered. The 1st-degree and 5th-degree polynomial regression of
 the \textit{Rhodobacter sphaeroides} fragment library filters nearly
-as many bases as each other. However, if we examine the specific
-reads filtered, the 5th-degree polynomial regression discards
-approximately two thirds less reads than the 1st-degree polynomial
-regression that are kept by the uncompressed reads (4,640 and
-12,066 reads, respectively) (Supplemental Table 2).
+as many bases as each other. However, if we examine the specific reads
+filtered, the 5th-degree polynomial regression discards approximately
+two thirds less reads than the 1st-degree polynomial regression that
+are kept by the uncompressed reads (4,640 and 12,066 reads,
+respectively) (Supplemental Table 2).
 
@@
 %inversions.
-Among the compression methods, the 256-, 64-, 128-profile encoding had
-the highest ranks, followed by \textsc{q}ual\textsc{c}omp with rates 10 bits/read, 30
-bits/read, 100 bits/read, then 7th-degree polynomial regression,
-followed by the \textsc{q}ual\textsc{c}omp with rate 6 bits/read, the 2-bin encoding,
-and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.
+Among the compression methods, the profile encoding performed best,
+polynomial regression modeling with degree $>$ 7 performed worst, and
+other methods were intermediate (Fig.~\ref{fig:assembly_ranks}).
 
@@
 values. All compression methods other than 0-degree polynomial
-regression produce an N50 ranging from 3.17--3.22 Mbps (see
-Supplemental Table 3). Despite the similar contiguity statistics,
-the different compression methods vary noticeably in the amount of
-\textsc{snp}s. The order of polynomial has an inverse relationship
-with the amount of \textsc{snp}s detected. The 2-bin and profile
-methods detected the least amount of \textsc{snp}s compared to the
-reference genome, outperforming the assembly using the original
-quality values. A more in-depth evaluation is needed to determine
-whether these compression methods are missing actual \textsc{snp}s.
+regression produce an N50 ranging from 3.17--3.22 Mbp (see
+Supplemental Table 3). Despite the similar contiguity statistics, the
+different compression methods vary noticeably in the amount of
+\textsc{snp}s. The 2-bin and profile methods exhibited the fewest
+\textsc{snp}s compared to the reference genome, thus outperforming the
+assembly using the original quality values in this characteristic. A
+more in-depth evaluation is needed to determine whether these
+compression methods are missing actual \textsc{snp}s.
 
@@
 bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
-assembly using uncompressed reads has an error rate of roughly 8.75
-errors per 100 kb of assembled sequence, while the 256-profile
-encoding has an error rate of 8.02 errors per 100 kbp (Supplemental
-Table 3).
+assembly using the original uncompressed reads has an error rate of
+roughly 8.75 errors per 100 kb of assembled sequence, while the
+256-profile encoding has an error rate of 8.02 errors per 100 kbp
+(Supplemental Table 3).
 
-In general, the greater the polynomial order, the better overall
+In general, the greater the polynomial degree, the better overall
 assembly; however, the 5th-degree polynomial regression performs
@@
 3rd-degree polynomial model. The 1st- and 0-degree polynomial
-regression methods perform poor in all metrics except assembled bases.
-One explanation is that the high error portions of reads are being
-marked as high quality, so \textsc{allpaths-lg} is unable to trim or
-error correct the sequences. Assembled sequences that overlap maybe
-unable to align across the errors at the end of the reads, artificially
-inflating the assembled genome size.
+regression methods perform poorly in all metrics except assembled
+bases.  One explanation is that the high error portions of reads are
+being marked as high quality, so \textsc{allpaths-lg} is unable to
+trim or error correct the sequences. Assembled sequences that overlap
+maybe unable to align across the errors at the end of the reads,
+artificially inflating the assembled genome size.
 
-Among the different \textsc{q}ual\textsc{c}omp rate parameters, the 10 bits/read rate
-ranked highest overall, outperforming the other rate parameters in
-terms of corrected N50, least missing reference bases, \textsc{snp}s,
-and indels $>$5bp. With the exception of the 6 bits/read rate, the
-assemblies decrease in rank with the increase in the rate parameter in
-terms of corrected N50, and least missing reference bases. This trend
-runs counter to the the decrease in \textsc{mse} of the different
-rates.
+Among the different \textsc{q}ual\textsc{c}omp rate parameters, the 10
+bits/read rate ranked highest overall, outperforming the other rate
+parameters in terms of corrected N50, least missing reference bases,
+\textsc{snp}s, and indels $>$5bp. With the exception of the 6
+bits/read rate, the assemblies decrease in rank with the increase in
+the rate parameter in terms of corrected N50, and least missing
+reference bases. This trend runs counter to the the decrease in
+\textsc{mse} of the different rates.
 
-\begin{figure}[!tbp]%figure2
+\begin{figure}[!tbp]
 %\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
@@
 
-Certain short read alignment tools use the quality value
-information when finding potential alignments. Bowtie2 (version 2.2.3)
-was used to evaluate the different decompressed \textsc{fastq}
-files. Bowtie2 uses quality values written in the \textsc{fastq} files
-when mapping reads against a reference genome. The original
-uncompressed and decompressed \textsc{fastq} files were mapped with
-Bowtie2 against \textit{Rhodobacter sphaeroides} reference genome. The
-generated \textsc{sam} file for each compressing approach were
-compared with the uncompressed \textsc{sam} file. The total, shared
-and unique proportional numbers of mapped reads are calculated with
-respect to the uncompressed \textsc{sam} matches as shown in table
+Certain short read alignment tools use the quality value information
+when finding potential alignments. Bowtie2 (version 2.2.3) was used to
+evaluate the different decompressed \textsc{fastq} files. Bowtie2 uses
+quality values written in the \textsc{fastq} files when mapping reads
+against a reference genome. The original uncompressed and decompressed
+\textsc{fastq} files were mapped with Bowtie2 against
+\textit{Rhodobacter sphaeroides} reference genome. The generated
+\textsc{sam} file for each compressing approach were compared with the
+uncompressed \textsc{sam} file. The total, shared and unique
+proportional numbers of mapped reads are calculated with respect to
+the uncompressed \textsc{sam} matches as shown in table
 \ref{tab:aligner}. Additionally, to monitor the effect of quality
@@
 
-We evaluate the compression methods using two approaches. In the
-first approach, we order the compression methods based on how similar
-the alignment results are using the original uncompressed quality values, i.e., the
-amount of reads aligned by both the uncompressed and compressed
-methods plus the amount of reads unaligned by both methods minus the
-amount of reads uniquely aligned by the uncompressed and compression
-method. In the second approach, we order the compression methods by
-total proportion of aligned reads.
+We evaluate the compression methods using two approaches. In the first
+approach, we order the compression methods based on how similar the
+alignment results are using the original uncompressed quality values,
+i.e., the amount of reads aligned by both the uncompressed and
+compressed methods plus the amount of reads unaligned by both methods
+minus the amount of reads uniquely aligned by the uncompressed and
+compression method. In the second approach, we order the compression
+methods by total proportion of aligned reads.
 
 The best compression method in terms of similarity with the
-uncompressed reads is \textsc{q}ual\textsc{c}omp with rate 100 bits/read, followed by
-\textsc{q}ual\textsc{c}omp with rate 30 bits/read, 256-, and 128-profile encoding, then
-2-bin encoding, 64-profile encoding, \textsc{q}ual\textsc{c}omp with rate 10 bits/read,
-7th-degree polynomial regression, \textsc{q}ual\textsc{c}omp with rate 6 bits/read, and
-finally, 5th-degree through 0-degree polynomial regression.
+uncompressed reads is \textsc{q}ual\textsc{c}omp with rate 100
+bits/read, followed by \textsc{q}ual\textsc{c}omp with rate 30
+bits/read, 256-, and 128-profile encoding, then 2-bin encoding,
+64-profile encoding, \textsc{q}ual\textsc{c}omp with rate 10
+bits/read, 7th-degree polynomial regression,
+\textsc{q}ual\textsc{c}omp with rate 6 bits/read, and finally,
+5th-degree through 0-degree polynomial regression.
 
 Ranking the compression methods by overall alignment rate produces an
-identical ordering as above. Aside from 0-degree polynomial
-regression (83.1\%), all other compression methods have an alignment
-rate between 87\% and 86.1\%. The alignment rate of the uncompressed
-reads is 87\%.
+identical ordering as above. Aside from 0-degree polynomial regression
+(83.1\%), all other compression methods have an alignment rate between
+87\% and 86.1\%. The alignment rate of the uncompressed reads is 87\%.
 
@@
 that are originally mapped, but unmapped by the compressed methods.
-\textsc{q}ual\textsc{c}omp with rate 100 bits/read results in the fewest missing
-original read alignments (159). Increasing the regression model
-polynomial degree results in a decreasing amount of reads that are
-originally mapped, but unmapped by the regression model (40,931 to
-1,134 reads for 0-degree and 7th-degree, respectively). There is no
-such trend for reads that are mapped only by the regression model.
+\textsc{q}ual\textsc{c}omp with rate 100 bits/read results in the
+fewest missing original read alignments (159). Increasing the
+regression model polynomial degree results in a decreasing amount of
+reads that are originally mapped, but unmapped by the regression model
+(40,931 to 1,134 reads for 0-degree and 7th-degree,
+respectively). There is no such trend for reads that are mapped only
+by the regression model.
 
-It is important to highlight the result that the 2-bin method aligns a greater portion of reads than the various regression models.   
-If a poor quality base is flanked by high quality sequence, then the low degree polynomial regression models tend to smooth out the base quality, erroneously marking the low quality base as higher quality.
-During alignment, Bowtie2 penalizes mismatches at high quality bases more so than mismatches at lower quality bases.   
-Thus, the polynomial regression models incur a high penalty for these bases, resulting in less alignments, despite having a better \textsc{mse} than 2-bin in most cases.
+It is important to highlight the result that the 2-bin method aligns a
+greater portion of reads than the various regression models.  If a
+poor quality base is flanked by high quality sequence, then the low
+degree polynomial regression models tend to smooth out the base
+quality, erroneously marking the low quality base as higher quality.
+During alignment, Bowtie2 penalizes mismatches at high quality bases
+more so than mismatches at lower quality bases.  Thus, the polynomial
+regression models incur a high penalty for these bases, resulting in
+less alignments, despite having a better \textsc{mse} than 2-bin in
+most cases.
 
@@
 experiment with different cutoffs for the \emph{okay} value and
-determine if the additional storage is noticeable in subsequent
-analyses.
+determine if the additional storage costs are associated with positive
+effects on subsequent analyses.
 
@@
 
-The downside of modeling quality sequences using polynomial regression is that the model often requires a high number of degrees to achieve the same \textsc{mse} as the profile and \textsc{QualComp} methods.
-However, storing a high number of coefficients requires more storage than losslessly compressing the original data.
-In order to increase the compressibility of modeling, we can attempt to store the \emph{profiles} of certain polynomial functions.
-In other words, we can use a spline (a function that is piecewise-defined by polynomial functions) to represent a given quality sequence.
-Similar to our profile-encoding method, the user can specify how many polynomial functions they wish to store.
-Then a quality sequence can be divided evenly into a given number of segments, and each segment can be annotated with a polynomial function profile that closely matches its quality sequence.
+The downside of modeling quality sequences using polynomial regression
+is that the model often requires a high number of degrees to achieve
+the same \textsc{mse} as the profile and \textsc{QualComp} methods.
+However, storing a high number of coefficients requires more storage
+than losslessly compressing the original data.  In order to increase
+the compressibility of modeling, we can attempt to store the
+\emph{profiles} of certain polynomial functions.  In other words, we
+can use a spline (a function that is piecewise-defined by polynomial
+functions) to represent a given quality sequence.  Similar to our
+profile-encoding method, the user can specify how many polynomial
+functions they wish to store.  Then a quality sequence can be divided
+evenly into a given number of segments, and each segment can be
+annotated with a polynomial function profile that closely matches its
+quality sequence.
 
@@
 We have simply provided here the initial steps in analyzing the effect
-of lossy compression on quality sequences using two
-high-coverage bacterial, human, and mouse data sets.
-A more direct comparison against additional lossy compression tools, such as
+of lossy compression on quality values using high-coverage bacterial,
+human, and mouse data sets.  A more direct comparison against
+additional lossy compression tools, such as
 \textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
-performed. Furthermore, other types of sequencing data can be
-analyzed apart from the Illumina data examined here. For example, the
-PacBio sequencing instruments produce very long reads (with average
-read lengths on the order of 10 kbp), but with the trade-off of having
-a high error-rate ($\sim$15\%). Unlike the class of quality values we
+performed. Furthermore, other types of sequencing data can be analyzed
+apart from the Illumina data examined here. For example, the PacBio
+sequencing instruments produce very long reads (with average read
+lengths on the order of 10 kbp), but with the trade-off of having a
+high error-rate ($\sim$15\%). Unlike the class of quality values we
 have examined here, the distribution of erroneous bases is relatively
@@
 \section*{Acknowledgement}
-We thank the 2014 Bioinformatics
-Exchange for Students and Teachers (\textsc{best}) Summer School, 
-funded by the offices of the Dean of The Graduate School at University
-of Maryland and the Rektor of University of T\"{u}bingen, where this 
-research was initiated.
+This research was initiated at the 2014 Bioinformatics Exchange for
+Students and Teachers (\textsc{best}) Summer School, which was funded
+by the offices of the Dean of The Graduate School at University of
+Maryland and the Rektor of University of T\"{u}bingen.
 

### 2015-04-24 manuscript/q-compression.tex
Rewrote some of the preprocessing section.

@@
 We evaluate effects of information loss from quality value compression
-on quality control steps of read filtering and trimming, which were
+on quality control steps of trimming and read filtering, which were
 performed using Sickle~\citep{sickle}, and make comparison to
 uncompressed data.
+Sickle starts at the ends of the sequence and uses a sliding window ($0.1 \times$ the read length) to find locations where the window's average quality falls below a given threshold (20 by default).
+The sequence is then trimmed from this position to the respective end.
+%uses a variable sliding window approach where base qualities are averaged over a window ($0.1 \times$ the read length) and trims from the position where the window's average quality falls below a given threshold (20 by default).
+If the trimmed sequence's length is less than a certain threshold (20 by default), then the sequence is discarded.
 
@@
 
-The majority of compression methods retain more base-pairs after
-preprocessing than the uncompressed sequences (Figure
+Preprocessing involves two steps: \emph{trimming} the poor quality regions of the sequences and \emph{discarding} sequences that are deemed to be poor quality overall.
+After trimming, the majority of compression methods retain more base-pairs than the uncompressed sequences (Figure
 \ref{fig:preprocessing}). In general, as a given compression model
@@
 The 2-bin approach is the only compression method that results in a
-higher number of filtered base-pairs across all data sets. Sickle uses
+higher number of trimmed base-pairs across all data sets. Sickle uses
 a sliding window approach to smooth the read quality values before it
-trims the sequence based on a specific quality threshold. In the 2-bin
+trims. In the 2-bin
 approach, there is not an even distribution of values per bin.  In
@@
 quality threshold when compressed, resulting in an increased number of
-filtered bases.
+trimmed bases.
 
@@
 filtering threshold, then no base-pairs are trimmed. Only reads that
-are comprised of mainly low quality values will be filtered.
+are comprised of mainly low quality values will be discarded.
 
-It is important to highlight that the even though a compression method
-may result in the same number of filtered base calls as the
-uncompressed sequences, it does not mean the \emph{same} base calls
-were filtered. The 1st-degree and 5th-degree polynomial regression of
-the \textit{Rhodobacter sphaeroides} fragment library filters nearly
+It is important to highlight that even though a compression method
+may result in the same number of trimmed bases as the
+uncompressed sequences, it does not mean the \emph{same} bases
+were trimmed. The 1st-degree and 5th-degree polynomial regression of
+the \textit{Rhodobacter sphaeroides} fragment library filter nearly
 as many bases as each other. However, if we examine the specific reads
-filtered, the 5th-degree polynomial regression discards approximately
+discarded, the 5th-degree polynomial regression discards approximately
 two thirds less reads than the 1st-degree polynomial regression that
@@
 \end{figure}
+
+\begin{table*}[!tbhp]
+\centering
+\caption[]{Mapping results of decompressed \textsc{fastq} files
+  against \textit{Rhodobacter sphaeroides} reference genome using
+  Bowtie2. Numbers corresponds to the proportion of mapped reads with
+  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
+  percentage of mapped reads by both the uncompressed and decompressed
+  data. ``Uncompressed only'' denotes the percentage of reads mapped
+  from the uncompressed data that are not mapped after
+  decompression. ``Compressed only'' denotes the percentage of reads
+  mapped from the decompressed data that were not mapped before
+  compression.}
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
+& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
+\cline{2-12}
+& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
+{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
+\cline{2-12}
+%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
+& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
+& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
+\cline{2-12}
+& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
+{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
+\cline{2-12}
+%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
+& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
+& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
+\cline{2-12}
+& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
+{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
+\cline{2-12}
+%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
+& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
+
+\end{tabular}
+
+\label{tab:aligner}
+\end{table*}
 
@@
 % \end{table*}
-
-\begin{table*}[!tbhp]
-\centering
-\caption[]{Mapping results of decompressed \textsc{fastq} files
-  against \textit{Rhodobacter sphaeroides} reference genome using
-  Bowtie2. Numbers corresponds to the proportion of mapped reads with
-  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
-  percentage of mapped reads by both the uncompressed and decompressed
-  data. ``Uncompressed only'' denotes the percentage of reads mapped
-  from the uncompressed data that are not mapped after
-  decompression. ``Compressed only'' denotes the percentage of reads
-  mapped from the decompressed data that were not mapped before
-  compression.}
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
-& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
-\cline{2-12}
-& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
-{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
-\cline{2-12}
-%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
-& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
-\end{tabular}
-
-\bigskip
-
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
-& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
-\cline{2-12}
-& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
-{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
-\cline{2-12}
-%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
-& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
-\end{tabular}
-
-\bigskip
-
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
-&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
-& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
-\cline{2-12}
-& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
-{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
-\cline{2-12}
-%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
-& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
-
-\end{tabular}
-
-\label{tab:aligner}
-\end{table*}
 

### 2015-04-25 manuscript/q-compression.tex
Numerous small changes, particularly on table and table header to make things more accurate and consistent.

@@
   for Bioinformatics and Computational Biology, University of
-  Maryland, College Park, 20742 USA}
+  Maryland, College Park, Maryland, 20742 USA}
 %$\dagger$authors contributed equally
@@
 performed using Sickle~\citep{sickle}, and make comparison to
-uncompressed data.
-Sickle starts at the ends of the sequence and uses a sliding window ($0.1 \times$ the read length) to find locations where the window's average quality falls below a given threshold (20 by default).
-The sequence is then trimmed from this position to the respective end.
-%uses a variable sliding window approach where base qualities are averaged over a window ($0.1 \times$ the read length) and trims from the position where the window's average quality falls below a given threshold (20 by default).
-If the trimmed sequence's length is less than a certain threshold (20 by default), then the sequence is discarded.
+uncompressed data. Sickle starts at the ends of the sequence and uses
+a sliding window ($0.1 \times$ the read length) to find locations
+where the window's average quality falls below a given threshold (20
+by default). The sequence is then trimmed from this position to the
+respective end.
+%uses a variable sliding window approach where base qualities are
+%averaged over a window ($0.1 \times$ the read length) and trims from
+%the position where the window's average quality falls below a given
+%threshold (20 by default).
+If the trimmed sequence length is less than a certain threshold (20 by
+default), then the sequence is discarded.
 
@@
 
-Preprocessing involves two steps: \emph{trimming} the poor quality regions of the sequences and \emph{discarding} sequences that are deemed to be poor quality overall.
-After trimming, the majority of compression methods retain more base-pairs than the uncompressed sequences (Figure
-\ref{fig:preprocessing}). In general, as a given compression model
-increases in complexity, i.e., as the number of profiles, polynomial
-degrees, or rate increases, the amount of base-pairs kept approaches
-the number base-pairs kept using the uncompressed sequences. The
-compression methods on the \textit{Mus musculus} data set have the
-greatest proportion of retained base-pairs compared to the
-uncompressed sequences. The \textit{Escherichia coli} str. K-12 MG1655
-MiSeq data set has the smallest range.
+Preprocessing involves two steps: \emph{trimming} the poor quality
+regions of the sequences and \emph{discarding} sequences that are
+deemed to be poor quality overall. After trimming, the majority of
+compression methods retain more base-pairs than the uncompressed
+sequences (Figure \ref{fig:preprocessing}). In general, as a given
+compression model increases in complexity, i.e., as the number of
+profiles, polynomial degrees, or rate increases, the amount of
+base-pairs kept approaches the number base-pairs kept using the
+uncompressed sequences. The compression methods on the \textit{Mus
+  musculus} data set have the greatest proportion of retained
+base-pairs compared to the uncompressed sequences. The
+\textit{Escherichia coli} str. K-12 MG1655 MiSeq data set has the
+smallest range.
 
@@
 a sliding window approach to smooth the read quality values before it
-trims. In the 2-bin
-approach, there is not an even distribution of values per bin.  In
-other words, \emph{bad} quality values may range from 0-33, whereas
-\emph{good} values may only range from 34-40. Thus, mid-range quality
-values that are above the threshold (20 by default) are set below the
-quality threshold when compressed, resulting in an increased number of
-trimmed bases.
+trims. In the 2-bin approach, there is not an even distribution of
+values per bin. In other words, \emph{bad} quality values may range
+from 0-33, whereas \emph{good} values may only range from 34-40. Thus,
+mid-range quality values that are above the threshold (20 by default)
+are set below the quality threshold when compressed, resulting in an
+increased number of trimmed bases.
 
@@
 
-It is important to highlight that even though a compression method
-may result in the same number of trimmed bases as the
-uncompressed sequences, it does not mean the \emph{same} bases
-were trimmed. The 1st-degree and 5th-degree polynomial regression of
-the \textit{Rhodobacter sphaeroides} fragment library filter nearly
-as many bases as each other. However, if we examine the specific reads
+It is important to highlight that even though a compression method may
+result in the same number of trimmed bases as the uncompressed
+sequences, it does not mean the \emph{same} bases were trimmed. The
+1st-degree and 5th-degree polynomial regression of the
+\textit{Rhodobacter sphaeroides} fragment library filter nearly as
+many bases as each other. However, if we examine the specific reads
 discarded, the 5th-degree polynomial regression discards approximately
@@
 \centering
-\caption[]{Mapping results of decompressed \textsc{fastq} files
-  against \textit{Rhodobacter sphaeroides} reference genome using
-  Bowtie2. Numbers corresponds to the proportion of mapped reads with
-  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
-  percentage of mapped reads by both the uncompressed and decompressed
-  data. ``Uncompressed only'' denotes the percentage of reads mapped
-  from the uncompressed data that are not mapped after
-  decompression. ``Compressed only'' denotes the percentage of reads
-  mapped from the decompressed data that were not mapped before
-  compression.}
+\caption[]{Comparison of mapping for original reads and reads with
+  compressed/decompressed quality values. Reads and reference genome
+  are for \textit{Rhodobacter sphaeroides}, and mapping was performed
+  using Bowtie2.}
+\begin{small}
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
+ & & \multicolumn{2}{c|}{max-qual} & \multicolumn{2}{c|}{min-qual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{regression (0)} & \multicolumn{2}{c}{regression (1)} \\
 & & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
@@
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
+ & & \multicolumn{2}{c|}{regression (3)} & \multicolumn{2}{c|}{regression (5)} & \multicolumn{2}{c|}{regression (7)} & \multicolumn{2}{c|}{profile (64)} & \multicolumn{2}{c}{profile (128)} \\
 & & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
@@
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
-&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
+&  & \multicolumn{2}{c|}{profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
 & &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
@@
 & proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
-
 \end{tabular}
+\end{small}
 

### 2015-04-25 src/parse_compression_results_all.R
Added some of the plotting code (WIP).

@@ -1,0 +1,113 @@
+library(ggplot2)
+
+tables = c("~/cbcb2/q-compression/results/rhodo3/results/frag_2.fastq.compression",
+           #"~/cbcb2/q-compression/results/staph2/results/frag_1.fastq.compression",
+           "~/cbcb2/q-compression//results/ecoli2/results/MiSeq_Ecoli_MG1655_110721_PF_R1.fastq.compression",
+           "~/cbcb2/q-compression//results/hg14/results/frag_1.fastq.compression",
+           "~/cbcb2/q-compression/results/mouse/results//SRR032209.fastq.compression")
+
+dataset_names = c("Rhodobacter sphaeroides",
+#"S. aureus",
+"Escherichia coli",
+"Homo sapiens",
+"Mus musculus")
+
+x_breaks <- c(.02,.03,.04,.05,.06,.07,.08,.09,.1,.2,.3,.4,.5,.6,.7,.8,.9,1.0,2,3,4,5,6,7,8,9,10)
+names = c("2B", "R0", "R1","R3","R5","R7","P64","P128","P256","Q6","Q10","Q30","Q100")
+pch_comp = c(1,5,5,5,5,5,2,2,2,0,0,0,0)
+pch = c(16,18,18,18,18,18,17,17,17,15,15,15,15)
+y_breaks <- c(2,3,4,5,6,7,8,9,10,20,30,40,50,60,70,80,90,100,200,300)
+
+dfr_tables <- c()
+original_tables <- c()
+counter <- 1
+
+for (table in tables) {
+  organism <- read.table(table, row.names = 1, header = T)
+  remove = c("original", "maxqual", "minqual")
+# remove = c("maxqual", "minqual")
+  original_value <- organism[rownames(organism) %in% "original", ]
+  organism <- organism[!rownames(organism) %in% remove, ]
+  
+  y <- organism$mse
+  x_bzip2 <- organism$bits_bp
+ 
+#   x_bzip2 <- apply(cbind(organism$comp_bzip2, organism$orig_bzip2), 1, min, na.rm=TRUE)
+#   x_bzip2 <- (x_bzip2 * 8) / organism$bases
+#   
+#   x_comp <- organism$comp_size
+#   x_comp <- (x_comp * 8) / organism$bases
+#   
+#   dfr <- data.frame(x = x_comp, y = y, pch = pch_comp, names=names, dataset=dataset_names[[counter]])
+#   dfr_tables <- rbind(dfr_tables,dfr)
+  
[…]

### 2015-04-25 src/parse_preprocessing_all.R

@@ -1,0 +1,62 @@
+library(ggplot2)
+library(gplots)
+library(gtools)
+library(xtable)
+library(scales)
+#install.packages("gplots")
+
+tables = c("~/cbcb2/q-compression/results/rhodo3/results/frag_1.fastq.preprocessing",
+           #"~/cbcb2/q-compression/results/staph2/results/frag_1.fastq.preprocessing",
+           "~/cbcb2/q-compression/results/ecoli2/results/MiSeq_Ecoli_MG1655_110721_PF_R1.fastq.preprocessing",
+           "~/cbcb2/q-compression/results/hg14/results/frag_1.fastq.preprocessing",
+           "~/cbcb2/q-compression/results/mouse/results//SRR032209.fastq.preprocessing")
+
+dataset_names = c("Rhodobacter sphaeroides",
+                  #"S. aureus",
+                  "Escherichia coli",
+                  "Homo sapiens",
+                  "Mus musculus")
+
+dfr_tables <- c()
+original_tables <- c()
+counter <- 1
+
+for (table in tables) {
+  table <- read.table(table, header = T, row.names = 1)
+  original <- table[rownames(table) %in% "original", ]$bases_kept
+  table <-  table[!rownames(table) %in% c("original","minqual","maxqual"), ]
+  table[,5] <- table[,5] - original
+  
+  dfr <- data.frame(y=seq(1:length(table[,1])), x = table[,5], dataset = dataset_names[[counter]])
+  dfr_tables <- rbind(dfr_tables, dfr)
+  
+  counter <- counter + 1
+}
+
+rowLabels = c("2-bin", "Regression (0)", "Regression (1)", "Regression (3)", "Regression (5)",
+              "Regression (7)", "Profile (64)", "Profile (128)", "Profile (256)", "QualComp (6)", "QualComp (10)",
+              "QualComp (30)", "QualComp (100)")
+
+library(grid)
+
+
+pdf("preprocessing_results.pdf", width=2, height=4)
+p <- ggplot(dfr_tables, aes(x=jitter(x,factor = 1), y=y, color = dataset)) + geom_point() + scale_y_discrete(labels=rowLabels) + 
+  theme_bw() +
+  xlab("Proportion of bases kept compared to original") +
+  ylab("") + 
+  theme_bw() + theme(legend.position="bottom", legend.key = element_blank(), legend.title=element_blank(),
+                     legend.text = element_text(face="italic", size=6),
[…]

### 2015-04-25 manuscript/q-compression.tex
Text edit.

@@
 Among the compression methods, the profile encoding performed best,
-polynomial regression modeling with degree $>$ 7 performed worst, and
+polynomial regression modeling with degree $<=$ 7 performed worst, and
 other methods were intermediate (Fig.~\ref{fig:assembly_ranks}).

### 2015-04-25 src/parse_compression_results_all.R
Some small changes to some R code for figures to make labeling more consistent with text. Note the new figures have not yet been generated.

@@ -95,8 +95,8 @@ p <- ggplot(dfr_tables, aes(x = x, y = y, label=names, col=dataset, shape=pch)) 
   scale_x_log10(breaks = x_breaks, labels = x_breaks, limits = c(.027,11)) +
   geom_point(size=5, data = original_tables, aes(x = x, y = y, pch=c(8,8,8,8), label=c("","","",""))) +
   geom_point(data = data.frame(x=c(8),y=c(0)), size=5, aes(x=x, y=y, shape=c(8), label=c("")), color="black") +
-  xlab("Quality bits per bp") + 
-  ylab("Mean Squared Error") +
+  xlab("bits/base-call") + 
+  ylab("mean squared error") +
   geom_point(size=2) + 
   geom_text(hjust=-.3, vjust=-0.3, size=2, show_guide = FALSE) + 
   theme_bw() + theme(legend.position="bottom", legend.key = element_blank(), legend.title=element_blank(),

### 2015-04-25 src/parse_preprocessing_all.R

@@ -33,8 +33,8 @@ for (table in tables) {
   counter <- counter + 1
 }
 
-rowLabels = c("2-bin", "Regression (0)", "Regression (1)", "Regression (3)", "Regression (5)",
-              "Regression (7)", "Profile (64)", "Profile (128)", "Profile (256)", "QualComp (6)", "QualComp (10)",
+rowLabels = c("2-bin", "regression (0)", "regression (1)", "regression (3)", "regression (5)",
+              "regression (7)", "profile (64)", "profile (128)", "profile (256)", "QualComp (6)", "QualComp (10)",
               "QualComp (30)", "QualComp (100)")
 
 library(grid)

### 2015-04-27 src/parse_compression_results_all.R
Corrected plot labels to be consistent with text.

@@ -50,7 +50,7 @@ for (table in tables) {
   counter <- counter + 1
 }
 
-pdf("compression_results", width=10, height=5)
+pdf("compression_results.pdf", width=10, height=5)
 
 # Hack to print the asterisks on the axis.
 library(grid)

### 2015-04-27 src/parse_preprocessing_all.R

@@ -40,7 +40,7 @@ rowLabels = c("2-bin", "regression (0)", "regression (1)", "regression (3)", "re
 library(grid)
 
 
-pdf("preprocessing_results.pdf", width=2, height=4)
+pdf("preprocessing_results.pdf", width=3.5, height=3.5)
 p <- ggplot(dfr_tables, aes(x=jitter(x,factor = 1), y=y, color = dataset)) + geom_point() + scale_y_discrete(labels=rowLabels) + 
   theme_bw() +
   xlab("Proportion of bases kept compared to original") +

### 2015-05-01 manuscript/q-compression.tex
Cleaned up text a bit.  Added more to discussion.

@@
 Despite the fact that polynomial functions have significantly lower
-number of parameters (i.e. one to six coefficients) than a read-length
+number of parameters (i.e. one to six coefficients in our approach) than a read-length
 string of raw quality values, the necessity of using floating point
@@
 and then poses an optimization problem of how to allot these bits for
-a given position while minimizing the \textsc{mse}. The quality values
+a given position within the read while minimizing the \textsc{mse}. The quality values
 can be clustered before hand to produce more accurate models.
@@
 proportional numbers of mapped reads are calculated with respect to
-the uncompressed \textsc{sam} matches as shown in table
+the uncompressed \textsc{sam} matches as shown in Table
 \ref{tab:aligner}. Additionally, to monitor the effect of quality
@@
 
+It is important to highlight that the downstream applications dictate which lossy compression methods are acceptable.
+As highlighted earlier, 2-bin encoding is able to retain information about low quality bases that occur within stretches of high quality bases, whereas the polynomial regression methods tend to smooth these qualities.
+In addition to aligning less reads that the 2-bin encoding, the polynomial regression methods may also be hindering downstream tools that use the alignment results.
+A \textsc{snp} detection pipeline will commonly use the quality values to record if a potential difference from the reference is the result of genetic variation or sequencing error.
+Erroneously marking a base as higher quality, may lead to the base being incorrectly labeled as a \textsc{snp}.
+2-bin encoding is acceptable for alignment; however, it performs quite poorly in terms of assembly.
+In the case of assembly, the 7th-degree polynomial regression method outperforms 2-bin encoding in nearly all metrics.
+
 \subsection{Extension of 2-bin encoding}
@@
 have examined here, the distribution of erroneous bases is relatively
-uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
+uniform~\citep{Ferrarini:2013vf}. The assembly complexity of bacterial
 genomes can be greatly simplified, producing near complete genome
@@
 
+\section{Availability}
+Implementations of the compression methods written in Python and R and pipeline to reproduce the results are available at \url{https://github.com/cmhill/q-compression}
+
 \section*{Acknowledgement}

### 2015-05-01 manuscript/q-compression.tex
some small edits

@@
 the effects of lossy compression on identifying variants within a data
-set. We build on these prior works and access the effects of
-% quantifying compression efficiency, we also assess the effects of
-quality value information loss resulting from compression on
-additional subsequent genomic analyses including read preprocessing
-(filtering and trimming), genome assembly, and read mapping.
+set. We build on these prior works, and access the effects of quality
+value information loss resulting from compression on subsequent
+genomic analyses including read preprocessing (filtering and
+trimming), genome assembly, and read mapping.
 
@@
 Despite the fact that polynomial functions have significantly lower
-number of parameters (i.e. one to six coefficients in our approach) than a read-length
-string of raw quality values, the necessity of using floating point
-numbers to store coefficients greatly limits the compression potential
-of the method. In order to get meaningful compression on
-single-precision four-byte floating point numbers, one would have to
-relax on the least-squares approximation constraint to obtain
-compressible values on the byte level which is outside the scope of
-this study.
+number of parameters (i.e. one to six coefficients in our approach)
+than a read-length string of raw quality values, the necessity of
+using floating point numbers to store coefficients greatly limits the
+compression potential of the method. In order to get meaningful
+compression on single-precision four-byte floating point numbers, one
+would have to relax on the least-squares approximation constraint to
+obtain compressible values on the byte level which is outside the
+scope of this study.
 
@@
 
-\textsc{QualComp} is a lossy compression tool that models quality
-values as a multivariate Gaussian distribution, computing the mean and
-covariance for each position in the read\citep{Ochoa:2013rt}. Once the
-model parameters are calculated, they are stored by the decoder to
-later reconstruct a \emph{representative} quality value.
-\textsc{QualComp} takes as input a user-specified rate (bits per read)
-and then poses an optimization problem of how to allot these bits for
-a given position within the read while minimizing the \textsc{mse}. The quality values
-can be clustered before hand to produce more accurate models.
-
 \subsubsection{Profiling}
@@
 \begin{figure*}[!tb]%figure2
-%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
 \centerline{\includegraphics[width=7in]{compression_results.pdf}}
@@
 
+\subsection{Comparison to other methods}
+
+For comparison to other developed methods we use \textsc{QualComp}, a
+lossy compression tool that models quality values as a multivariate
+Gaussian distribution, computing the mean and covariance for each
+position in the read\citep{Ochoa:2013rt}. Once the model parameters
+are calculated, they are stored by the decoder to later reconstruct a
+\emph{representative} quality value.  \textsc{QualComp} takes as input
+a user-specified rate (bits/read) and then poses an optimization
+problem of how to allot these bits for a given position within the
+read while minimizing the \textsc{mse}. The quality values can be
+clustered before hand to produce more accurate models.
+
 \subsection{Performance evaluation}
@@
 by default). The sequence is then trimmed from this position to the
-respective end.
-%uses a variable sliding window approach where base qualities are
-%averaged over a window ($0.1 \times$ the read length) and trims from
-%the position where the window's average quality falls below a given
-%threshold (20 by default).
-If the trimmed sequence length is less than a certain threshold (20 by
-default), then the sequence is discarded.
+respective end. If the trimmed sequence length is less than a certain
+threshold (20 by default), then the sequence is discarded.
 
@@
 
-No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
-outperforms all others in all metrics (Table
-\ref{fig:assembly_ranks}).
-%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
-%using the uncompressed reads outperforms all compression methods in
-%terms of \textsc{lap}, NG50, relocations, translations, and
-%inversions.
-Among the compression methods, the profile encoding performed best,
-polynomial regression modeling with degree $<=$ 7 performed worst, and
-other methods were intermediate (Fig.~\ref{fig:assembly_ranks}).
+No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set was
+best in all metrics (Table \ref{fig:assembly_ranks}). Among the
+compression methods, the profile encoding performed best, polynomial
+regression modeling performed worst, and other methods were
+intermediate (Fig.~\ref{fig:assembly_ranks}).
 
@@
 \begin{figure}[!tbp]
-%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
 \centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
@@
 
-It is important to highlight that the downstream applications dictate which lossy compression methods are acceptable.
-As highlighted earlier, 2-bin encoding is able to retain information about low quality bases that occur within stretches of high quality bases, whereas the polynomial regression methods tend to smooth these qualities.
-In addition to aligning less reads that the 2-bin encoding, the polynomial regression methods may also be hindering downstream tools that use the alignment results.
-A \textsc{snp} detection pipeline will commonly use the quality values to record if a potential difference from the reference is the result of genetic variation or sequencing error.
-Erroneously marking a base as higher quality, may lead to the base being incorrectly labeled as a \textsc{snp}.
-2-bin encoding is acceptable for alignment; however, it performs quite poorly in terms of assembly.
-In the case of assembly, the 7th-degree polynomial regression method outperforms 2-bin encoding in nearly all metrics.
+It is important to highlight that the downstream applications dictate
+which lossy compression methods are acceptable. As highlighted
+earlier, 2-bin encoding is able to retain information about low
+quality bases that occur within stretches of high quality bases,
+whereas the polynomial regression methods tend to smooth these
+qualities.  In addition to aligning less reads that the 2-bin
+encoding, the polynomial regression methods may also be hindering
+downstream tools that use the alignment results. A \textsc{snp}
+detection pipeline will commonly use the quality values to record if a
+potential difference from the reference is the result of genetic
+variation or sequencing error.  Erroneously marking a base as higher
+quality, may lead to the base being incorrectly labeled as a
+\textsc{snp}.  2-bin encoding is acceptable for alignment; however, it
+performs quite poorly in terms of assembly. In the case of assembly,
+the 7th-degree polynomial regression method outperforms 2-bin encoding
+in nearly all metrics.
 
@@
 \section{Availability}
-Implementations of the compression methods written in Python and R and pipeline to reproduce the results are available at \url{https://github.com/cmhill/q-compression}
+Implementations of the compression methods written in Python and R and
+pipeline to reproduce the results are available at
+\url{https://github.com/cmhill/q-compression}
 

### 2015-05-01 manuscript/q-compression.tex
More edits on text to make writing more concise and to clean up tex file.

@@
   Maryland, College Park, Maryland, 20742 USA}
-%$\dagger$authors contributed equally
 \history{Received on XXXXX; revised on XXXXX; accepted on XXXX}
-% Hack to display authors contribution.
 \editor{Associate Editor: XXXXXXX}
@@
 
-\begin{figure}[!tpb]%figure2
+\begin{figure}[!tpb]
 \centerline{\includegraphics[width=3.35in]{profiles_128.png}}
@@
 data with the ordered nucleotides (1 to read length) representing the
-abscissa, and quality values representing the ordinate. In this
-research we model read quality values as polynomial functions obtained
-with least-squares fitting, as one approach to compression read
-quality values by modeling.
-
-Despite the fact that polynomial functions have significantly lower
-number of parameters (i.e. one to six coefficients in our approach)
-than a read-length string of raw quality values, the necessity of
-using floating point numbers to store coefficients greatly limits the
-compression potential of the method. In order to get meaningful
-compression on single-precision four-byte floating point numbers, one
-would have to relax on the least-squares approximation constraint to
-obtain compressible values on the byte level which is outside the
-scope of this study.
+abscissa, and quality values representing the ordinate. We model read
+quality values as polynomial functions obtained with least-squares
+fitting, as one approach to compression read quality values by
+modeling.  Although polynomial functions have significantly lower
+number of parameters than a read-length string of raw quality values
+(e.g., one to six coefficients in our approach), the use of floating
+point numbers to store coefficients greatly limits the compression
+potential of the method. The coefficients could be represented with
+fewer bits, perhaps with some loss of precision, but we have not
+considered alternative representations in this study.
 
@@
 
-\begin{figure*}[!tb]%figure2
+\begin{figure*}[!tb]
 \centerline{\includegraphics[width=7in]{compression_results.pdf}}
@@
 As large sets of quality strings show similar trends of quality over
-their length, it makes sense to identify such common patterns in the
-data and use them as reference profiles to approximate individual
-sequences of quality values. Such patterns can be readily determined
-by clustering data points (i.e. quality vectors) and using the
-resulting cluster centers as representative profiles.
-
-$k$-means clustering is a vector quantization method, partitioning a
-set of samples into $k$ sets that minimize within-cluster sum of
-squares~\citep{macqueen1967some}. Using a random subset of read
-quality values, a compression method can use the computed cluster
-centers as read quality profiles. As the problem is NP-hard, we use a
-heuristic iterative refinement approach by quickly converging to a
-locally optimal minimum provided by R~\citep{hartigan1979algorithm}.
-
-First, the method samples an adjustable number of reads randomly from
-the file to be used as a training set. Quality values are represented
-by vectors containing their \textsc{phred} scores corresponding to
-each position along the read. Subsequently, $k$-means clustering is
-performed on the training set until convergence. The obtained cluster
-centers will be the quality profile prototypes for the dataset.
-
-Once the $k$ quality profiles are determined, all reads are passed
-through the trained $k$-means predictor, with the nearest quality
-profile in Euclidean space being assigned to every read as their
-compressed representation.
-
-The compressed quality file therefore consists of an index enumerating
-the $k$ quality profiles, and a binary part containing the assigned
-quality profile index for each read.
-
-Although this approach is not able to capture randomly occurring
-outlier quality values, it ensures that the overall trends in quality
-value patterns are retained. Quality profiles capture different
-overall qualities and different drop-off positions and gradients. An
-example of 128 quality profiles are shown on Figure
-\ref{fig:profiles_128}.
+their length, it may be possible to identify such common patterns in
+the data and use them as reference profiles to approximate individual
+sequences of quality values. Here we use $k$-means clustering, a
+vector quantization method, which partitions a set of samples into $k$
+sets that minimize within-cluster sum of
+squares~\citep{macqueen1967some}. We sampled 10K reads at random and
+computed cluster centers as read quality profiles using heuristic
+iterative refinement approach by quickly converging to a locally
+optimal minimum~\citep{hartigan1979algorithm}.  The obtained cluster
+were then used quality profile prototypes for the dataset.  All reads
+are evaluated with the nearest quality profile in Euclidean space
+being assigned to every read as their compressed representation.  The
+compressed quality file therefore consists of an index enumerating the
+$k$ quality profiles, and a binary part containing the assigned
+quality profile index for each read.  Although this approach is not
+able to capture some read specific differences in quality values, it
+does approximates the overall patterns in quality values. An example
+of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.
 
@@
 NC\_007494.1, NC\_009007.1, NC\_009008.1).
-
-% Will be moved to supplemental.
-%\textit{Stapylococcus aureus} USA300, which are generated from a
-%fragment library (insert size of 180 nt; 1,294,104 paired-end reads)
-%and short-jump library (insert size of 3,500 nt; 3,494,070 reads).
-%The corresponding reference sequence was obtained from the NCBI RefSeq
-%database (NC\_010063.1, NC\_010079.1, NC\_012417.1).
 
@@
 
-\begin{figure}[!tbp]%figure2
-%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
+\begin{figure}[!tbp]
 \centerline{\includegraphics[width=3.65in]{preprocessing_results.pdf}}
@@
 \cline{2-12}
-%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
 & proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
@@
 \cline{2-12}
-%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
 & proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
@@
 \cline{2-12}
-%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
 & proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
@@
 quality results in the lowest proportion of mapped reads 72.8\%.
-
-% \begin{table*}[!tbhp]
-% \centering
-% \caption[]{Quality filtering/trimming for the uncompressed and
-%   decompressed \textsc{fastq} files. Numbers correspond to the number
-%   of reads that are kept or discarded after applying compression with
-%   respect to the original, uncompressed values.}
-% \label{tab:quality_control}
-% \begin{tabular}{lr|cc|cc|cc|cc|cc}
-%  & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
-%  & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
-%   \cline{2-12}
-% & kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
-% {\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
-% \cline{2-12}
-%  & proportion & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
-% \end{tabular}
-
-% \bigskip
-
-% \begin{tabular}{lr|cc|cc|cc|cc|cc}
-%  & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
-%  & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
-% \cline{2-12}
-% & kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
-% {\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
-% \cline{2-12}
-% & proportion & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
-% \end{tabular}
-
-% \bigskip
-
-% \begin{tabular}{lr|cc|cc|cc|cc|cc}
-% &  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
-%  & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
-% \cline{2-12}
-% & kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
-% {\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
-% \cline{2-12}
-% & proportion & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
-% \end{tabular}
-% \end{table*}
 
@@
 
-\subsection{Future of lossy compression in bioinformatics analyses}
-
-We have simply provided here the initial steps in analyzing the effect
-of lossy compression on quality values using high-coverage bacterial,
-human, and mouse data sets. A more direct comparison against
-additional lossy compression tools, such as
-\textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
-performed. Furthermore, other types of sequencing data can be analyzed
-apart from the Illumina data examined here. For example, the PacBio
-sequencing instruments produce very long reads (with average read
-lengths on the order of 10 kbp), but with the trade-off of having a
-high error-rate ($\sim$15\%). Unlike the class of quality values we
-have examined here, the distribution of erroneous bases is relatively
-uniform~\citep{Ferrarini:2013vf}. The assembly complexity of bacterial
-genomes can be greatly simplified, producing near complete genome
-assemblies, by utilizing a single run of these long
-reads~\citep{Koren:2013ye}. If long read sequencing technologies such
-as PacBio become more widely adopted, it would be of huge benefit to
-examine the potential of lossy compression algorithms on not only the
-quality values, but the biological sequencing data themselves.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-05-01 manuscript/q-compression.tex
More revisions to the text, and fixed an error in a reference entry.

@@
 
-As large sets of quality strings show similar trends of quality over
-their length, it may be possible to identify such common patterns in
-the data and use them as reference profiles to approximate individual
-sequences of quality values. Here we use $k$-means clustering, a
-vector quantization method, which partitions a set of samples into $k$
-sets that minimize within-cluster sum of
-squares~\citep{macqueen1967some}. We sampled 10K reads at random and
+Sets of quality strings show similar trends of quality over their
+length, and it is possible to identify common patterns in the data and
+use them as reference profiles to approximate individual sequences of
+quality values. Here we use $k$-means clustering, a vector
+quantization method that partitions a set of samples into $k$ sets
+that minimize within-cluster sum of
+squares~\citep{macqueen1967some}. We sampled 10$K$ reads at random and
 computed cluster centers as read quality profiles using heuristic
 iterative refinement approach by quickly converging to a locally
-optimal minimum~\citep{hartigan1979algorithm}.  The obtained cluster
-were then used quality profile prototypes for the dataset.  All reads
-are evaluated with the nearest quality profile in Euclidean space
-being assigned to every read as their compressed representation.  The
+optimal minimum~\citep{hartigan1979algorithm}.  All reads are
+evaluated with the nearest quality profile in Euclidean space being
+assigned to every read as their compressed representation.  The
 compressed quality file therefore consists of an index enumerating the
@@
 We compare the \textsc{mse} versus bits/base-call of the
-\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
-chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
+\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
+\textit{Escherichia coli}, and
 \textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
 only include the fragment libraries for the \textit{Rhodobacter
-  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
+  sphaeroides}, and \textit{Homo sapiens} data
 sets, but the additional short-jump library results are available in
@@
 performs slightly worse in terms of \textsc{mse}. In the
-\textit{Rhodobacter sphaeroides} 2.4.1 fragment library,
+\textit{Rhodobacter sphaeroides} fragment library,
 \textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has
@@
 
-No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set was
+No assembly of the \textit{Rhodobacter sphaeroides} data set was
 best in all metrics (Table \ref{fig:assembly_ranks}). Among the
@@
 
-It is important to highlight the result that the 2-bin method aligns a
+Note that the 2-bin method aligns a
 greater portion of reads than the various regression models. If a
@@
 
-It is important to highlight that the downstream applications dictate
-which lossy compression methods are acceptable. As highlighted
-earlier, 2-bin encoding is able to retain information about low
+It is important to note that the downstream applications dictate
+which lossy compression methods are acceptable. As noted
+above, 2-bin encoding is able to retain information about low
 quality bases that occur within stretches of high quality bases,

### 2015-05-02 manuscript/q-compression.tex
More polishing text and making the document more concise.

@@
 For comparison to other developed methods we use \textsc{QualComp}, a
-lossy compression tool that models quality values as a multivariate
-Gaussian distribution, computing the mean and covariance for each
-position in the read\citep{Ochoa:2013rt}. Once the model parameters
-are calculated, they are stored by the decoder to later reconstruct a
-\emph{representative} quality value.  \textsc{QualComp} takes as input
-a user-specified rate (bits/read) and then poses an optimization
-problem of how to allot these bits for a given position within the
-read while minimizing the \textsc{mse}. The quality values can be
-clustered before hand to produce more accurate models.
+lossy compression tool~\citep{Ochoa:2013rt}. The program models
+quality values as a multivariate Gaussian distribution, computing the
+mean and covariance for each position in the read, and these varlues
+are stored by the decoder to later reconstruct a \emph{representative}
+quality value. \textsc{QualComp} takes as input a user-specified rate
+(bits/read), and these bits are allotted among positions within the
+read to minimize the \textsc{mse}. The quality values can be clustered
+before hand to produce more accurate models.
 
@@
 \textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
-\textit{Escherichia coli}, and
-\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
-only include the fragment libraries for the \textit{Rhodobacter
-  sphaeroides}, and \textit{Homo sapiens} data
-sets, but the additional short-jump library results are available in
-the Supplemental Table 1. Storing the uncompressed quality values
-requires 1 byte per base-call because they are stored in
-\textsc{ascii} format and is denoted by the dotted black asterisk in
-the figure. The lossless compression of each data set using
-\textsc{bz}ip2 ranges from 2.19 - 3.10 bits/base-call and is denoted
-by the colored asterisks on the abscissa. The compression methods tend
-to cluster together across the different data sets. Across all data
-sets, the 0-degree polynomial regression, profile encodings, and
-\textsc{q}ual\textsc{c}omp have the lowest bits/base-call.
+\textit{Escherichia coli}, and \textit{Mus musculus} data sets (Figure
+\ref{fig:mse_vs_bpbp}). We only include the fragment libraries for the
+\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
+but the additional short-jump library results are available in the
+Supplemental Table 1. Storing the uncompressed quality values requires
+1 byte per base-call because they are stored in \textsc{ascii} format
+and is denoted by the dotted black asterisk in the figure. The
+lossless compression of each data set using \textsc{bz}ip2 ranges from
+2.19 - 3.10 bits/base-call and is denoted by the colored asterisks on
+the abscissa. The compression methods tend to cluster together across
+the different data sets. Across all data sets, the 0-degree polynomial
+regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
+lowest bits/base-call.
 
@@
 
-No assembly of the \textit{Rhodobacter sphaeroides} data set was
-best in all metrics (Table \ref{fig:assembly_ranks}). Among the
-compression methods, the profile encoding performed best, polynomial
-regression modeling performed worst, and other methods were
-intermediate (Fig.~\ref{fig:assembly_ranks}).
+No assembly of the \textit{Rhodobacter sphaeroides} data set was best
+in all metrics (Table \ref{fig:assembly_ranks}). Among the compression
+methods, the profile encoding performed best, polynomial regression
+modeling performed worst, and other methods were intermediate
+(Fig.~\ref{fig:assembly_ranks}).
 
@@
 
-Note that the 2-bin method aligns a
-greater portion of reads than the various regression models. If a
-poor quality base is flanked by high quality sequence, then the low
-degree polynomial regression models tend to smooth out the base
-quality, erroneously marking the low quality base as higher quality.
-During alignment, Bowtie2 penalizes mismatches at high quality bases
-more so than mismatches at lower quality bases. Thus, the polynomial
-regression models incur a high penalty for these bases, resulting in
-less alignments, despite having a better \textsc{mse} than 2-bin in
-most cases.
+Note that the 2-bin method aligns a greater portion of reads than the
+various regression models. If a poor quality base is flanked by high
+quality sequence, then the low degree polynomial regression models
+tend to smooth out the base quality, erroneously marking the low
+quality base as higher quality.  During alignment, Bowtie2 penalizes
+mismatches at high quality bases more so than mismatches at lower
+quality bases. Thus, the polynomial regression models incur a high
+penalty for these bases, resulting in less alignments, despite having
+a better \textsc{mse} than 2-bin in most cases.
 
@@
 
-\subsection{Lossy compression acceptable for subsequent biological analyses}
+\subsection{Lossy compression is acceptable for some subsequent  
+analyses}
 
-The primary concern of using lossy compression methods is naturally
-the extent of information loss, that we quantified by \textsc{mse} in
-this study. \textsc{mse} and compressibility provide information in
-the theoretical context to the methods, but they are not the end-all
-of evaluation criteria. The performance of compressed datasets in
-different subsequent analyses and applications are just as
-important. Our benchmarks showed that some of the compression methods
-with high error rates are still practical for certain kinds of
-applications. Many subsequent tools proved to have enough additional
-redundancy built-in to handle such loss in information. Passing the
-decompressed quality values through quality control software shows
-that most methods filter nearly as many bases as using original
-quality sequences. Assemblers performing sequence alignment use
-percent similarity scores that are typically robust to standard
-sequencing errors.
+It is common to evaluate lossy compression methods in terms of
+information loss, and some measure of compression effectiveness.
+However, from a pragmatic perspective, it is the effects on the
+results of subsequent analyses that are most important. Our results
+showed that some of the compression methods with relatively high
+information lossy (\textsc{mse}) are still practical for some
+applications. Many bioinformatics tools proved to be robust to
+moderate information loss. Passing the decompressed quality values
+through quality control software shows that most methods filter nearly
+as many bases as using original quality sequences. Assemblers
+performing sequence alignment use percent similarity scores that are
+typically robust to standard sequencing errors.
 
-It is important to note that the downstream applications dictate
-which lossy compression methods are acceptable. As noted
-above, 2-bin encoding is able to retain information about low
-quality bases that occur within stretches of high quality bases,
-whereas the polynomial regression methods tend to smooth these
-qualities. In addition to aligning less reads that the 2-bin
-encoding, the polynomial regression methods may also be hindering
-downstream tools that use the alignment results. A \textsc{snp}
-detection pipeline will commonly use the quality values to record if a
-potential difference from the reference is the result of genetic
-variation or sequencing error.  Erroneously marking a base as higher
-quality, may lead to the base being incorrectly labeled as a
+Downstream applications dictate which lossy compression methods are
+acceptable. As noted above, 2-bin encoding is able to retain
+information about low quality bases that occur within stretches of
+high quality bases, whereas the polynomial regression methods tend to
+smooth these qualities. In addition to aligning less reads that the
+2-bin encoding, the polynomial regression methods may also be
+hindering downstream tools that use the alignment results. A
+\textsc{snp} detection pipeline will commonly use the quality values
+to record if a potential difference from the reference is the result
+of genetic variation or sequencing error. Erroneously marking a base
+as higher quality, may lead to the base being incorrectly labeled as a
 \textsc{snp}.  2-bin encoding is acceptable for alignment; however, it
-performs quite poorly in terms of assembly. In the case of assembly,
-the 7th-degree polynomial regression method outperforms 2-bin encoding
-in nearly all metrics.
-
-\subsection{Extension of 2-bin encoding}
-
-2-bin encoding has the nice property of being simple to compute and
-has good bits/base-call values. The 2-bin encoding suffers from having
-a high \textsc{mse}, but fortunately, we have shown that in the case
-of genome assembly, 2-bin encoding outperforms all polynomial
-regressions encodings with degree less than 3. 2-bin encoding of the
-fragment and short-jump libraries of \textit{Rhodobacter sphaeroides}
-have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-degree
-polynomial regression encodings, respectively. This further highlights
-the importance of using additional contextual information of the
-subsequent analyses when evaluating compressed quality values.
-
-A potential extension to 2-bin encoding is to incorporate an
-additional bin (\emph{okay}). The \emph{okay} value can be used where
-the base qualities fall within a 2-bin range. Because the distribution
-of quality values is skewed towards higher quality, we need to
-experiment with different cutoffs for the \emph{okay} value and
-determine if the additional storage costs are associated with positive
-effects on subsequent analyses.
-
-\subsection{Extension of polynomial regression}
-
-The downside of modeling quality sequences using polynomial regression
-is that the model often requires a high number of degrees to achieve
-the same \textsc{mse} as the profile and \textsc{QualComp} methods.
-However, storing a high number of coefficients requires more storage
-than losslessly compressing the original data. In order to increase
-the compressibility of modeling, we can attempt to store the
-\emph{profiles} of certain polynomial functions. In other words, we
-can use a spline (a function that is piecewise-defined by polynomial
-functions) to represent a given quality sequence. Similar to our
-profile-encoding method, the user can specify how many polynomial
-functions they wish to store. Then a quality sequence can be divided
-evenly into a given number of segments, and each segment can be
-annotated with a polynomial function profile that closely matches its
-quality sequence.
+performs quite poorly in terms of assembly, and 7th-degree polynomial
+regression modeling outperforms it in nearly all metrics.
 

### 2015-05-02 manuscript/q-compression.tex
Still more revising ...

@@
 
-\begin{figure*}[!tb]
-\centerline{\includegraphics[width=7in]{compression_results.pdf}}
-\caption{The relationship of bits/base-call and mean squared error for
-  quality value compression methods applied to four data sets:
-  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
-  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
-  \textit{Mus musculus}. Point labels correspond to different
-  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
-  profiles; R$n$, modeling with polynomial regression of degree $n$;
-  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
-  $n$. Asterisks denote the corresponding lossless compression using
-  \textsc{bz}ip2, with the black asterisk corresponding to original
-  data.}
-\label{fig:mse_vs_bpbp}
-\end{figure*}
-
 Sets of quality strings show similar trends of quality over their
@@
 
+\begin{figure*}[!tb]
+\centerline{\includegraphics[width=7in]{compression_results.pdf}}
+\caption{The relationship of bits/base-call and mean squared error for
+  quality value compression methods applied to four data sets:
+  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
+  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
+  \textit{Mus musculus}. Point labels correspond to different
+  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
+  profiles; R$n$, modeling with polynomial regression of degree $n$;
+  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
+  $n$. Asterisks denote the corresponding lossless compression using
+  \textsc{bz}ip2, with the black asterisk corresponding to original
+  data.}
+\label{fig:mse_vs_bpbp}
+\end{figure*}
+
 We compare the \textsc{mse} versus bits/base-call of the
@@
 lossless compression of each data set using \textsc{bz}ip2 ranges from
-2.19 - 3.10 bits/base-call and is denoted by the colored asterisks on
+2.19--3.10 bits/base-call and is denoted by the colored asterisks on
 the abscissa. The compression methods tend to cluster together across
@@
 \textsc{q}ual\textsc{c}omp with the rate parameter set to 100
-bits/read has the lowest \textsc{mse}, but requires 10-15x more
-storage than the profile encoding methods for only a 2-3x reduction in
-\textsc{mse}. When \textsc{q}ual\textsc{c}omp's rate parameter is set
-to match the profile encoding methods, \textsc{q}ual\textsc{c}omp
-performs slightly worse in terms of \textsc{mse}. In the
-\textit{Rhodobacter sphaeroides} fragment library,
-\textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has
-a \textsc{mse} of 17.29. Using 256-profile encoding requires less
-storage (0.079 bits/bp) and has a lower \textsc{mse} (11.85).
+bits/read has the lowest \textsc{mse}, but requires
+$\sim$10--15$\times$ more storage than the profile encoding methods
+for only a $\sim$2--3$\times$ reduction in \textsc{mse}. When
+\textsc{q}ual\textsc{c}omp's rate parameter is set to match the
+profile encoding methods, \textsc{q}ual\textsc{c}omp performs slightly
+worse in terms of \textsc{mse}. In the \textit{Rhodobacter
+  sphaeroides} fragment library, \textsc{q}ual\textsc{c}omp with rate
+10 bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29. Using
+256-profile encoding requires less storage (0.079 bits/bp) and has a
+lower \textsc{mse} (11.85).
 
@@
 
-\begin{figure}[!tbp]
-\centerline{\includegraphics[width=3.65in]{preprocessing_results.pdf}}
-\caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
-  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
-  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
-    musculus} data sets. Sequences were trimmed using Sickle. The
-  total amount of bases filtered by each compression method is
-  compared with the amount of bases filtered using the uncompressed
-  sequences.}
-  \label{fig:preprocessing}
-\end{figure}
+As an additional point of reference regarding compression
+effectiveness the recently published
+Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved best
+case compression of 0.254 bits/base-call, and a mean compression of
+1.841 bits/base-call; both values higher than all of the profile
+results and for some results of other methods presented here
+(Figure~\ref{fig:mse_vs_bpbp}).
 
@@
 base-pairs compared to the uncompressed sequences. The
-\textit{Escherichia coli} str. K-12 MG1655 MiSeq data set has the
-smallest range.
+\textit{Escherichia coli} MiSeq data set has the smallest range.
 
@@
 values per bin. In other words, \emph{bad} quality values may range
-from 0-33, whereas \emph{good} values may only range from 34-40. Thus,
-mid-range quality values that are above the threshold (20 by default)
-are set below the quality threshold when compressed, resulting in an
-increased number of trimmed bases.
+from 0--33, whereas \emph{good} values may only range from
+34--40. Thus, mid-range quality values that are above the threshold
+(20 by default) are set below the quality threshold when compressed,
+resulting in an increased number of trimmed bases.
 
@@
 respectively) (Supplemental Table 2).
+
+\begin{figure}[!tbp]
+\centerline{\includegraphics[width=3.65in]{preprocessing_results.pdf}}
+\caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
+  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
+  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
+    musculus} data sets. Sequences were trimmed using Sickle. The
+  total amount of bases filtered by each compression method is
+  compared with the amount of bases filtered using the uncompressed
+  sequences.}
+  \label{fig:preprocessing}
+\end{figure}
 
@@
 Perhaps one of the greatest potential benefits of compressing quality
-values is the potential to perform quality control and possibly other
-operations directly on the compressed representations of the
-data. This is easiest to to consider for profile-based
-compression. The $k$ profiles can be evaluated for (pre-)processing
-operations such as filtering and trimming, and the operations
-transitively applied to the entire set of reads, thus saving
-substantial computation associated with evaluating the full set of
-reads.
+values is the ability to perform quality control and possibly other
+operations directly on the compressed representations of the data. For
+example, with profile-based compression the $k$ profiles can be
+evaluated for (pre-)processing operations such as filtering and
+trimming, and the operations transitively applied to the entire set of
+reads, thus saving substantial computation.
 
@@
 values and their effect on subsequent analyses. Although most previous
-examinations on lossy compression primarily focused on information
-loss, we have shown that typically used bioinformatics software today
-have additional built-in sensitivity to handle significant loss of
-information in the compressed quality values.
+examinations on lossy compression focused primarily on information
+loss and compression effectiveness, we show introduce several general
+approaches to quality value compression and show that common
+bioinformatic analyses are robust substantial loss of information in
+quality values.
 

### 2015-05-03 manuscript/q-compression.tex
Continued revisions ...

@@
 performed using Sickle~\citep{sickle}, and make comparison to
-uncompressed data. Sickle starts at the ends of the sequence and uses
+uncompressed data. Sickle starts at the ends of the read and uses
 a sliding window ($0.1 \times$ the read length) to find locations
-where the window's average quality falls below a given threshold (20
+where the mean quality in the window falls below a given threshold (20
 by default). The sequence is then trimmed from this position to the
@@
 \textit{Escherichia coli}, and \textit{Mus musculus} data sets (Figure
-\ref{fig:mse_vs_bpbp}). We only include the fragment libraries for the
+\ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
 \textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
-but the additional short-jump library results are available in the
-Supplemental Table 1. Storing the uncompressed quality values requires
-1 byte per base-call because they are stored in \textsc{ascii} format
-and is denoted by the dotted black asterisk in the figure. The
-lossless compression of each data set using \textsc{bz}ip2 ranges from
+but the short-jump library results are available in the Supplemental
+Table 1. Storing the uncompressed quality values requires 1
+byte/base-call because they are stored in \textsc{ascii} format and is
+denoted by the dotted black asterisk in the figure. The lossless
+compression of each data set using \textsc{bz}ip2 ranges from
 2.19--3.10 bits/base-call and is denoted by the colored asterisks on
@@
 $\sim$10--15$\times$ more storage than the profile encoding methods
-for only a $\sim$2--3$\times$ reduction in \textsc{mse}. When
-\textsc{q}ual\textsc{c}omp's rate parameter is set to match the
-profile encoding methods, \textsc{q}ual\textsc{c}omp performs slightly
-worse in terms of \textsc{mse}. In the \textit{Rhodobacter
-  sphaeroides} fragment library, \textsc{q}ual\textsc{c}omp with rate
-10 bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29. Using
-256-profile encoding requires less storage (0.079 bits/bp) and has a
-lower \textsc{mse} (11.85).
+for only a $\sim$2--3$\times$ reduction in \textsc{mse}. When rate
+parameter of \textsc{q}ual\textsc{c}omp is set to match the profile
+encoding methods, \textsc{q}ual\textsc{c}omp performs slightly worse
+in terms of \textsc{mse}. For the \textit{Rhodobacter sphaeroides}
+fragment library, \textsc{q}ual\textsc{c}omp with rate 10 bits/read
+(0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
+encoding only requires 0.079 bits/bp and has a \textsc{mse} of 11.85.
 

### 2015-05-04 manuscript/q-compression.tex
Added back multi-bin encoding.  Updated preprocessing figure and author list.

@@
 \title[lossy-compression]{Lossy compression of DNA sequencing quality
-  data} \author[El Hadidi \textit{et~al.}]{Mohamed El
-  Hadidi\,$^{1}$\footnote{authors contributed equally}, Christopher
-  M. Hill\,$^{2}$\footnotemark[1], Andr\'{a}s
-  Szolek\,$^{3}$\footnotemark[1], and Michael
-  P. Cummings\,$^4$\footnote{to whom correspondence should be
-    addressed}} \address{$^{1}$Department of Algorithms in
+  data} \author[Hill \textit{et~al.}]{Christopher
+  M. Hill\,$^{1}$, Andr\'{a}s Szolek\,$^{2}$, Mohamed El
+  Hadidi\,$^{3}$, and Michael P. Cummings\,$^4$\footnote{to whom correspondence should be
+    addressed}} \address{$^{1}$Department of Computer Science, University of Maryland,
+  College Park, Maryland, 20742 USA\\ $^{2}$Department of Applied
+  Bioinformatics, Center for Bioinformatics, Quantitative Biology
+  Center, and Department of Computer Science, University of
+  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany\\ 
+  $^{3}$Department of Algorithms in
   Bioinformatics, Center for Bioinformatics, University of
   T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany
-  \\ $^{2}$Department of Computer Science, University of Maryland,
-  College Park, Maryland, 20742 USA\\ $^{3}$Department of Applied
-  Bioinformatics, Center for Bioinformatics, Quantitative Biology
-  Center, and Department of Computer Science, University of
-  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany\\ $^{4}$Center
-  for Bioinformatics and Computational Biology, University of
+  \\ $^{4}$ Center for Bioinformatics and Computational Biology, University of
   Maryland, College Park, Maryland, 20742 USA}
@@
 \end{figure}
+
+While we focus only on two bins in this work, a greater number of bins can be used in practice.
+For example, an additional third bin (\emph{okay}) can be used for base qualities near the border between bins in the 2-bin encoding.
+Examining the trade-off of additional bins and their resulting compressibility and affect on downstream analyses is outside the scope of this work.
 

### 2015-05-04 manuscript/q-compression.tex
Added spline regression.

@@
 costs that are a function of the number of reads encoded.
+
+Profiles can be obtained from techniques other than $k$-means clustering.
+%While we focus on profiles obtained from $k$-means clustering, other profiling techniques can be used in practice.
+We can store the profiles of polynomial functions to increase the compressibility of polynomial regression modeling.
+%For example, in order to increase the compressibility of polynomial regression modeling, we can store the profiles of polynomial functions.
+In other words, we can use a spline (a function that is piecewise-defined by polynomial functions) to represent a given quality sequence.
+Exploring other potential applications of profile-based compression is outside the scope of this work.
 

### 2015-05-07 manuscript/q-compression.tex
Incorporated Adam's suggestions.

@@
 individual base calls. As the cost of sequencing continues to
-decrease, the rate of sequencing data production is increasing,
+decrease, the rate of sequence data production is increasing,
 requiring efficient ways of storing and transferring this vast amount
-of data. Most methods on sequencing data compression focus on
+of data. Most methods of sequencing data compression focus on
 compressing nucleotide information without any loss of information.
@@
 
-Existing methods for compressing quality sequences mostly focused on
-minimizing the loss of information with less emphasis on effects on
+Existing methods for compressing quality sequences mostly focus on
+minimizing the loss of information with less emphasis on the effects on
 subsequent analyses. In this paper, we evaluate several different
@@
 Lossy compression of quality information can greatly decrease storage
-and memory requirements with little discernible effects on subsequent
+and memory requirements with little discernible effect on subsequent
 analysis results. The three compression strategies in this study were
-able to produce similar results to those obtained with uncompressed
-quality sequences in terms of quality control, genome assembly, and
-alignment of short read to a reference sequence.
+able to produce results similar to those obtained with uncompressed
+quality sequences when performing quality control, genome assembly, and
+alignment of short reads to a reference sequence.
 
@@
 inclusion of quality values, and read-level naming and
-metadata. Because of this abundance effective compression of read data
-has the potential for substantial improvement in data storage and
+metadata. Because of this abundance, effective compression of read data
+has the potential to substantially improve data storage and
 transfer efficiency.
@@
 files~\citep{Cock:2010ve}, a very common format for sequence read
-data. At the level of sequence read the probability of error for each
-base-call is typically represented by \textsc{phred} quality value,
+data. At the level of the sequence read, the probability of error for each
+base-call is typically represented by a \textsc{phred} quality value,
 which is defined as $Q =
 -10\,log_{10}P$~\citep{Ewing:1998ly}. Depending on the sequencing
-technology these quality values can range from 0 to 93, and are
+technology, these quality values can range from 0 to 93, and are
 represented with the \textsc{ascii} characters 33 to 126 (with some
@@
 prior to subsequent analysis steps. Quality control based on quality
-values generally includes two operations: \textit{i}.~filtering, the
-elimination of reads that on the whole do not meet arbitrary quality
-standards, which reduces the total number of reads; and
-\textit{ii}.~trimming of low quality base-calls from reads, which
+values generally includes two operations: \textit{i}.~filtering, which removes reads
+that on the whole do not meet arbitrary quality
+standards, thus reducing the total number of reads; and
+\textit{ii}.~trimming of low-quality base-calls from reads, which
 reduces the number total number of bases. Quality values can be used
@@
 assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
-alignment software, such as Bowtie2~\citep{Langmead:2012rw}, use
+alignment programs, such as Bowtie2~\citep{Langmead:2012rw}, use
 quality values to weight mismatches between read and reference
@@
 and identified \textsc{snp}s with high-quality calls are deemed more
-reliable than those with low-quality calls, particularly in low
-coverage regions.
+reliable than those with low-quality calls, particularly in low-coverage regions.
 
@@
 sequencing technologies. Sequence reads generated by instruments such
-as an Illumina HiSeq, the focus of this research, are characterized by
+as an Illumina HiSeq, the focus of this research due to its ubiquitousness in bioinformatics research, are characterized by
 having relatively few insertion and deletion errors, but substitution
@@
 length (e.g., error rates increase up to $\sim$16$\times$ at the
-3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low quality tails
+3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low-quality tails
 at the 3$^{\prime}$ end~\citep{Minoche:2011km}).
@@
 is largely pragmatic with a focus on the use of quality values in
-subsequent analyses. From this perspective some loss of information is
+subsequent analyses. From this perspective, some loss of information is
 deemed acceptable if the inferences from analyses are relatively
@@
 associated with Illumina instruments, with the objective to provide
-some perspective on several strategies rather than to develop a robust
+some perspective on several strategies rather than to develop robust
 high-quality software for use. Recognizing these properties of
@@
 classes of lossy compression methods -- binning, modeling, and
-profiling -- and consider an exemplar of each
+profiling -- and our consideration of an exemplar of each
 class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated
 the effects of lossy compression on identifying variants within a data
-set. We build on these prior works, and access the effects of quality
+set. We build on these prior works, and assess the effects of quality
 value information loss resulting from compression on subsequent
-genomic analyses including read preprocessing (filtering and
+genomic analyses, including read preprocessing (filtering and
 trimming), genome assembly, and read mapping.
@@
 Quality values can be binned, and the minimum number of bins that
-allows for a any distinction among quality values is two, i.e., two
-categories ``good'' and ``bad'' quality. We implement 2-bin encoding
+allows for a any distinction among quality values is two; i.e., two
+categories of quality: ``good'' and ``bad''. We implement 2-bin encoding
 by setting a quality value threshold empirically determined by the
@@
 
-With 2-bin encoding binary encoding is possible, allowing us to use a
+With 2-bin encoding, binary encoding is possible, allowing us to use a
 single bit to represent the quality of a base instead of the standard
@@
 Truncating treats a user-defined number of highest quality values as a
-single bin. LogBinning works similar to UniBinning, except with the
+single bin. LogBinning works similarly to UniBinning, except it uses the
 \emph{log} of the error probability distribution, which effectively
 bins the \textsc{ascii} quality values evenly. Our 2-bin encoding is
-a combination of LogBinning and Truncating in that we are placing the
+a combination of LogBinning and Truncating in that we place the
 highest quality values (as defined above) of the log of the error
@@
 While we focus only on two bins in this work, a greater number of bins can be used in practice.
-For example, an additional third bin (\emph{okay}) can be used for base qualities near the border between bins in the 2-bin encoding.
-Examining the trade-off of additional bins and their resulting compressibility and affect on downstream analyses is outside the scope of this work.
+For example, an additional third bin (good, bad, and \emph{okay}) can be used for base qualities near the border between bins in the 2-bin encoding.
+Examining the trade-off of additional bins and their resulting compressibility and effect on downstream analyses is outside the scope of this work.
 
@@
 quality values as polynomial functions obtained with least-squares
-fitting, as one approach to compression read quality values by
-modeling.  Although polynomial functions have significantly lower
-number of parameters than a read-length string of raw quality values
+fitting, as one approach to the compression of read quality values by
+modeling.  Although polynomial functions have significantly fewer
+parameters than a read-length string of raw quality values
 (e.g., one to six coefficients in our approach), the use of floating
@@
 general terms, include no fixed costs, and marginal costs that are a
-function of the number of reads times the cost representing the model
+function of the number of reads times the cost of representing the model
 parameters.
@@
 squares~\citep{macqueen1967some}. We sampled 10$K$ reads at random and
-computed cluster centers as read quality profiles using heuristic
-iterative refinement approach by quickly converging to a locally
+computed cluster centers as read quality profiles using a heuristic
+iterative refinement approach that quickly converges to a locally
 optimal minimum~\citep{hartigan1979algorithm}.  All reads are
@@
 quality profile index for each read.  Although this approach is not
-able to capture some read specific differences in quality values, it
-does approximates the overall patterns in quality values. An example
-of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.
+able to capture some read-specific differences in quality values, it
+does approximate the overall patterns in quality values. An example
+of 128 quality profiles is shown in Figure \ref{fig:profiles_128}.
 
@@
 which are taken from data from the \textsc{gage} (Genome Assembly
-Gold-Standard Evaluations)~\citep{Salzberg:2012rc} except as
+Gold-Standard Evaluations)~\citep{Salzberg:2012rc}, except as
 noted. These data sets are as follows.
@@
 and short-jump library (insert size of 3,500 nt; 2,050,868 reads). The
-corresponding reference sequence was obtained from the NCBI RefSeq
+corresponding reference sequence was obtained from the \textsc{ncbi} RefSeq
 database (NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1,
@@
 22,669,408 reads). The corresponding reference sequence was obtained
-from the NCBI RefSeq database (NC\_000014.8).
+from the \textsc{ncbi} RefSeq database (NC\_000014.8).
 
@@
 1,145,8940 paired-end reads). The corresponding reference sequence was
-obtained from the NCBI RefSeq database (NC\_000913.2).
+obtained from the \textsc{ncbi} RefSeq database (NC\_000913.2).
 
@@
 (bits/read), and these bits are allotted among positions within the
-read to minimize the \textsc{mse}. The quality values can be clustered
-before hand to produce more accurate models.
+read to minimize the average error. The quality values can be clustered
+beforehand to produce more accurate models.
 
@@
 size after the contigs are broken apart at locations of errors. The
-\textsc{lap} score can be viewed as a log likelihood score, where a
+\textsc{lap} score can be viewed as a log-likelihood score, where a
 value closer to 0 is better. We use a script provided by \textsc{gage}
@@
 to facilitate comparison. For the genome assembly we used software
-that makes use quality values in the assembly process:
+that makes use of quality values in the assembly process:
 \textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
@@
 \textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
-but the short-jump library results are available in the Supplemental
+but the short-jump library results are available in Supplemental
 Table 1. Storing the uncompressed quality values requires 1
-byte/base-call because they are stored in \textsc{ascii} format and is
+byte/base-call because they are stored in \textsc{ascii} format, which is
 denoted by the dotted black asterisk in the figure. The lossless
@@
 in terms of \textsc{mse}. For the \textit{Rhodobacter sphaeroides}
-fragment library, \textsc{q}ual\textsc{c}omp with rate 10 bits/read
+fragment library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
 (0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
@@
 and the \textsc{mse} decreases at an exponential rate. The 7th-degree
-polynomial regression has the highest bits/base-call and in the
+polynomial regression has the highest bits/base-call in the
 \textit{Mus musculus} data set, and requires more storage than the
-\textsc{ascii} original quality values. A 7th-degree polynomial
+original \textsc{ascii} quality values. A 7th-degree polynomial
 requires storing eight floating point values, resulting in 32 bytes
@@
 regression is storing six more bytes than necessary for lossless
-encoding the quality data.
+encoding of the quality data.
 
 As an additional point of reference regarding compression
-effectiveness the recently published
-Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved best
-case compression of 0.254 bits/base-call, and a mean compression of
-1.841 bits/base-call; both values higher than all of the profile
+effectiveness, the recently published
+Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved best-case
+compression of 0.254 bits/base-call, and a mean compression of
+1.841 bits/base-call; both values are higher than all of the profile
 results and for some results of other methods presented here
@@
 
-Preprocessing involves two steps: \emph{trimming} the poor quality
-regions of the sequences and \emph{discarding} sequences that are
-deemed to be poor quality overall. After trimming, the majority of
+Preprocessing involves two steps: \emph{trimming} the poor-quality
+regions of the sequences, and \emph{discarding} sequences that are
+deemed to be poor-quality overall. After trimming, the majority of
 compression methods retain more base-pairs than the uncompressed
 sequences (Figure \ref{fig:preprocessing}). In general, as a given
-compression model increases in complexity, i.e., as the number of
-profiles, polynomial degrees, or rate increases, the amount of
-base-pairs kept approaches the number base-pairs kept using the
+compression model increases in complexity --- i.e., as the number of
+profiles, polynomial degrees, or rate increases --- the amount of
+base-pairs kept approaches the number of base-pairs kept using the
 uncompressed sequences. The compression methods on the \textit{Mus
@@
 discarded, the 5th-degree polynomial regression discards approximately
-two thirds less reads than the 1st-degree polynomial regression that
+two-thirds less reads than the 1st-degree polynomial regression that
 are kept by the uncompressed reads (4,640 and 12,066 reads,
@@
 produce the best assembly in terms of any of the metrics. The
-uncompressed reads scores worse than the top overall assembly
+uncompressed reads score worse than the top overall assembly
 (256-profile encoding) in terms of assembled bases, missing reference
@@
 
-In general, the greater the polynomial degree, the better overall
+In general, the greater the polynomial degree, the better the overall
 assembly; however, the 5th-degree polynomial regression performs
@@
 regression methods perform poorly in all metrics except assembled
-bases. One explanation is that the high error portions of reads are
+bases. One explanation is that the high-error portions of reads are
 being marked as high quality, so \textsc{allpaths-lg} is unable to
-trim or error correct the sequences. Assembled sequences that overlap
-maybe unable to align across the errors at the end of the reads,
+trim or error-correct the sequences. Assembled sequences that overlap
+may be unable to align across the errors at the end of the reads,
 artificially inflating the assembled genome size.
@@
 bits/read rate ranked highest overall, outperforming the other rate
-parameters in terms of corrected N50, least missing reference bases,
+parameters in terms of corrected N50, fewest missing reference bases,
 \textsc{snp}s, and indels $>$5bp. With the exception of the 6
 bits/read rate, the assemblies decrease in rank with the increase in
-the rate parameter in terms of corrected N50, and least missing
-reference bases. This trend runs counter to the the decrease in
+the rate parameter in terms of corrected N50, and fewest missing
+reference bases. This trend runs counter to the decrease in
 \textsc{mse} of the different rates.
@@
 against a reference genome. The original uncompressed and decompressed
-\textsc{fastq} files were mapped with Bowtie2 against
+\textsc{fastq} files were mapped with Bowtie2 against the
 \textit{Rhodobacter sphaeroides} reference genome. The generated
-\textsc{sam} file for each compressing approach were compared with the
+\textsc{sam} file for each compression approach was compared with the
 uncompressed \textsc{sam} file. The total, shared and unique
@@
 approach, we order the compression methods based on how similar the
-alignment results are using the original uncompressed quality values,
+alignment results are using the original uncompressed quality values --- 
 i.e., the amount of reads aligned by both the uncompressed and
@@
 bits/read, followed by \textsc{q}ual\textsc{c}omp with rate 30
-bits/read, 256-, and 128-profile encoding, then 2-bin encoding,
+bits/read, 256-profile encoding, 128-profile encoding, 2-bin encoding,
 64-profile encoding, \textsc{q}ual\textsc{c}omp with rate 10
@@
 Note that the 2-bin method aligns a greater portion of reads than the
-various regression models. If a poor quality base is flanked by high
-quality sequence, then the low degree polynomial regression models
-tend to smooth out the base quality, erroneously marking the low
-quality base as higher quality.  During alignment, Bowtie2 penalizes
-mismatches at high quality bases more so than mismatches at lower
-quality bases. Thus, the polynomial regression models incur a high
-penalty for these bases, resulting in less alignments, despite having
+various regression models. If a poor-quality base is flanked by high-quality sequence, 
+then the low-degree polynomial regression models
+tend to smooth out the base quality, erroneously marking the low-quality
+base as higher quality.  During alignment, Bowtie2 penalizes
+mismatches at high-quality bases more so than mismatches at lower-quality
+bases. Thus, the polynomial regression models incur a high
+penalty for these bases, resulting in fewer alignments, despite having
 a better \textsc{mse} than 2-bin in most cases.
@@
 Setting all bases as minimum quality results in the highest proportion
-of mapped reads 88.1\%. Conversely, setting all bases as maximum
-quality results in the lowest proportion of mapped reads 72.8\%.
+of mapped reads (88.1\%). Conversely, setting all bases as maximum
+quality results in the lowest proportion of mapped reads (72.8\%).
 
@@
 acceptable. As noted above, 2-bin encoding is able to retain
-information about low quality bases that occur within stretches of
-high quality bases, whereas the polynomial regression methods tend to
-smooth these qualities. In addition to aligning less reads that the
+information about low-quality bases that occur within stretches of
+high-quality bases, whereas the polynomial regression methods tend to
+smooth these qualities. In addition to aligning fewer reads than the
 2-bin encoding, the polynomial regression methods may also be
@@
 of genetic variation or sequencing error. Erroneously marking a base
-as higher quality, may lead to the base being incorrectly labeled as a
+as higher quality may lead to the base being incorrectly labeled as a
 \textsc{snp}. 2-bin encoding is acceptable for alignment; however, it
@@
 operations directly on the compressed representations of the data. For
-example, with profile-based compression the $k$ profiles can be
+example, with profile-based compression, the $k$ profiles can be
 evaluated for (pre-)processing operations such as filtering and
@@
 
-In this paper we have examined lossy compression on sequence quality
+In this paper we have examined several general approaches for lossy compression of sequence quality
 values and their effect on subsequent analyses. Although most previous
-examinations on lossy compression focused primarily on information
-loss and compression effectiveness, we show introduce several general
-approaches to quality value compression and show that common
-bioinformatic analyses are robust substantial loss of information in
+examinations of lossy compression focused primarily on information
+loss and compression effectiveness, we show that common
+bioinformatic analyses are robust to substantial loss of information in
 quality values.
@@
 \section{Availability}
-Implementations of the compression methods written in Python and R and
-pipeline to reproduce the results are available at
+Implementations of the compression methods written in Python and R, and
+a pipeline to reproduce the results are available at
 \url{https://github.com/cmhill/q-compression}

### 2015-05-07 manuscript/q-compression.tex
trivial edits

@@
 \title[lossy-compression]{Lossy compression of DNA sequencing quality
-  data} \author[Hill \textit{et~al.}]{Christopher
-  M. Hill\,$^{1}$, Andr\'{a}s Szolek\,$^{2}$, Mohamed El
-  Hadidi\,$^{3}$, and Michael P. Cummings\,$^4$\footnote{to whom correspondence should be
-    addressed}} \address{$^{1}$Department of Computer Science, University of Maryland,
-  College Park, Maryland, 20742 USA\\ $^{2}$Department of Applied
-  Bioinformatics, Center for Bioinformatics, Quantitative Biology
-  Center, and Department of Computer Science, University of
-  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany\\ 
-  $^{3}$Department of Algorithms in
+  data}
+
+\author[Hill \textit{et~al.}]{Christopher M. Hill\,$^{1}$, Andr\'{a}s
+  Szolek\,$^{2}$, Mohamed El Hadidi\,$^{3}$, and Michael
+  P. Cummings\,$^4$\footnote{to whom correspondence should be
+    addressed}} \address{$^{1}$Department of Computer Science,
+  University of Maryland, College Park, Maryland, 20742
+  USA\\ $^{2}$Department of Applied Bioinformatics, Center for
+  Bioinformatics, Quantitative Biology Center, and Department of
+  Computer Science, University of T\"{u}bingen, Sand 14, 72076
+  T\"{u}bingen, Germany\\ $^{3}$Department of Algorithms in
   Bioinformatics, Center for Bioinformatics, University of
-  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany
-  \\ $^{4}$ Center for Bioinformatics and Computational Biology, University of
+  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany \\ $^{4}$ Center
+  for Bioinformatics and Computational Biology, University of
   Maryland, College Park, Maryland, 20742 USA}
+
 \history{Received on XXXXX; revised on XXXXX; accepted on XXXX}
+
 \editor{Associate Editor: XXXXXXX}
@@
 Existing methods for compressing quality sequences mostly focus on
-minimizing the loss of information with less emphasis on the effects on
-subsequent analyses. In this paper, we evaluate several different
+minimizing the loss of information with less emphasis on the effects
+on subsequent analyses. In this paper, we evaluate several different
 compression methods for quality values that compromise accuracy for
@@
 able to produce results similar to those obtained with uncompressed
-quality sequences when performing quality control, genome assembly, and
-alignment of short reads to a reference sequence.
+quality sequences when performing quality control, genome assembly,
+and alignment of short reads to a reference sequence.
 
@@
 inclusion of quality values, and read-level naming and
-metadata. Because of this abundance, effective compression of read data
-has the potential to substantially improve data storage and
+metadata. Because of this abundance, effective compression of read
+data has the potential to substantially improve data storage and
 transfer efficiency.
@@
 files~\citep{Cock:2010ve}, a very common format for sequence read
-data. At the level of the sequence read, the probability of error for each
-base-call is typically represented by a \textsc{phred} quality value,
-which is defined as $Q =
+data. At the level of the sequence read, the probability of error for
+each base-call is typically represented by a \textsc{phred} quality
+value, which is defined as $Q =
 -10\,log_{10}P$~\citep{Ewing:1998ly}. Depending on the sequencing
@@
 prior to subsequent analysis steps. Quality control based on quality
-values generally includes two operations: \textit{i}.~filtering, which removes reads
-that on the whole do not meet arbitrary quality
+values generally includes two operations: \textit{i}.~filtering, which
+removes reads that on the whole do not meet arbitrary quality
 standards, thus reducing the total number of reads; and
@@
 and identified \textsc{snp}s with high-quality calls are deemed more
-reliable than those with low-quality calls, particularly in low-coverage regions.
+reliable than those with low-quality calls, particularly in
+low-coverage regions.
 
@@
 sequencing technologies. Sequence reads generated by instruments such
-as an Illumina HiSeq, the focus of this research due to its ubiquitousness in bioinformatics research, are characterized by
-having relatively few insertion and deletion errors, but substitution
+as an Illumina HiSeq, the focus of this research due to its
+ubiquitousness in bioinformatics research, are characterized by having
+relatively few insertion and deletion errors, but substitution
 (miscall) errors are much more frequent and have context-specific
@@
 is largely pragmatic with a focus on the use of quality values in
-subsequent analyses. From this perspective, some loss of information is
-deemed acceptable if the inferences from analyses are relatively
+subsequent analyses. From this perspective, some loss of information
+is deemed acceptable if the inferences from analyses are relatively
 unaffected. Here we describe our research investigating lossy
@@
 allows for a any distinction among quality values is two; i.e., two
-categories of quality: ``good'' and ``bad''. We implement 2-bin encoding
-by setting a quality value threshold empirically determined by the
-distribution of quality values across reads. Base-calls are marked
-``bad'' if their quality value falls below the first quartile minus
-1.5 $\times$ the interquartile range (IQR), which is the difference
-between the first and third quartile. 1.5 $\times$ IQR is the value
-used by Tukey's box plot~\citep{mcgill1978variations}. The main
-benefit of this approach is that it is completely data-dependent, and
-no assumptions regarding the distribution of the quality values need
-to be made.
+categories of quality: ``good'' and ``bad''. We implement 2-bin
+encoding by setting a quality value threshold empirically determined
+by the distribution of quality values across reads. Base-calls are
+marked ``bad'' if their quality value falls below the first quartile
+minus 1.5 $\times$ the interquartile range (IQR), which is the
+difference between the first and third quartile. 1.5 $\times$ IQR is
+the value used by Tukey's box plot~\citep{mcgill1978variations}. The
+main benefit of this approach is that it is completely data-dependent,
+and no assumptions regarding the distribution of the quality values
+need to be made.
 
@@
 Truncating treats a user-defined number of highest quality values as a
-single bin. LogBinning works similarly to UniBinning, except it uses the
-\emph{log} of the error probability distribution, which effectively
-bins the \textsc{ascii} quality values evenly. Our 2-bin encoding is
-a combination of LogBinning and Truncating in that we place the
-highest quality values (as defined above) of the log of the error
-probability distribution into a single bin.
+single bin. LogBinning works similarly to UniBinning, except it uses
+the \emph{log} of the error probability distribution, which
+effectively bins the \textsc{ascii} quality values evenly. Our 2-bin
+encoding is a combination of LogBinning and Truncating in that we
+place the highest quality values (as defined above) of the log of the
+error probability distribution into a single bin.
 
@@
 
-While we focus only on two bins in this work, a greater number of bins can be used in practice.
-For example, an additional third bin (good, bad, and \emph{okay}) can be used for base qualities near the border between bins in the 2-bin encoding.
-Examining the trade-off of additional bins and their resulting compressibility and effect on downstream analyses is outside the scope of this work.
+While we focus only on two bins in this work, a greater number of bins
+can be used in practice. For example, an additional third bin (good,
+bad, and \emph{okay}) can be used for base qualities near the border
+between bins in the 2-bin encoding. Examining the trade-off of
+additional bins and their resulting compressibility and effect on
+downstream analyses is outside the scope of this work.
 
@@
 modeling.  Although polynomial functions have significantly fewer
-parameters than a read-length string of raw quality values
-(e.g., one to six coefficients in our approach), the use of floating
-point numbers to store coefficients greatly limits the compression
-potential of the method. The coefficients could be represented with
-fewer bits, perhaps with some loss of precision, but we have not
-considered alternative representations in this study.
+parameters than a read-length string of raw quality values (e.g., one
+to six coefficients in our approach), the use of floating point
+numbers to store coefficients greatly limits the compression potential
+of the method. The coefficients could be represented with fewer bits,
+perhaps with some loss of precision, but we have not considered
+alternative representations in this study.
 
@@
 general terms, include no fixed costs, and marginal costs that are a
-function of the number of reads times the cost of representing the model
-parameters.
+function of the number of reads times the cost of representing the
+model parameters.
 
@@
 able to capture some read-specific differences in quality values, it
-does approximate the overall patterns in quality values. An example
-of 128 quality profiles is shown in Figure \ref{fig:profiles_128}.
+does approximate the overall patterns in quality values. An example of
+128 quality profiles is shown in Figure \ref{fig:profiles_128}.
 
@@
 
-Profiles can be obtained from techniques other than $k$-means clustering.
-%While we focus on profiles obtained from $k$-means clustering, other profiling techniques can be used in practice.
-We can store the profiles of polynomial functions to increase the compressibility of polynomial regression modeling.
-%For example, in order to increase the compressibility of polynomial regression modeling, we can store the profiles of polynomial functions.
-In other words, we can use a spline (a function that is piecewise-defined by polynomial functions) to represent a given quality sequence.
-Exploring other potential applications of profile-based compression is outside the scope of this work.
+Profiles can be obtained from techniques other than $k$-means
+clustering. We can store the profiles of polynomial functions to
+increase the compressibility of polynomial regression modeling.  In
+other words, we can use a spline (a function that is piecewise-defined
+by polynomial functions) to represent a given quality sequence.
+However, we have not explored any of these other approaches here.
 
@@
 and short-jump library (insert size of 3,500 nt; 2,050,868 reads). The
-corresponding reference sequence was obtained from the \textsc{ncbi} RefSeq
-database (NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1,
-NC\_007494.1, NC\_009007.1, NC\_009008.1).
+corresponding reference sequence was obtained from the \textsc{ncbi}
+RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1,
+NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
 
@@
 (bits/read), and these bits are allotted among positions within the
-read to minimize the average error. The quality values can be clustered
-beforehand to produce more accurate models.
+read to minimize the average error. The quality values can be
+clustered beforehand to produce more accurate models.
 
@@
 performed using Sickle~\citep{sickle}, and make comparison to
-uncompressed data. Sickle starts at the ends of the read and uses
-a sliding window ($0.1 \times$ the read length) to find locations
-where the mean quality in the window falls below a given threshold (20
-by default). The sequence is then trimmed from this position to the
+uncompressed data. Sickle starts at the ends of the read and uses a
+sliding window ($0.1 \times$ the read length) to find locations where
+the mean quality in the window falls below a given threshold (20 by
+default). The sequence is then trimmed from this position to the
 respective end. If the trimmed sequence length is less than a certain
@@
 \textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
-but the short-jump library results are available in Supplemental
-Table 1. Storing the uncompressed quality values requires 1
-byte/base-call because they are stored in \textsc{ascii} format, which is
-denoted by the dotted black asterisk in the figure. The lossless
-compression of each data set using \textsc{bz}ip2 ranges from
-2.19--3.10 bits/base-call and is denoted by the colored asterisks on
-the abscissa. The compression methods tend to cluster together across
-the different data sets. Across all data sets, the 0-degree polynomial
+but the short-jump library results are available in Supplemental Table
+1. Storing the uncompressed quality values requires 1 byte/base-call
+because they are stored in \textsc{ascii} format, which is denoted by
+the dotted black asterisk in the figure. The lossless compression of
+each data set using \textsc{bz}ip2 ranges from 2.19--3.10
+bits/base-call and is denoted by the colored asterisks on the
+abscissa. The compression methods tend to cluster together across the
+different data sets. Across all data sets, the 0-degree polynomial
 regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
@@
 in terms of \textsc{mse}. For the \textit{Rhodobacter sphaeroides}
-fragment library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
-(0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
-encoding only requires 0.079 bits/bp and has a \textsc{mse} of 11.85.
+fragment library, \textsc{q}ual\textsc{c}omp with rate set to 10
+bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29, whereas
+256-profile encoding only requires 0.079 bits/bp and has a
+\textsc{mse} of 11.85.
 
@@
 effectiveness, the recently published
-Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved best-case
-compression of 0.254 bits/base-call, and a mean compression of
-1.841 bits/base-call; both values are higher than all of the profile
-results and for some results of other methods presented here
+Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
+best-case compression of 0.254 bits/base-call, and a mean compression
+of 1.841 bits/base-call; both values are higher than all of the
+profile results and for some results of other methods presented here
 (Figure~\ref{fig:mse_vs_bpbp}).
@@
 approach, we order the compression methods based on how similar the
-alignment results are using the original uncompressed quality values --- 
-i.e., the amount of reads aligned by both the uncompressed and
+alignment results are using the original uncompressed quality values
+--- i.e., the amount of reads aligned by both the uncompressed and
 compressed methods plus the amount of reads unaligned by both methods
@@
 Note that the 2-bin method aligns a greater portion of reads than the
-various regression models. If a poor-quality base is flanked by high-quality sequence, 
-then the low-degree polynomial regression models
-tend to smooth out the base quality, erroneously marking the low-quality
-base as higher quality.  During alignment, Bowtie2 penalizes
-mismatches at high-quality bases more so than mismatches at lower-quality
-bases. Thus, the polynomial regression models incur a high
-penalty for these bases, resulting in fewer alignments, despite having
-a better \textsc{mse} than 2-bin in most cases.
+various regression models. If a poor-quality base is flanked by
+high-quality sequence, then the low-degree polynomial regression
+models tend to smooth out the base quality, erroneously marking the
+low-quality base as higher quality.  During alignment, Bowtie2
+penalizes mismatches at high-quality bases more so than mismatches at
+lower-quality bases. Thus, the polynomial regression models incur a
+high penalty for these bases, resulting in fewer alignments, despite
+having a better \textsc{mse} than 2-bin in most cases.
 
@@
 
-In this paper we have examined several general approaches for lossy compression of sequence quality
-values and their effect on subsequent analyses. Although most previous
-examinations of lossy compression focused primarily on information
-loss and compression effectiveness, we show that common
-bioinformatic analyses are robust to substantial loss of information in
-quality values.
+In this paper we have examined several general approaches for lossy
+compression of sequence quality values and their effect on subsequent
+analyses. Although most previous examinations of lossy compression
+focused primarily on information loss and compression effectiveness,
+we show that common bioinformatic analyses are robust to substantial
+loss of information in quality values.
 
 \section{Availability}
-Implementations of the compression methods written in Python and R, and
-a pipeline to reproduce the results are available at
+Implementations of the compression methods written in Python and R,
+and a pipeline to reproduce the results are available at
 \url{https://github.com/cmhill/q-compression}

### 2015-05-11 manuscript/q-compression.tex
Lots of polishing of the text. We are very close to a submittable version with this draft.

@@
 
-The \textsc{fastq} file format has become the \emph{de facto} standard
-for storing next-generation sequencing data, containing nucleotide
-information along with a quantitative measure of the reliability of
-individual base calls. As the cost of sequencing continues to
-decrease, the rate of sequence data production is increasing,
-requiring efficient ways of storing and transferring this vast amount
-of data. Most methods of sequencing data compression focus on
-compressing nucleotide information without any loss of information.
-Quality data, however, have different properties than nucleotide data,
-and methods compressing nucleotide sequences efficiently do not
-perform as well on quality sequences. Furthermore, while lossless
-representation is necessary for nucleotide sequences, it is not an
-essential requirement for quality values.
+As the cost of sequencing continues to decrease, the rate of sequence
+data production is increasing, placing greater demands on storing and
+transferring these vast amount of data. Most methods of sequencing
+data compression focus on compressing nucleotide information without
+any loss of information.  Quality data, however, have different
+properties than nucleotide data, and methods compressing nucleotide
+sequences efficiently do not perform as well on quality
+sequences. Furthermore, while lossless representation is necessary for
+nucleotide sequences, it is not an essential requirement for quality
+values.
 
@@
 \textit{ii}.~trimming of low-quality base-calls from reads, which
-reduces the number total number of bases. Quality values can be used
-by genome assembly software to produce better
+reduces the length of reads trimmed. Quality values can be used by
+genome assembly software to produce better
 assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
@@
 biases, and other idiosyncratic properties of distinct high-throughput
-sequencing technologies. Sequence reads generated by instruments such
-as an Illumina HiSeq, the focus of this research due to its
-ubiquitousness in bioinformatics research, are characterized by having
-relatively few insertion and deletion errors, but substitution
-(miscall) errors are much more frequent and have context-specific
-patterns. These errors are non-uniformly distributed over the read
-length (e.g., error rates increase up to $\sim$16$\times$ at the
-3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low-quality tails
-at the 3$^{\prime}$ end~\citep{Minoche:2011km}).
+sequencing technologies.
 
@@
 is deemed acceptable if the inferences from analyses are relatively
-unaffected. Here we describe our research investigating lossy
-compression of sequence read quality values, specifically those
-associated with Illumina instruments, with the objective to provide
-some perspective on several strategies rather than to develop robust
-high-quality software for use. Recognizing these properties of
-Illumina sequence reads motivates our exploration of three general
-classes of lossy compression methods -- binning, modeling, and
-profiling -- and our consideration of an exemplar of each
-class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated
-the effects of lossy compression on identifying variants within a data
-set. We build on these prior works, and assess the effects of quality
+unaffected. Sequence reads generated by instruments such as an
+Illumina HiSeq, the focus of this research due to its preponderance in
+genomics research, are characterized by having relatively few
+insertion and deletion errors, but substitution (miscall) errors are
+much more frequent and have context-specific patterns. These errors
+are non-uniformly distributed over the read length (e.g., error rates
+increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
+67.9\% of reads have low-quality tails at the 3$^{\prime}$
+end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
+sequence reads motivates our exploration of three general classes of
+lossy compression methods -- binning, modeling, and profiling -- and
+our consideration of an exemplar of each class.~\cite{Canovas:2014fr}
+and~\cite{janin2013adaptive} evaluated the effects of lossy
+compression on identifying variants within a data set. Here we
+describe our research investigating lossy compression of sequence read
+quality values, specifically those associated with Illumina
+instruments, with the objective to provide some perspective on several
+strategies rather than to develop robust high-quality software for
+use. We build on these prior works, and assess the effects of quality
 value information loss resulting from compression on subsequent
@@
 the \emph{log} of the error probability distribution, which
-effectively bins the \textsc{ascii} quality values evenly. Our 2-bin
-encoding is a combination of LogBinning and Truncating in that we
-place the highest quality values (as defined above) of the log of the
-error probability distribution into a single bin.
+effectively bins the \textsc{ascii} quality values evenly. The 2-bin
+encoding examined here can be viewed as something like a combination
+of LogBinning and Truncating in that we place the highest quality
+values (as defined above) of the log of the error probability
+distribution into a single bin.
 
@@
 can be used in practice. For example, an additional third bin (good,
-bad, and \emph{okay}) can be used for base qualities near the border
+bad, and intermediate) can be used for base qualities near the border
 between bins in the 2-bin encoding. Examining the trade-off of
 additional bins and their resulting compressibility and effect on
-downstream analyses is outside the scope of this work.
+downstream analyses are not considered here.
 
@@
 
-Sets of quality strings show similar trends of quality over their
-length, and it is possible to identify common patterns in the data and
-use them as reference profiles to approximate individual sequences of
-quality values. Here we use $k$-means clustering, a vector
-quantization method that partitions a set of samples into $k$ sets
-that minimize within-cluster sum of
-squares~\citep{macqueen1967some}. We sampled 10$K$ reads at random and
-computed cluster centers as read quality profiles using a heuristic
-iterative refinement approach that quickly converges to a locally
-optimal minimum~\citep{hartigan1979algorithm}.  All reads are
-evaluated with the nearest quality profile in Euclidean space being
+Sets of strings representing quality values show similar trends of
+over their length, and it is possible to identify common patterns in
+the data and use them as reference profiles to approximate individual
+sequences of quality values. Here we use $k$-means clustering, a
+vector quantization method that partitions a set of samples into $k$
+sets that minimize within-cluster sum of
+squares~\citep{macqueen1967some}. We sampled 1 $\times$ 10$^{4}$ reads
+at random and computed cluster centers as read quality profiles using
+a heuristic iterative refinement approach that quickly converges to a
+locally optimal minimum~\citep{hartigan1979algorithm}. All reads are
+evaluated and the nearest quality profile in Euclidean space being
 assigned to every read as their compressed representation.  The
@@
 set of reads to which they apply. Additionally there are marginal
-costs that are a function of the number of reads encoded.
+costs that are a function of the number of reads encoded times the
+cost of the encoding.  Profiling like modeling, but not binning, has
+the advantageous property that cost improves with read length.
 
 Profiles can be obtained from techniques other than $k$-means
-clustering. We can store the profiles of polynomial functions to
-increase the compressibility of polynomial regression modeling.  In
-other words, we can use a spline (a function that is piecewise-defined
-by polynomial functions) to represent a given quality sequence.
-However, we have not explored any of these other approaches here.
+clustering. For example, we could store the profiles of polynomial
+functions to increase the compressibility of polynomial regression
+modeling.  In other words, we could use a spline (a function that is
+piecewise-defined by polynomial functions) to represent a given
+quality sequence.  However, we have not explored these other
+approaches here.
 
@@
 
-We used several Illumina sequence read data sets in this research,
-which are taken from data from the \textsc{gage} (Genome Assembly
-Gold-Standard Evaluations)~\citep{Salzberg:2012rc}, except as
-noted. These data sets are as follows.
+We used several Illumina sequence read data sets taken from data from
+the \textsc{gage} (Genome Assembly Gold-Standard
+Evaluations)~\citep{Salzberg:2012rc}, except as noted. These data sets
+are as follows.
 
@@
 from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
-which are generated from a fragment library (insert size of 180bp nt;
+which are generated from a fragment library (insert size of 180 nt;
 1,145,8940 paired-end reads). The corresponding reference sequence was
@@
 quality value. \textsc{QualComp} takes as input a user-specified rate
-(bits/read), and these bits are allotted among positions within the
+(bits/read), and these bits are apportioned among positions within the
 read to minimize the average error. The quality values can be
@@
 the mean quality in the window falls below a given threshold (20 by
-default). The sequence is then trimmed from this position to the
-respective end. If the trimmed sequence length is less than a certain
-threshold (20 by default), then the sequence is discarded.
+default). The sequence is then trimmed from this position to the read
+end. If the trimmed sequence length is less than a certain threshold
+(20 by default), then the sequence is discarded.
 
@@
 \textsc{lap} score can be viewed as a log-likelihood score, where a
-value closer to 0 is better. We use a script provided by \textsc{gage}
+larger value is better. We use a script provided by \textsc{gage}
 reference-based evaluation to count single nucleotide polymorphisms
@@
 
-We compare the \textsc{mse} versus bits/base-call of the
+We compare the quality-value \textsc{mse} versus bits/base-call of the
 \textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
-\textit{Escherichia coli}, and \textit{Mus musculus} data sets (Figure
-\ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
-\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
-but the short-jump library results are available in Supplemental Table
-1. Storing the uncompressed quality values requires 1 byte/base-call
-because they are stored in \textsc{ascii} format, which is denoted by
-the dotted black asterisk in the figure. The lossless compression of
-each data set using \textsc{bz}ip2 ranges from 2.19--3.10
-bits/base-call and is denoted by the colored asterisks on the
-abscissa. The compression methods tend to cluster together across the
-different data sets. Across all data sets, the 0-degree polynomial
+\textit{Escherichia coli}, and \textit{Mus musculus} data sets for
+different compressed representations, which include both fixed and
+marginal costs (Figure \ref{fig:mse_vs_bpbp}). Here we present the
+fragment libraries for the \textit{Rhodobacter sphaeroides}, and
+\textit{Homo sapiens} data sets; the short-jump library results are
+available in Supplemental Table 1. Storing the uncompressed quality
+values requires 1 byte/base-call because they are stored in
+\textsc{ascii} format, which is denoted by the dotted black asterisk
+in the figure. The lossless compression of each data set using
+\textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call and is denoted by
+the colored asterisks on the abscissa; these values provide a
+reference point using a widely available compression program. Values
+from the same class of compression method tend to cluster together
+across the different data sets, with the 0-degree polynomial
 regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
@@
 bits/read has the lowest \textsc{mse}, but requires
-$\sim$10--15$\times$ more storage than the profile encoding methods
-for only a $\sim$2--3$\times$ reduction in \textsc{mse}. When rate
+$\sim$10--15$\times$ more memory than the profile encoding methods for
+only a $\sim$2--3$\times$ reduction in \textsc{mse}. When rate
 parameter of \textsc{q}ual\textsc{c}omp is set to match the profile
@@
 
-Preprocessing involves two steps: \emph{trimming} the poor-quality
-regions of the sequences, and \emph{discarding} sequences that are
-deemed to be poor-quality overall. After trimming, the majority of
-compression methods retain more base-pairs than the uncompressed
+Preprocessing involves two steps: \emph{discarding} reads that are
+deemed to be poor-quality overall, and \emph{trimming} the
+poor-quality regions of the reads. After trimming, the majority of
+compression methods retain more base-calls than the uncompressed
 sequences (Figure \ref{fig:preprocessing}). In general, as a given
-compression model increases in complexity --- i.e., as the number of
+compression method increases in complexity --- i.e., as the number of
 profiles, polynomial degrees, or rate increases --- the amount of
-base-pairs kept approaches the number of base-pairs kept using the
-uncompressed sequences. The compression methods on the \textit{Mus
-  musculus} data set have the greatest proportion of retained
-base-pairs compared to the uncompressed sequences. The
-\textit{Escherichia coli} MiSeq data set has the smallest range.
+base-calls retained more closely approximates the number of base-calls
+retained using the uncompressed sequences. The compression methods on
+the \textit{Mus musculus} data set had the greatest proportion of
+retained base-calls compared to the uncompressed sequences. The
+\textit{Escherichia coli} MiSeq data set had the smallest range.
 
 The 2-bin approach is the only compression method that results in a
-higher number of trimmed base-pairs across all data sets. Sickle uses
-a sliding window approach to smooth the read quality values before it
-trims. In the 2-bin approach, there is not an even distribution of
-values per bin. In other words, \emph{bad} quality values may range
-from 0--33, whereas \emph{good} values may only range from
-34--40. Thus, mid-range quality values that are above the threshold
-(20 by default) are set below the quality threshold when compressed,
-resulting in an increased number of trimmed bases.
+higher number of trimmed base-calls compared to the original
+uncompressed reads across all data sets. Sickle uses a sliding window
+approach to smooth the read quality values before it trims. In the
+2-bin approach, there is an uneven distribution of values per bin. In
+other words, \emph{bad} quality values may range from 0--33, whereas
+\emph{good} values may only range from 34--40. Thus, mid-range quality
+values that are above the threshold (20 by default) are set below the
+quality threshold when compressed, resulting in an increased number of
+trimmed bases.
 
 The 0-degree polynomial regression results in the highest proportion
-of bases kept. If the mean quality value of the read is above the
-filtering threshold, then no base-pairs are trimmed. Only reads that
-are comprised of mainly low quality values will be discarded.
+of base-calls kept. If the mean quality value of the read is above the
+filtering threshold, then no base-calls are trimmed. Only reads that
+are comprised of mostly low quality values will be discarded.
 
 It is important to highlight that even though a compression method may
-result in the same number of trimmed bases as the uncompressed
-sequences, it does not mean the \emph{same} bases were trimmed. The
-1st-degree and 5th-degree polynomial regression of the
-\textit{Rhodobacter sphaeroides} fragment library filter nearly as
-many bases as each other. However, if we examine the specific reads
-discarded, the 5th-degree polynomial regression discards approximately
-two-thirds less reads than the 1st-degree polynomial regression that
-are kept by the uncompressed reads (4,640 and 12,066 reads,
-respectively) (Supplemental Table 2).
+result in the same number of trimmed base-calls as the uncompressed
+sequences, it does not mean the \emph{same} base-calls were retained.
+For example, pre-processing of the 1st-degree and 5th-degree
+polynomial regression model compressed reads of the
+\textit{Rhodobacter sphaeroides} fragment library retain approximately
+the same number of base-calls. However, if we examine the specific
+reads discarded, the 5th-degree model discards approximately
+two-thirds less reads than the 1st-degree model (4,640 and 12,066
+reads, respectively) (Supplemental Table 2).
 
@@
 
-No assembly of the \textit{Rhodobacter sphaeroides} data set was best
-in all metrics (Table \ref{fig:assembly_ranks}). Among the compression
-methods, the profile encoding performed best, polynomial regression
-modeling performed worst, and other methods were intermediate
+No compression method resulted in the uniformly best assembly of the
+\textit{Rhodobacter sphaeroides} data set in all metrics (Table
+\ref{fig:assembly_ranks}). Among the compression methods, the profile
+encoding performed best, polynomial regression modeling performed
+worst, and other methods were intermediate
 (Fig.~\ref{fig:assembly_ranks}).
@@
 assembly using the original uncompressed reads has an error rate of
-roughly 8.75 errors per 100 kb of assembled sequence, while the
-256-profile encoding has an error rate of 8.02 errors per 100 kbp
-(Supplemental Table 3).
+$\sim$8.75 errors/100 kb of assembled sequence, while the 256-profile
+encoding has an error rate of $\sim$8.02 errors/100 kbp (Supplemental
+Table 3).
 
@@
 
-Ranking the compression methods by overall alignment rate produces an
-identical ordering as above. Aside from 0-degree polynomial regression
-(83.1\%), all other compression methods have an alignment rate between
-87\% and 86.1\%. The alignment rate of the uncompressed reads is 87\%.
+Ranking the compression methods by overall proportion of reads aligned
+produces an identical ordering as above. Aside from 0-degree
+polynomial regression (0.831), all other compression methods have a
+read alignment rate between 0.87 and 0.861. The proportion of reads
+aligned for the uncompressed reads is 0.87.
 
@@
 Setting all bases as minimum quality results in the highest proportion
-of mapped reads (88.1\%). Conversely, setting all bases as maximum
-quality results in the lowest proportion of mapped reads (72.8\%).
+of mapped reads (0.881). Conversely, setting all bases as maximum
+quality results in the lowest proportion of mapped reads (0.728).
 
@@
 information loss, and some measure of compression effectiveness.
-However, from a pragmatic perspective, it is the effects on the
-results of subsequent analyses that are most important. Our results
-showed that some of the compression methods with relatively high
-information lossy (\textsc{mse}) are still practical for some
+However, from a pragmatic perspective, it is the consequences for
+subsequent analyses that are most important. Our results showed that
+some of the compression methods with relatively high information lossy
+(\textsc{mse}) are practical, and possibly advantageous, for some
 applications. Many bioinformatics tools proved to be robust to
@@
 through quality control software shows that most methods filter nearly
-as many bases as using original quality sequences. Assemblers
+as many bases as using original quality sequences. Assembly programs
 performing sequence alignment use percent similarity scores that are
@@
 performs quite poorly in terms of assembly, and 7th-degree polynomial
-regression modeling outperforms it in nearly all metrics.
+regression modeling outperforms it in nearly all metrics. Profiling
+performs quite well in all use cases.
 
@@
 
-In this paper we have examined several general approaches for lossy
+We have examined several simple and general approaches for lossy
 compression of sequence quality values and their effect on subsequent
@@
 we show that common bioinformatic analyses are robust to substantial
-loss of information in quality values.
+loss of information in quality values. Some methods, such as
+profiling, have yield excellent compression and provide excellent
+results in subsequent analyses, and in some cases surpassing results
+from original uncompressed data.
 
@@
 \section*{Acknowledgement}
-This research was initiated at the 2014 Bioinformatics Exchange for
+This project was initiated at the 2014 Bioinformatics Exchange for
 Students and Teachers (\textsc{best}) Summer School, which was funded
-by the offices of the Dean of The Graduate School at University of
-Maryland and the Rektor of University of T\"{u}bingen.
+by the offices of the Dean of The Graduate School, University of
+Maryland, and the Rektor of the University of T\"{u}bingen. Additional
+funding included an International Graduate Research Fellowship from
+The Graduate School, University of Maryland, to CMH, a Global
+Partnerships-Faculty Travel Grant from the Office of International
+Affairs to MPC, and other funding from the University of T\"{u}bingen.
 

### 2015-05-12 manuscript/q-compression.tex
a few more small changes

@@
 sequences efficiently do not perform as well on quality
-sequences. Furthermore, while lossless representation is necessary for
-nucleotide sequences, it is not an essential requirement for quality
-values.
+sequences. Furthermore, although lossless representation is necessary
+for nucleotide sequences, it is not an essential requirement for
+quality values.
 
@@
 
-While we focus only on two bins in this work, a greater number of bins
-can be used in practice. For example, an additional third bin (good,
-bad, and intermediate) can be used for base qualities near the border
+Although we focus only on two bins in this work, a greater number of
+bins can be used in practice, albeit with some higher cost in memory.
+For example, an additional third bin, thus giving bins of good, bad,
+and intermediate, could be used for base qualities near the border
 between bins in the 2-bin encoding. Examining the trade-off of
@@
 assembly using the original uncompressed reads has an error rate of
-$\sim$8.75 errors/100 kb of assembled sequence, while the 256-profile
+$\sim$8.75 errors/100 kb of assembled sequence, and the 256-profile
 encoding has an error rate of $\sim$8.02 errors/100 kbp (Supplemental

### 2015-05-12 manuscript/q-compression.tex
more tweaks

@@
 which are generated from a fragment library (insert size of 180 nt;
-1,145,8940 paired-end reads). The corresponding reference sequence was
-obtained from the \textsc{ncbi} RefSeq database (NC\_000913.2).
+1,145,8940 paired-end reads; read length 300 nt). The corresponding
+reference sequence was obtained from the \textsc{ncbi} RefSeq database
+(NC\_000913.2).
 
@@
 \url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}, which
-consisted of 18,828,274 reads of length 36.
+consisted of 18,828,274 reads of length 36 nt.
 
@@
 \caption{The relationship of bits/base-call and mean squared error for
-  quality value compression methods applied to four data sets:
-  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
+  quality value compression methods and subsequently further
+  compressed using a general algorithm via \textsc{bz}ip2 to four data
+  sets: \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
   chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
@@
 different compressed representations, which include both fixed and
-marginal costs (Figure \ref{fig:mse_vs_bpbp}). Here we present the
-fragment libraries for the \textit{Rhodobacter sphaeroides}, and
-\textit{Homo sapiens} data sets; the short-jump library results are
-available in Supplemental Table 1. Storing the uncompressed quality
-values requires 1 byte/base-call because they are stored in
-\textsc{ascii} format, which is denoted by the dotted black asterisk
-in the figure. The lossless compression of each data set using
-\textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call and is denoted by
-the colored asterisks on the abscissa; these values provide a
-reference point using a widely available compression program. Values
-from the same class of compression method tend to cluster together
-across the different data sets, with the 0-degree polynomial
-regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
-lowest bits/base-call.
+marginal costs, and subsequently further compressed using a general
+algorithm via \textsc{bz}ip2 (Figure \ref{fig:mse_vs_bpbp}). Here we
+present the fragment libraries for the \textit{Rhodobacter
+  sphaeroides}, and \textit{Homo sapiens} data sets; the short-jump
+library results are available in Supplemental Table 1. Storing the
+uncompressed quality values requires 1 byte/base-call because they are
+stored in \textsc{ascii} format, and the lossless compression of each
+data set using \textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call;
+these values provide a reference point using a widely available
+compression program. Values from the same class of compression method
+tend to cluster together across the different data sets, with the
+0-degree polynomial regression, profile encodings, and
+\textsc{q}ual\textsc{c}omp have the lowest bits/base-call.
 

### 2015-05-12 manuscript/q-compression.tex
updated and corrected read lengths for data sets

@@
 fragment library (insert size of 180 nt; 2,050,868 paired-end reads)
-and short-jump library (insert size of 3,500 nt; 2,050,868 reads). The
-corresponding reference sequence was obtained from the \textsc{ncbi}
-RefSeq database (NC\_007488.1, NC\_007489.1, NC\_007490.1,
-NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
+and short-jump library (insert size of 3,500 nt; 2,050,868 reads; read
+length 101 nt). The corresponding reference sequence was obtained from
+the \textsc{ncbi} RefSeq database (NC\_007488.1, NC\_007489.1,
+NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
 
@@
 and short-jump library (insert sizes ranging from 2283-2803 nt;
-22,669,408 reads). The corresponding reference sequence was obtained
-from the \textsc{ncbi} RefSeq database (NC\_000014.8).
+22,669,408 reads; read length 101 nt). The corresponding reference
+sequence was obtained from the \textsc{ncbi} RefSeq database
+(NC\_000014.8).
 
@@
 which are generated from a fragment library (insert size of 180 nt;
-1,145,8940 paired-end reads; read length 300 nt). The corresponding
+1,145,8940 paired-end reads; read length 151 nt). The corresponding
 reference sequence was obtained from the \textsc{ncbi} RefSeq database
@@
 \textit{Mus musculus} data was downloaded from
-\url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}, which
-consisted of 18,828,274 reads of length 36 nt.
+\url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}
+(18,828,274 reads; length 36 nt).
 

### 2015-05-12 manuscript/plot_qual_regression.R
Added plotting code for regression.

@@ -1,0 +1,54 @@
+# quals = "@@CFFFDDHHGHHJIJJIJIJJJIIJJJJGIJJJFGIGHHHFFFFEEEE@BDDDDDDDDDBDBDEDDDCACCDACCDDDDBDDDDDDDDDDBDC??CDDDDDBBBB@<CDDDDDDCCDACCC>ACCDDDDDCDAAACB8<55@BDD7<?B?"
+# quals = "CCCFFFFFHHHHHJJJEFHGGIJIJGHJJJJJIJJJJJJIIJJJJJJJJJHHHHFFFFEECCCCDDCDDDCDDDBDDDDDDDCCDDCDBDDDDDBBB0<@CDEEDBBB@@0<5?>@C1?AA8<ABCDB?A?CC>:BCDD+4<13>A:3:>>"
+# quals = "@@CDFFFFHGGGHIIGIIICADHBHIIIIIDHGHGIIIIIGIIIIIIIIIGIIIIIIII8BEBDDDFFFDDDB;BDDDCCDCDCDDDCCDCDDD@BB>?CA>C@BCDDBD8?>CCD<<AAB>>9>0<@5902:144>>@9<&)8&&9(>(("
+# quals = "CCCFFFFFHHHGHJJJJHJIJJIJJIJJJIIJJJIGIJJJJJJJJJJJJJJJJJJJIJIJF=;=;;.;?=',(,(53(5(,((,(,5<C(>34(3&))55??B&))&)0(+((((0))00((((+((+&&&)&+(+((+((4+4((((()&"
+# quals = "@@@FFFDFHHHFHGBGIIJJDIIIHJJJJJ@FHIGIJIJGIJJIIHGCHFFFFDD5&)&007((+((((((((+(+><)(29(+4(((+4?(&)))0(((((44(()5&&&&&&((((++4(+(+44:9(&++3((+&)&&&++&&&)(()"
+# quals = "@@CFFFFDHDFFDHGIIIHGJJGHIIJFGGGGEHHIIGIFFEHFFEFDDDD8?:A@CDCDC8ADBCDDDCDDCDDDDBDCDDDCCCD?BB<:BC@C<>9809BBBDBCD@B9<@BDBD((>:99B<&0<9>9&88<B3)09>7)05)(42<"
[…]

### 2015-05-13 manuscript/q-compression.tex
Added reference for Burrows-Wheeler algorithm, which is the algorithm in BZip2. Revised the description of economic costs for all methods, and added cost description for QualComp. Described the advantageous properties of profile-based compression with regard to increases in either or both read length and number of reads. Changed figure legend to reflect the use of arrows. Corrected a spelling error, and made some other small revisions.

@@
 The economic costs of memory use for binning, in general terms,
-include no fixed costs, and marginal costs that are a function of the
+include no fixed costs, and marginal costs that are the product of the
 number of base-call quality values times the cost of the encoding.
@@
 Although we focus only on two bins in this work, a greater number of
-bins can be used in practice, albeit with some higher cost in memory.
-For example, an additional third bin, thus giving bins of good, bad,
-and intermediate, could be used for base qualities near the border
-between bins in the 2-bin encoding. Examining the trade-off of
-additional bins and their resulting compressibility and effect on
-downstream analyses are not considered here.
+bins can be used in practice, albeit with a higher cost in memory
+requirement.  For example, an additional third bin, thus giving bins
+of good, bad, and intermediate, could be used for base qualities near
+the border between bins in the 2-bin encoding. Additional bins beyond
+two, and their resulting compressibility and effect on downstream
+analyses, are not considered here.
 
@@
 The economic costs of memory use for model-based compression, in
-general terms, include no fixed costs, and marginal costs that are a
-function of the number of reads times the cost of representing the
+general terms, include no fixed costs, and marginal costs that are the
+product of the number of reads times the cost of representing the
 model parameters.
@@
 general terms, include fixed costs associated with representing the
-profiles, which is a function of the number of profiles times the cost
-of encoding them, and these fixed costs are amortized over the entire
-set of reads to which they apply. Additionally there are marginal
-costs that are a function of the number of reads encoded times the
-cost of the encoding.  Profiling like modeling, but not binning, has
-the advantageous property that cost improves with read length.
+profiles, which is the product of the number of profiles times the
+cost of encoding them, and these fixed costs are amortized over the
+entire set of reads to which they apply. Additionally there are
+marginal costs that are the product of the number of reads encoded
+times the cost of the encoding. Profiling like modeling, but not
+binning, has the advantageous property that marginal costs are the
+reciprocal of the read length. Thus the costs of memory use for
+profile-based compression decrease proportionally with increases in
+either or both the number of reads and read length.
 
@@
 piecewise-defined by polynomial functions) to represent a given
-quality sequence.  However, we have not explored these other
+quality sequence. However, we have not explored this or other
 approaches here.
@@
 quality values as a multivariate Gaussian distribution, computing the
-mean and covariance for each position in the read, and these varlues
+mean and covariance for each position in the read, and these values
 are stored by the decoder to later reconstruct a \emph{representative}
@@
 
+The economic costs of memory use for compression with
+\textsc{QualComp}, in in our interpretation, appear to have no fixed
+costs, and the marginal costs that are the product of the number of
+reads times the cost of representing the model parameters.
+
 \subsection{Performance evaluation}
 
-As a measure of compression effectiveness we use bits/base-call, and
-define it as the size of the compressed representation of quality
-values (in bits) divided by the number of quality values represented.
-As a measure of information loss we use mean squared error
-(\textsc{mse}) as a loss function, and define it as
+As a measure of compression effectiveness, which reflects the sum of
+fixed and marginal costs after subsequent compression using the
+Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, we use
+bits/base-call, define it as the size of the compressed representation
+of quality values (in bits) divided by the number of quality values
+represented. As a measure of information loss we use mean squared
+error (\textsc{mse}) as a loss function, and define it as
 $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
@@
 \caption{The relationship of bits/base-call and mean squared error for
-  quality value compression methods and subsequently further
-  compressed using a general algorithm via \textsc{bz}ip2 to four data
-  sets: \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
+  quality value compression methods applied to four data sets:
+  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
   chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
@@
   profiles; R$n$, modeling with polynomial regression of degree $n$;
-  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
-  $n$. Asterisks denote the corresponding lossless compression using
-  \textsc{bz}ip2, with the black asterisk corresponding to original
-  data.}
+  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
+  denote the corresponding lossless compression using \textsc{bz}ip2,
+  with the black arrow corresponding to original data.}
 \label{fig:mse_vs_bpbp}
@@
 \textit{Escherichia coli}, and \textit{Mus musculus} data sets for
-different compressed representations, which include both fixed and
-marginal costs, and subsequently further compressed using a general
-algorithm via \textsc{bz}ip2 (Figure \ref{fig:mse_vs_bpbp}). Here we
-present the fragment libraries for the \textit{Rhodobacter
-  sphaeroides}, and \textit{Homo sapiens} data sets; the short-jump
-library results are available in Supplemental Table 1. Storing the
-uncompressed quality values requires 1 byte/base-call because they are
-stored in \textsc{ascii} format, and the lossless compression of each
-data set using \textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call;
-these values provide a reference point using a widely available
-compression program. Values from the same class of compression method
-tend to cluster together across the different data sets, with the
-0-degree polynomial regression, profile encodings, and
-\textsc{q}ual\textsc{c}omp have the lowest bits/base-call.
+different compressed representations (Figure
+\ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
+\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets;
+the short-jump library results are available in Supplemental Table
+1. Storing the uncompressed quality values requires 1 byte/base-call
+because they are stored in \textsc{ascii} format, and the lossless
+compression of each data set using \textsc{bz}ip2 ranges from
+2.19--3.10 bits/base-call; these values provide a reference point
+using a widely available compression program. Values from the same
+class of compression method tend to cluster together across the
+different data sets, with the 0-degree polynomial regression, profile
+encodings, and \textsc{q}ual\textsc{c}omp have the lowest
+bits/base-call.
 

### 2015-05-13 manuscript/q-compression.tex
removed extraneous word

@@
 bins can be used in practice, albeit with a higher cost in memory
-requirement.  For example, an additional third bin, thus giving bins
+requirement.  For example, an additional bin, thus giving bins
 of good, bad, and intermediate, could be used for base qualities near

### 2015-05-13 manuscript/q-compression.tex
Added text edits.

@@
 Quality values can be binned, and the minimum number of bins that
-allows for a any distinction among quality values is two; i.e., two
+allows for any distinction among quality values is two; i.e., two
 categories of quality: ``good'' and ``bad''. We implement 2-bin
@@
 The economic costs of memory use for compression with
-\textsc{QualComp}, in in our interpretation, appear to have no fixed
-costs, and the marginal costs that are the product of the number of
+\textsc{QualComp}, in our interpretation, appear to have optional fixed
+costs for storing the cluster representatives, and the marginal costs that are the product of the number of
 reads times the cost of representing the model parameters.
@@
 Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
-best-case compression of 0.254 bits/base-call, and a mean compression
+best-case compression of 0.254 bits/base-call on \textit{Homo sapiens} chromosome 21 data (read lengths ranging from 50-110 nt), and a mean compression
 of 1.841 bits/base-call; both values are higher than all of the
@@
 retained using the uncompressed sequences. The compression methods on
-the \textit{Mus musculus} data set had the greatest proportion of
+the \textit{Mus musculus} data set has the greatest proportion of
 retained base-calls compared to the uncompressed sequences. The
-\textit{Escherichia coli} MiSeq data set had the smallest range.
+\textit{Escherichia coli} MiSeq data set has the smallest range.
 

### 2015-05-13 src/parse_compression_results_all.R
Changed the figure for compression results to remove the plot border. See if it works.

@@ -89,7 +89,7 @@ gglabcol <-
 
 ### End Hack
 
-p <- ggplot(dfr_tables, aes(x = x, y = y, label=names, col=dataset, shape=pch)) + 
+p <- ggplot(dfr_tables, aes(x = x, y = y, label=names, col=dataset, shape=pch, bty="n")) + 
   scale_shape_identity() +
   scale_y_log10(breaks = y_breaks, labels = y_breaks, limits = c(5,190)) +
   scale_x_log10(breaks = x_breaks, labels = x_breaks, limits = c(.027,11)) +

### 2015-05-13 src/parse_compression_results_all.R
Removed border from plots.

@@ -1,113 +1,135 @@
-library(ggplot2)
-
-tables = c("~/cbcb2/q-compression/results/rhodo3/results/frag_2.fastq.compression",
-           #"~/cbcb2/q-compression/results/staph2/results/frag_1.fastq.compression",
-           "~/cbcb2/q-compression//results/ecoli2/results/MiSeq_Ecoli_MG1655_110721_PF_R1.fastq.compression",
-           "~/cbcb2/q-compression//results/hg14/results/frag_1.fastq.compression",
-           "~/cbcb2/q-compression/results/mouse/results//SRR032209.fastq.compression")
-
-dataset_names = c("Rhodobacter sphaeroides",
-#"S. aureus",
-"Escherichia coli",
-"Homo sapiens",
-"Mus musculus")
-
-x_breaks <- c(.02,.03,.04,.05,.06,.07,.08,.09,.1,.2,.3,.4,.5,.6,.7,.8,.9,1.0,2,3,4,5,6,7,8,9,10)
-names = c("2B", "R0", "R1","R3","R5","R7","P64","P128","P256","Q6","Q10","Q30","Q100")
-pch_comp = c(1,5,5,5,5,5,2,2,2,0,0,0,0)
-pch = c(16,18,18,18,18,18,17,17,17,15,15,15,15)
-y_breaks <- c(2,3,4,5,6,7,8,9,10,20,30,40,50,60,70,80,90,100,200,300)
-
-dfr_tables <- c()
-original_tables <- c()
-counter <- 1
-
-for (table in tables) {
-  organism <- read.table(table, row.names = 1, header = T)
-  remove = c("original", "maxqual", "minqual")
-# remove = c("maxqual", "minqual")
-  original_value <- organism[rownames(organism) %in% "original", ]
-  organism <- organism[!rownames(organism) %in% remove, ]
+  library(ggplot2)
   
-  y <- organism$mse
-  x_bzip2 <- organism$bits_bp
- 
-#   x_bzip2 <- apply(cbind(organism$comp_bzip2, organism$orig_bzip2), 1, min, na.rm=TRUE)
-#   x_bzip2 <- (x_bzip2 * 8) / organism$bases
-#   
-#   x_comp <- organism$comp_size
-#   x_comp <- (x_comp * 8) / organism$bases
-#   
-#   dfr <- data.frame(x = x_comp, y = y, pch = pch_comp, names=names, dataset=dataset_names[[counter]])
[…]

### 2015-05-13 manuscript/q-compression.tex
Numerous small changes to make our use of terminology more consistent.

@@
 sequences efficiently do not perform as well on quality
-sequences. Furthermore, although lossless representation is necessary
-for nucleotide sequences, it is not an essential requirement for
-quality values.
+values. Furthermore, although lossless representation is necessary for
+nucleotide sequences, it is not an essential requirement for quality
+values.
 
-Existing methods for compressing quality sequences mostly focus on
+Existing methods for compressing quality values mostly focus on
 minimizing the loss of information with less emphasis on the effects
@@
 Lossy compression of quality information can greatly decrease storage
-and memory requirements with little discernible effect on subsequent
+and memory requirements with little negative effect on subsequent
 analysis results. The three compression strategies in this study were
 able to produce results similar to those obtained with uncompressed
-quality sequences when performing quality control, genome assembly,
-and alignment of short reads to a reference sequence.
+quality values when performing quality control, genome assembly, and
+alignment of short reads to a reference sequence.
 
@@
 bins can be used in practice, albeit with a higher cost in memory
-requirement.  For example, an additional bin, thus giving bins
-of good, bad, and intermediate, could be used for base qualities near
-the border between bins in the 2-bin encoding. Additional bins beyond
-two, and their resulting compressibility and effect on downstream
-analyses, are not considered here.
+requirement.  For example, an additional bin, thus giving bins of
+good, bad, and intermediate, could be used for base qualities near the
+border between bins in the 2-bin encoding. Additional bins beyond two,
+and their resulting compressibility and effect on downstream analyses,
+are not considered here.
 
@@
 modeling. In other words, we could use a spline (a function that is
-piecewise-defined by polynomial functions) to represent a given
-quality sequence. However, we have not explored this or other
-approaches here.
+piecewise-defined by polynomial functions) to represent quality
+values. However, we have not explored this or other approaches here.
 
@@
 The economic costs of memory use for compression with
-\textsc{QualComp}, in our interpretation, appear to have optional fixed
-costs for storing the cluster representatives, and the marginal costs that are the product of the number of
-reads times the cost of representing the model parameters.
+\textsc{QualComp}, in our interpretation, appear to have fixed costs
+for storing the cluster representatives, and the marginal costs that
+are the product of the number of reads times the cost of representing
+the model parameters.
 
@@
 $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
-sequences, $Q_i'$ is the compressed/decompressed quality value, and
-$Q_i$ is the original quality value associated with sequence position
-$i$.
+quality values, $Q_i'$ is the compressed/decompressed quality value,
+and $Q_i$ is the original quality value associated with position $i$.
 
@@
 requires storing eight floating point values, resulting in 32 bytes
-per sequence of quality values. The read length of the \textit{Mus
-  musculus} data set is only 26, so the 7th-degree polynomial
-regression is storing six more bytes than necessary for lossless
-encoding of the quality data.
+per read. The read length of the \textit{Mus musculus} data set is
+only 26, so the 7th-degree polynomial regression is storing six more
+bytes than necessary for lossless encoding of the quality data.
 
@@
 Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
-best-case compression of 0.254 bits/base-call on \textit{Homo sapiens} chromosome 21 data (read lengths ranging from 50-110 nt), and a mean compression
-of 1.841 bits/base-call; both values are higher than all of the
-profile results and for some results of other methods presented here
-(Figure~\ref{fig:mse_vs_bpbp}).
+best-case compression of 0.254 bits/base-call on \textit{Homo sapiens}
+chromosome 21 data (read lengths ranging from 50-110 nt), and a mean
+compression of 1.841 bits/base-call; both values are higher than all
+of the profile results and for some results of other methods presented
+here (Figure~\ref{fig:mse_vs_bpbp}).
 
@@
 poor-quality regions of the reads. After trimming, the majority of
-compression methods retain more base-calls than the uncompressed
-sequences (Figure \ref{fig:preprocessing}). In general, as a given
+compression methods retain more base-calls than the original quality
+values (Figure \ref{fig:preprocessing}). In general, as a given
 compression method increases in complexity --- i.e., as the number of
@@
 base-calls retained more closely approximates the number of base-calls
-retained using the uncompressed sequences. The compression methods on
+retained using the original quality values. The compression methods on
 the \textit{Mus musculus} data set has the greatest proportion of
-retained base-calls compared to the uncompressed sequences. The
+retained base-calls compared to the original quality values. The
 \textit{Escherichia coli} MiSeq data set has the smallest range.
@@
 result in the same number of trimmed base-calls as the uncompressed
-sequences, it does not mean the \emph{same} base-calls were retained.
-For example, pre-processing of the 1st-degree and 5th-degree
-polynomial regression model compressed reads of the
+quality values, it does not mean the \emph{same} base-calls were
+retained.  For example, pre-processing of the 1st-degree and
+5th-degree polynomial regression model compressed reads of the
 \textit{Rhodobacter sphaeroides} fragment library retain approximately
@@
   and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
-    musculus} data sets. Sequences were trimmed using Sickle. The
-  total amount of bases filtered by each compression method is
-  compared with the amount of bases filtered using the uncompressed
-  sequences.}
+    musculus} data sets. Reads were trimmed using Sickle. The total
+  number of bases filtered by each compression method is compared with
+  the number of bases filtered using the uncompressed quality values.}
   \label{fig:preprocessing}
@@
 Note that the 2-bin method aligns a greater portion of reads than the
-various regression models. If a poor-quality base is flanked by
-high-quality sequence, then the low-degree polynomial regression
-models tend to smooth out the base quality, erroneously marking the
-low-quality base as higher quality.  During alignment, Bowtie2
-penalizes mismatches at high-quality bases more so than mismatches at
-lower-quality bases. Thus, the polynomial regression models incur a
-high penalty for these bases, resulting in fewer alignments, despite
-having a better \textsc{mse} than 2-bin in most cases.
+various regression models. If a poor-quality base-call is flanked by
+high-quality base-calls, then the low-degree polynomial regression
+models tend to smooth out the quality valus, erroneously marking the
+low-quality base-calls as higher quality.  During alignment, Bowtie2
+penalizes mismatches at high-quality base-calls more so than
+mismatches at lower-quality base-calls. Thus, the polynomial
+regression models incur a high penalty for these base-calls, resulting
+in fewer alignments, despite having a better \textsc{mse} than 2-bin
+in most cases.
 
-Setting all bases as minimum quality results in the highest proportion
-of mapped reads (0.881). Conversely, setting all bases as maximum
-quality results in the lowest proportion of mapped reads (0.728).
+Setting all base-calls as minimum quality results in the highest
+proportion of mapped reads (0.881). Conversely, setting all base-calls
+as maximum quality results in the lowest proportion of mapped reads
+(0.728).
 
@@
 through quality control software shows that most methods filter nearly
-as many bases as using original quality sequences. Assembly programs
+as many bases as using original quality values. Assembly programs
 performing sequence alignment use percent similarity scores that are
@@
 acceptable. As noted above, 2-bin encoding is able to retain
-information about low-quality bases that occur within stretches of
-high-quality bases, whereas the polynomial regression methods tend to
-smooth these qualities. In addition to aligning fewer reads than the
-2-bin encoding, the polynomial regression methods may also be
-hindering downstream tools that use the alignment results. A
+information about low-quality base-calls that occur within stretches
+of high-quality base-calls, whereas the polynomial regression methods
+tend to smooth these quality values. In addition to aligning fewer
+reads than the 2-bin encoding, the polynomial regression methods may
+also be hindering downstream tools that use the alignment results. A
 \textsc{snp} detection pipeline will commonly use the quality values
@@
 We have examined several simple and general approaches for lossy
-compression of sequence quality values and their effect on subsequent
+compression of read quality values and their effect on subsequent
 analyses. Although most previous examinations of lossy compression

### 2015-05-13 manuscript/q-compression.tex
Surprisingly large number of manuscript-improving edits.

@@
 compression methods for quality values that compromise accuracy for
-better storage efficiency, and their resulting impact on common
+better storage efficiency, and assess the resulting impacts on common
 bioinformatic analyses using sequence read data.
@@
 analysis results. The three compression strategies in this study were
-able to produce results similar to those obtained with uncompressed
-quality values when performing quality control, genome assembly, and
-alignment of short reads to a reference sequence.
+able to produce results similar to those obtained with the original
+uncompressed quality values when performing quality control, genome
+assembly, and alignment of short reads to a reference sequence.
 
@@
 on lossless compression of base calls~\cite[reviewed
-  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,
+  in][]{Deorowicz:2013hq,Giancarlo:2009fk,Giancarlo:2014rw,
   Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
@@
 Illumina HiSeq, the focus of this research due to its preponderance in
-genomics research, are characterized by having relatively few
-insertion and deletion errors, but substitution (miscall) errors are
-much more frequent and have context-specific patterns. These errors
-are non-uniformly distributed over the read length (e.g., error rates
+genomics, are characterized by having relatively few insertion and
+deletion errors, but substitution (miscall) errors are much more
+frequent and have context-specific patterns. These errors are
+non-uniformly distributed over the read length (e.g., error rates
 increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
@@
 strategies rather than to develop robust high-quality software for
-use. We build on these prior works, and assess the effects of quality
-value information loss resulting from compression on subsequent
-genomic analyses, including read preprocessing (filtering and
-trimming), genome assembly, and read mapping.
+use. We assess the effects of quality value information loss resulting
+from compression on subsequent genomic analyses, including read
+preprocessing (filtering and trimming), genome assembly, and read
+mapping.
 
@@
 of LogBinning and Truncating in that we place the highest quality
-values (as defined above) of the log of the error probability
-distribution into a single bin.
+values (as defined above) from the distribution of log error
+probability values into a single bin.
 
@@
 bins can be used in practice, albeit with a higher cost in memory
-requirement.  For example, an additional bin, thus giving bins of
-good, bad, and intermediate, could be used for base qualities near the
-border between bins in the 2-bin encoding. Additional bins beyond two,
-and their resulting compressibility and effect on downstream analyses,
-are not considered here.
+usage. For example, an additional bin, thus giving bins of good, bad,
+and intermediate, could be used for quality values near the border
+between bins in the 2-bin encoding. Additional bins beyond two, and
+their resulting compressibility and effect on downstream analyses, are
+not considered here.
 
@@
 numbers to store coefficients greatly limits the compression potential
-of the method. The coefficients could be represented with fewer bits,
+of the models. The coefficients could be represented with fewer bits,
 perhaps with some loss of precision, but we have not considered
@@
 \textit{Rhodobacter sphaeroides} 2.4.1, which are generated from a
-fragment library (insert size of 180 nt; 2,050,868 paired-end reads)
-and short-jump library (insert size of 3,500 nt; 2,050,868 reads; read
-length 101 nt). The corresponding reference sequence was obtained from
-the \textsc{ncbi} RefSeq database (NC\_007488.1, NC\_007489.1,
-NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).
+fragment library (insert size of 180 bp; 2,050,868 paired-end reads;
+read length 101 nt) and short-jump library (insert size of 3,500 bp;
+2,050,868 reads; read length 101 nt). The corresponding reference
+sequence was obtained from the \textsc{ncbi} RefSeq database
+(NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1, NC\_007494.1,
+NC\_009007.1, NC\_009008.1).
 
 \textit{Homo sapiens} chromosome 14 data, which are generated from a
-fragment library (insert size of 155 nt; 36,504,800 paired-end reads)
-and short-jump library (insert sizes ranging from 2283-2803 nt;
+fragment library (insert size of 155 bp; 36,504,800 paired-end reads)
+and short-jump library (insert sizes ranging from 2283-2803 bp;
 22,669,408 reads; read length 101 nt). The corresponding reference
@@
 from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
-which are generated from a fragment library (insert size of 180 nt;
+which are generated from a fragment library (insert size of 180 bp;
 1,145,8940 paired-end reads; read length 151 nt). The corresponding
@@
 Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, we use
-bits/base-call, define it as the size of the compressed representation
+bits/base-call, defined as the size of the compressed representation
 of quality values (in bits) divided by the number of quality values
@@
 on quality control steps of trimming and read filtering, which were
-performed using Sickle~\citep{sickle}, and make comparison to
+performed using Sickle~\citep{sickle}, and make comparison to original
 uncompressed data. Sickle starts at the ends of the read and uses a
-sliding window ($0.1 \times$ the read length) to find locations where
+sliding window (0.1 $\times$ the read length) to find locations where
 the mean quality in the window falls below a given threshold (20 by
@@
 \textit{Escherichia coli}, and \textit{Mus musculus} data sets for
-different compressed representations (Figure
+quality values resulting from the compression methods examined (Figure
 \ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
 \textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets;
-the short-jump library results are available in Supplemental Table
-1. Storing the uncompressed quality values requires 1 byte/base-call
-because they are stored in \textsc{ascii} format, and the lossless
-compression of each data set using \textsc{bz}ip2 ranges from
-2.19--3.10 bits/base-call; these values provide a reference point
-using a widely available compression program. Values from the same
-class of compression method tend to cluster together across the
-different data sets, with the 0-degree polynomial regression, profile
-encodings, and \textsc{q}ual\textsc{c}omp have the lowest
-bits/base-call.
+the corresponding short-jump library results are available in
+Supplemental Table 1. Storing the original uncompressed quality values
+requires 1 byte/base-call because they are stored in \textsc{ascii}
+format, and the lossless compression of each original data set using
+\textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call; these values
+provide a reference point using a widely available general compression
+program. Values from the same class of compression method tend to
+cluster together across the different data sets, with the 0-degree
+polynomial regression, profile encodings, and
+\textsc{q}ual\textsc{c}omp yielding the lowest bits/base-call values.
 
@@
 $\sim$10--15$\times$ more memory than the profile encoding methods for
-only a $\sim$2--3$\times$ reduction in \textsc{mse}. When rate
-parameter of \textsc{q}ual\textsc{c}omp is set to match the profile
-encoding methods, \textsc{q}ual\textsc{c}omp performs slightly worse
-in terms of \textsc{mse}. For the \textit{Rhodobacter sphaeroides}
-fragment library, \textsc{q}ual\textsc{c}omp with rate set to 10
-bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29, whereas
-256-profile encoding only requires 0.079 bits/bp and has a
-\textsc{mse} of 11.85.
+only a $\sim$2--3$\times$ reduction in error. When the rate parameter
+of \textsc{q}ual\textsc{c}omp is set to match the profile encoding
+methods, \textsc{q}ual\textsc{c}omp performs slightly worse in terms
+of \textsc{mse}. For the \textit{Rhodobacter sphaeroides} fragment
+library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
+(0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
+encoding only requires 0.079 bits/bp and has a \textsc{mse} of 11.85.
 
@@
 and the \textsc{mse} decreases at an exponential rate. The 7th-degree
-polynomial regression has the highest bits/base-call in the
-\textit{Mus musculus} data set, and requires more storage than the
-original \textsc{ascii} quality values. A 7th-degree polynomial
-requires storing eight floating point values, resulting in 32 bytes
-per read. The read length of the \textit{Mus musculus} data set is
-only 26, so the 7th-degree polynomial regression is storing six more
-bytes than necessary for lossless encoding of the quality data.
+polynomial regression results in the largest bits/base-call, as it
+requires 32 bytes per read before subsequent compression with
+\textsc{bz}ip2. For the \textit{Mus musculus} data set, which has read
+lengths of only 36 nt, the memory required for storage, even after
+subsequent compression with \textsc{bz}ip2 exceeds that for the
+original \textsc{ascii} quality values.
 
@@
 best-case compression of 0.254 bits/base-call on \textit{Homo sapiens}
-chromosome 21 data (read lengths ranging from 50-110 nt), and a mean
-compression of 1.841 bits/base-call; both values are higher than all
+chromosome 21 data (read lengths ranging from 50--110 nt), and a mean
+compression of 1.841 bits/base-call; both values are larger than all
 of the profile results and for some results of other methods presented
@@
 poor-quality regions of the reads. After trimming, the majority of
-compression methods retain more base-calls than the original quality
-values (Figure \ref{fig:preprocessing}). In general, as a given
-compression method increases in complexity --- i.e., as the number of
-profiles, polynomial degrees, or rate increases --- the amount of
-base-calls retained more closely approximates the number of base-calls
-retained using the original quality values. The compression methods on
-the \textit{Mus musculus} data set has the greatest proportion of
-retained base-calls compared to the original quality values. The
-\textit{Escherichia coli} MiSeq data set has the smallest range.
+compression methods resulted in retaining more base-calls than the
+original quality values (Figure \ref{fig:preprocessing}). In general,
+as a given compression method increases in complexity --- i.e., as the
+number of profiles, polynomial degrees, or rate increases --- the
+amount of base-calls retained more closely approximates the number of
+base-calls retained using the original quality values. The compression
+methods on the \textit{Mus musculus} data set has the greatest
+proportion of retained base-calls compared to the original quality
+values. The \textit{Escherichia coli} MiSeq data set has the smallest
+range.
 
 The 2-bin approach is the only compression method that results in a
-higher number of trimmed base-calls compared to the original
+larger number of trimmed base-calls compared to the original
 uncompressed reads across all data sets. Sickle uses a sliding window
@@
 two-thirds less reads than the 1st-degree model (4,640 and 12,066
-reads, respectively) (Supplemental Table 2).
+reads, respectively; Supplemental Table 2), which means there are
+differences in the base-calls trimmed.
 
@@
 No compression method resulted in the uniformly best assembly of the
-\textit{Rhodobacter sphaeroides} data set in all metrics (Table
-\ref{fig:assembly_ranks}). Among the compression methods, the profile
-encoding performed best, polynomial regression modeling performed
-worst, and other methods were intermediate
-(Fig.~\ref{fig:assembly_ranks}).
+\textit{Rhodobacter sphaeroides} data set in all metrics, although
+several methods resulted in better assemblies than the original
+uncompressed data (Table \ref{fig:assembly_ranks}). Among the
+compression methods, the profile encoding performed best, polynomial
+regression modeling performed worst, and other methods were
+intermediate (Fig.~\ref{fig:assembly_ranks}).
 
 The lossy compression methods largely preserve the contiguity found in
-the assembly produced using the reads with unmodified quality
+the assembly produced using the reads with the original quality
 values. All compression methods other than 0-degree polynomial
@@
 Supplemental Table 3). Despite the similar contiguity statistics, the
-different compression methods vary noticeably in the amount of
+different compression methods vary markedly in the number of
 \textsc{snp}s. The 2-bin and profile methods exhibited the fewest
@@
 
-It is important to highlight that using uncompressed reads does not
-produce the best assembly in terms of any of the metrics. The
-uncompressed reads score worse than the top overall assembly
-(256-profile encoding) in terms of assembled bases, missing reference
-bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
-assembly using the original uncompressed reads has an error rate of
-$\sim$8.75 errors/100 kb of assembled sequence, and the 256-profile
-encoding has an error rate of $\sim$8.02 errors/100 kbp (Supplemental
-Table 3).
+It is important to highlight that using the original uncompressed
+quality values does not produce the best assembly in terms of any of
+the metrics. The assembly based on the original uncompressed quality
+values score worse than the top overall assembly (256-profile
+encoding) for number of assembled bases, missing reference bases, N50,
+\textsc{snp}s, indels $>$5bp, and relocations. The assembly using the
+original uncompressed quality values has an error rate of $\sim$8.75
+errors/100 kb of assembled sequence, and the 256-profile encoding has
+an error rate of $\sim$8.02 errors/100 kbp (Supplemental Table 3).
 
@@
 overall ranking of the 5th-degree polynomial slightly below that for
-3rd-degree polynomial model. The 1st- and 0-degree polynomial
+the 3rd-degree polynomial model. The 1st- and 0-degree polynomial
 regression methods perform poorly in all metrics except assembled
-bases. One explanation is that the high-error portions of reads are
-being marked as high quality, so \textsc{allpaths-lg} is unable to
-trim or error-correct the sequences. Assembled sequences that overlap
-may be unable to align across the errors at the end of the reads,
-artificially inflating the assembled genome size.
+bases. One explanation for this observation is that the high-error
+portions of reads are being marked as high quality, so
+\textsc{allpaths-lg} is unable to trim or error-correct the
+reads. Assembled sequences that overlap may be unable to align across
+the errors at the end of the reads, artificially inflating the
+assembly size.
 
@@
 bits/read rate, the assemblies decrease in rank with the increase in
-the rate parameter in terms of corrected N50, and fewest missing
-reference bases. This trend runs counter to the decrease in
-\textsc{mse} of the different rates.
+the rate parameter for corrected N50, and fewest missing reference
+bases. This trend runs counter to the decrease in \textsc{mse} of the
+different rates.
 
@@
 
-Certain short read alignment tools use the quality value information
-when finding potential alignments. Bowtie2 (version 2.2.3) was used to
-evaluate the different decompressed \textsc{fastq} files. Bowtie2 uses
-quality values written in the \textsc{fastq} files when mapping reads
-against a reference genome. The original uncompressed and decompressed
-\textsc{fastq} files were mapped with Bowtie2 against the
-\textit{Rhodobacter sphaeroides} reference genome. The generated
-\textsc{sam} file for each compression approach was compared with the
-uncompressed \textsc{sam} file. The total, shared and unique
-proportional numbers of mapped reads are calculated with respect to
-the uncompressed \textsc{sam} matches as shown in Table
-\ref{tab:aligner}. Additionally, to monitor the effect of quality
-values on mapping in general, Bowtie2 was adjusted so that the maximum
-and minimum mismatch penalty were equivalent to maximum and minimum
-quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).
+Some short read alignment tools, such as Bowtie2 (version 2.2.3),
+which was used here, utilize quality value information when evaluating
+potential alignments. The reads from with original uncompressed and
+compressed/decompressed quality vlaues were mapped with Bowtie2 to the
+\textit{Rhodobacter sphaeroides} reference genome. The total, shared,
+and unique proportion of mapped reads are calculated with respect to
+the results for the the original uncompressed quality values as shown
+in Table \ref{tab:aligner}. Additionally, to assess the effect of
+quality values on mapping in general, Bowtie2 was adjusted so that the
+maximum and minimum mismatch penalty were equivalent to maximum and
+minimum quality scores (with parameters: --mp 6,6 and --mp 2,2
+respectively).
 
@@
 alignment results are using the original uncompressed quality values
---- i.e., the amount of reads aligned by both the uncompressed and
-compressed methods plus the amount of reads unaligned by both methods
-minus the amount of reads uniquely aligned by the uncompressed and
-compression method. In the second approach, we order the compression
-methods by total proportion of aligned reads.
+--- i.e., the number of reads aligned using the original uncompressed
+quality values and quality values resulting from compression, plus the
+number of reads unaligned using the original uncompressed quality
+values and quality values resulting from compression minus the amount
+of reads uniquely aligned using the original uncompressed quality
+values and quality values resulting from compression. In the second
+approach, we order the compression methods by total proportion of
+aligned reads.
 
@@
 Most of the compression methods did not vary greatly in terms of the
-number of reads that were mapped \emph{only} by the compression
-method; however, there is a sizable difference in the amount of reads
-that are originally mapped, but unmapped by the compressed methods.
-\textsc{q}ual\textsc{c}omp with rate 100 bits/read results in the
-fewest missing original read alignments (159). Increasing the
-regression model polynomial degree results in a decreasing amount of
-reads that are originally mapped, but unmapped by the regression model
-(40,931 to 1,134 reads for 0-degree and 7th-degree,
-respectively). There is no such trend for reads that are mapped only
-by the regression model.
+number of reads that were mapped \emph{only} using quality values
+resulting from compression; however, there is a sizable difference in
+the amount of reads that are originally mapped, but unmapped by the
+compressed methods.  \textsc{q}ual\textsc{c}omp with rate 100
+bits/read results in the fewest missing original read alignments
+(159). Increasing the regression model polynomial degree results in a
+decreasing amount of reads that are originally mapped, but unmapped by
+the regression model (40,931 and 1,134 reads for 0-degree and
+7th-degree, respectively). There is no such trend for reads that are
+mapped only by the regression model.
 
@@
 regression modeling outperforms it in nearly all metrics. Profiling
-performs quite well in all use cases.
+performs quite well in all use cases, and has the best performance in
+terms of bits/base-call versus error.
 
@@
 profiling, have yield excellent compression and provide excellent
-results in subsequent analyses, and in some cases surpassing results
-from original uncompressed data.
+results in subsequent analyses, and in some cases surpass performance
+of the original uncompressed data.
 
@@
 Partnerships-Faculty Travel Grant from the Office of International
-Affairs to MPC, and other funding from the University of T\"{u}bingen.
+Affairs to MPC, and funding from the University of T\"{u}bingen.
 

### 2015-05-13 manuscript/q-compression.tex
Added eps

--- 2015-05-13 manuscript/q-compression.tex
@@
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=3.35in]{profiles_128.png}}
+\centerline{\includegraphics[width=3.35in]{profiles_128.eps}}
 \caption{Quality profiles obtained by $k$-means clustering on the
@@
 \begin{figure*}[!tb]
-\centerline{\includegraphics[width=7in]{compression_results.pdf}}
+\centerline{\includegraphics[width=7in]{compression_results.eps}}
 \caption{The relationship of bits/base-call and mean squared error for
@@
 per read. The read length of the \textit{Mus musculus} data set is
-only 26, so the 7th-degree polynomial regression is storing six more
+only 36 nt, so the 7th-degree polynomial regression is storing six more
 bytes than necessary for lossless encoding of the quality data.
@@
 \begin{figure}[!tbp]
-\centerline{\includegraphics[width=3.65in]{preprocessing_results.pdf}}
+\centerline{\includegraphics[width=3.65in]{preprocessing_results.eps}}
 \caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
@@
 \begin{figure}[!tbp]
-\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
+\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.eps}}
 \caption{Rankings of compression methods based on \textit{Rhodobacter

### 2015-05-14 manuscript/q-compression.tex
bits/pb -> bits/base-call

--- 2015-05-13 manuscript/q-compression.tex
@@
 and no assumptions regarding the distribution of the quality values
-need to be made.
+need to be made. We adopt similar ``good''/``bad'' quality value assignments as the preprocessing tool Sickle~\citep{sickle}, which are 40 and 10 respectively.
 
@@
 library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
-(0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
-encoding only requires 0.079 bits/bp and has a \textsc{mse} of 11.85.
+(0.099 bits/base-call) has a \textsc{mse} of 17.29, whereas 256-profile
+encoding only requires 0.079 bits/base-call and has a \textsc{mse} of 11.85.
 

### 2015-05-14 manuscript/q-compression.tex
Added double-precision floating-point.

@@
 parameters than a read-length string of raw quality values (e.g., one
-to six coefficients in our approach), the use of floating point
+to six coefficients in our approach), the use of double-precision floating-point
 numbers to store coefficients greatly limits the compression potential
@@
 library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
-(0.099 bits/base-call) has a \textsc{mse} of 17.29, whereas 256-profile
-encoding only requires 0.079 bits/base-call and has a \textsc{mse} of 11.85.
+(0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
+encoding only requires 0.079 bits/bp and has a \textsc{mse} of 11.85.
 

### 2015-05-14 manuscript/q-compression.tex
Revert "Added double-precision floating-point."
This reverts commit e882a7695b2951ac2fd01f730ac17cdc05be4c80.

--- 2015-05-14 manuscript/q-compression.tex
@@
 sequences efficiently do not perform as well on quality
-values. Furthermore, although lossless representation is necessary for
-nucleotide sequences, it is not an essential requirement for quality
-values.
+values. Furthermore, although lossless representation might be
+necessary for nucleotide sequences, it is not an essential requirement
+for quality values.
 
-Existing methods for compressing quality values mostly focus on
+Previous studies of quality value compression have mostly focus on
 minimizing the loss of information with less emphasis on the effects
-on subsequent analyses. In this paper, we evaluate several different
-compression methods for quality values that compromise accuracy for
-better storage efficiency, and assess the resulting impacts on common
-bioinformatic analyses using sequence read data.
+on bioinformatic analyses. In this paper, we evaluate several
+different compression methods for quality values, and assess the
+resulting impacts on common bioinformatic analyses using sequence read
+data: quality control, genome assembly, and alignment of short reads
+to a reference sequence.
 
 \section{Results:}
-Lossy compression of quality information can greatly decrease storage
-and memory requirements with little negative effect on subsequent
-analysis results. The three compression strategies in this study were
-able to produce results similar to those obtained with the original
-uncompressed quality values when performing quality control, genome
-assembly, and alignment of short reads to a reference sequence.
+Lossy compression of quality information can greatly decrease memory
+requirements with little or no negative, and our results demonstrate
+that some compression methods can result in transformed quality values
+that are quite useful, and in some cases advantageous, compared to
+original uncompressed values.
 
@@
 is largely pragmatic with a focus on the use of quality values in
-subsequent analyses. From this perspective, some loss of information
-is deemed acceptable if the inferences from analyses are relatively
-unaffected. Sequence reads generated by instruments such as an
-Illumina HiSeq, the focus of this research due to its preponderance in
-genomics, are characterized by having relatively few insertion and
+bioinformatic analyses. From this perspective, some loss of
+information is deemed acceptable if the inferences from analyses are
+relatively unaffected. Sequence reads generated by instruments such as
+an Illumina HiSeq, the focus of this research due to its preponderance
+in genomics, are characterized by having relatively few insertion and
 deletion errors, but substitution (miscall) errors are much more
@@
 use. We assess the effects of quality value information loss resulting
-from compression on subsequent genomic analyses, including read
-preprocessing (filtering and trimming), genome assembly, and read
+from compression on \textsc{dna} sequence data analyses, including
+read preprocessing (filtering and trimming), genome assembly, and read
 mapping.
@@
 minus 1.5 $\times$ the interquartile range (IQR), which is the
-difference between the first and third quartile. 1.5 $\times$ IQR is
+difference between the first and third quartile; 1.5 $\times$ IQR is
 the value used by Tukey's box plot~\citep{mcgill1978variations}. The
@@
 and no assumptions regarding the distribution of the quality values
-need to be made. We adopt similar ``good''/``bad'' quality value assignments as the preprocessing tool Sickle~\citep{sickle}, which are 40 and 10 respectively.
+need to be made. We adopt the quality value assignments of the read
+preprocessing tool Sickle~\citep{sickle}, which correspond to the 40
+for ``good'' and 10 for ``bad''.
 
@@
 8 bits used to store quality values in \textsc{ascii}. An additional
-benefit of 2-bin encoding is the potential for increased adjacency of
+possible benefit of 2-bin encoding is for increased adjacency of
 identical values and repeating patterns, properties that may increase
@@
 algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
-  DBLP:journals/tit/ZivL78}.
+  DBLP:journals/tit/ZivL78}, although this potential benefit is not
+evaluated in this study.
 
@@
   of quality profiles is their drop-off position and average overall
-  quality. One can also see sporadic low-position values in a handful
-  of profiles, likely capturing intermittent problems in the
-  sequencing process affecting thousands of reads at a
-  time.}\label{fig:profiles_128}
+  quality. Occaional low values occur early in some profiles, likely
+  relecting intermittent problems in the sequencing process affecting
+  many reads at a time.}\label{fig:profiles_128}
 \end{figure}
@@
 library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
-(0.099 bits/base-call) has a \textsc{mse} of 17.29, whereas 256-profile
-encoding only requires 0.079 bits/base-call and has a \textsc{mse} of 11.85.
+(0.099 bits/base-call) has a \textsc{mse} of 17.29, whereas
+256-profile encoding only requires 0.079 bits/base-call and has a
+\textsc{mse} of 11.85.
 
@@
 
-\subsection{Lossy compression is acceptable for some subsequent  
-analyses}
+We have examined several simple and general approaches for lossy
+compression of read quality values, and their effect on bioinformatic
+analyses. Our results demonstrate that some compression methods can
+result in transformed quality values that are quite useful, and in
+some cases advantageous, compared to original uncompressed values.
 
-It is common to evaluate lossy compression methods in terms of
-information loss, and some measure of compression effectiveness.
-However, from a pragmatic perspective, it is the consequences for
-subsequent analyses that are most important. Our results showed that
-some of the compression methods with relatively high information lossy
-(\textsc{mse}) are practical, and possibly advantageous, for some
-applications. Many bioinformatics tools proved to be robust to
-moderate information loss. Passing the decompressed quality values
-through quality control software shows that most methods filter nearly
-as many bases as using original quality values. Assembly programs
-performing sequence alignment use percent similarity scores that are
-typically robust to standard sequencing errors.
+Downstream applications dictate the relative performance lossy
+compression methods.  Many bioinformatics tools proved to be robust to
+moderate information loss in base-call quality. The use of compression
+transformed quality values in quality control demonstrate that several
+compression methods result in approximately the same total number
+base-calls passing through the combined filtering and trimming
+processes as when using original quality values, though the absolute
+and relative number of reads filtered and base-calls trimmed may
+differ. Genome assembly appears robust to standard sequencing errors,
+and compression transformed quality values, particularly those
+associated with profiling-based compression, result in assemblies
+better than those based on original uncompressed quality values by
+most measures.
 
-Downstream applications dictate which lossy compression methods are
-acceptable. As noted above, 2-bin encoding is able to retain
-information about low-quality base-calls that occur within stretches
-of high-quality base-calls, whereas the polynomial regression methods
-tend to smooth these quality values. In addition to aligning fewer
-reads than the 2-bin encoding, the polynomial regression methods may
-also be hindering downstream tools that use the alignment results. A
-\textsc{snp} detection pipeline will commonly use the quality values
-to record if a potential difference from the reference is the result
-of genetic variation or sequencing error. Erroneously marking a base
-as higher quality may lead to the base being incorrectly labeled as a
-\textsc{snp}. 2-bin encoding is acceptable for alignment; however, it
-performs quite poorly in terms of assembly, and 7th-degree polynomial
-regression modeling outperforms it in nearly all metrics. Profiling
-performs quite well in all use cases, and has the best performance in
-terms of bits/base-call versus error.
-
-\subsection{Potential for operations on compressed data}
-
-Perhaps one of the greatest potential benefits of compressing quality
-values is the ability to perform quality control and possibly other
-operations directly on the compressed representations of the data. For
-example, with profile-based compression, the $k$ profiles can be
+Among the potential benefits of compressing quality values is the
+ability to perform quality control and possibly other operations
+directly on the compressed representations of the data. For example,
+with profile-based compression, each of the $k$ profiles can be
 evaluated for (pre-)processing operations such as filtering and
@@
 
-\section{Conclusion}
-
-We have examined several simple and general approaches for lossy
-compression of read quality values and their effect on subsequent
-analyses. Although most previous examinations of lossy compression
-focused primarily on information loss and compression effectiveness,
-we show that common bioinformatic analyses are robust to substantial
-loss of information in quality values. Some methods, such as
-profiling, have yield excellent compression and provide excellent
-results in subsequent analyses, and in some cases surpass performance
-of the original uncompressed data.
+Quality values for base-calls in \textsc{dna} sequencing are not
+sacrosanct; they are representations of estimates for measurement
+(detection) error determined by various mechanisms including signal
+processing characteristics and algorithms. The true worth of quality
+values lies in the utility they provide in bioinformatic analyses. In
+some cases, as demonstrated here, this value is enhanced through the
+quality values transformed through compression processes. Thus
+appropriate compression of quality values can both reduce associated
+memory cost and improve analysis results.
 
 \section{Availability}
-Implementations of the compression methods written in Python and R,
-and a pipeline to reproduce the results are available at
-\url{https://github.com/cmhill/q-compression}
+Implementations of the compression methods written in Python and
+R~\citep{R-Core-Team:2014yq}, and a pipeline to reproduce the results
+are available at \url{https://github.com/cmhill/q-compression}
 

### 2015-05-14 manuscript/q-compression.tex
32 bytes => 64 bytes.
6 coefficients => 8 coefficients.

--- 2015-05-14 manuscript/q-compression.tex
@@
 parameters than a read-length string of raw quality values (e.g., one
-to six coefficients in our approach), the use of floating point
+to eight coefficients in our approach), the use of double-precision floating-point
 numbers to store coefficients greatly limits the compression potential
@@
 polynomial regression results in the largest bits/base-call, as it
-requires 32 bytes per read before subsequent compression with
+requires 64 bytes per read before subsequent compression with
 \textsc{bz}ip2. For the \textit{Mus musculus} data set, which has read

### 2015-05-15 manuscript/q-compression.tex
revised a sentence in the discussion
refinement of the discussion section

@@
 parameters than a read-length string of raw quality values (e.g., one
-to eight coefficients in our approach), the use of double-precision floating-point
-numbers to store coefficients greatly limits the compression potential
-of the models. The coefficients could be represented with fewer bits,
-perhaps with some loss of precision, but we have not considered
-alternative representations in this study.
+to eight coefficients in our approach), the use of double-precision
+floating-point numbers to store coefficients greatly limits the
+compression potential of the models. The coefficients could be
+represented with fewer bits, perhaps with some loss of precision, but
+we have not considered alternative representations in this study.
 
@@
 analyses. Our results demonstrate that some compression methods can
-result in transformed quality values that are quite useful, and in
-some cases advantageous, compared to original uncompressed values.
+result in quality values that are quite useful, and in some cases
+advantageous, compared to original uncompressed values.
 
 Downstream applications dictate the relative performance lossy
-compression methods.  Many bioinformatics tools proved to be robust to
-moderate information loss in base-call quality. The use of compression
-transformed quality values in quality control demonstrate that several
-compression methods result in approximately the same total number
-base-calls passing through the combined filtering and trimming
-processes as when using original quality values, though the absolute
-and relative number of reads filtered and base-calls trimmed may
-differ. Genome assembly appears robust to standard sequencing errors,
-and compression transformed quality values, particularly those
-associated with profiling-based compression, result in assemblies
-better than those based on original uncompressed quality values by
-most measures.
+compression methods.  Some bioinformatics tools proved to be robust to
+moderate information loss in base-call quality. The use of quality
+values modified via several compression/decompression processes result
+in approximately the same total number base-calls passing through the
+combined filtering and trimming processes as when using original
+quality values, though the absolute and relative number of reads
+filtered and base-calls trimmed may differ. Genome assembly appears
+robust to standard sequencing errors, and compression transformed
+quality values, particularly those associated with profiling-based
+compression, result in assemblies better than those based on original
+uncompressed quality values by most measures.
 
@@
 values lies in the utility they provide in bioinformatic analyses. In
-some cases, as demonstrated here, this value is enhanced through the
-quality values transformed through compression processes. Thus
-appropriate compression of quality values can both reduce associated
-memory cost and improve analysis results.
+some cases, as demonstrated here, this value is enhanced through
+transformation associated with compression. Thus appropriate
+compression of quality values can both reduce associated memory cost
+and improve analysis results.
 

### 2015-05-15 manuscript/q-compression.tex
Removed a citation and reference that appeared to be a preprint of paper that we also cite.

@@
 addressed compression of quality
-values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,
+values~\cite[e.g.,][]{Canovas:2014fr,Hach:2012ys,
   janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,

### 2015-05-17 manuscript/q-compression.tex
Refined costs of profile method more explicitly, added subsection to explicitly describe our use of BZip2 for secondary compression, and mentioned other approaches not explored (to show we thought of them, to point to work that could be done if someone did implement these efficiently, and to head off potential review criticism that we did not explore these). Also fix one of the references.

@@
 profiles, which is the product of the number of profiles times the
-cost of encoding them, and these fixed costs are amortized over the
-entire set of reads to which they apply. Additionally there are
+cost of encoding them. These fixed costs are amortized over the entire
+set of reads to which they apply, and thus on a read basis vary as a
+reciprocal function of the number of reads. Additionally there are
 marginal costs that are the product of the number of reads encoded
@@
 
+\subsection{Subsequent (secondary) compression}
+
+We secondarily compress all otherwise compressed data sets using the
+Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, and it is
+the results of these subsequent compressions that we report. Although
+other algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
+  DBLP:journals/tit/ZivL78} might be particularly effective for
+subsequent compression for some compressed data representations
+examined, we not explored such possibilities here, and instead
+restrict ourselves to \textsc{bz}ip2 because it is commonly used and
+readily available. Similarly, it may very well be possible to obtain
+increased secondary compression by reordering (e.g., sorting)
+compressed representations (e.g., sets of bin designations, model, or
+profiles), but these operations have also not been explored here.
+
 \subsection{Performance evaluation}
@@
 As a measure of compression effectiveness, which reflects the sum of
-fixed and marginal costs after subsequent compression using the
-Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, we use
+fixed and marginal costs, we use
 bits/base-call, defined as the size of the compressed representation

### 2015-07-17 bioinformatics_manuscript/q-compression.tex
renamed previous manuscript directory, created new manuscript directory for BMC, populated it with the BMC LaTeX template directory and moved compression.bib

--- 2015-05-17 manuscript/q-compression.tex
@@
   Bioinformatics, Center for Bioinformatics, University of
-  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany \\ $^{4}$ Center
+  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany \\ $^{4}$Center
   for Bioinformatics and Computational Biology, University of
@@
 Lossy compression of quality information can greatly decrease memory
-requirements with little or no negative, and our results demonstrate
+requirements, and our results demonstrate
 that some compression methods can result in transformed quality values
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 Quality values for base-calls in \textsc{dna} sequencing are not
@@
 \bibliographystyle{natbib}
-\bibliographystyle{achemnat}
-\bibliographystyle{plainnat}
-\bibliographystyle{abbrv}
-\bibliographystyle{bioinformatics}
+%\bibliographystyle{achemnat}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{bioinformatics}
 %
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
 %

### 2015-08-29 bmc_manuscript/manuscipt.tex
Initial work to move previous manuscript to BMC Bioinformatics template. More work is needed.

--- 2015-07-17 bioinformatics_manuscript/q-compression.tex
@@
-\documentclass{bioinfo}
+%% BioMed_Central_Tex_Template_v1.06
+%%                                      %
+%  bmc_article.tex            ver: 1.06 %
+%                                       %
 
-%\usepackage{doi}
-\usepackage{url}
+%%IMPORTANT: do not delete the first line of this template
+%%It must be present to enable the BMC Submission system to
+%%recognise this template!!
 
-\copyrightyear{2015}
-\pubyear{2015}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                     %%
+%%  LaTeX template for BioMed Central  %%
+%%     journal article submissions     %%
+%%                                     %%
+%%          <8 June 2012>              %%
+%%                                     %%
+%%                                     %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%
+%% For instructions on how to fill out this Tex template           %%
+%% document please refer to Readme.html and the instructions for   %%
+%% authors page on the biomed central website                      %%
+%% http://www.biomedcentral.com/info/authors/                      %%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%% BioMed Central currently use the MikTex distribution of         %%
+%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
+%% http://www.miktex.org                                           %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%% additional documentclass options:
+%  [doublespacing]
+%  [linenumbers]   - put the line numbers on margins
+
+%%% loading packages, author definitions
+
+%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+\documentclass{bmcart}
+
+%%% Load packages
+%\usepackage{amsthm,amsmath}
+%\RequirePackage{natbib}
+%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
+%\RequirePackage{hyperref}
+\usepackage[utf8]{inputenc} %unicode support
+%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                             %%
+%%  If you wish to display your graphics for   %%
+%%  your own use using includegraphic or       %%
+%%  includegraphics, then comment out the      %%
+%%  following two lines of code.               %%
+%%  NB: These line *must* be included when     %%
+%%  submitting to BMC.                         %%
+%%  All figure files must be submitted as      %%
+%%  separate graphics through the BMC          %%
+%%  submission process, not included in the    %%
+%%  submitted article.                         %%
+%%                                             %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\def\includegraphic{}
+\def\includegraphics{}
+
+%%% Put your definitions there:
+\startlocaldefs
+\endlocaldefs
+
+\usepackage{hyperref}
+
+%%% Begin ...
 \begin{document}
-\firstpage{1}
 
-\title[lossy-compression]{Lossy compression of DNA sequencing quality
+%%% Start of article front matter
+\begin{frontmatter}
+
+\begin{fmbox}
+\dochead{Research}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\title{Lossy compression of DNA sequencing quality
   data}
 
-\author[Hill \textit{et~al.}]{Christopher M. Hill\,$^{1}$, Andr\'{a}s
-  Szolek\,$^{2}$, Mohamed El Hadidi\,$^{3}$, and Michael
-  P. Cummings\,$^4$\footnote{to whom correspondence should be
-    addressed}} \address{$^{1}$Department of Computer Science,
-  University of Maryland, College Park, Maryland, 20742
-  USA\\ $^{2}$Department of Applied Bioinformatics, Center for
-  Bioinformatics, Quantitative Biology Center, and Department of
-  Computer Science, University of T\"{u}bingen, Sand 14, 72076
-  T\"{u}bingen, Germany\\ $^{3}$Department of Algorithms in
-  Bioinformatics, Center for Bioinformatics, University of
-  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany \\ $^{4}$Center
-  for Bioinformatics and Computational Biology, University of
-  Maryland, College Park, Maryland, 20742 USA}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors here                   %%
+%%                                          %%
+%% Specify information, if available,       %%
+%% in the form:                             %%
+%%   <key>={<id1>,<id2>}                    %%
+%%   <key>=                                 %%
+%% Comment or delete the keys which are     %%
+%% not used. Repeat \author command as much %%
+%% as required.                             %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXX}
+\author[
+   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
+   noteref={n1},                        % id's of article notes, if any
+   email={jane.e.doe@cambridge.co.uk}   % email address
+]{\inits{CMH}\fnm{Christopher M} \snm{Hill}}
+\author[
+   addressref={aff3,aff5},
+   email={john.RS.Smith@cambridge.co.uk}
+]{\inits{AS}\fnm{Andr\'{a}s} \snm{Szolek}}
+\author[
+   addressref={aff4,aff5},
+   email={john.RS.Smith@cambridge.co.uk}
+]{\inits{MEH}\fnm{Mohamed} \snm{El Hadidi}}
+\author[
+   addressref={aff2},
+   corref={aff2},                       % id of corresponding address, if any
+   email={mike@umiac.umd.edu}
+]{\inits{MPC}\fnm{Michael P} \snm{Cummings}}
 
-\editor{Associate Editor: XXXXXXX}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors' addresses here        %%
+%%                                          %%
+%% Repeat \address commands as much as      %%
+%% required.                                %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\maketitle
+\address[id=aff1]{%                           % unique id
+  \orgname{Department of Computer Science, University of Maryland}, % university, etc
+  \city{College Park, Maryland},                              % city
+  \postcode{20742}                                % post or zip code
+  \cny{USA}                                    % country
+}
+\address[id=aff2]{%                           % unique id
+  \orgname{Center for Bioinformatics and Computational Biology, University of Maryland}, % university, etc
+  \city{College Park, Maryland},                              % city
+  \postcode{20742}                                % post or zip code
+  \cny{USA}                                    % country
+}
+\address[id=aff3]{%
+  \orgname{Department of Applied Bioinformatics, Center for
+  Bioinformatics, Quantitative Biology Center, University of T\"{u}bingen},
+  \street{Sand 14},
+  \postcode{72076}
+  \city{T\"{u}bingen},
+  \cny{Germany}
+}
+\address[id=aff4]{%
+  \orgname{Department of Algorithms in Bioinformatics, Center for
+  Bioinformatics, Quantitative Biology Center, University of T\"{u}bingen},
+  \street{Sand 14},
+  \postcode{72076}
+  \city{T\"{u}bingen},
+  \cny{Germany}
+}
+\address[id=aff5]{%
+  \orgname{Department of Computer Science, University of T\"{u}bingen},
+  \street{Sand 14},
+  \postcode{72076}
+  \city{T\"{u}bingen},
+  \cny{Germany}
+}
 
-\begin{abstract}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter short notes here                   %%
+%%                                          %%
+%% Short notes will be after addresses      %%
+%% on first page.                           %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Motivation:}
+\begin{artnotes}
+%\note{Sample of title note}     % note to the article
+\note[id=n1]{Current address: \ldots} % note, connected to author
+\end{artnotes}
 
-As the cost of sequencing continues to decrease, the rate of sequence
-data production is increasing, placing greater demands on storing and
-transferring these vast amount of data. Most methods of sequencing
-data compression focus on compressing nucleotide information without
-any loss of information. Quality data, however, have different
-properties than nucleotide data, and methods compressing nucleotide
-sequences efficiently do not perform as well on quality
-values. Furthermore, although lossless representation might be
-necessary for nucleotide sequences, it is not an essential requirement
-for quality values.
+\end{fmbox}% comment this for two column layout
 
-Previous studies of quality value compression have mostly focus on
-minimizing the loss of information with less emphasis on the effects
-on bioinformatic analyses. In this paper, we evaluate several
-different compression methods for quality values, and assess the
-resulting impacts on common bioinformatic analyses using sequence read
-data: quality control, genome assembly, and alignment of short reads
-to a reference sequence.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Abstract begins here                 %%
+%%                                          %%
+%% Please refer to the Instructions for     %%
+%% authors on http://www.biomedcentral.com  %%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Results:}
-Lossy compression of quality information can greatly decrease memory
-requirements, and our results demonstrate
-that some compression methods can result in transformed quality values
-that are quite useful, and in some cases advantageous, compared to
-original uncompressed values.
+\begin{abstractbox}
 
-\section{Contact:} \href{mike@umiacs.umd.edu}{mike@umiacs.umd.edu}
+\begin{abstract} % abstract
+\parttitle{First part title} %if any
+Text for this section.
+
+\parttitle{Second part title} %if any
+Text for this section.
 \end{abstract}
 
-\section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The keywords begin here                  %%
+%%                                          %%
+%% Put each keyword in separate \kwd{}.     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\begin{keyword}
+\kwd{sample}
+\kwd{article}
+\kwd{author}
+\end{keyword}
+
+% MSC classifications codes, if any
+%\begin{keyword}[class=AMS]
+%\kwd[Primary ]{}
+%\kwd{}
+%\kwd[; secondary ]{}
+%\end{keyword}
+
+\end{abstractbox}
+%
+%\end{fmbox}% uncomment this for twcolumn layout
+
+\end{frontmatter}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Main Body begins here                %%
+%%                                          %%
+%% Please refer to the instructions for     %%
+%% authors on:                              %%
+%% http://www.biomedcentral.com/info/authors%%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%% See the Results and Discussion section   %%
+%% for details on how to create sub-sections%%
+%%                                          %%
+%% use \cite{...} to cite references        %%
+%%  \cite{koon} and                         %%
+%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
+%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
+% <put your article body there>
+
+%%%%%%%%%%%%%%%%
+%% Background %%
+%%
+\section*{Background}
 Read data from high-throughput sequencing constitutes the largest
@@
 Quality values comprise a standard component of \textsc{fastq}
-files~\citep{Cock:2010ve}, a very common format for sequence read
+files~\cite{Cock:2010ve}, a very common format for sequence read
 data. At the level of the sequence read, the probability of error for
@@
 value, which is defined as $Q =
--10\,log_{10}P$~\citep{Ewing:1998ly}. Depending on the sequencing
+-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
 technology, these quality values can range from 0 to 93, and are
@@
 genome assembly software to produce better
-assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
-alignment programs, such as Bowtie2~\citep{Langmead:2012rw}, use
+assemblies~\cite{Bryant:2009uq,Gnerre:2011kx}. Short-read
+alignment programs, such as Bowtie2~\cite{Langmead:2012rw}, use
 quality values to weight mismatches between read and reference
 sequences. Software for detecting single nucleotide polymorphisms
-(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
+(\textsc{snp}s) can use quality values~\cite{McKenna:2010bh},
 and identified \textsc{snp}s with high-quality calls are deemed more
@@
 Previous literature on sequence data compression has largely focused
-on lossless compression of base calls~\cite[reviewed
-  in][]{Deorowicz:2013hq,Giancarlo:2009fk,Giancarlo:2014rw,
+on lossless compression of base calls~\cite{Deorowicz:2013hq,
+Giancarlo:2009fk,Giancarlo:2014rw,
   Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
 addressed compression of quality
-values~\cite[e.g.,][]{Canovas:2014fr,Hach:2012ys,
+values~\cite{Canovas:2014fr,Hach:2012ys,
   janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
@@
 67.9\% of reads have low-quality tails at the 3$^{\prime}$
-end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
+end~\cite{Minoche:2011km}). Recognizing these properties of Illumina
 sequence reads motivates our exploration of three general classes of
@@
 
-\begin{methods}
-\section{Methods}
+\section*{Methods}
 
-\subsection{Compression strategies}
+\subsection*{Compression strategies}
 
-\subsubsection{Binning}
+\subsubsection*{Binning}
 
@@
 marked ``bad'' if their quality value falls below the first quartile
-minus 1.5 $\times$ the interquartile range (IQR), which is the
-difference between the first and third quartile; 1.5 $\times$ IQR is
-the value used by Tukey's box plot~\citep{mcgill1978variations}. The
+minus 1.5 $\times$ the interquartile range (\textsc{iqr}), which is the
+difference between the first and third quartile; 1.5 $\times$ \textsc{iqr} is
+the value used by Tukey's box plot~\cite{mcgill1978variations}. The
 main benefit of this approach is that it is completely data-dependent,
@@
 need to be made. We adopt the quality value assignments of the read
-preprocessing tool Sickle~\citep{sickle}, which correspond to the 40
+preprocessing tool Sickle~\cite{sickle}, which correspond to the 40
 for ``good'' and 10 for ``bad''.
@@
 effectiveness of subsequent compression using established
-algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
+algorithms~\cite{HUFFMAN:1952nr,Ziv77auniversal,
   DBLP:journals/tit/ZivL78}, although this potential benefit is not
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=3.35in]{profiles_128.eps}}
-\caption{Quality profiles obtained by $k$-means clustering on the
-  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
-  set using $k$ = 128, with each row corresponding to a quality
-  profile. Dark to light colors represent low to high quality
-  values. It is readily visible that the two most distinctive features
-  of quality profiles is their drop-off position and average overall
-  quality. Occaional low values occur early in some profiles, likely
-  relecting intermittent problems in the sequencing process affecting
-  many reads at a time.}\label{fig:profiles_128}
-\end{figure}
-
 Although we focus only on two bins in this work, a greater number of
@@
 
-\subsubsection{Modeling}
+\subsubsection*{Modeling}
 
@@
 
-\subsubsection{Profiling}
+\subsubsection*{Profiling}
 
@@
 sets that minimize within-cluster sum of
-squares~\citep{macqueen1967some}. We sampled 1 $\times$ 10$^{4}$ reads
+squares~\cite{macqueen1967some}. We sampled 1 $\times$ 10$^{4}$ reads
 at random and computed cluster centers as read quality profiles using
 a heuristic iterative refinement approach that quickly converges to a
-locally optimal minimum~\citep{hartigan1979algorithm}. All reads are
+locally optimal minimum~\cite{hartigan1979algorithm}. All reads are
 evaluated and the nearest quality profile in Euclidean space being
@@
 
-\subsection{Data sets}
+\subsection*{Data sets}
 
@@
 the \textsc{gage} (Genome Assembly Gold-Standard
-Evaluations)~\citep{Salzberg:2012rc}, except as noted. These data sets
+Evaluations)~\cite{Salzberg:2012rc}, except as noted. These data sets
 are as follows.
@@
 
-\subsection{Comparison to other methods}
+\subsection*{Comparison to other methods}
 
 For comparison to other developed methods we use \textsc{QualComp}, a
-lossy compression tool~\citep{Ochoa:2013rt}. The program models
+lossy compression tool~\cite{Ochoa:2013rt}. The program models
 quality values as a multivariate Gaussian distribution, computing the
@@
 
-\subsection{Subsequent (secondary) compression}
+\subsection*{Subsequent (secondary) compression}
 
 We secondarily compress all otherwise compressed data sets using the
-Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, and it is
+Burrows-Wheeler algorithm~\cite{bwt} via \textsc{bz}ip2, and it is
 the results of these subsequent compressions that we report. Although
-other algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
+other algorithms~\cite{HUFFMAN:1952nr,Ziv77auniversal,
   DBLP:journals/tit/ZivL78} might be particularly effective for
@@
 
-\subsection{Performance evaluation}
+\subsection*{Performance evaluation}
 
@@
 on quality control steps of trimming and read filtering, which were
-performed using Sickle~\citep{sickle}, and make comparison to original
+performed using Sickle~\cite{sickle}, and make comparison to original
 uncompressed data. Sickle starts at the ends of the read and uses a
@@
 statistics, log average read probability
-(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
+(\textsc{lap})~\cite{Ghodsi:2013hb}, and a collection of
 reference-based metrics. The contiguity statistics include N50, which
@@
 that makes use of quality values in the assembly process:
-\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
+\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
 settings and 32 threads.
 
-\end{methods}
+\section*{Results}
 
-\section{Results}
-
-\subsection{Compression effectiveness versus information loss}
-
-\begin{figure*}[!tb]
-\centerline{\includegraphics[width=7in]{compression_results.eps}}
-\caption{The relationship of bits/base-call and mean squared error for
-  quality value compression methods applied to four data sets:
-  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
-  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
-  \textit{Mus musculus}. Point labels correspond to different
-  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
-  profiles; R$n$, modeling with polynomial regression of degree $n$;
-  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
-  denote the corresponding lossless compression using \textsc{bz}ip2,
-  with the black arrow corresponding to original data.}
-\label{fig:mse_vs_bpbp}
-\end{figure*}
+\subsection*{Compression effectiveness versus information loss}
 
@@
 effectiveness, the recently published
-Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
+Read-Quality-Sparsifier~\cite{DBLP:conf/recomb/YuYB14} achieved
 best-case compression of 0.254 bits/base-call on \textit{Homo sapiens}
@@
 
-\subsection{Effects on sequence read preprocessing}
+\subsection*{Effects on sequence read preprocessing}
 
@@
 
-\begin{figure}[!tbp]
-\centerline{\includegraphics[width=3.65in]{preprocessing_results.eps}}
-\caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
-  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
-  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
-    musculus} data sets. Reads were trimmed using Sickle. The total
-  number of bases filtered by each compression method is compared with
-  the number of bases filtered using the uncompressed quality values.}
-  \label{fig:preprocessing}
-\end{figure}
-
-\subsection{Effects on genome assembly}
+\subsection*{Effects on genome assembly}
 
@@
 
-\begin{figure}[!tbp]
-\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.eps}}
-\caption{Rankings of compression methods based on \textit{Rhodobacter
-    sphaeroides} assembly attributes sorted by overall
-  rank. Assemblies were constructed using \textsc{allpaths-lg}.
-  Rankings above the median value are in cyan, those below the median
-  value in magenta.}
-  \label{fig:assembly_ranks}
-\end{figure}
-
-\begin{table*}[!tbhp]
-\centering
-\caption[]{Comparison of mapping for original reads and reads with
-  compressed/decompressed quality values. Reads and reference genome
-  are for \textit{Rhodobacter sphaeroides}, and mapping was performed
-  using Bowtie2.}
-\begin{small}
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{max-qual} & \multicolumn{2}{c|}{min-qual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{regression (0)} & \multicolumn{2}{c}{regression (1)} \\
-& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
-\cline{2-12}
-& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
-{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
-\cline{2-12}
-& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
-\end{tabular}
-
-\bigskip
-
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
- & & \multicolumn{2}{c|}{regression (3)} & \multicolumn{2}{c|}{regression (5)} & \multicolumn{2}{c|}{regression (7)} & \multicolumn{2}{c|}{profile (64)} & \multicolumn{2}{c}{profile (128)} \\
-& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
-\cline{2-12}
-& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
-{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
-\cline{2-12}
-& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
-\end{tabular}
-
-\bigskip
-
-\begin{tabular}{lr|cc|cc|cc|cc|cc}
-&  & \multicolumn{2}{c|}{profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
-& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
-\cline{2-12}
-& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
-{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
-\cline{2-12}
-& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
-\end{tabular}
-\end{small}
-
-\label{tab:aligner}
-\end{table*}
-
-\subsection{Effects on read mapping}
+\subsection*{Effects on read mapping}
 
@@
 
-\section{Discussion}
+\section*{Conclusions}
 
@@
 
-\section{Availability}
-Implementations of the compression methods written in Python and
-R~\citep{R-Core-Team:2014yq}, and a pipeline to reproduce the results
-are available at \url{https://github.com/cmhill/q-compression}
+\section*{Section title}
+Text for this section \ldots
+\subsection*{Sub-heading for section}
+Text for this sub-heading \ldots
+\subsubsection*{Sub-sub heading for section}
+Text for this sub-sub-heading \ldots
+\paragraph*{Sub-sub-sub heading for section}
+Text for this sub-sub-sub-heading \ldots
+In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
+addition, we examine a common modeling assumption and note the
+importance of considering the tails of the extinction time $T_x$ in
+studies of escape dynamics.
+We will first consider the expected resistant population at $vT_x$ for
+some $v>0$, (and temporarily assume $\alpha=0$)
+%
+\[
+ E \bigl[Z_1(vT_x) \bigr]= E
+\biggl[\mu T_x\int_0^{v\wedge
+1}Z_0(uT_x)
+\exp \bigl(\lambda_1T_x(v-u) \bigr)\,du \biggr].
+\]
+%
+If we assume that sensitive cells follow a deterministic decay
+$Z_0(t)=xe^{\lambda_0 t}$ and approximate their extinction time as
+$T_x\approx-\frac{1}{\lambda_0}\log x$, then we can heuristically
+estimate the expected value as
+%
+\begin{eqnarray}\label{eqexpmuts}
+E\bigl[Z_1(vT_x)\bigr] &=& \frac{\mu}{r}\log x
+\int_0^{v\wedge1}x^{1-u}x^{({\lambda_1}/{r})(v-u)}\,du
+\nonumber\\
+&=& \frac{\mu}{r}x^{1-{\lambda_1}/{\lambda_0}v}\log x\int_0^{v\wedge
+1}x^{-u(1+{\lambda_1}/{r})}\,du
+\nonumber\\
+&=& \frac{\mu}{\lambda_1-\lambda_0}x^{1+{\lambda_1}/{r}v} \biggl(1-\exp \biggl[-(v\wedge1) \biggl(1+
+\frac{\lambda_1}{r}\biggr)\log x \biggr] \biggr).
+\end{eqnarray}
+%
+Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).
+%\nocite{oreg,schn,pond,smith,marg,hunn,advi,koha,mouse}
 
-\section*{Acknowledgement}
-This project was initiated at the 2014 Bioinformatics Exchange for
-Students and Teachers (\textsc{best}) Summer School, which was funded
-by the offices of the Dean of The Graduate School, University of
-Maryland, and the Rektor of the University of T\"{u}bingen. Additional
-funding included an International Graduate Research Fellowship from
-The Graduate School, University of Maryland, to CMH, a Global
-Partnerships-Faculty Travel Grant from the Office of International
-Affairs to MPC, and funding from the University of T\"{u}bingen.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Backmatter begins here                   %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-%
-%\bibliographystyle{plain}
-%
+\begin{backmatter}
 
-\bibliography{compression}
+\section*{Competing interests}
+  The authors declare that they have no competing interests.
 
+\section*{Author's contributions}
+    Text for this section \ldots
+
+\section*{Acknowledgements}
+  Text for this section \ldots
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                  The Bibliography                       %%
+%%                                                         %%
+%%  Bmc_mathpys.bst  will be used to                       %%
+%%  create a .BBL file for submission.                     %%
+%%  After submission of the .TEX file,                     %%
+%%  you will be prompted to submit your .BBL file.         %%
+%%                                                         %%
+%%                                                         %%
+%%  Note that the displayed Bibliography will not          %%
+%%  necessarily be rendered by Latex exactly as specified  %%
+%%  in the online Instructions for Authors.                %%
+%%                                                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% if your bibliography is in bibtex format, use those commands:
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{compression}      % Bibliography file (usually '*.bib' )
+% for author-year bibliography (bmc-mathphys or spbasic)
+% a) write to bib file (bmc-mathphys only)
+% @settings{label, options="nameyear"}
+% b) uncomment next line
+%\nocite{label}
+
+% or include bibliography directly:
+% \begin{thebibliography}
+% \bibitem{b1}
+% \end{thebibliography}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Figures                       %%
+%%                               %%
+%% NB: this is for captions and  %%
+%% Titles. All graphics must be  %%
+%% submitted separately and NOT  %%
+%% included in the Tex document  %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%
+%% Do not use \listoffigures as most will included as separate files
+
+\section*{Figures}
+\begin{figure}[h!]
+\centerline{\includegraphics[width=3.35in]{profiles128.eps}}
+\caption{\csentence{Quality profiles obtained by $k$-means clustering.} 
+  Example for the
+  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
+  set using $k$ = 128. Each row corresponding to a quality
+  profile. Dark to light colors represent low to high quality
+  values. It is readily visible that the two most distinctive features
+  of quality profiles is their drop-off position and average overall
+  quality. Occasional low values occur early in some profiles, likely
+  reflecting intermittent problems in the sequencing process affecting
+  many reads at a time.}
+  \label{fig:profiles_128}
+\end{figure}
+
+\begin{figure*}[h!]
+\centerline{\includegraphics[width=7in]{compression.results.eps}}
+\caption{\csentence{The relationship of bits/base-call and mean squared error for
+  quality value compression methods applied to four data sets.}
+  The data sets are \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
+  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
+  \textit{Mus musculus}. Point labels correspond to different
+  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
+  profiles; R$n$, modeling with polynomial regression of degree $n$;
+  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
+  denote the corresponding lossless compression using \textsc{bz}ip2,
+  with the black arrow corresponding to original data.}
+\label{fig:mse_vs_bpbp}
+\end{figure*}
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=3.65in]{preprocessing.results.eps}}
+\caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
+  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
+  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
+    musculus} data sets. Reads were trimmed using Sickle. The total
+  number of bases filtered by each compression method is compared with
+  the number of bases filtered using the uncompressed quality values.}
+  \label{fig:preprocessing}
+\end{figure}
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=3.65in]{rhodo.assembly.results.eps}}
+\caption{\csentence{Rankings of compression methods based on assembly attributes}
+ The read data are for \textit{Rhodobacter
+    sphaeroides} 2.4.1, and the results sorted by overall
+  rank. Assemblies were constructed using \textsc{allpaths-lg}.
+  Rankings above the median value are in cyan, those below the median
+  value in magenta.}
+  \label{fig:assembly_ranks}
+\end{figure}
+
+  \begin{figure}[h!]
+  \caption{\csentence{Sample figure title.}
+      A short description of the figure content
+      should go here.}
+      \end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{Sample figure title.}
+      Figure legend text.}
+      \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Tables                        %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Use of \listoftables is discouraged.
+%%
+\section*{Tables}
+
+\begin{table*}[!tbhp]
+\centering
+\caption[]{Comparison of mapping for original reads and reads with
+  compressed/decompressed quality values. Reads and reference genome
+  are for \textit{Rhodobacter sphaeroides}, and mapping was performed
+  using Bowtie2.}
+\begin{small}
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{max-qual} & \multicolumn{2}{c|}{min-qual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{regression (0)} & \multicolumn{2}{c}{regression (1)} \\
+& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
+\cline{2-12}
+& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
+{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
+\cline{2-12}
+& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{regression (3)} & \multicolumn{2}{c|}{regression (5)} & \multicolumn{2}{c|}{regression (7)} & \multicolumn{2}{c|}{profile (64)} & \multicolumn{2}{c}{profile (128)} \\
+& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
+\cline{2-12}
+& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
+{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
+\cline{2-12}
+& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+&  & \multicolumn{2}{c|}{profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
+& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
+\cline{2-12}
+& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
+{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
+\cline{2-12}
+& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
+\end{tabular}
+\end{small}
+
+\label{tab:aligner}
+\end{table*}
+
+\begin{table}[h!]
+\caption{Sample table title. This is where the description of the table should go.}
+      \begin{tabular}{cccc}
+        \hline
+           & B1  &B2   & B3\\ \hline
+        A1 & 0.1 & 0.2 & 0.3\\
+        A2 & ... & ..  & .\\
+        A3 & ..  & .   & .\\ \hline
+      \end{tabular}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Additional Files              %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Additional Files}
+  \subsection*{Additional file 1 --- Sample additional file title}
+    Additional file descriptions text (including details of how to
+    view the file, if it is in a non-standard format or the file extension).  This might
+    refer to a multi-page table or a figure.
+
+  \subsection*{Additional file 2 --- Sample additional file title}
+    Additional file descriptions text.
+
+\end{backmatter}
 \end{document}

### 2015-08-30 bmc_manuscript/manuscipt.tex
Completed movement of the Bioinformatics manuscript material to BMC Bioinformatics template. Also addressed all of the minor comments of reviewer 2.

@@
    addressref={aff3,aff5},
-   email={john.RS.Smith@cambridge.co.uk}
+   email={szolek@informatik.uni-tuebingen.de}
 ]{\inits{AS}\fnm{Andr\'{a}s} \snm{Szolek}}
@@
    addressref={aff4,aff5},
-   email={john.RS.Smith@cambridge.co.uk}
+   email={mohamed.el-hadidi@uni-tuebingen.de}
 ]{\inits{MEH}\fnm{Mohamed} \snm{El Hadidi}}
@@
 genome assembly software to produce better
-assemblies~\cite{Bryant:2009uq,Gnerre:2011kx}. Short-read
-alignment programs, such as Bowtie2~\cite{Langmead:2012rw}, use
-quality values to weight mismatches between read and reference
-sequences. Software for detecting single nucleotide polymorphisms
-(\textsc{snp}s) can use quality values~\cite{McKenna:2010bh},
-and identified \textsc{snp}s with high-quality calls are deemed more
-reliable than those with low-quality calls, particularly in
-low-coverage regions.
+assemblies~\cite{Bryant:2009uq,Gnerre:2011kx}. Short-read alignment
+programs, such as Bowtie2~\cite{Langmead:2012rw}, use quality values
+to weight mismatches between read and reference sequences. Software
+for detecting single nucleotide polymorphisms (\textsc{snp}s) can use
+quality values~\cite{McKenna:2010bh}, and identified \textsc{snp}s
+with high-quality calls are deemed more reliable than those with
+low-quality calls, particularly in low-coverage regions.
 
@@
 on lossless compression of base calls~\cite{Deorowicz:2013hq,
-Giancarlo:2009fk,Giancarlo:2014rw,
-  Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
-addressed compression of quality
+  Giancarlo:2009fk,Giancarlo:2014rw, Nalbantoglu:2010uq,Zhu:2013qr},
+although some recent work has addressed compression of quality
 values~\cite{Canovas:2014fr,Hach:2012ys,
@@
 marked ``bad'' if their quality value falls below the first quartile
-minus 1.5 $\times$ the interquartile range (\textsc{iqr}), which is the
-difference between the first and third quartile; 1.5 $\times$ \textsc{iqr} is
-the value used by Tukey's box plot~\cite{mcgill1978variations}. The
-main benefit of this approach is that it is completely data-dependent,
-and no assumptions regarding the distribution of the quality values
-need to be made. We adopt the quality value assignments of the read
-preprocessing tool Sickle~\cite{sickle}, which correspond to the 40
-for ``good'' and 10 for ``bad''.
+minus 1.5 $\times$ the interquartile range (\textsc{iqr}), which is
+the difference between the first and third quartile; 1.5 $\times$
+\textsc{iqr} is the value used by Tukey's box
+plot~\cite{mcgill1978variations}. The main benefit of this approach is
+that it is completely data-dependent, and no assumptions regarding the
+distribution of the quality values need to be made. We adopt the
+quality value assignments of the read preprocessing tool
+Sickle~\cite{sickle}, which correspond to the 40 for ``good'' and 10
+for ``bad''.
 
@@
 
-The economic costs of memory use for binning, in general terms,
-include no fixed costs, and marginal costs that are the product of the
-number of base-call quality values times the cost of the encoding.
+The economic costs of file size for binning, in general terms, include
+no fixed costs, and marginal costs that are the product of the number
+of base-call quality values times the cost of the encoding.
 
@@
 Although we focus only on two bins in this work, a greater number of
-bins can be used in practice, albeit with a higher cost in memory
-usage. For example, an additional bin, thus giving bins of good, bad,
-and intermediate, could be used for quality values near the border
-between bins in the 2-bin encoding. Additional bins beyond two, and
-their resulting compressibility and effect on downstream analyses, are
-not considered here.
+bins can be used in practice, albeit with a higher cost in file size.
+For example, an additional bin, thus giving bins of good, bad, and
+intermediate, could be used for quality values near the border between
+bins in the 2-bin encoding. Additional bins beyond two, and their
+resulting compressibility and effect on downstream analyses, are not
+considered here.
 
@@
 
-The economic costs of memory use for model-based compression, in
+The economic costs of file size for model-based compression, in
 general terms, include no fixed costs, and marginal costs that are the
@@
 
-Sets of strings representing quality values show similar trends of
-over their length, and it is possible to identify common patterns in
-the data and use them as reference profiles to approximate individual
+Sets of strings representing quality values show similar trends over
+their length, and it is possible to identify common patterns in the
+data and use them as reference profiles to approximate individual
 sequences of quality values. Here we use $k$-means clustering, a
@@
 
-The economic costs of memory use for profile-based compression, in
+The economic costs of file size for profile-based compression, in
 general terms, include fixed costs associated with representing the
@@
 binning, has the advantageous property that marginal costs are the
-reciprocal of the read length. Thus the costs of memory use for
+reciprocal of the read length. Thus the costs of file size for
 profile-based compression decrease proportionally with increases in
@@
 For comparison to other developed methods we use \textsc{QualComp}, a
-lossy compression tool~\cite{Ochoa:2013rt}. The program models
-quality values as a multivariate Gaussian distribution, computing the
-mean and covariance for each position in the read, and these values
-are stored by the decoder to later reconstruct a \emph{representative}
-quality value. \textsc{QualComp} takes as input a user-specified rate
+lossy compression tool~\cite{Ochoa:2013rt}. The program models quality
+values as a multivariate Gaussian distribution, computing the mean and
+covariance for each position in the read, and these values are stored
+by the decoder to later reconstruct a \emph{representative} quality
+value. \textsc{QualComp} takes as input a user-specified rate
 (bits/read), and these bits are apportioned among positions within the
@@
 
-The economic costs of memory use for compression with
+The economic costs of file size for compression with
 \textsc{QualComp}, in our interpretation, appear to have fixed costs
@@
 We secondarily compress all otherwise compressed data sets using the
-Burrows-Wheeler algorithm~\cite{bwt} via \textsc{bz}ip2, and it is
-the results of these subsequent compressions that we report. Although
+Burrows-Wheeler algorithm~\cite{bwt} via \textsc{bz}ip2, and it is the
+results of these subsequent compressions that we report. Although
 other algorithms~\cite{HUFFMAN:1952nr,Ziv77auniversal,
@@
 As a measure of compression effectiveness, which reflects the sum of
-fixed and marginal costs, we use
-bits/base-call, defined as the size of the compressed representation
-of quality values (in bits) divided by the number of quality values
-represented. As a measure of information loss we use mean squared
-error (\textsc{mse}) as a loss function, and define it as
-$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
-quality values, $Q_i'$ is the compressed/decompressed quality value,
-and $Q_i$ is the original quality value associated with position $i$.
+fixed and marginal costs, we use bits/base-call, defined as the size
+of the compressed representation of quality values (in bits) divided
+by the number of quality values represented. As a measure of
+information loss we use mean squared error (\textsc{mse}) as a loss
+function, and define it as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$,
+where $n$ is the number of quality values, $Q_i'$ is the
+compressed/decompressed quality value, and $Q_i$ is the original
+quality value associated with position $i$.
 
@@
 
-\section*{Results}
+\section*{Results and discussion}
 
@@
 bits/read has the lowest \textsc{mse}, but requires
-$\sim$10--15$\times$ more memory than the profile encoding methods for
-only a $\sim$2--3$\times$ reduction in error. When the rate parameter
-of \textsc{q}ual\textsc{c}omp is set to match the profile encoding
-methods, \textsc{q}ual\textsc{c}omp performs slightly worse in terms
-of \textsc{mse}. For the \textit{Rhodobacter sphaeroides} fragment
-library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
-(0.099 bits/base-call) has a \textsc{mse} of 17.29, whereas
-256-profile encoding only requires 0.079 bits/base-call and has a
-\textsc{mse} of 11.85.
+$\sim$10--15$\times$ larger file size than the profile encoding
+methods for only a $\sim$2--3$\times$ reduction in error. When the
+rate parameter of \textsc{q}ual\textsc{c}omp is set to match the
+profile encoding methods, \textsc{q}ual\textsc{c}omp performs slightly
+worse in terms of \textsc{mse}. For the \textit{Rhodobacter
+  sphaeroides} fragment library, \textsc{q}ual\textsc{c}omp with rate
+set to 10 bits/read (0.099 bits/base-call) has a \textsc{mse} of
+17.29, whereas 256-profile encoding only requires 0.079 bits/base-call
+and has a \textsc{mse} of 11.85.
 
@@
 \textsc{bz}ip2. For the \textit{Mus musculus} data set, which has read
-lengths of only 36 nt, the memory required for storage, even after
+lengths of only 36 nt, the file size, even after
 subsequent compression with \textsc{bz}ip2 exceeds that for the
@@
 Some short read alignment tools, such as Bowtie2 (version 2.2.3),
-which was used here, utilize quality value information when evaluating
-potential alignments. The reads from with original uncompressed and
-compressed/decompressed quality vlaues were mapped with Bowtie2 to the
-\textit{Rhodobacter sphaeroides} reference genome. The total, shared,
-and unique proportion of mapped reads are calculated with respect to
-the results for the the original uncompressed quality values as shown
-in Table \ref{tab:aligner}. Additionally, to assess the effect of
-quality values on mapping in general, Bowtie2 was adjusted so that the
-maximum and minimum mismatch penalty were equivalent to maximum and
-minimum quality scores (with parameters: --mp 6,6 and --mp 2,2
-respectively).
-
-We evaluate the compression methods using two approaches. In the first
-approach, we order the compression methods based on how similar the
-alignment results are using the original uncompressed quality values
---- i.e., the number of reads aligned using the original uncompressed
-quality values and quality values resulting from compression, plus the
-number of reads unaligned using the original uncompressed quality
-values and quality values resulting from compression minus the amount
-of reads uniquely aligned using the original uncompressed quality
-values and quality values resulting from compression. In the second
-approach, we order the compression methods by total proportion of
-aligned reads.
+which was used here, utilize quality value information whenover
+evaluating potential alignments. The reads from with original
+uncompressed and compressed/decompressed quality values were mapped
+with Bowtie2 to the \textit{Rhodobacter sphaeroides} reference
+genome. The total, shared, and unique proportion of mapped reads are
+calculated with respect to the results for the original uncompressed
+quality values as shown in Table \ref{tab:aligner}. Additionally, to
+assess the effect of quality values on mapping in general, Bowtie2 was
+adjusted so that the maximum and minimum mismatch penalty were
+equivalent to maximum and minimum quality scores (with parameters:
+--mp 6,6 and --mp 2,2 respectively).
+
+We evaluate the compression methods to reflect that comparisons among
+mapping identical reads with different quality values may vary in two
+mutually exclusive ways: reads map to the reference, and reads do not
+map to a reference. Thus comparing mapping using original quality
+values and reads using quality values modified via lossy compression,
+we have a $2 \times 2$ contingency table consisting of reads that map
+for both original and compressed, reads that map for original and do
+not map for compressed, reads that do not map for original and map for
+compressed, and reads that do not map for both original and
+compressed.
 
@@
 transformation associated with compression. Thus appropriate
-compression of quality values can both reduce associated memory cost
+compression of quality values can both reduce associated file size cost
 and improve analysis results.
-
-\section*{Section title}
-Text for this section \ldots
-\subsection*{Sub-heading for section}
-Text for this sub-heading \ldots
-\subsubsection*{Sub-sub heading for section}
-Text for this sub-sub-heading \ldots
-\paragraph*{Sub-sub-sub heading for section}
-Text for this sub-sub-sub-heading \ldots
-In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
-addition, we examine a common modeling assumption and note the
-importance of considering the tails of the extinction time $T_x$ in
-studies of escape dynamics.
-We will first consider the expected resistant population at $vT_x$ for
-some $v>0$, (and temporarily assume $\alpha=0$)
-%
-\[
- E \bigl[Z_1(vT_x) \bigr]= E
-\biggl[\mu T_x\int_0^{v\wedge
-1}Z_0(uT_x)
-\exp \bigl(\lambda_1T_x(v-u) \bigr)\,du \biggr].
-\]
-%
-If we assume that sensitive cells follow a deterministic decay
-$Z_0(t)=xe^{\lambda_0 t}$ and approximate their extinction time as
-$T_x\approx-\frac{1}{\lambda_0}\log x$, then we can heuristically
-estimate the expected value as
-%
-\begin{eqnarray}\label{eqexpmuts}
-E\bigl[Z_1(vT_x)\bigr] &=& \frac{\mu}{r}\log x
-\int_0^{v\wedge1}x^{1-u}x^{({\lambda_1}/{r})(v-u)}\,du
-\nonumber\\
-&=& \frac{\mu}{r}x^{1-{\lambda_1}/{\lambda_0}v}\log x\int_0^{v\wedge
-1}x^{-u(1+{\lambda_1}/{r})}\,du
-\nonumber\\
-&=& \frac{\mu}{\lambda_1-\lambda_0}x^{1+{\lambda_1}/{r}v} \biggl(1-\exp \biggl[-(v\wedge1) \biggl(1+
-\frac{\lambda_1}{r}\biggr)\log x \biggr] \biggr).
-\end{eqnarray}
-%
-Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).
-%\nocite{oreg,schn,pond,smith,marg,hunn,advi,koha,mouse}
 
@@
 \section*{Acknowledgements}
-  Text for this section \ldots
+This project was initiated at the 2014 Bioinformatics Exchange for
+Students and Teachers (\textsc{best}) Summer School, which was funded
+by the offices of the Dean of The Graduate School, University of
+Maryland, and the Rektor of the University of T\"{u}bingen. Additional
+funding included an International Graduate Research Fellowship from
+The Graduate School, University of Maryland, to CMH, a Global
+Partnerships-Faculty Travel Grant from the Office of International
+Affairs to MPC, and funding from the University of T\"{u}bingen.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=3.35in]{profiles128.eps}}
-\caption{\csentence{Quality profiles obtained by $k$-means clustering.} 
-  Example for the
-  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
-  set using $k$ = 128. Each row corresponding to a quality
-  profile. Dark to light colors represent low to high quality
-  values. It is readily visible that the two most distinctive features
-  of quality profiles is their drop-off position and average overall
-  quality. Occasional low values occur early in some profiles, likely
-  reflecting intermittent problems in the sequencing process affecting
-  many reads at a time.}
+% profiles128.eps 
+\caption{\csentence{Quality profiles obtained by $k$-means
+    clustering.} Example for the fragment library from
+  \textit{Rhodobacter sphaeroides} 2.4.1 data set using $k$ =
+  128. Each row corresponding to a quality profile. Dark to light
+  colors represent low to high quality values. It is readily visible
+  that the two most distinctive features of quality profiles is their
+  drop-off position and average overall quality. Occasional low values
+  occur early in some profiles, likely reflecting intermittent
+  problems in the sequencing process affecting many reads at a time.}
   \label{fig:profiles_128}
@@
 \begin{figure*}[h!]
-\centerline{\includegraphics[width=7in]{compression.results.eps}}
-\caption{\csentence{The relationship of bits/base-call and mean squared error for
-  quality value compression methods applied to four data sets.}
-  The data sets are \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
-  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
-  \textit{Mus musculus}. Point labels correspond to different
-  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
-  profiles; R$n$, modeling with polynomial regression of degree $n$;
-  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
-  denote the corresponding lossless compression using \textsc{bz}ip2,
-  with the black arrow corresponding to original data.}
+% compression.results.eps
+\caption{\csentence{The relationship of bits/base-call and mean
+    squared error for quality value compression methods applied to
+    four data sets.} The data sets are \textit{Rhodobacter
+    sphaeroides} 2.4.1; \textit{Homo sapiens} chromosome 14;
+  \textit{Escherichia coli} str. K-12 MG1655; and \textit{Mus
+    musculus}. Point labels correspond to different compression
+  methods: 2B, 2-bin encoding; P$n$, profiling with $n$ profiles;
+  R$n$, modeling with polynomial regression of degree $n$; Q$n$,
+  \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows denote
+  the corresponding lossless compression using \textsc{bz}ip2, with
+  the black arrow corresponding to original data.}
 \label{fig:mse_vs_bpbp}
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=3.65in]{preprocessing.results.eps}}
-\caption{Preprocessing results of \textit{Rhodobacter sphaeroides}
-  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
-  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
+% preprocessing.results.eps
+\caption{\csentence{Read preprocessing results using compressed
+    quality values compared original uncompressed quality values.}
+  The reads are for the \textit{Rhodobacter sphaeroides} 2.4.1, and
+  \textit{Homo sapiens} chromosome 14 fragment libraries, and
+  \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
     musculus} data sets. Reads were trimmed using Sickle. The total
@@
 \begin{figure}[h!]
-\centerline{\includegraphics[width=3.65in]{rhodo.assembly.results.eps}}
-\caption{\csentence{Rankings of compression methods based on assembly attributes}
- The read data are for \textit{Rhodobacter
+% rhodo.assembly.results.eps
+\caption{\csentence{Rankings of compression methods based on assembly
+    attributes.} The read data are for \textit{Rhodobacter
     sphaeroides} 2.4.1, and the results sorted by overall
@@
 
-  \begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      A short description of the figure content
-      should go here.}
-      \end{figure}
-
-\begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      Figure legend text.}
-      \end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{table*}[!tbhp]
-\centering
-\caption[]{Comparison of mapping for original reads and reads with
+\caption{Comparison of mapping for original reads and reads with
   compressed/decompressed quality values. Reads and reference genome
@@
   using Bowtie2.}
-\begin{small}
+\begin{tiny}
 \begin{tabular}{lr|cc|cc|cc|cc|cc}
@@
 \end{tabular}
-\end{small}
+\end{tiny}
 
@@
 \end{table*}
-
-\begin{table}[h!]
-\caption{Sample table title. This is where the description of the table should go.}
-      \begin{tabular}{cccc}
-        \hline
-           & B1  &B2   & B3\\ \hline
-        A1 & 0.1 & 0.2 & 0.3\\
-        A2 & ... & ..  & .\\
-        A3 & ..  & .   & .\\ \hline
-      \end{tabular}
-\end{table}
 

### 2016-04-12 src/compress.py
Added RQS to pipeline.

@@ -103,6 +103,8 @@ def compress(options):
     QUALCOMP_COMPRESS_CMD = "./runCompress.sh -i [READ] -c [CLUSTERS] -r [RATE]"
     QUALCOMP_DECOMPRESS_CMD = "./runDecompress.sh -p [DIR] -c [CLUSTERS] -r [RATE]"
 
+    RQS_COMPRESS_CMD = "./src/run_rqs.sh [READ] [OUTPUT]"
+
     # Store which compression directories we created.
     options.compressed_dirs = []
     options.compressed_dirs.append('original')
@@ -142,99 +144,134 @@ def compress(options):
         #continue
 
         # Polynomial regression.
-        for degree in options.poly_degrees.split(','):
-            ensure_dir(options.output_dir + '/degree_' + degree + '/')
+        if options.poly_degrees:
+            for degree in options.poly_degrees.split(','):
+                ensure_dir(options.output_dir + '/degree_' + degree + '/')
 
-            if 'degree_' + degree not in options.compressed_dirs:
-                options.compressed_dirs.append('degree_' + degree)
+                if 'degree_' + degree not in options.compressed_dirs:
+                    options.compressed_dirs.append('degree_' + degree)
 
-            #continue
+                #continue
 
-            call_arr = POLY_REGRESSION_CMD.replace('[READ]', reads_filename)\
-                    .replace('[OUTPUT]', options.output_dir + '/degree_' + degree + '/' + os.path.basename(reads_filename))\
-                    .replace('[DEGREE]', degree)\
-                    .replace('[COMPRESSED_FILE]', options.output_dir + '/degree_' + degree +'/' + os.path.basename(reads_filename) + '.comp')\
-                    .replace('[NUM_THREADS]', options.threads).split()
+                call_arr = POLY_REGRESSION_CMD.replace('[READ]', reads_filename)\
+                        .replace('[OUTPUT]', options.output_dir + '/degree_' + degree + '/' + os.path.basename(reads_filename))\
+                        .replace('[DEGREE]', degree)\
+                        .replace('[COMPRESSED_FILE]', options.output_dir + '/degree_' + degree +'/' + os.path.basename(reads_filename) + '.comp')\
+                        .replace('[NUM_THREADS]', options.threads).split()
 
-            out_cmd("", std_err_file.name, call_arr)
-            call(call_arr, stderr=std_err_file)
+                out_cmd("", std_err_file.name, call_arr)
[…]

### 2016-04-12 src/run_rqs.sh

@@ -1,0 +1,39 @@
+#!/bin/bash
+# echo First convert FASTQ files to a fake SAM format:
+$RQS/fastq2fsam.sh $1
+output=$2/`basename $1`.fsam
+mv $1.fsam $output
+
+# echo Generate our dictionary with read multiplicity r=2:
+#../generate_dict 2 dict.db $1.fsam
+$RQS/generate_dict 2 $2/dict.db $output
+
+# echo In this example, we compress the corpus we used to generate the dictionary:
+$RQS/sparsify $2/dict.db $output
+
+# echo After sparsification, we still need to cut-off high quality scores:
+# echo "Let's choose a threshold of 'I', or Phred quality 40 under most encodings"
+# echo "../threshold 'I' *.fsam.filtered"
+$RQS/threshold 'I' $output.filtered
+
+# echo Compression is now done, but let\'s see how well we did using BZIP2:
+# echo "We'll cut out column 11 (the quality scores) and pipe it through BZIP2"
+# echo and then compute the bits per quality value needed.
+for filename in $output.filtered.reduced
+do
+        file=`basename $filename`
+        cut -f11 $filename > $2/$file.qual
+        orig_size=`wc -c < $2/$file.qual`
+        orig_lines=`wc -l < $2/$file.qual`
+        orig_size=`echo "$orig_size - $orig_lines" | bc`
+        bzip2 -f $2/$file.qual
+        new_size=`wc -c < $2/$file.qual.bz2`
+        #rm $file.qual.bz2
+        echo -e $file:'\t' `echo "scale=4; 1/( $orig_size / ( $new_size * 8)) " | bc` bits / quality score
+done
+
+# echo "Although we're basically done, you might want to convert your files back"
+# echo "from this fake SAM format to a FASTQ file:"
+# echo "../fsam2fastq.sh *.filtered.reduced"
+$RQS/fsam2fastq.sh $output.filtered.reduced
+mv $output.filtered.reduced.fastq $2/`basename $1`

### 2016-04-12 bmc_manuscript/bmc_article.tex

--- 2015-07-17 bioinformatics_manuscript/q-compression.tex
@@
-\documentclass{bioinfo}
+%% BioMed_Central_Tex_Template_v1.06
+%%                                      %
+%  bmc_article.tex            ver: 1.06 %
+%                                       %
 
-%\usepackage{doi}
+%%IMPORTANT: do not delete the first line of this template
+%%It must be present to enable the BMC Submission system to
+%%recognise this template!!
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                     %%
+%%  LaTeX template for BioMed Central  %%
+%%     journal article submissions     %%
+%%                                     %%
+%%          <8 June 2012>              %%
+%%                                     %%
+%%                                     %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%
+%% For instructions on how to fill out this Tex template           %%
+%% document please refer to Readme.html and the instructions for   %%
+%% authors page on the biomed central website                      %%
+%% http://www.biomedcentral.com/info/authors/                      %%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%% BioMed Central currently use the MikTex distribution of         %%
+%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
+%% http://www.miktex.org                                           %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%% additional documentclass options:
+%  [doublespacing]
+%  [linenumbers]   - put the line numbers on margins
+
+%%% loading packages, author definitions
+
+%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+\documentclass{bmcart}
+
+%%% Load packages
+%\usepackage{amsthm,amsmath}
+\RequirePackage{natbib}
+%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
+%\RequirePackage{hyperref}
+%\usepackage[utf8]{inputenc} %unicode support
+%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
 \usepackage{url}
 
-\copyrightyear{2015}
-\pubyear{2015}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                             %%
+%%  If you wish to display your graphics for   %%
+%%  your own use using includegraphic or       %%
+%%  includegraphics, then comment out the      %%
+%%  following two lines of code.               %%
+%%  NB: These line *must* be included when     %%
+%%  submitting to BMC.                         %%
+%%  All figure files must be submitted as      %%
+%%  separate graphics through the BMC          %%
+%%  submission process, not included in the    %%
+%%  submitted article.                         %%
+%%                                             %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%\def\includegraphic{}
+%\def\includegraphics{}
+\usepackage{graphicx}
+
+%%% Put your definitions there:
+\startlocaldefs
+\endlocaldefs
+
+%%% Begin ...
 \begin{document}
-\firstpage{1}
 
-\title[lossy-compression]{Lossy compression of DNA sequencing quality
+%%% Start of article front matter
+\begin{frontmatter}
+
+\begin{fmbox}
+\dochead{Research}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\title{Lossy compression of DNA sequencing quality
   data}
 
-\author[Hill \textit{et~al.}]{Christopher M. Hill\,$^{1}$, Andr\'{a}s
-  Szolek\,$^{2}$, Mohamed El Hadidi\,$^{3}$, and Michael
-  P. Cummings\,$^4$\footnote{to whom correspondence should be
-    addressed}} \address{$^{1}$Department of Computer Science,
-  University of Maryland, College Park, Maryland, 20742
-  USA\\ $^{2}$Department of Applied Bioinformatics, Center for
-  Bioinformatics, Quantitative Biology Center, and Department of
-  Computer Science, University of T\"{u}bingen, Sand 14, 72076
-  T\"{u}bingen, Germany\\ $^{3}$Department of Algorithms in
-  Bioinformatics, Center for Bioinformatics, University of
-  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany \\ $^{4}$Center
-  for Bioinformatics and Computational Biology, University of
-  Maryland, College Park, Maryland, 20742 USA}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors here                   %%
+%%                                          %%
+%% Specify information, if available,       %%
+%% in the form:                             %%
+%%   <key>={<id1>,<id2>}                    %%
+%%   <key>=                                 %%
+%% Comment or delete the keys which are     %%
+%% not used. Repeat \author command as much %%
+%% as required.                             %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXX}
+\author[
+   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
+   %corref={aff1},                       % id of corresponding address, if any
+   %noteref={n1},                        % id's of article notes, if any
+   email={chrismh@uw.edu}   % email address
+]{\inits{CMH}\fnm{Christopher M.} \snm{Hill}}
+\author[
+   addressref={aff2}                   % id's of addresses, e.g. {aff1,aff2}
+   %corref={aff1},                       % id of corresponding address, if any
+   %noteref={n1},                        % id's of article notes, if any
+   %email={}   % email address
+]{\inits{AS}\fnm{Andr\'{a}s} \snm{Szolek}}
+\author[
+   addressref={aff3}                   % id's of addresses, e.g. {aff1,aff2}
+   %corref={aff1},                       % id of corresponding address, if any
+   %noteref={n1},                        % id's of article notes, if any
+   %email={}   % email address
+]{\inits{MH}\fnm{Mohamed} \snm{El Hadidi}}
+\author[
+   addressref={aff4},
+   corref={aff4},
+   email={mike@umiacs.umd.edu}
+]{\inits{MPC}\fnm{Michael P.} \snm{Cummings}}
 
-\editor{Associate Editor: XXXXXXX}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors' addresses here        %%
+%%                                          %%
+%% Repeat \address commands as much as      %%
+%% required.                                %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\maketitle
+\address[id=aff1]{%                           % unique id
+  \orgname{Department of Genome Science, University of Washington School of Medicine},                    %
+  \postcode{98195},                                % post or zip code
+  \city{Seattle},                              % city
+  \cny{USA}                                    % country
+}
+\address[id=aff2]{%
+  \orgname{Department of Applied Bioinformatics, Center for Bioinformatics, Quantitative Biology Center, and Department of Computer Science, University of T\"{u}bingen},
+  \street{Sand 14},
+  \postcode{72076},
+  \city{T\"{u}bingen},
+  \cny{Germany}
+}
+\address[id=aff3]{%
+  \orgname{Department of Algorithms in Bioinformatics, Center for Bioinformatics, University of T\"{u}bingen},
+  \street{Sand 14},
+  \postcode{72076},
+  \city{T\"{u}bingen},
+  \cny{Germany}
+}
+\address[id=aff4]{%                           % unique id
+  \orgname{Center for Bioinformatics and Computational Biology, University of Maryland},                  %
+  \postcode{20742},                                % post or zip code
+  \city{College Park},                              % city
+  \cny{USA}                                    % country
+}
 
-\begin{abstract}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter short notes here                   %%
+%%                                          %%
+%% Short notes will be after addresses      %%
+%% on first page.                           %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Motivation:}
+\begin{artnotes}
+%\note{Sample of title note}     % note to the article
+%\note[id=n1]{Equal contributor} % note, connected to author
+\end{artnotes}
 
+\end{fmbox}% comment this for two column layout
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Abstract begins here                 %%
+%%                                          %%
+%% Please refer to the Instructions for     %%
+%% authors on http://www.biomedcentral.com  %%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{abstractbox}
+
+\begin{abstract} % abstract
+\parttitle{Motivation:} %if any
 As the cost of sequencing continues to decrease, the rate of sequence
@@
 
-\section{Results:}
+\parttitle{Results:} %if any
 Lossy compression of quality information can greatly decrease memory
@@
 
-\section{Contact:} \href{mike@umiacs.umd.edu}{mike@umiacs.umd.edu}
 \end{abstract}
 
-\section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The keywords begin here                  %%
+%%                                          %%
+%% Put each keyword in separate \kwd{}.     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\begin{keyword}
+\kwd{lossy compression}
+\kwd{sequencing}
+\end{keyword}
+
+% MSC classifications codes, if any
+%\begin{keyword}[class=AMS]
+%\kwd[Primary ]{}
+%\kwd{}
+%\kwd[; secondary ]{}
+%\end{keyword}
+
+\end{abstractbox}
+%
+%\end{fmbox}% uncomment this for twcolumn layout
+
+\end{frontmatter}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Main Body begins here                %%
+%%                                          %%
+%% Please refer to the instructions for     %%
+%% authors on:                              %%
+%% http://www.biomedcentral.com/info/authors%%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%% See the Results and Discussion section   %%
+%% for details on how to create sub-sections%%
+%%                                          %%
+%% use \cite{...} to cite references        %%
+%%  \cite{koon} and                         %%
+%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
+%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
+% <put your article body there>
+
+%%%%%%%%%%%%%%%%
+%% Background %%
+%%
+\section*{Background}
 Read data from high-throughput sequencing constitutes the largest
@@
 
-\subsection{Comparison to other methods}
+\subsection*{Comparison to other methods}
 
@@
 
-\subsection{Subsequent (secondary) compression}
+\subsection*{Subsequent (secondary) compression}
 
@@
 
-\end{methods}
+\section*{Results}
 
-\section{Results}
-
-\subsection{Compression effectiveness versus information loss}
-
-\begin{figure*}[!tb]
-\centerline{\includegraphics[width=7in]{compression_results.eps}}
-\caption{The relationship of bits/base-call and mean squared error for
-  quality value compression methods applied to four data sets:
-  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
-  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
-  \textit{Mus musculus}. Point labels correspond to different
-  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
-  profiles; R$n$, modeling with polynomial regression of degree $n$;
-  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
-  denote the corresponding lossless compression using \textsc{bz}ip2,
-  with the black arrow corresponding to original data.}
-\label{fig:mse_vs_bpbp}
-\end{figure*}
+\subsection*{Compression effectiveness versus information loss}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-\section{Availability}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Backmatter begins here                   %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{backmatter}
+
+\section*{Competing interests}
+  The authors declare that they have no competing interests.
+
+\section*{Author's contributions}
+    Text for this section \ldots
+
+\section*{Availability}
 Implementations of the compression methods written in Python and
@@
 
-\section*{Acknowledgement}
+\section*{Acknowledgements}
 This project was initiated at the 2014 Bioinformatics Exchange for
@@
 Affairs to MPC, and funding from the University of T\"{u}bingen.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                  The Bibliography                       %%
+%%                                                         %%
+%%  Bmc_mathpys.bst  will be used to                       %%
+%%  create a .BBL file for submission.                     %%
+%%  After submission of the .TEX file,                     %%
+%%  you will be prompted to submit your .BBL file.         %%
+%%                                                         %%
+%%                                                         %%
+%%  Note that the displayed Bibliography will not          %%
+%%  necessarily be rendered by Latex exactly as specified  %%
+%%  in the online Instructions for Authors.                %%
+%%                                                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-%
-%\bibliographystyle{plain}
-%
+% if your bibliography is in bibtex format, use those commands:
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{compression.bib}      % Bibliography file (usually '*.bib' )
+% for author-year bibliography (bmc-mathphys or spbasic)
+% a) write to bib file (bmc-mathphys only)
+% @settings{label, options="nameyear"}
+% b) uncomment next line
+%\nocite{label}
 
-\bibliography{compression}
+% or include bibliography directly:
+% \begin{thebibliography}
+% \bibitem{b1}
+% \end{thebibliography}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Figures                       %%
+%%                               %%
+%% NB: this is for captions and  %%
+%% Titles. All graphics must be  %%
+%% submitted separately and NOT  %%
+%% included in the Tex document  %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%
+%% Do not use \listoffigures as most will included as separate files
+
+\section*{Figures}
+  \begin{figure}[h!]
+  \centerline{\includegraphics[width=3.35in]{profiles_128.eps}}
+  \caption{\csentence{Quality profiles obtained by $k$-means clustering.}
+  Quality profiles obtained by $k$-means clustering on the
+    fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
+    set using $k$ = 128, with each row corresponding to a quality
+    profile. Dark to light colors represent low to high quality
+    values. It is readily visible that the two most distinctive features
+    of quality profiles is their drop-off position and average overall
+    quality. Occaional low values occur early in some profiles, likely
+    relecting intermittent problems in the sequencing process affecting
+    many reads at a time.}\label{fig:profiles_128}
+      \end{figure}
+
+\begin{figure}[h!]
+  \centerline{\includegraphics[width=7in]{compression_results.eps}}
+  \caption{\csentence{The relationship of bits/base-call and mean squared error for quality value compression methods applied to four data sets:
+    \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
+    chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
+    \textit{Mus musculus}.} Point labels correspond to different
+    compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
+    profiles; R$n$, modeling with polynomial regression of degree $n$;
+    Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
+    denote the corresponding lossless compression using \textsc{bz}ip2,
+    with the black arrow corresponding to original data.}
+  \label{fig:mse_vs_bpbp}
+      \end{figure}
+
+\begin{figure}[h!]
+  \centerline{\includegraphics[width=3.65in]{preprocessing_results.eps}}
+  \caption{\csentence{Preprocessing results of \textit{Rhodobacter sphaeroides}
+    2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
+    and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
+      musculus} data sets.} Reads were trimmed using Sickle. The total
+    number of bases filtered by each compression method is compared with
+    the number of bases filtered using the uncompressed quality values.}
+    \label{fig:preprocessing}
+      \end{figure}
+
+\begin{figure}[h!]
+  \centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.eps}}
+  \caption{\csentence{Rankings of compression methods based on \textit{Rhodobacter
+      sphaeroides} assembly attributes sorted by overall
+    rank.} Assemblies were constructed using \textsc{allpaths-lg}.
+    Rankings above the median value are in cyan, those below the median
+    value in magenta.}
+    \label{fig:assembly_ranks}
+    \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Tables                        %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Use of \listoftables is discouraged.
+%%
+\section*{Tables}
+\begin{table}[t]
+\centering
+\caption[]{Comparison of mapping for original reads and reads with
+  compressed/decompressed quality values. Reads and reference genome
+  are for \textit{Rhodobacter sphaeroides}, and mapping was performed
+  using Bowtie2.}
+\begin{tiny}
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{max-qual} & \multicolumn{2}{c|}{min-qual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{regression (0)} & \multicolumn{2}{c}{regression (1)} \\
+& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\
+\cline{2-12}
+& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\
+{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\
+\cline{2-12}
+& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+ & & \multicolumn{2}{c|}{regression (3)} & \multicolumn{2}{c|}{regression (5)} & \multicolumn{2}{c|}{regression (7)} & \multicolumn{2}{c|}{profile (64)} & \multicolumn{2}{c}{profile (128)} \\
+& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\
+\cline{2-12}
+& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\
+{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\
+\cline{2-12}
+& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
+\end{tabular}
+
+\bigskip
+
+\begin{tabular}{lr|cc|cc|cc|cc|cc}
+&  & \multicolumn{2}{c|}{profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
+& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\
+\cline{2-12}
+& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\
+{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\
+\cline{2-12}
+& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
+\end{tabular}
+\end{tiny}
+
+\label{tab:aligner}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Additional Files              %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Additional Files}
+  \subsection*{Additional file 1 --- Sample additional file title}
+    Additional file descriptions text (including details of how to
+    view the file, if it is in a non-standard format or the file extension).  This might
+    refer to a multi-page table or a figure.
+
+  \subsection*{Additional file 2 --- Sample additional file title}
+    Additional file descriptions text.
+
+\end{backmatter}
 \end{document}

### 2016-04-12 bmc_manuscript/manuscipt.tex
Created a merged version of the two manuscripts that Chris and I independently created.

@@
    addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
-   noteref={n1},                        % id's of article notes, if any
-   email={jane.e.doe@cambridge.co.uk}   % email address
+%   noteref={n1},                        % id's of article notes, if any
+   email={chrismh@uw.edu}   % email address
 ]{\inits{CMH}\fnm{Christopher M} \snm{Hill}}
 \author[
-   addressref={aff3,aff5},
+   addressref={aff4,aff6},
    email={szolek@informatik.uni-tuebingen.de}
@@
 \author[
-   addressref={aff4,aff5},
+   addressref={aff5,aff6},
    email={mohamed.el-hadidi@uni-tuebingen.de}
@@
 \author[
-   addressref={aff2},
+   addressref={aff3},
    corref={aff2},                       % id of corresponding address, if any
@@
 \address[id=aff2]{%                           % unique id
+  \orgname{Department of Genome Science, University of Washington School of Medicine},                    %
+  \city{Seattle, Washington},                              % city
+  \postcode{98195},                                % post or zip code
+  \cny{USA}                                    % country
+}
+\address[id=aff3]{%                           % unique id
   \orgname{Center for Bioinformatics and Computational Biology, University of Maryland}, % university, etc
@@
 }
-\address[id=aff3]{%
+\address[id=aff4]{%
   \orgname{Department of Applied Bioinformatics, Center for
@@
 }
-\address[id=aff4]{%
+\address[id=aff5]{%
   \orgname{Department of Algorithms in Bioinformatics, Center for
@@
 }
-\address[id=aff5]{%
+\address[id=aff6]{%
   \orgname{Department of Computer Science, University of T\"{u}bingen},
@@
 
-\begin{artnotes}
+%\begin{artnotes}
 %\note{Sample of title note}     % note to the article
-\note[id=n1]{Current address: \ldots} % note, connected to author
-\end{artnotes}
+%\note[id=n1]{Current address: \ldots} % note, connected to author
+%\end{artnotes}
 
@@
 \begin{abstract} % abstract
-\parttitle{First part title} %if any
-Text for this section.
-
-\parttitle{Second part title} %if any
-Text for this section.
+\parttitle{Motivation:} %if any
+As the cost of sequencing continues to decrease, the rate of sequence
+data production is increasing, placing greater demands on storing and
+transferring these vast amount of data. Most methods of sequencing
+data compression focus on compressing nucleotide information without
+any loss of information. Quality data, however, have different
+properties than nucleotide data, and methods compressing nucleotide
+sequences efficiently do not perform as well on quality
+values. Furthermore, although lossless representation might be
+necessary for nucleotide sequences, it is not an essential requirement
+for quality values.
+
+Previous studies of quality value compression have mostly focus on
+minimizing the loss of information with less emphasis on the effects
+on bioinformatic analyses. In this paper, we evaluate several
+different compression methods for quality values, and assess the
+resulting impacts on common bioinformatic analyses using sequence read
+data: quality control, genome assembly, and alignment of short reads
+to a reference sequence.
+
+\parttitle{Results:} %if any
+Lossy compression of quality information can greatly decrease memory
+requirements, and our results demonstrate
+that some compression methods can result in transformed quality values
+that are quite useful, and in some cases advantageous, compared to
+original uncompressed values.
 \end{abstract}
@@
   janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
-  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}. Among the
+  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression,Malysa01102015}. Among the
 several challenges for compression of read data is dealing with
@@
 \bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+%\bibliographystyle{vancouver} % Style BST file (bmc-mathphys, vancouver, spbasic).
 \bibliography{compression}      % Bibliography file (usually '*.bib' )

### 2016-04-12 bmc_manuscript/manuscript.tex
Added natbib to manuscript.
A few little tweaks.

--- 2016-04-12 bmc_manuscript/manuscipt.tex
@@
 %\usepackage{amsthm,amsmath}
-%\RequirePackage{natbib}
+\RequirePackage{natbib}
 %\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
@@
 genome assembly software to produce better
-assemblies~\cite{Bryant:2009uq,Gnerre:2011kx}. Short-read alignment
-programs, such as Bowtie2~\cite{Langmead:2012rw}, use quality values
-to weight mismatches between read and reference sequences. Software
-for detecting single nucleotide polymorphisms (\textsc{snp}s) can use
-quality values~\cite{McKenna:2010bh}, and identified \textsc{snp}s
-with high-quality calls are deemed more reliable than those with
-low-quality calls, particularly in low-coverage regions.
+assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
+alignment programs, such as Bowtie2~\cite{Langmead:2012rw}, use
+quality values to weight mismatches between read and reference
+sequences. Software for detecting single nucleotide polymorphisms
+(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
+and identified \textsc{snp}s with high-quality calls are deemed more
+reliable than those with low-quality calls, particularly in
+low-coverage regions.
 
 Previous literature on sequence data compression has largely focused
-on lossless compression of base calls~\cite{Deorowicz:2013hq,
-  Giancarlo:2009fk,Giancarlo:2014rw, Nalbantoglu:2010uq,Zhu:2013qr},
-although some recent work has addressed compression of quality
-values~\cite{Canovas:2014fr,Hach:2012ys,
+on lossless compression of base calls~\cite[reviewed
+  in][]{Deorowicz:2013hq, Giancarlo:2009fk,Giancarlo:2014rw,
+  Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
+addressed compression of quality
+values~\cite[e.g.,][]{Canovas:2014fr,Hach:2012ys,
   janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
-  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression,Malysa01102015}. Among the
-several challenges for compression of read data is dealing with
-different error profiles resulting from differences in underlying
+  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression,Malysa01102015}.
+Among the several challenges for compression of read data is dealing
+with different error profiles resulting from differences in underlying
 chemistries, signal detection and processing mechanisms, inherent
@@
 lossy compression methods -- binning, modeling, and profiling -- and
-our consideration of an exemplar of each class.~\cite{Canovas:2014fr}
-and~\cite{janin2013adaptive} evaluated the effects of lossy
-compression on identifying variants within a data set. Here we
+our consideration of an exemplar of each class. Some studies have
+evaluated the effects of lossy compression on identifying variants
+within a data set~\cite{Canovas:2014fr,janin2013adaptive}. Here we
 describe our research investigating lossy compression of sequence read
@@
 results of these subsequent compressions that we report. Although
-other algorithms~\cite{HUFFMAN:1952nr,Ziv77auniversal,
+other algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
   DBLP:journals/tit/ZivL78} might be particularly effective for
@@
 \textsc{bz}ip2. For the \textit{Mus musculus} data set, which has read
-lengths of only 36 nt, the file size, even after
-subsequent compression with \textsc{bz}ip2 exceeds that for the
-original \textsc{ascii} quality values.
+lengths of only 36 nt, the file size, even after subsequent
+compression with \textsc{bz}ip2 exceeds that for the original
+\textsc{ascii} quality values.
 
@@
 regression produce an N50 ranging from 3.17--3.22 Mbp (see
-Supplemental Table 3). Despite the similar contiguity statistics, the
+vSupplemental Table 3). Despite the similar contiguity statistics, the
 different compression methods vary markedly in the number of
@@
 the amount of reads that are originally mapped, but unmapped by the
-compressed methods.  \textsc{q}ual\textsc{c}omp with rate 100
-bits/read results in the fewest missing original read alignments
+compressed methods. \textsc{q}ual\textsc{c}omp with rate 100 bits/read
+results in the fewest missing original read alignments
 (159). Increasing the regression model polynomial degree results in a
@@
 transformation associated with compression. Thus appropriate
-compression of quality values can both reduce associated file size cost
-and improve analysis results.
+compression of quality values can both reduce associated file size
+cost and improve analysis results.
 

### 2016-04-13 src/compress.py
Added QVZ to pipeline.

@@ -13,7 +13,6 @@ import random
 import re
 import shlex
 import shutil
-import subprocess
 import sys
 import tempfile
 import time
@@ -104,6 +103,10 @@ def compress(options):
     QUALCOMP_DECOMPRESS_CMD = "./runDecompress.sh -p [DIR] -c [CLUSTERS] -r [RATE]"
 
     RQS_COMPRESS_CMD = "./src/run_rqs.sh [READ] [OUTPUT]"
+
+    #qvz -c 3 -r .10 -v test_results4/original/frag_1.fastq.quals tmp/test_1_c3_r.10
+    QVZ_COMPRESS_CMD = "[QVZ]/qvz -c [CLUSTERS] -r [RATE] -v [READ] [OUTPUT]"
+    QVZ_DECOMPRESS_CMD = "[QVZ]/qvz -x -v [INPUT] [OUTPUT]"
 
     # Store which compression directories we created.
     options.compressed_dirs = []
@@ -241,6 +244,48 @@ def compress(options):
                 out_cmd(bin_file.name, std_err_file.name, call_arr)
                 call(call_arr, stdout=bin_file, stderr=std_err_file)
 
+        # Compress using QualComp.
+        if options.qvz_rates:
+            for rate in options.qvz_rates.split(','):
+                ensure_dir(options.output_dir + '/qvz_r' + rate + '/')
+
+                if 'qvz_r' + rate not in options.compressed_dirs:
+                    options.compressed_dirs.append('qvz_r' + rate)
+
+                """
+                QVZ_COMPRESS_CMD = "[QVZ]/qvz -c [CLUSTERS] -r [RATE] -v [READ] [OUTPUT]"
+                QVZ_DECOMPRESS_CMD = "[QVZ]/qvz -x -v [INPUT] [OUTPUT]"
+                """
+
+                # We first need to get the quality values only. TODO(cmhill): Very hacky.
+                with open(reads_filename) as fin, open(options.output_dir + '/qvz_r' + rate + '/orig_' +  os.path.basename(reads_filename) + '.quals', 'w') as fout:
+                    fout.writelines(islice(fin, 3, None, 4))
+
+                call_arr = QVZ_COMPRESS_CMD.replace('[READ]', options.output_dir + '/qvz_r' + rate + '/orig_' +  os.path.basename(reads_filename) + '.quals')\
+                        .replace('[QVZ]', os.environ["QVZ"])\
+                        .replace('[CLUSTERS]', options.clusters)\
[…]

### 2016-04-18 bmc_manuscript/manuscript.tex
Fixed a reference, revised some text, and added a note (in red) where more text is needed.

@@
 Lossy compression of quality information can greatly decrease memory
-requirements, and our results demonstrate
-that some compression methods can result in transformed quality values
-that are quite useful, and in some cases advantageous, compared to
-original uncompressed values.
+requirements, and our results demonstrate that some compression
+methods can result in transformed quality values that are quite
+useful, and in some cases advantageous, compared to original
+uncompressed values.
 \end{abstract}
@@
 
-\cite{Wan:2012kq} provide three similar lossy compression strategies
-based on binning the base error probability distribution: UniBinning,
-Truncating, and LogBinning. UniBinning evenly splits the error
-probability distribution into a user-defined number of partitions.
-Truncating treats a user-defined number of highest quality values as a
-single bin. LogBinning works similarly to UniBinning, except it uses
-the \emph{log} of the error probability distribution, which
-effectively bins the \textsc{ascii} quality values evenly. The 2-bin
-encoding examined here can be viewed as something like a combination
-of LogBinning and Truncating in that we place the highest quality
-values (as defined above) from the distribution of log error
-probability values into a single bin.
+Previous work by others \cite{Wan:2012kq} describe three similar lossy
+compression strategies based on binning the base error probability
+distribution: UniBinning, Truncating, and LogBinning. UniBinning
+evenly splits the error probability distribution into a user-defined
+number of partitions.  Truncating treats a user-defined number of
+highest quality values as a single bin. LogBinning works similarly to
+UniBinning, except it uses the \emph{log} of the error probability
+distribution, which effectively bins the \textsc{ascii} quality values
+evenly. The 2-bin encoding examined here can be viewed as something
+like a combination of LogBinning and Truncating in that we place the
+highest quality values (as defined above) from the distribution of log
+error probability values into a single bin.
 
@@
 
-For comparison to other developed methods we use \textsc{QualComp}, a
-lossy compression tool~\cite{Ochoa:2013rt}. The program models quality
-values as a multivariate Gaussian distribution, computing the mean and
-covariance for each position in the read, and these values are stored
-by the decoder to later reconstruct a \emph{representative} quality
-value. \textsc{QualComp} takes as input a user-specified rate
-(bits/read), and these bits are apportioned among positions within the
-read to minimize the average error. The quality values can be
-clustered beforehand to produce more accurate models.
+For comparison to other developed methods for lossy compression of
+quality values we analyzed the reference data with
+\textsc{QualComp}~\cite{Ochoa:2013rt},
+\textsc{rsa}~\cite{DBLP:conf/recomb/YuYB14}, and
+\textsc{qvz}~\cite{Malysa01102015}.
+
+\textsc{QualComp} models quality values as a multivariate Gaussian
+distribution, computing the mean and covariance for each position in
+the read, and these values are stored by the decoder to later
+reconstruct a \emph{representative} quality value. \textsc{QualComp}
+takes as input a user-specified rate (bits/read), and these bits are
+apportioned among positions within the read to minimize the average
+error. The quality values can be clustered beforehand to produce more
+accurate models.
 
@@
 the model parameters.
+
+\textcolor{red}{Add brief description of \textsc{rsa} and
+  \textsc{qvz}, including economic costs.}
 

### 2016-04-25 src/compress.py
Added SNP pipeline.

@@ -287,7 +287,6 @@ def compress(options):
                 call(call_arr, stdout=decompressed_file, stderr=std_err_file)
 
         # Compress with RQS.
-        #for degree in options.poly_degrees.split(','):
         ensure_dir(options.output_dir + '/rqs/')
 
         if 'rqs' not in options.compressed_dirs:
@@ -523,7 +522,7 @@ def align_reads(options):
     call(call_arr, stdout=FNULL, stderr=std_err_file)
 
     # Align the reads.
-    BOWTIE2_CMD = "bowtie2 -x " + options.output_dir + "/align/reference -p " + options.threads + " --reorder -U [READ] --al [ALIGNED]"
+    BOWTIE2_CMD = "bowtie2 -x " + options.output_dir + "/align/reference -p " + options.threads + " --reorder -U [READ] --al [ALIGNED] -S [SAM]"
 
     # Have to do max/min alignment
 
@@ -535,13 +534,56 @@ def align_reads(options):
             alignment_file = open(alignment_filename, 'w')
 
             call_arr = BOWTIE2_CMD.replace('[READ]', options.output_dir + '/' + compression_method + '/' + os.path.basename(reads_filename))\
-                    .replace('[ALIGNED]', options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.aligned').split()
+                    .replace('[ALIGNED]', options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.aligned')\
+                    .replace('[SAM]', options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.sam').split()
             out_cmd(FNULL.name, alignment_filename, call_arr)
             call(call_arr, stdout=FNULL, stderr=alignment_file)
 
             # Print out the headers of the aligned sequences.
             with open(options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.aligned', 'r') as aligned_file, open(options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.headers', 'w') as headers:
                 headers.writelines(islice(aligned_file, 0, None, 4))
+
+def call_snps(options):
+    """
+    Call SNPs using decompressed reads.
+    """
+
+    std_err_file = open('snp.log', 'a')
+
+    ensure_dir(options.output_dir + "/snp/")
[…]

### 2016-04-25 src/compress.py
Fixed typo.

@@ -920,8 +920,8 @@ def get_options():
     #parser.add_option("--clusters", dest="clusters", help="QualComp parameter for setting number of clusters.", default="3")
 
     # QVZ-specific compression options.
-    parser.add_option("--qvz-rates", dest="qvz_rates", help="QualComp parameter for setting the bits/reads.", default=".30")
-    parser.add_option("--qvz-clusters", dest="qvz_clusters", help="QualComp parameter for setting number of clusters.", default="3")
+    parser.add_option("--qvz-rates", dest="qvz_rates", help="QVZ parameter for setting the bits/reads.", default=".30")
+    parser.add_option("--qvz-clusters", dest="qvz_clusters", help="QVZ parameter for setting number of clusters.", default="3")
 
     # Max, min quality value compression options.
     parser.add_option("--max-qv", dest="max_quality", help="Use this value for max quality value compression.", default="40")

### 2016-04-25 bmc_manuscript/manuscript.tex

@@
    email={chrismh@uw.edu}   % email address
-]{\inits{CMH}\fnm{Christopher M} \snm{Hill}}
+]{\inits{CMH}\fnm{Christopher M.} \snm{Hill}}
 \author[
@@
    email={mike@umiac.umd.edu}
-]{\inits{MPC}\fnm{Michael P} \snm{Cummings}}
+]{\inits{MPC}\fnm{Michael P.} \snm{Cummings}}
 
@@
 \textsc{QualComp}~\cite{Ochoa:2013rt},
-\textsc{rsa}~\cite{DBLP:conf/recomb/YuYB14}, and
+\textsc{rqs}~\cite{DBLP:conf/recomb/YuYB14}, and
 \textsc{qvz}~\cite{Malysa01102015}.
@@
 
-\textcolor{red}{Add brief description of \textsc{rsa} and
+\textcolor{red}{Add brief description of \textsc{rqs} and
   \textsc{qvz}, including economic costs.}
+
+\textsc{qvz} extends the idea of using rate-distortion theory introduced in \textsc{QualComp}, allowing additional distortion metrics besides \textsc{mse} to be minimized with respect to user-specified rate (bits/read).
+
+\textsc{rqs} first constructs a dictionary of commonly occurring $k$-mers from either the sequencing data set they seek to compress or a larger population-sized sequencing data set.
+During compression, $k$-mers within reads that have a small Hamming distance from a dictionary $k$-mer are identified.
+These positions may correspond to SNPs or sequencing error.
+Positions within the $k$-mer that differ from the nearest $k$-mer in the dictionary are marked.
+Lastly, the quality values of the remaining positions are smoothed.
 

### 2016-04-26 src/compress.py
Fixed missing results due to unsorted sam in  SNP pipeline.

@@ -551,6 +551,7 @@ def call_snps(options):
     std_err_file = open('snp.log', 'a')
 
     ensure_dir(options.output_dir + "/snp/")
+    SAMTOOLS_SORT_CMD = "samtools sort -T ./ -o [SORTED_SAM] -O sam [SAM]"
     PILEUP_CMD = "samtools mpileup -ugf [REFERENCE] [SAM]"
     SNP_CMD = "bcftools call -vV indels -mO v -o [OUTPUT_VCF] [PILEUP]"
 
@@ -559,10 +560,15 @@ def call_snps(options):
 
             pileup_filename = options.output_dir + '/snp/' + compression_method + '/' + os.path.basename(reads_filename) + '.pileup'
             ensure_dir(pileup_filename)
-            pileup_file = open(pileup_filename, 'w')
 
+            call_arr = SAMTOOLS_SORT_CMD.replace('[SAM]', options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.sam')\
+                    .replace('[SORTED_SAM]', options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.sorted.sam').split()
+            out_cmd(std_err_file.name, std_err_file.name, call_arr)
+            call(call_arr, stdout=std_err_file, stderr=std_err_file)
+
+            pileup_file = open(pileup_filename, 'w')
             call_arr = PILEUP_CMD.replace('[REFERENCE]', options.reference_fasta)\
-                    .replace('[SAM]', options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.sam').split()
+                    .replace('[SAM]', options.output_dir + '/align/' + compression_method + '/' + os.path.basename(reads_filename) + '.sorted.sam').split()
             out_cmd(pileup_file.name, std_err_file.name, call_arr)
             call(call_arr, stdout=pileup_file, stderr=std_err_file)
 

### 2016-04-26 bmc_manuscript/manuscript.tex
Added text describing the economic costs for QVZ and RQS. Other small revisions. Need to confirm if RQS smooths or discards quality values for positions in k-mers outside the chosen Hamming distance of dictionary k-mers.

@@
 quality values we analyzed the reference data with
-\textsc{QualComp}~\cite{Ochoa:2013rt},
-\textsc{rqs}~\cite{DBLP:conf/recomb/YuYB14}, and
-\textsc{qvz}~\cite{Malysa01102015}.
+\textsc{QualComp}~\cite{Ochoa:2013rt}, Read Quality Sparsifier
+(\textsc{rqs})~\cite{DBLP:conf/recomb/YuYB14}, and Quality Values Zip
+(\textsc{qvz)}~\cite{Malysa01102015}.
 
@@
 
-The economic costs of file size for compression with
-\textsc{QualComp}, in our interpretation, appear to have fixed costs
+\textsc{qvz} extends the idea of using rate-distortion theory
+introduced in \textsc{QualComp}, allowing distortion metrics other
+than \textsc{mse} to be minimized with respect to user-specified rate
+(bits/read).
+
+The economic costs of file size for compression with \textsc{QualComp}
+or \textsc{qvz}, in our interpretation, appear to be the fixed costs
 for storing the cluster representatives, and the marginal costs that
@@
 
-\textcolor{red}{Add brief description of \textsc{rqs} and
-  \textsc{qvz}, including economic costs.}
-
-\textsc{qvz} extends the idea of using rate-distortion theory introduced in \textsc{QualComp}, allowing additional distortion metrics besides \textsc{mse} to be minimized with respect to user-specified rate (bits/read).
-
-\textsc{rqs} first constructs a dictionary of commonly occurring $k$-mers from either the sequencing data set they seek to compress or a larger population-sized sequencing data set.
-During compression, $k$-mers within reads that have a small Hamming distance from a dictionary $k$-mer are identified.
-These positions may correspond to SNPs or sequencing error.
-Positions within the $k$-mer that differ from the nearest $k$-mer in the dictionary are marked.
-Lastly, the quality values of the remaining positions are smoothed.
+\textsc{rqs} first constructs a dictionary of commonly occurring
+$k$-mers from either the sequencing data set they seek to compress or
+a larger population-sized sequencing data set. During compression,
+$k$-mers within reads that have a small Hamming distance from a
+dictionary $k$-mer are identified. The explicit assumption is that
+positions differing from the $k$-mers in the dictionary more likely
+correspond to SNPs or sequencing errors. These positions that differ
+from the nearest $k$-mer in the dictionary are marked. Lastly, the
+quality values of the remaining positions are smoothed (discarded? -
+see paper).
+
+The economic costs of file size for compression with \textsc{rqs}, in
+our interpretation, appear to be the fixed costs of the dictionary of
+commonly occurring $k$-mers, and the marginal costs that are the
+product of the number of positions of $k$-mers within the chosen
+Hamming distance from dictionary $k$-mers times the cost of
+representing these positions. It should be noted that the fixed costs
+of the dictionary of commonly occurring $k$-mers can be amortized over
+reads for which the dictionary may be appropriately used (e.g., reads
+from the same species).
 
@@
 increased secondary compression by reordering (e.g., sorting)
-compressed representations (e.g., sets of bin designations, model, or
+compressed representations (e.g., sets of bin designations, models, or
 profiles), but these operations have also not been explored here.

### 2016-04-26 bmc_manuscript/manuscript.tex
A bit more polishing of the text for economics of RQS compression.

@@
 Hamming distance from dictionary $k$-mers times the cost of
-representing these positions. It should be noted that the fixed costs
-of the dictionary of commonly occurring $k$-mers can be amortized over
-reads for which the dictionary may be appropriately used (e.g., reads
-from the same species).
+representing these positions. As with compression based on profiles,
+amortization applies to fixed costs of compression with
+\textsc{rqs}. Here the fixed costs of the dictionary of commonly
+occurring $k$-mers can be amortized over reads for which the
+dictionary may be appropriately used (e.g., reads from the same
+species).
 

### 2016-05-05 src/evaluate_loss.py
Added L1 and Lorentzian metrics (Need to move to a separate R file).

@@ -1,3 +1,4 @@
+import math
 import numpy as np
 from collections import Counter
 import sys
@@ -13,12 +14,23 @@ def evaluate_loss(qstring1, qstring2):
     q1, q2 = np.array(qual_ord(qstring1)), np.array(qual_ord(qstring2))
     diff = q1 - q2
     mse = (diff ** 2).mean()
-    return mse, Counter(q1 - q2)
+
+    # L1
+    l1 = np.linalg.norm((q1 - q2), ord=1)
+
+    # Lorentzian
+    lorentzian = math.log(1 + l1, 2)
+
+    return mse, l1, lorentzian
+
 
 file1 = sys.argv[1]
 file2 = sys.argv[2]
 
 mses = []
+L1s = []
+lorentzians = []
+
 diff_histogram = Counter([])
 
 IS_FASTQ = False
@@ -35,12 +47,17 @@ with open(file1, 'r') as in1:
             j += 1
             if (j % 50) != 0:  # only sample every 50th read. Otherwise quite slow
                 continue
-            mse, diffcounts = evaluate_loss(line1.rstrip(), line2.rstrip())
+
+            mse, L1, lorentzian = evaluate_loss(line1.rstrip(), line2.rstrip())
             mses.append(mse)
-            diff_histogram += diffcounts
+            L1s.append(L1)
+            lorentzians.append(lorentzian)
+            #diff_histogram += diffcounts
 
 mses = np.array(mses)
-print mses.mean(),'\t',mses.std()
+L1s = np.array(L1s)
+lorentzians = np.array(lorentzians)
+
+print mses.mean(),'\t',mses.std(),'\t',L1s.mean(),'\t',L1s.std(),'\t',lorentzians.mean(),'\t',lorentzians.std()
 #hist = np.histogram(mses, bins=50)
 #print 'error histogram', diff_histogram
-

### 2016-05-05 bmc_manuscript/manuscript.tex

@@
 no fixed costs, and marginal costs that are the product of the number
-of base-call quality values times the cost of the encoding.
+of base-call quality values multiplied by the cost of the encoding.
 
@@
 general terms, include no fixed costs, and marginal costs that are the
-product of the number of reads times the cost of representing the
-model parameters.
+product of the number of reads multiplied by the cost of representing
+the model parameters.
 
@@
 general terms, include fixed costs associated with representing the
-profiles, which is the product of the number of profiles times the
-cost of encoding them. These fixed costs are amortized over the entire
-set of reads to which they apply, and thus on a read basis vary as a
-reciprocal function of the number of reads. Additionally there are
-marginal costs that are the product of the number of reads encoded
-times the cost of the encoding. Profiling like modeling, but not
-binning, has the advantageous property that marginal costs are the
+profiles, which is the product of the number of profiles multiplied by
+the cost of encoding them. These fixed costs are amortized over the
+entire set of reads to which they apply, and thus on a read basis vary
+as a reciprocal function of the number of reads. Additionally there
+are marginal costs that are the product of the number of reads encoded
+multiplied by the cost of the encoding. Profiling like modeling, but
+not binning, has the advantageous property that marginal costs are the
 reciprocal of the read length. Thus the costs of file size for
@@
 quality values we analyzed the reference data with
-\textsc{QualComp}~\cite{Ochoa:2013rt}, Read Quality Sparsifier
-(\textsc{rqs})~\cite{DBLP:conf/recomb/YuYB14}, and Quality Values Zip
-(\textsc{qvz)}~\cite{Malysa01102015}.
+\textsc{QualComp}~\cite{Ochoa:2013rt}, and Quality Values Zip
+(\textsc{qvz)}~\cite{Malysa01102015}, and Read Quality Sparsifier
+(\textsc{rqs})~\cite{DBLP:conf/recomb/YuYB14}
 
@@
 for storing the cluster representatives, and the marginal costs that
-are the product of the number of reads times the cost of representing
-the model parameters.
-
-\textsc{rqs} first constructs a dictionary of commonly occurring
-$k$-mers from either the sequencing data set they seek to compress or
-a larger population-sized sequencing data set. During compression,
-$k$-mers within reads that have a small Hamming distance from a
-dictionary $k$-mer are identified. The explicit assumption is that
-positions differing from the $k$-mers in the dictionary more likely
-correspond to SNPs or sequencing errors. These positions that differ
-from the nearest $k$-mer in the dictionary are marked. Lastly, the
-quality values of the remaining positions are smoothed (discarded? -
-see paper).
+are the product of the number of reads multiplied by the cost of
+representing the model parameters.
+
+Compression with \textsc{rqs} first begins with constructing a
+dictionary of commonly occurring $k$-mers from either the sequencing
+data set to be compressed or a larger population-sized sequencing
+data set. During the second phase, sparsification, $k$-mers within
+reads that have a small Hamming distance from a dictionary $k$-mer are
+identified. Base-calls within these $k$-mers are deemed high
+confidence and the associated quality values discarded, as are quality
+values above a chosen threshold for base-calls in other $k$-mers. The
+explicit assumption is that positions differing from the $k$-mers in
+the dictionary more likely correspond to SNPs or sequencing errors. At
+a subsequent step the discarded quality values are replaced with the
+threshold value, in a manner somewhat similar to the binning strategy
+described previously.
 
@@
 commonly occurring $k$-mers, and the marginal costs that are the
-product of the number of positions of $k$-mers within the chosen
-Hamming distance from dictionary $k$-mers times the cost of
-representing these positions. As with compression based on profiles,
-amortization applies to fixed costs of compression with
-\textsc{rqs}. Here the fixed costs of the dictionary of commonly
-occurring $k$-mers can be amortized over reads for which the
-dictionary may be appropriately used (e.g., reads from the same
-species).
+product of the number of base-calls represented as high-quality or
+other value multiplied by the cost of representing these values. As
+with compression based on profiles, amortization applies to fixed
+costs of compression with \textsc{rqs}. Here the fixed costs of the
+dictionary of commonly occurring $k$-mers can be amortized over reads
+for which the dictionary may be appropriately used (e.g., reads from
+the same species).
 

### 2016-05-05 src/compress.py
Fixed SNP calling bug.

@@ -584,12 +584,18 @@ def call_snps(options):
                 while line.startswith('#'):
                     line = vcf_file.readline()
 
+                snps = []
                 while line:
                     tuple = line.split('\t')
                     if len(tuple)>1:
-                        snps_file.write(str(tuple[1] + '\n'))
+                        snps.append(str(tuple[0] + '\t' + tuple[1] + '\n'))
+                        #snps_file.write(str(tuple[0] + '\t' + tuple[1] + '\n'))
 
                     line = vcf_file.readline()
+
+                snps.sort()
+                for snp in snps:
+                    snps_file.write(snp)
 
 
 def post_process_results(options):

### 2016-05-05 src/compress.py
Added L1 and Lorentzian metrics to results table.

@@ -636,7 +636,7 @@ def process_compression_stats(options):
 
 
         results_file = open(options.output_dir + "/results/" + os.path.basename(reads_filename) + '.compression', 'w')
-        results_file.write("compression\tbases\torig_size\torig_bzip2\tcomp_size\tcomp_bzip2\tmse\tbits_bp\n")
+        results_file.write("compression\tbases\torig_size\torig_bzip2\tcomp_size\tcomp_bzip2\tmse\tL1\tlorentzian\tbits_bp\n")
         for compression_method in options.compressed_dirs:
 
             filename = options.output_dir + '/' + compression_method + '/' + os.path.basename(reads_filename)
@@ -663,7 +663,8 @@ def process_compression_stats(options):
                 results += "NA\tNA\t"
 
             # Get the MSE.
-            results += str(grab_value_from_file(filename + '.mse')) + '\t'
+            distortions = grab_list_of_values_from_file(filename + '.mse')
+            results += distortions[0] + '\t' + distortions[2] + '\t' + distortions[4] + '\t'
 
             # Print the bits/bp.
             results += str((compressed_size * 8) / float(bases)) + '\n'
@@ -972,6 +973,14 @@ def grab_value_from_file(filename):
     Return the value from the first line of a file.
     """
     return open(filename, 'r').readline().strip().split()[0]
+
+
+def grab_list_of_values_from_file(filename):
+    """
+    Return the value from the first line of a file.
+    """
+    return open(filename, 'r').readline().strip().split()
+
 
 def num_lines(filename):
     """

### 2016-05-06 bmc_manuscript/manuscript.tex
added definitions of additional loss functions

@@
 of the compressed representation of quality values (in bits) divided
-by the number of quality values represented. As a measure of
-information loss we use mean squared error (\textsc{mse}) as a loss
-function, and define it as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$,
-where $n$ is the number of quality values, $Q_i'$ is the
+by the number of quality values represented. As measures of
+information loss we use three loss functions: mean squared error
+(\textsc{mse}), defined as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$;
+mean absolute deviation (\textsc{mad}), defined as
+$\frac{1}{n}\sum_{i=1}^{n}{|Q_i'-Q_i|}$; and mean Lorentz distortion,
+defined as $\frac{1}{n}\sum_{i=1}^{n}{log_{2}(1 + |Q_i'-Q_i|)}$, where
+$n$ is the number of quality values, $Q_i'$ is the
 compressed/decompressed quality value, and $Q_i$ is the original

### 2016-05-26 src/compress.py
Added ability to change max quality value.

@@ -94,9 +94,9 @@ def compress(options):
 
     # Basic command line scripts to run the individual compression schemes.
     GB_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -c 2 -b 0 -i [COMPRESSED_FILE]"
-    MAX_VALUE_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -g 40 -b 40 -c 2 -i [COMPRESSED_FILE]"
+    MAX_VALUE_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -g [MAX_QV] -b 40 -c 2 -i [COMPRESSED_FILE]"
     MIN_VALUE_COMPRESSION_CMD = "./src/good_bad_coding.py -r [READ] -g 0 -b 0 -c 2 -i [COMPRESSED_FILE]"
-    POLY_REGRESSION_CMD = "Rscript src/poly_regression_parallel.R [READ] [OUTPUT] [DEGREE] [COMPRESSED_FILE] [NUM_THREADS]"
+    POLY_REGRESSION_CMD = "Rscript src/poly_regression_parallel.R [READ] [OUTPUT] [DEGREE] [COMPRESSED_FILE] [NUM_THREADS] [MAX_QV]"
     PROFILE_COMPRESSION_CMD = "Rscript src/profile_parallel.R [READ] [OUTPUT] [TRAINING_SIZE] [NUM_PROFILES] [COMPRESSED_FILE] [NUM_THREADS]"
 
     QUALCOMP_COMPRESS_CMD = "./runCompress.sh -i [READ] -c [CLUSTERS] -r [RATE]"
@@ -131,7 +131,8 @@ def compress(options):
 
         # Max/min quality value compression. We can use good_bad.py script to do this.
         call_arr = MAX_VALUE_COMPRESSION_CMD.replace('[READ]', reads_filename)\
-                .replace('[COMPRESSED_FILE]', options.output_dir + '/maxqual/' + os.path.basename(reads_filename) + '.comp').split()
+                .replace('[COMPRESSED_FILE]', options.output_dir + '/maxqual/' + os.path.basename(reads_filename) + '.comp')\
+                .replace('[MAX_QV]', options.max_quality).split()
         output_fp = open(options.output_dir + '/maxqual/' + os.path.basename(reads_filename), 'w')
 
[…]

### 2016-05-26 src/parse_preprocessing_all.R

@@ -8,7 +8,7 @@ library(scales)
 tables = c("~/cbcb2/q-compression/results/rhodo3/results/frag_1.fastq.preprocessing",
            #"~/cbcb2/q-compression/results/staph2/results/frag_1.fastq.preprocessing",
            "~/cbcb2/q-compression/results/ecoli2/results/MiSeq_Ecoli_MG1655_110721_PF_R1.fastq.preprocessing",
-           "~/cbcb2/q-compression/results/hg14/results/frag_1.fastq.preprocessing",
+           "~/cbcb2/q-compression/results2/hg14_results/frag_1.fastq.preprocessing",
            "~/cbcb2/q-compression/results/mouse/results//SRR032209.fastq.preprocessing")
 
 dataset_names = c("Rhodobacter sphaeroides",
@@ -25,7 +25,7 @@ for (table in tables) {
   table <- read.table(table, header = T, row.names = 1)
   original <- table[rownames(table) %in% "original", ]$bases_kept
   table <-  table[!rownames(table) %in% c("original","minqual","maxqual"), ]
-  table[,5] <- table[,5] - original
+  table[,5] <- table[,5] / original
   
   dfr <- data.frame(y=seq(1:length(table[,1])), x = table[,5], dataset = dataset_names[[counter]])
   dfr_tables <- rbind(dfr_tables, dfr)
@@ -40,12 +40,14 @@ rowLabels = c("2-bin", "regression (0)", "regression (1)", "regression (3)", "re
 library(grid)
 
 
-pdf("preprocessing_results.pdf", width=3.5, height=3.5)
+setEPS()
+postscript("preprocessing_results.eps",width=3.5, height=3.5)
+#pdf("preprocessing_results.pdf", width=3.5, height=3.5)
 p <- ggplot(dfr_tables, aes(x=jitter(x,factor = 1), y=y, color = dataset)) + geom_point() + scale_y_discrete(labels=rowLabels) + 
   theme_bw() +
   xlab("Proportion of bases kept compared to original") +
   ylab("") + 
-  theme_bw() + theme(legend.position="bottom", legend.key = element_blank(), legend.title=element_blank(),
+  theme_minimal() + theme(legend.position="bottom", legend.key = element_blank(), legend.title=element_blank(),
                      legend.text = element_text(face="italic", size=6),
                      axis.text.y = element_text(size=6),
                      axis.title.x = element_text(size=6),

### 2016-05-26 src/poly_regression_parallel.R

@@ -16,8 +16,9 @@ degree <- args[3]
 
 # Set the min-max quality values.
 min_max <- function(x) {
-  if (x > 73)
-    return(73)
+  max_value <- 33 + strtoi(args[6])
+  if (x > max_value)
+    return(max_value)
   else if (x < 33)
     return(33)
   else
@@ -32,7 +33,7 @@ poly_regression <- function(quals) {
   if (strtoi(degree) > 0) {
   # Fit the polynomial function.
     fit = lm(quals ~ poly(x, strtoi(degree)))
-  
+
     # Predict the quality values using the above equation.
     predicted_quals = rawToChar(as.raw(unlist(lapply(round(predict(fit)), min_max))))
     return(predicted_quals)
@@ -46,11 +47,11 @@ poly_regression_coef <- function(quals) {
   # Get the quality values. Use a hardcoded quality offset.
   quals = quals + 33
   x = seq(1,length(quals))
-  
+
   if (strtoi(degree) > 0) {
     # Fit the polynomial function.
     fit = lm(quals ~ poly(x, strtoi(degree)))
-    
+
     # Return the coefficients.
     return(as.vector(fit$coefficients))
   } else {
@@ -91,18 +92,18 @@ while (length(oneLine <- readLines(con, n = 1, warn = FALSE)) > 0) {
   # Write the header...
   writeLines(oneLine, con=output_file)
   oneLine <- readLines(con, n = 1, warn = FALSE)
-  
+
   # ... then the DNA sequence ...
   writeLines(oneLine, con=output_file)
   oneLine <- readLines(con, n = 1, warn = FALSE)
-  
+
   # ... then the + ...
   writeLines(oneLine, con=output_file)
   oneLine <- readLines(con, n = 1, warn = FALSE)
-  
+
   # ... and finally the new quality values.
   writeLines(results[[counter]], con=output_file)
   counter <- counter + 1
-} 
+}
 
 close(con)

### 2016-05-26 bmc_manuscript/manuscript.tex
BibTex: added original reference for quality values; fixed some formatting; added publisher location manuscript: cite original reference for quality values and edited related text; better describe and name third loss function based on my understanding

--- 2016-05-05 bmc_manuscript/manuscript.tex
@@
 
-Quality values comprise a standard component of \textsc{fastq}
-files~\cite{Cock:2010ve}, a very common format for sequence read
-data. At the level of the sequence read, the probability of error for
-each base-call is typically represented by a \textsc{phred} quality
-value, which is defined as $Q =
--10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
+Advantageous use of quality values has been long
+recognize~\cite{Bonfield:1995kq}, and quality values can be used
+throughout bioinformatics pipelines.  Quality values comprise a
+standard component of \textsc{fastq} files~\cite{Cock:2010ve}, a very
+common format for sequence read data. At the level of the sequence
+read, the probability of error for each base-call is typically
+represented by a \textsc{phred} quality value, which is defined as $Q
+= -10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
 technology, these quality values can range from 0 to 93, and are
@@
 
-Quality values can be used throughout bioinformatics pipelines. Among
-the most fundamental uses of sequence quality values is as part of the
-quality assessment and quality control (\textsc{qa/qc}) processes
-prior to subsequent analysis steps. Quality control based on quality
-values generally includes two operations: \textit{i}.~filtering, which
-removes reads that on the whole do not meet arbitrary quality
-standards, thus reducing the total number of reads; and
-\textit{ii}.~trimming of low-quality base-calls from reads, which
-reduces the length of reads trimmed. Quality values can be used by
-genome assembly software to produce better
+Among the most fundamental uses of sequence quality values is as part
+of the quality assessment and quality control (\textsc{qa/qc})
+processes prior to subsequent analysis steps. Quality control based on
+quality values generally includes two operations:
+\textit{i}.~filtering, which removes reads that on the whole do not
+meet arbitrary quality standards, thus reducing the total number of
+reads; and \textit{ii}.~trimming of low-quality base-calls from reads,
+which reduces the length of reads trimmed. Quality values can be used
+by genome assembly software to produce better
 assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
@@
 of the compressed representation of quality values (in bits) divided
-by the number of quality values represented. As a measure of
-information loss we use mean squared error (\textsc{mse}) as a loss
-function, and define it as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$,
-where $n$ is the number of quality values, $Q_i'$ is the
-compressed/decompressed quality value, and $Q_i$ is the original
-quality value associated with position $i$.
+by the number of quality values represented. As measures of
+information loss we use three loss functions following previous
+work~\cite{Malysa01102015}: mean squared error (\textsc{mse}), defined
+as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$; mean absolute deviation
+(\textsc{mad}), defined as $\frac{1}{n}\sum_{i=1}^{n}{|Q_i'-Q_i|}$;
+and mean of the values derived from the Cauchy--Lorentz distribution
+(mean Cauchy--Lorentz loss), defined as
+$\frac{1}{n}\sum_{i=1}^{n}{log_{2}(1 + |Q_i'-Q_i|)}$, where $n$ is the
+number of quality values, $Q_i'$ is the compressed/decompressed
+quality value, and $Q_i$ is the original quality value associated with
+position $i$.
 

### 2016-05-26 bmc_manuscript/manuscript.tex
small tweak

@@
 (\textsc{mad}), defined as $\frac{1}{n}\sum_{i=1}^{n}{|Q_i'-Q_i|}$;
-and mean of the values derived from the Cauchy--Lorentz distribution
-(mean Cauchy--Lorentz loss), defined as
+and mean of the values derived from part of the formula defining the
+Cauchy--Lorentz distribution (mean Cauchy--Lorentz loss), defined as
 $\frac{1}{n}\sum_{i=1}^{n}{log_{2}(1 + |Q_i'-Q_i|)}$, where $n$ is the

### 2016-10-03 peerj_manuscript/manuscript.tex
Added peerj format.

--- 2016-05-26 bmc_manuscript/manuscript.tex
@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
+%% Submissions for peer-review must enable line-numbering
+%% using the lineno option in the \documentclass command.
+%%
+%% Preprints and camera-ready submissions do not need
+%% line numbers, and should have this option removed.
+%%
+%% Please note that the line numbering option requires
+%% version 1.1 or newer of the wlpeerj.cls file, and
+%% the corresponding author info requires v1.2
 
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
+\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
+% \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                     %%
-%%  LaTeX template for BioMed Central  %%
-%%     journal article submissions     %%
-%%                                     %%
-%%          <8 June 2012>              %%
-%%                                     %%
-%%                                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% For instructions on how to fill out this Tex template           %%
-%% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%
-%% http://www.biomedcentral.com/info/authors/                      %%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%% BioMed Central currently use the MikTex distribution of         %%
-%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
-%% http://www.miktex.org                                           %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
-
-%%% loading packages, author definitions
-
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmcart}
-
-%%% Load packages
-%\usepackage{amsthm,amsmath}
-\RequirePackage{natbib}
-%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-%\RequirePackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                             %%
-%%  If you wish to display your graphics for   %%
-%%  your own use using includegraphic or       %%
-%%  includegraphics, then comment out the      %%
-%%  following two lines of code.               %%
-%%  NB: These line *must* be included when     %%
-%%  submitting to BMC.                         %%
-%%  All figure files must be submitted as      %%
-%%  separate graphics through the BMC          %%
-%%  submission process, not included in the    %%
-%%  submitted article.                         %%
-%%                                             %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\usepackage{hyperref}
 \def\includegraphic{}
 \def\includegraphics{}
-
-%%% Put your definitions there:
-\startlocaldefs
-\endlocaldefs
-
-\usepackage{hyperref}
-
-%%% Begin ...
-\begin{document}
-
-%%% Start of article front matter
-\begin{frontmatter}
-
-\begin{fmbox}
-\dochead{Research}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\author[1,2]{Christopher M. Hill}
+\author[3]{Andr\'{a}s Szolek}
+\author[4]{Mohamed El Hadidi}
+\author[2]{Michael P. Cummings}
+\affil[1]{Department of Computer Science, University of Maryland, College Park, Maryland, 20742 USA}
+\affil[2]{Center for Bioinformatics and Computational Biology, University of Maryland, College Park, Maryland, 20742 USA}
+\affil[3]{Department of Applied Bioinformatics, Center for Bioinformatics, Quantitative Biology Center, University of T\"{u}bingen, Sand 14, 72076, T\"{u}bingen, Germany}
+\affil[4]{Department of Algorithms in Bioinformatics, Center for Bioinformatics, Quantitative Biology Center, University of T\"{u}bingen, Sand 14, 72076, T\"{u}bingen, Germany}
+\corrauthor[4]{Michael P. Cummings}{mike@umiac.umd.edu}
 
-\author[
-   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
-%   noteref={n1},                        % id's of article notes, if any
-   email={chrismh@uw.edu}   % email address
-]{\inits{CMH}\fnm{Christopher M.} \snm{Hill}}
-\author[
-   addressref={aff4,aff6},
-   email={szolek@informatik.uni-tuebingen.de}
-]{\inits{AS}\fnm{Andr\'{a}s} \snm{Szolek}}
-\author[
-   addressref={aff5,aff6},
-   email={mohamed.el-hadidi@uni-tuebingen.de}
-]{\inits{MEH}\fnm{Mohamed} \snm{El Hadidi}}
-\author[
-   addressref={aff3},
-   corref={aff2},                       % id of corresponding address, if any
-   email={mike@umiac.umd.edu}
-]{\inits{MPC}\fnm{Michael P.} \snm{Cummings}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\address[id=aff1]{%                           % unique id
-  \orgname{Department of Computer Science, University of Maryland}, % university, etc
-  \city{College Park, Maryland},                              % city
-  \postcode{20742}                                % post or zip code
-  \cny{USA}                                    % country
-}
-\address[id=aff2]{%                           % unique id
-  \orgname{Department of Genome Science, University of Washington School of Medicine},                    %
-  \city{Seattle, Washington},                              % city
-  \postcode{98195},                                % post or zip code
-  \cny{USA}                                    % country
-}
-\address[id=aff3]{%                           % unique id
-  \orgname{Center for Bioinformatics and Computational Biology, University of Maryland}, % university, etc
-  \city{College Park, Maryland},                              % city
-  \postcode{20742}                                % post or zip code
-  \cny{USA}                                    % country
-}
-\address[id=aff4]{%
-  \orgname{Department of Applied Bioinformatics, Center for
-  Bioinformatics, Quantitative Biology Center, University of T\"{u}bingen},
-  \street{Sand 14},
-  \postcode{72076}
-  \city{T\"{u}bingen},
-  \cny{Germany}
-}
-\address[id=aff5]{%
-  \orgname{Department of Algorithms in Bioinformatics, Center for
-  Bioinformatics, Quantitative Biology Center, University of T\"{u}bingen},
-  \street{Sand 14},
-  \postcode{72076}
-  \city{T\"{u}bingen},
-  \cny{Germany}
-}
-\address[id=aff6]{%
-  \orgname{Department of Computer Science, University of T\"{u}bingen},
-  \street{Sand 14},
-  \postcode{72076}
-  \city{T\"{u}bingen},
-  \cny{Germany}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter short notes here                   %%
-%%                                          %%
-%% Short notes will be after addresses      %%
-%% on first page.                           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{artnotes}
-%\note{Sample of title note}     % note to the article
-%\note[id=n1]{Current address: \ldots} % note, connected to author
-%\end{artnotes}
-
-\end{fmbox}% comment this for two column layout
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% Please refer to the Instructions for     %%
-%% authors on http://www.biomedcentral.com  %%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstractbox}
-
-\begin{abstract} % abstract
-\parttitle{Motivation:} %if any
+\begin{abstract}
   As the cost of sequencing continues to decrease, the rate of sequence
@@
 
-\parttitle{Results:} %if any
   Lossy compression of quality information can greatly decrease memory
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
 
-\begin{keyword}
-\kwd{sample}
-\kwd{article}
-\kwd{author}
-\end{keyword}
+\flushbottom
+\maketitle
+\thispagestyle{empty}
 
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-
-\end{abstractbox}
-%
-%\end{fmbox}% uncomment this for twcolumn layout
-
-\end{frontmatter}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Main Body begins here                %%
-%%                                          %%
-%% Please refer to the instructions for     %%
-%% authors on:                              %%
-%% http://www.biomedcentral.com/info/authors%%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%% See the Results and Discussion section   %%
-%% for details on how to create sub-sections%%
-%%                                          %%
-%% use \cite{...} to cite references        %%
-%%  \cite{koon} and                         %%
-%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
-%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
-% <put your article body there>
-
-%%%%%%%%%%%%%%%%
-%% Background %%
-%%
-\section*{Background}
+\section*{Introduction}
 Read data from high-throughput sequencing constitutes the largest
@@
 mapping.
-
 \section*{Methods}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Backmatter begins here                   %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgments}
 
-\begin{backmatter}
-
-\section*{Competing interests}
-  The authors declare that they have no competing interests.
-
-\section*{Author's contributions}
-    Text for this section \ldots
-
-\section*{Acknowledgements}
 This project was initiated at the 2014 Bioinformatics Exchange for
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-%\bibliographystyle{vancouver} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{compression}      % Bibliography file (usually '*.bib' )
-% for author-year bibliography (bmc-mathphys or spbasic)
-% a) write to bib file (bmc-mathphys only)
-% @settings{label, options="nameyear"}
-% b) uncomment next line
-%\nocite{label}
-
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
+\bibliography{compression}
 
@@
 % profiles128.eps
-\caption{\csentence{Quality profiles obtained by $k$-means
-    clustering.} Example for the fragment library from
+\caption{Quality profiles obtained by $k$-means
+    clustering. Example for the fragment library from
   \textit{Rhodobacter sphaeroides} 2.4.1 data set using $k$ =
@@
 % compression.results.eps
-\caption{\csentence{The relationship of bits/base-call and mean
+\caption{The relationship of bits/base-call and mean
     squared error for quality value compression methods applied to
-    four data sets.} The data sets are \textit{Rhodobacter
+    four data sets. The data sets are \textit{Rhodobacter
     sphaeroides} 2.4.1; \textit{Homo sapiens} chromosome 14;
@@
 % preprocessing.results.eps
-\caption{\csentence{Read preprocessing results using compressed
-    quality values compared original uncompressed quality values.}
+\caption{Read preprocessing results using compressed
+    quality values compared original uncompressed quality values.
   The reads are for the \textit{Rhodobacter sphaeroides} 2.4.1, and
@@
 % rhodo.assembly.results.eps
-\caption{\csentence{Rankings of compression methods based on assembly
-    attributes.} The read data are for \textit{Rhodobacter
+\caption{Rankings of compression methods based on assembly
+    attributes. The read data are for \textit{Rhodobacter
     sphaeroides} 2.4.1, and the results sorted by overall
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
-
-\end{backmatter}
 \end{document}

### 2018-03-03 peerj_manuscript/manuscript.tex
removed some blank space
changed many citations from \cite to \citep, which is more appropriate for PeerJ and how we actually cite references

@@
 Advantageous use of quality values has been long
-recognize~\cite{Bonfield:1995kq}, and quality values can be used
+recognize~\citep{Bonfield:1995kq}, and quality values can be used
 throughout bioinformatics pipelines.  Quality values comprise a
-standard component of \textsc{fastq} files~\cite{Cock:2010ve}, a very
+standard component of \textsc{fastq} files~\citep{Cock:2010ve}, a very
 common format for sequence read data. At the level of the sequence
@@
 represented by a \textsc{phred} quality value, which is defined as $Q
-= -10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
+= -10\,log_{10}P$~\citep{Ewing:1998ly}. Depending on the sequencing
 technology, these quality values can range from 0 to 93, and are
@@
 assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
-alignment programs, such as Bowtie2~\cite{Langmead:2012rw}, use
+alignment programs, such as Bowtie2~\citep{Langmead:2012rw}, use
 quality values to weight mismatches between read and reference
@@
 67.9\% of reads have low-quality tails at the 3$^{\prime}$
-end~\cite{Minoche:2011km}). Recognizing these properties of Illumina
+end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
 sequence reads motivates our exploration of three general classes of
@@
 evaluated the effects of lossy compression on identifying variants
-within a data set~\cite{Canovas:2014fr,janin2013adaptive}. Here we
+within a data set~\citep{Canovas:2014fr,janin2013adaptive}. Here we
 describe our research investigating lossy compression of sequence read
@@
 \textsc{iqr} is the value used by Tukey's box
-plot~\cite{mcgill1978variations}. The main benefit of this approach is
+plot~\citep{mcgill1978variations}. The main benefit of this approach is
 that it is completely data-dependent, and no assumptions regarding the
@@
 quality value assignments of the read preprocessing tool
-Sickle~\cite{sickle}, which correspond to the 40 for ``good'' and 10
+Sickle~\citep{sickle}, which correspond to the 40 for ``good'' and 10
 for ``bad''.
@@
 effectiveness of subsequent compression using established
-algorithms~\cite{HUFFMAN:1952nr,Ziv77auniversal,
+algorithms~\citep{HUFFMAN:1952nr,Ziv77auniversal,
   DBLP:journals/tit/ZivL78}, although this potential benefit is not
@@
 
-Previous work by others \cite{Wan:2012kq} describe three similar lossy
+Previous work by others \citep{Wan:2012kq} describe three similar lossy
 compression strategies based on binning the base error probability
@@
 sets that minimize within-cluster sum of
-squares~\cite{macqueen1967some}. We sampled 1 $\times$ 10$^{4}$ reads
+squares~\citep{macqueen1967some}. We sampled 1 $\times$ 10$^{4}$ reads
 at random and computed cluster centers as read quality profiles using
 a heuristic iterative refinement approach that quickly converges to a
-locally optimal minimum~\cite{hartigan1979algorithm}. All reads are
+locally optimal minimum~\citep{hartigan1979algorithm}. All reads are
 evaluated and the nearest quality profile in Euclidean space being
@@
 the \textsc{gage} (Genome Assembly Gold-Standard
-Evaluations)~\cite{Salzberg:2012rc}, except as noted. These data sets
+Evaluations)~\citep{Salzberg:2012rc}, except as noted. These data sets
 are as follows.
@@
 quality values we analyzed the reference data with
-\textsc{QualComp}~\cite{Ochoa:2013rt}, and Quality Values Zip
-(\textsc{qvz)}~\cite{Malysa01102015}, and Read Quality Sparsifier
-(\textsc{rqs})~\cite{DBLP:conf/recomb/YuYB14}
+\textsc{QualComp}~\citep{Ochoa:2013rt}, and Quality Values Zip
+(\textsc{qvz)}~\citep{Malysa01102015}, and Read Quality Sparsifier
+(\textsc{rqs})~\citep{DBLP:conf/recomb/YuYB14}
 
@@
 We secondarily compress all otherwise compressed data sets using the
-Burrows-Wheeler algorithm~\cite{bwt} via \textsc{bz}ip2, and it is the
+Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, and it is the
 results of these subsequent compressions that we report. Although
@@
 information loss we use three loss functions following previous
-work~\cite{Malysa01102015}: mean squared error (\textsc{mse}), defined
+work~\citep{Malysa01102015}: mean squared error (\textsc{mse}), defined
 as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$; mean absolute deviation
@@
 on quality control steps of trimming and read filtering, which were
-performed using Sickle~\cite{sickle}, and make comparison to original
+performed using Sickle~\citep{sickle}, and make comparison to original
 uncompressed data. Sickle starts at the ends of the read and uses a
@@
 statistics, log average read probability
-(\textsc{lap})~\cite{Ghodsi:2013hb}, and a collection of
+(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
 reference-based metrics. The contiguity statistics include N50, which
@@
 that makes use of quality values in the assembly process:
-\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
+\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
 settings and 32 threads.
@@
 effectiveness, the recently published
-Read-Quality-Sparsifier~\cite{DBLP:conf/recomb/YuYB14} achieved
+Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
 best-case compression of 0.254 bits/base-call on \textit{Homo sapiens}
